---
title: "Self-organising maps with the wines data"
subtitle: "STA5077Z demonstration"
author: "Miguel Rodó"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

This demonstration applies a self-organising map (SOM) to the `wines`
dataset supplied with the `kohonen` package.

The data contain chemical and spectroscopic measurements on 177 wines from
three grape cultivars grown in the Piedmont region of Italy. There are 13
measured variables. The accompanying object `vintages` contains the known
cultivar labels.

The SOM will be trained using only the 13 measurements. The cultivar labels
will be added afterwards to help us assess and interpret the unsupervised
map.

The main aims are to:

1. inspect and standardise the variables;
2. fit a hexagonal SOM;
3. examine training progress and node occupancy;
4. display the observations on the fitted map;
5. interpret component planes and the U-matrix;
6. calculate quantisation and topographic error; and
7. add post hoc cluster boundaries to the map.

## Setup

Install the required package once if necessary.

```{r}
#| label: install-kohonen
#| eval: false

install.packages("kohonen")
```

Load the package and set a reproducible seed.

```{r}
#| label: setup

library(kohonen)

set.seed(5077)

knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "80%"
)

map_palette <- function(n) {
  hcl.colors(n, palette = "YlOrRd", rev = TRUE)
}
```

## Load and inspect the data

Loading `wines` also loads the associated factor `vintages`.

```{r}
#| label: load-wines-data

data("wines", package = "kohonen")

wine_x <- as.matrix(wines)
wine_class <- factor(vintages)

dim(wine_x)
head(wine_x)
table(wine_class)
```

The class labels are **not** included among the SOM inputs.

```{r}
#| label: inspect-wine-variables

wine_summary <- data.frame(
  variable = colnames(wine_x),
  mean = apply(wine_x, 2, mean) |> signif(2),
  standard_deviation = apply(wine_x, 2, sd) |> signif(2),
  minimum = apply(wine_x, 2, min) |> signif(2),
  maximum = apply(wine_x, 2, max) |> signif(2),
  row.names = NULL
)
wine_summary
```

The variables are measured in different units and have very different
numerical spreads. Without standardisation, variables with large numerical
variation would dominate the Euclidean distances used by the SOM.

```{r}
#| label: fig-wine-raw-scales
#| fig-cap: "Raw standard deviations of the 13 wine variables, shown on a logarithmic scale."
#| fig-width: 9
#| fig-height: 5

barplot(
  log10(apply(wine_x, 2, sd)),
  names.arg = colnames(wine_x),
  las = 2,
  ylab = "log10(standard deviation)",
  main = "The original variables have very different scales"
)
```

Standardise each variable to have mean zero and standard deviation one.

```{r}
#| label: standardise-wine-data

wine_scaled <- scale(
  wine_x,
  center = TRUE,
  scale = TRUE
)

round(
  cbind(
    mean = apply(wine_scaled, 2, mean),
    standard_deviation = apply(wine_scaled, 2, sd)
  ),
  3
)
```

## A familiar comparison: PCA

Before fitting the SOM, display the observations using their first two
principal component scores.

The cultivar labels are used only for colouring the plot.

```{r}
#| label: fig-wine-pca
#| fig-cap: "The wines represented by their first two principal component scores."
#| fig-width: 7
#| fig-height: 6

wine_pca <- prcomp(
  wine_scaled,
  center = FALSE,
  scale. = FALSE
)

pca_variance <- 100 *
  wine_pca$sdev^2 /
  sum(wine_pca$sdev^2)

class_colours <- hcl.colors(
  nlevels(wine_class),
  palette = "Dark 3"
)

plot(
  wine_pca$x[, 1],
  wine_pca$x[, 2],
  col = class_colours[as.integer(wine_class)],
  pch = 19,
  xlab = paste0("PC1 (", round(pca_variance[1], 1), "%)"),
  ylab = paste0("PC2 (", round(pca_variance[2], 1), "%)"),
  main = "PCA score plot"
)

legend(
  "topright",
  legend = levels(wine_class),
  col = class_colours,
  pch = 19,
  bty = "n"
)
```

PCA gives every observation a pair of continuous coordinates. The SOM will
instead assign each wine to one of a fixed number of grid units.

## Fit the SOM

Use a $6\times4$ hexagonal grid, giving 24 prototype vectors.

Every prototype has 13 components, one for each standardised wine variable.

```{r}
#| label: fit-wine-som

wine_grid <- somgrid(
  xdim = 6,
  ydim = 4,
  topo = "hexagonal",
  neighbourhood.fct = "gaussian"
)

set.seed(5077)

wine_som <- som(
  X = wine_scaled,
  grid = wine_grid,
  rlen = 500,
  alpha = c(0.05, 0.01),
  keep.data = TRUE
)

wine_som
summary(wine_som)
```

Here:

- `xdim` and `ydim` determine the number of map units;
- `topo = "hexagonal"` gives most interior units six immediate neighbours;
- `rlen = 500` means 500 complete passes through the data;
- `alpha` decreases from 0.05 to 0.01 during training; and
- the neighbourhood radius is also reduced automatically during training.

## Training progress

The `changes` plot shows the mean distance from observations to their
closest codebook vectors during training.

It measures representation error during training. It does not directly show
how far the prototypes moved at each iteration, and it does not prove that a
global optimum has been found.

```{r}
#| label: fig-wine-training-progress
#| fig-cap: "Mean distance to the closest prototype during SOM training."
#| fig-width: 7
#| fig-height: 5

plot(
  wine_som,
  type = "changes",
  main = "Training progress"
)
```

The curve should generally settle as training proceeds. Small fluctuations
are not surprising because the online algorithm updates the map from
individual observations.

## Node counts

Each wine is represented by its best matching unit (BMU). The count plot
shows how many wines map to each unit.

```{r}
#| label: fig-wine-node-counts
#| fig-cap: "Number of wines assigned to each SOM unit."
#| fig-width: 8
#| fig-height: 6

plot(
  wine_som,
  type = "counts",
  shape = "straight",
  palette.name = map_palette,
  main = "Node counts"
)
```

A large count means that many observations share the same prototype. Empty
units are possible and are shown separately by the plotting method.

## Map the wines to the grid

The cultivar labels were not used to train the SOM. We now overlay them on
the fitted map.

The numbers and colours correspond to the three levels of `vintages`.

```{r}
#| label: fig-wine-mapping
#| fig-cap: "Wine observations mapped to their BMUs. Cultivar labels are overlaid after unsupervised training."
#| fig-width: 9
#| fig-height: 6

class_number <- as.integer(wine_class)

plot(
  wine_som,
  type = "mapping",
  labels = class_number,
  col = class_colours[class_number],
  shape = "straight",
  main = "Wine cultivars on the fitted SOM",
  keepMargins = TRUE
)

legend(
  "topright",
  legend = paste(
    seq_along(levels(wine_class)),
    levels(wine_class),
    sep = ": "
  ),
  text.col = class_colours,
  bty = "n"
)
```

Interpret the map using neighbourhoods rather than ordinary Cartesian
distance:

- wines mapped to the same unit share the same prototype;
- wines in adjacent units tend to have similar chemical profiles;
- the horizontal and vertical axes have no intrinsic substantive meaning;
- the map may be rotated or reflected without changing its interpretation.

The extent to which the three known cultivars occupy separate regions is a
useful external check, but it is not part of the unsupervised fitting
criterion.

## Cultivar composition of each unit

A unit may contain wines from one cultivar or a mixture of cultivars.

```{r}
#| label: unit-cultivar-counts

n_units <- nrow(getCodes(wine_som))

unit_class_table <- table(
  unit = factor(
    wine_som$unit.classif,
    levels = seq_len(n_units)
  ),
  cultivar = wine_class
)

unit_class_table
```

For visual interpretation, colour each unit by the cultivar that occurs most
frequently among the wines mapped to it. Grey units are empty.

This is a post hoc summary. It did not influence the SOM training.

```{r}
#| label: fig-wine-majority-cultivar
#| fig-cap: "SOM units coloured by their majority cultivar, with individual wine labels overlaid."
#| fig-width: 9
#| fig-height: 6

majority_class <- apply(
  unit_class_table,
  1,
  function(z) {
    if (sum(z) == 0) {
      return(NA_integer_)
    }

    which.max(z)
  }
)

unit_background <- rep("grey90", n_units)

unit_background[!is.na(majority_class)] <- adjustcolor(
  class_colours[majority_class[!is.na(majority_class)]],
  alpha.f = 0.35
)

plot(
  wine_som,
  type = "mapping",
  labels = class_number,
  col = class_colours[class_number],
  bgcol = unit_background,
  shape = "straight",
  border = "grey40",
  main = "Majority cultivar by SOM unit",
  keepMargins = TRUE
)

legend(
  "topright",
  legend = levels(wine_class),
  fill = adjustcolor(class_colours, alpha.f = 0.35),
  border = NA,
  bty = "n"
)
```

## Component planes

The fitted codebook matrix has one row per SOM unit and one column per
original variable.

```{r}
#| label: inspect-wine-codes

wine_codes <- getCodes(wine_som)

dim(wine_codes)
head(wine_codes)
```

A component plane colours each unit according to one component of its
prototype vector.

The following four variables are shown:

1. alcohol;
2. flavonoids;
3. colour intensity; and
4. proline.

Because the SOM was trained on standardised variables, the displayed values
are standardised prototype values.

```{r}
#| label: fig-wine-selected-component-planes
#| fig-cap: "Selected component planes. Each panel shows how one standardised prototype component changes across the SOM."
#| fig-width: 10
#| fig-height: 8

selected_components <- c(
  "Alcohol" = 1,
  "Flavonoids" = 7,
  "Colour intensity" = 10,
  "Proline" = 13
)

old_par <- par(
  mfrow = c(2, 2),
  mar = c(2.5, 2.5, 3, 3)
)

for (component_name in names(selected_components)) {
  j <- selected_components[[component_name]]

  plot(
    wine_som,
    type = "property",
    property = wine_codes[, j],
    shape = "straight",
    palette.name = map_palette,
    main = component_name
  )
}

par(old_par)
```

When interpreting component planes:

- nearby units with similar colours have similar prototype values for that
  variable;
- contrasting regions show gradients across the map;
- similar spatial patterns in two component planes suggest that the
  variables tend to vary together across the fitted prototypes; and
- opposite patterns suggest opposing gradients.

The orientation of a SOM is arbitrary. The substantive interpretation comes
from comparing regions and neighbouring units, not from assigning inherent
meaning to "top", "bottom", "left" or "right".

### All component planes

The complete set of component planes can be displayed as follows.

```{r}
#| label: fig-wine-all-component-planes
#| code-fold: true
#| code-summary: "Show code for all component planes"
#| fig-cap: "Component planes for all 13 wine variables."
#| fig-width: 12
#| fig-height: 12

old_par <- par(
  mfrow = c(4, 4),
  mar = c(1.5, 1.5, 2.3, 2)
)

for (j in seq_len(ncol(wine_codes))) {
  plot(
    wine_som,
    type = "property",
    property = wine_codes[, j],
    shape = "straight",
    palette.name = map_palette,
    main = colnames(wine_codes)[j],
    heatkey = FALSE
  )
}

plot.new()
text(0.5, 0.5, "Each panel shows\none prototype component")

par(old_par)
```

## U-matrix

The U-matrix summarises differences between neighbouring prototypes.

For every unit, the `kohonen` plot displays the **sum** of its distances to
all immediate neighbours in the original standardised feature space.

```{r}
#| label: fig-wine-u-matrix
#| fig-cap: "U-matrix for the wine SOM. Larger values indicate sharper changes between neighbouring prototypes."
#| fig-width: 8
#| fig-height: 6

u_values <- plot(
  wine_som,
  type = "dist.neighbours",
  shape = "straight",
  palette.name = map_palette,
  main = "U-matrix: distances to neighbouring prototypes"
)
```

Interpretation:

- low values indicate locally similar prototypes;
- high values indicate a sharper change across that part of the map;
- connected high-value regions may suggest possible boundaries; and
- the U-matrix is a diagnostic, not a formal clustering procedure.

Because boundary units have fewer neighbours than interior units, the
package's sum-based display is partly affected by the number of neighbours.
Its main use is comparative visual interpretation across the fitted map.

## Quantisation error

For observation $i$, the observation-level quantisation error is its
distance to its BMU:

$$
q_i
=
\left\|
\mathbf{x}_i-\mathbf{m}_{b(i)}
\right\|.
$$

The overall quantisation error is the mean of these distances over
observations:

$$
Q
=
\frac{1}{n}
\sum_{i=1}^n
q_i.
$$

```{r}
#| label: calculate-wine-quantisation-error

wine_distance_to_bmu <- dist2WU(wine_som)

overall_quantisation_error <- mean(wine_distance_to_bmu)

overall_quantisation_error
summary(wine_distance_to_bmu)
```

The average is over observations, not over map units. A heavily occupied
unit therefore contributes once for every wine assigned to it, while an
empty unit contributes nothing.

The `quality` plot instead shows the average BMU distance separately within
each occupied unit.

```{r}
#| label: fig-wine-unit-quality
#| fig-cap: "Mean distance to the BMU prototype among wines assigned to each unit."
#| fig-width: 8
#| fig-height: 6

plot(
  wine_som,
  type = "quality",
  shape = "straight",
  palette.name = map_palette,
  main = "Within-unit representation quality"
)
```

Large values indicate units whose mapped wines are relatively poorly
represented by their prototype.

## Topographic error

For each observation:

1. find the closest prototype;
2. find the second-closest prototype; and
3. check whether those two prototypes are immediate grid neighbours.

An observation contributes an error if its closest two prototypes are not
adjacent.

$$
\operatorname{TE}
=
\frac{1}{n}
\sum_{i=1}^n
I\left\{
b_1(i)\text{ and }b_2(i)\text{ are not adjacent}
\right\}.
$$

```{r}
#| label: calculate-wine-topographic-error

topographic_error <- function(data, som_model) {
  data <- as.matrix(data)
  codes <- getCodes(som_model)

  squared_distances <- vapply(
    seq_len(nrow(codes)),
    function(j) {
      rowSums(
        sweep(
          data,
          MARGIN = 2,
          STATS = codes[j, ],
          FUN = "-"
        )^2
      )
    },
    numeric(nrow(data))
  )

  nearest_two <- t(
    apply(
      squared_distances,
      1,
      order
    )
  )[, 1:2, drop = FALSE]

  map_distances <- unit.distances(som_model$grid)

  first_second_map_distance <- map_distances[
    cbind(
      nearest_two[, 1],
      nearest_two[, 2]
    )
  ]

  mean(first_second_map_distance > 1 + 1e-8)
}

wine_topographic_error <- topographic_error(
  wine_scaled,
  wine_som
)

wine_topographic_error
```

A value of zero would mean that every observation's two closest prototypes
are adjacent. Larger values indicate more frequent local breaks in the
organisation of the map.

Topographic error is binary at the observation level. A second BMU two grid
steps away and one several grid steps away both count as one error.

## Post hoc clustering of the prototypes

The SOM itself gives an organised grid of prototypes, not a final set of
discrete clusters.

We can cluster the 24 prototype vectors after training. Here, hierarchical
clustering is applied to the codebook vectors and cut into three groups.

```{r}
#| label: cluster-wine-prototypes

code_distances <- object.distances(
  wine_som,
  type = "codes"
)

code_hclust <- hclust(
  code_distances,
  method = "ward.D2"
)

unit_cluster <- cutree(
  code_hclust,
  k = 3
)

unit_cluster
```

The number of clusters is chosen here for demonstration and to facilitate
comparison with the three known cultivars. It was not selected by the SOM
algorithm.

Draw the cluster boundaries on top of the U-matrix.

```{r}
#| label: fig-wine-cluster-boundaries
#| fig-cap: "Hierarchical clusters of the SOM prototypes, displayed as boundaries on the U-matrix."
#| fig-width: 8
#| fig-height: 6

plot(
  wine_som,
  type = "dist.neighbours",
  shape = "straight",
  palette.name = map_palette,
  main = "Post hoc clusters of SOM prototypes"
)

add.cluster.boundaries(
  wine_som,
  clustering = unit_cluster,
  lwd = 4
)
```

These lines are reproducible post hoc cluster boundaries:

1. the SOM prototypes are clustered;
2. every map unit receives a cluster label; and
3. a boundary is drawn between adjacent units with different labels.

The boundaries are not produced by the basic SOM training algorithm.

Compare the post hoc prototype clusters with the known cultivars.

```{r}
#| label: compare-clusters-and-cultivars

wine_cluster <- unit_cluster[wine_som$unit.classif]

table(
  SOM_cluster = wine_cluster,
  cultivar = wine_class
)
```

Cluster numbers are arbitrary. The comparison should focus on whether each
cluster is dominated by one cultivar and where mixing occurs.

## What has the SOM shown?

The fitted map provides several related views of the same model:

- the mapping plot shows where individual wines are represented;
- the count plot shows how many wines use each prototype;
- component planes show what the prototypes represent chemically;
- the U-matrix shows where neighbouring prototypes differ sharply;
- the quality plot shows where observations are represented less well;
- quantisation error measures average representation accuracy;
- topographic error measures local failures of grid organisation; and
- post hoc clustering turns the continuous prototype organisation into
  discrete regions.

These views should be interpreted together. No single plot is sufficient on
its own.

## Optional extension: sensitivity to map size

A larger map contains more prototypes and will usually reduce quantisation
error, but may contain more empty units and may not improve topographic
organisation.

The following code compares three grid sizes.

```{r}
#| label: compare-wine-map-sizes
#| code-fold: true
#| code-summary: "Show map-size comparison"

fit_wine_som <- function(xdim, ydim, seed) {
  set.seed(seed)

  som(
    data = wine_scaled,
    grid = somgrid(
      xdim = xdim,
      ydim = ydim,
      topo = "hexagonal",
      neighbourhood.fct = "gaussian"
    ),
    rlen = 500,
    alpha = c(0.05, 0.01),
    keep.data = TRUE
  )
}

map_sizes <- list(
  "4 x 3" = c(4, 3),
  "6 x 4" = c(6, 4),
  "8 x 6" = c(8, 6)
)

map_size_results <- lapply(
  seq_along(map_sizes),
  function(i) {
    dimensions <- map_sizes[[i]]

    fit <- fit_wine_som(
      xdim = dimensions[1],
      ydim = dimensions[2],
      seed = 5077 + i
    )

    data.frame(
      map = names(map_sizes)[i],
      units = prod(dimensions),
      empty_units = sum(
        tabulate(
          fit$unit.classif,
          nbins = prod(dimensions)
        ) == 0
      ),
      quantisation_error = mean(dist2WU(fit)),
      topographic_error = topographic_error(
        wine_scaled,
        fit
      )
    )
  }
)

do.call(
  rbind,
  map_size_results
)
```

Do not select a map solely because it has the smallest quantisation error.
Map size should balance:

- representation accuracy;
- topographic organisation;
- the number of empty units;
- stability across initialisations; and
- interpretability.

## Discussion questions

1. Why was standardisation important for these data?
2. Which plot directly shows where individual wines map?
3. What additional information is supplied by the component planes?
4. What does a high U-matrix value mean?
5. How does the unit-level quality plot differ from the overall
   quantisation error?
6. Why does topographic error use the first and second BMUs?
7. Were the cultivar labels used to train this SOM?
8. Are the post hoc cluster boundaries part of the SOM algorithm?
9. Why should the horizontal and vertical directions of the map not be
   interpreted as ordinary variables?
10. How would increasing the number of units affect quantisation error and
    interpretability?

## Recap

The wine SOM converts 13-dimensional observations into positions on a fixed
two-dimensional grid:

$$
\mathbf{x}_i
\longmapsto
\text{BMU}
\longmapsto
\text{grid location}.
$$

The prototypes remain 13-dimensional. The two-dimensional grid organises
those prototypes so that neighbouring units tend to represent similar
chemical profiles.

The cultivar labels provide useful external information after fitting, but
the map itself is learnt without them.
