### The main idea

A self-organising map learns a set of representative vectors called **prototypes** or **codebook vectors**.

Each prototype has:

- a learnt position in the original data space; and
- a fixed position on a low-dimensional grid.

The map is trained so that nearby grid positions tend to represent similar data patterns.

### Why this belongs in dimensionality reduction

After training, each observation is represented by its nearest prototype.

$$
\mathbf{x}_i
\quad\longmapsto\quad
\text{nearest prototype}
\quad\longmapsto\quad
\text{grid location}.
$$

So a high-dimensional observation is represented by a low-dimensional grid location.

The representation is discrete rather than continuous.

### What a SOM is useful for

Self-organising maps are mainly exploratory tools.

They are used to:

- visualise high-dimensional observations on a two-dimensional grid;
- find broad gradients across the data;
- inspect possible groups or boundaries;
- summarise observations by prototypes; and
- study how original variables change across the map.

## Two spaces

A common source of confusion is that a SOM uses two different spaces.

### Data space

The observations live in the original data space:

$$
\mathbf{x}_i \in \mathbb{R}^p.
$$

Each SOM unit has a prototype vector:

$$
\mathbf{m}_j \in \mathbb{R}^p.
$$

The prototype has the same number of components as the original observations.

### Map space

Each unit also has a fixed grid location:

$$
\boldsymbol{\ell}_j = (\text{row}_j, \text{column}_j).
$$

For example, a $6 \times 6$ SOM has 36 grid positions.

The grid positions are fixed.

The prototypes are learnt.

### Keep these separate

The prototype tells us **what the unit represents**.

The grid location tells us **where the unit appears on the map**.

$$
\begin{array}{ccc}
\text{unit } j
&\longrightarrow&
\mathbf{m}_j \in \mathbb{R}^p \\
&\longrightarrow&
\boldsymbol{\ell}_j \in \mathbb{R}^2
\end{array}
$$

## A first visual example

Here is a simple simulated example.

The original observations have four measured variables, but the data were generated from two hidden coordinates.

```{r}
#| label: fig-som-toy-data
#| code-fold: true
#| code-summary: "Show simulation code"
#| fig-cap: "A toy dataset with four measured variables generated from two hidden coordinates."

som_data <- make_som_toy_data(n = 300, seed = 5077)

som_data |>
  ggplot(aes(x = u, y = v, colour = group)) +
  geom_point(alpha = 0.75, size = 2) +
  coord_equal() +
  cowplot::theme_cowplot() +
  labs(
    x = "Hidden coordinate u",
    y = "Hidden coordinate v",
    colour = "Region"
  )
```

In practice, we would not know these hidden coordinates.

The SOM will only see the four measured variables.

## Nearest-prototype representation

Suppose the SOM has $K$ prototypes:

$$
\mathbf{m}_1,\mathbf{m}_2,\ldots,\mathbf{m}_K.
$$

For each observation, we find the closest prototype:

$$
c(i)
=
\arg\min_j
\left\|\mathbf{x}_i-\mathbf{m}_j\right\|.
$$

This closest prototype is called the **best matching unit**, or **BMU**.

### The BMU uses data-space distance

The distance

$$
\left\|\mathbf{x}_i-\mathbf{m}_j\right\|
$$

is computed in the original $p$-dimensional data space.

This is not a distance on the SOM grid.

### Small numerical example

Suppose

$$
\mathbf{x}=(1,-2,2),
$$

and two prototypes are

$$
\mathbf{m}_1=(2,-1,3),
\qquad
\mathbf{m}_2=(-2,0,1).
$$

Then

$$
\begin{aligned}
\|\mathbf{x}-\mathbf{m}_1\|
&=
\sqrt{(1-2)^2+(-2+1)^2+(2-3)^2}
=\sqrt{3},\\
\|\mathbf{x}-\mathbf{m}_2\|
&=
\sqrt{(1+2)^2+(-2-0)^2+(2-1)^2}
=\sqrt{14}.
\end{aligned}
$$

So prototype 1 is the BMU.

## Why add a grid?

If the prototypes were unrelated, their labels would be arbitrary.

Prototype 7 would not have any special relationship to prototype 8.

A SOM places prototypes on a fixed grid and encourages neighbouring grid units to represent similar data patterns.

$$
\text{nearby on the grid}
\quad\Rightarrow\quad
\text{similar prototype vectors}.
$$

This is the topological idea behind the method.

### Buttons on a flexible sheet

A useful picture is to imagine the prototypes as buttons sewn onto a sheet.

The sheet begins as a regular grid.

During training, the buttons move through the high-dimensional data space, but neighbouring buttons are encouraged to move together.

The fitted SOM is therefore like a discrete low-dimensional surface sitting inside the original feature space.

## The SOM algorithm

The online SOM algorithm processes observations one at a time.

At each step, it performs three actions.

### Step 1: Find the BMU

For the current observation $\mathbf{x}_i$, find

$$
b
=
\arg\min_j
\left\|
\mathbf{x}_i-\mathbf{m}_j
\right\|.
$$

This is the competition step.

The prototypes compete to represent the observation.

### Step 2: Find the BMU's map neighbours

Now look at the BMU's position on the map grid.

Units close to the BMU on the grid should also learn from this observation.

The grid distance is

$$
\left\|
\boldsymbol{\ell}_b-\boldsymbol{\ell}_j
\right\|.
$$

This is a distance in map space.

### Step 3: Update the prototypes

Each prototype is moved towards the observation:

$$
\mathbf{m}_j^{\text{new}}
=
\mathbf{m}_j^{\text{old}}
+
\alpha_t h_{bj}(t)
\left(
\mathbf{x}_i-\mathbf{m}_j^{\text{old}}
\right).
$$

Here:

- $\alpha_t$ is the learning rate;
- $h_{bj}(t)$ is the neighbourhood weight; and
- $\mathbf{x}_i-\mathbf{m}_j^{\text{old}}$ points from the prototype towards the observation.

### Neighbourhood weights

A common choice is a Gaussian neighbourhood:

$$
h_{bj}(t)
=
\exp\left\{
-
\frac{
\left\|
\boldsymbol{\ell}_b-\boldsymbol{\ell}_j
\right\|^2
}{
2\sigma_t^2
}
\right\}.
$$

The closer unit $j$ is to the BMU on the grid, the larger $h_{bj}(t)$ is.

### Two distances, two roles

This is the most important technical distinction.

$$
\left\|\mathbf{x}_i-\mathbf{m}_j\right\|
$$

chooses the BMU in the original data space.

$$
\left\|\boldsymbol{\ell}_b-\boldsymbol{\ell}_j\right\|
$$

decides how strongly nearby grid units learn together.

## A small update by hand

Suppose the BMU is unit B.

Units A and C are immediate neighbours of B on a one-dimensional map.

We use:

$$
\alpha=0.45,
\qquad
h_B=1,
\qquad
h_A=h_C=0.5.
$$

```{r}
#| label: fig-som-manual-update
#| code-fold: true
#| code-summary: "Show update code"
#| fig-cap: "One SOM update. The BMU moves most strongly, while neighbouring units move partway towards the same observation."

update_tbl <- manual_update_example()
observation_tbl <- tibble(x = 0, y = 1.2)

update_tbl |>
  ggplot(aes(x = x, y = y)) +
  geom_path(
    data = update_tbl |> arrange(stage, unit),
    aes(group = stage, linetype = stage),
    linewidth = 0.8,
    alpha = 0.6
  ) +
  geom_point(
    aes(shape = stage),
    size = 3
  ) +
  geom_text_repel(aes(label = unit), seed = 5077) +
  geom_point(
    data = observation_tbl,
    aes(x = x, y = y),
    inherit.aes = FALSE,
    size = 4
  ) +
  annotate("text", x = 0.15, y = 1.25, label = "observation", hjust = 0) +
  coord_equal(xlim = c(-1.5, 1.5), ylim = c(-0.2, 1.4)) +
  cowplot::theme_cowplot() +
  labs(
    x = "Feature 1",
    y = "Feature 2",
    shape = NULL,
    linetype = NULL
  )
```

The BMU moves most.

Its neighbours also move, but less strongly.

Far-away units would barely move at all.

## Why learning rates and neighbourhoods shrink

Early in training, the neighbourhood radius is usually large.

This means many units move together.

That helps the map organise itself globally.

### Early training

Large $\sigma_t$:

- many prototypes are updated together;
- the map is smooth;
- the broad ordering of the map is established.

### Late training

Small $\sigma_t$:

- mainly the BMU and close neighbours are updated;
- prototypes can fit local detail;
- the map becomes more precise.

Usually both $\alpha_t$ and $\sigma_t$ decrease over time.

## Fitting the toy SOM

Now train an $8 \times 8$ SOM using the four measured variables.

```{r}
#| label: fit-som-toy
#| include: false

som_variables <- c("x1", "x2", "x3", "x4")
som_x <- som_data |>
  select(all_of(som_variables)) |>
  scale()

som_fit <- train_som(
  som_x,
  n_rows = 8,
  n_cols = 8,
  iterations = 2500,
  alpha_start = 0.35,
  alpha_end = 0.02,
  sigma_start = 4,
  sigma_end = 0.8,
  seed = 5077,
  trace_every = 100
)

som_assignment <- assign_bmus(som_x, som_fit) |>
  left_join(som_data, by = "observation") |>
  mutate(
    plot_col = col + runif(n(), min = -0.32, max = 0.32),
    plot_row = row + runif(n(), min = -0.32, max = 0.32)
  )

som_codes <- make_code_table(som_fit, variable_names = som_variables)
som_u_matrix <- make_u_matrix(som_fit)
```

### Training progress

```{r}
#| label: fig-som-training-progress
#| code-fold: true
#| code-summary: "Show training progress code"
#| fig-cap: "Quantisation error during SOM training. A continued decrease suggests that more training may still help."

som_fit$trace |>
  ggplot(aes(x = iteration, y = quantisation_error)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.7) +
  cowplot::theme_cowplot() +
  labs(
    x = "Iteration",
    y = "Quantisation error"
  )
```

The error measures how far observations are from their assigned prototypes.

It should usually decrease and then flatten out.

## The fitted map

Each observation is assigned to its BMU and placed at that BMU's grid location.

```{r}
#| label: fig-som-fitted-map
#| code-fold: true
#| code-summary: "Show fitted map code"
#| fig-cap: "Observations represented by their BMU locations on the fitted SOM grid. The colours use the hidden regions only for explanation."

som_assignment |>
  ggplot(aes(x = plot_col, y = plot_row, colour = group)) +
  geom_point(alpha = 0.75, size = 1.8) +
  scale_x_continuous(breaks = seq_len(8)) +
  scale_y_continuous(breaks = seq_len(8)) +
  coord_equal() +
  cowplot::theme_cowplot() +
  labs(
    x = "SOM column",
    y = "SOM row",
    colour = "Hidden region"
  )
```

The SOM has converted four-dimensional observations into positions on an $8 \times 8$ grid.

Nearby locations tend to represent similar observations.

## Node counts

A node count plot shows how many observations are assigned to each unit.

```{r}
#| label: fig-som-node-counts
#| code-fold: true
#| code-summary: "Show node-count code"
#| fig-cap: "Node counts for the fitted SOM. Empty or heavily occupied nodes are often informative."

som_assignment |>
  count(row, col, name = "node_count") |>
  right_join(som_fit$grid, by = c("row", "col")) |>
  mutate(node_count = replace_na(node_count, 0)) |>
  ggplot(aes(x = col, y = row, fill = node_count)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = node_count), size = 3) +
  scale_x_continuous(breaks = seq_len(8)) +
  scale_y_continuous(breaks = seq_len(8)) +
  coord_equal() +
  cowplot::theme_cowplot() +
  labs(
    x = "SOM column",
    y = "SOM row",
    fill = "Count"
  )
```

A large count means many observations are represented by the same prototype.

Empty nodes may indicate that the map is too large or that the data occupy only part of the grid.

## U-matrix

A U-matrix displays how different neighbouring prototypes are from one another.

For each grid unit, we compute its average distance to its immediate neighbours in data space.

```{r}
#| label: fig-som-u-matrix
#| code-fold: true
#| code-summary: "Show U-matrix code"
#| fig-cap: "U-matrix for the fitted SOM. Larger values indicate sharper changes between neighbouring prototypes."

som_u_matrix |>
  ggplot(aes(x = col, y = row, fill = u_distance)) +
  geom_tile(colour = "white") +
  scale_x_continuous(breaks = seq_len(8)) +
  scale_y_continuous(breaks = seq_len(8)) +
  coord_equal() +
  cowplot::theme_cowplot() +
  labs(
    x = "SOM column",
    y = "SOM row",
    fill = "Neighbour\ndistance"
  )
```

Large values may indicate boundaries between different regions of the map.

A U-matrix is not a formal clustering method.

It is a visual diagnostic for the fitted prototypes.

## Component planes

A component plane colours each SOM unit according to one component of its prototype vector.

For variable $r$, unit $j$ is coloured by $m_{jr}$.

```{r}
#| label: fig-som-component-planes
#| code-fold: true
#| code-summary: "Show component-plane code"
#| fig-cap: "Component planes for the fitted SOM. Each panel shows how one original variable changes across the map."
#| fig-height: 7

som_codes |>
  mutate(variable = factor(variable, levels = som_variables)) |>
  ggplot(aes(x = col, y = row, fill = prototype_value)) +
  geom_tile(colour = "white") +
  facet_wrap(~ variable, ncol = 2) +
  scale_x_continuous(breaks = seq_len(8)) +
  scale_y_continuous(breaks = seq_len(8)) +
  coord_equal() +
  cowplot::theme_cowplot() +
  labs(
    x = "SOM column",
    y = "SOM row",
    fill = "Prototype\nvalue"
  )
```

Component planes help us interpret the map.

They show which variables are high or low in each region.

They can also show whether two variables have similar or opposite patterns across the map.

## SOMs and k-means

The k-means connection is important, but we do not need to start there.

Both methods use prototypes.

Both assign observations to their nearest prototype.

The difference is what happens after the nearest prototype has been found.

### K-means-style update

If only the BMU is updated, then

$$
h_{bj}
=
\begin{cases}
1, & j=b,\\
0, & j\neq b.
\end{cases}
$$

The update becomes

$$
\mathbf{m}_b
\leftarrow
\mathbf{m}_b+
\alpha_t
\left(\mathbf{x}_i-\mathbf{m}_b\right).
$$

Only the winning prototype moves.

### SOM update

In a SOM, the BMU and its map neighbours move:

$$
\mathbf{m}_j
\leftarrow
\mathbf{m}_j+
\alpha_t h_{bj}(t)
\left(\mathbf{x}_i-\mathbf{m}_j\right).
$$

This keeps nearby grid units similar.

### The key comparison

| Method | Representation | Neighbourhood structure |
|---|---|---|
| K-means | nearest prototype | prototypes are independent |
| SOM | nearest prototype on a grid | neighbouring prototypes learn together |

So a SOM can be viewed as a constrained or regularised form of k-means.

But in this part of the course, the more useful framing is:

$$
\text{SOM}=
\text{prototype-based dimensionality reduction on a fixed grid}.
$$

## What the map does not guarantee

A fitted SOM is useful, but it should not be over-interpreted.

The map does not guarantee that:

- Euclidean distance on the grid equals Euclidean distance in the original data space;
- all original neighbourhoods are preserved;
- empty space on the grid means empty space in the original data;
- visual separation automatically means genuine clusters; or
- cluster size and density are faithfully represented.

A SOM tries to preserve local organisation.

It does not perfectly preserve all distances.

## Quality measures

A useful SOM should represent the data well and organise the prototypes coherently.

These are different goals.

### Quantisation error

Quantisation error measures how far observations are from their assigned prototypes:

$$
\frac{1}{n}
\sum_{i=1}^n
\left\|
\mathbf{x}_i-
\mathbf{m}_{c(i)}
\right\|.
$$

Lower values mean better prototype representation.

### Topographic error

For each observation:

1. find the closest prototype;
2. find the second-closest prototype;
3. check whether those two prototypes are adjacent on the grid.

A topographic error occurs when the two closest prototypes are not grid neighbours.

This indicates that the map may have broken local topology.

### The trade-off

A good SOM balances:

$$
\text{representation accuracy}
\qquad\text{and}\qquad
\text{topological organisation}.
$$

K-means will usually have smaller quantisation error if it uses the same number of prototypes.

That does not automatically make k-means better, because k-means does not produce an organised low-dimensional map.

## Practical choices

When fitting a SOM, we need to choose:

- whether variables should be scaled;
- the map size, such as $6 \times 6$ or $10 \times 10$;
- the grid shape, such as rectangular or hexagonal;
- the number of training iterations;
- the initial learning rate;
- the initial neighbourhood radius; and
- the decay schedule for learning rate and radius.

### Scaling

SOMs commonly use Euclidean distance.

If variables have very different scales, large-scale variables can dominate the map.

As with PCA and MDS, scaling is often essential.

### Map size

A small map gives a coarse summary.

A large map gives more detail, but may be harder to interpret.

A very large map may contain many empty or unstable units.

## A typical R workflow

The practical workflow is usually:

1. prepare and scale the data matrix;
2. choose the grid size;
3. train the SOM;
4. assign observations to BMUs;
5. inspect training progress;
6. inspect node counts, U-matrix and component planes; and
7. decide whether the map is useful for the scientific question.

### Example using the `kohonen` package

```{r}
#| label: code-kohonen-workflow
#| eval: false
#| code-fold: false

library(kohonen)

x_scaled <- scale(my_data)

grid <- somgrid(
  xdim = 8,
  ydim = 8,
  topo = "hexagonal"
)

fit <- som(
  data = x_scaled,
  grid = grid,
  rlen = 200,
  alpha = c(0.05, 0.01)
)

plot(fit, type = "changes")
plot(fit, type = "counts")
plot(fit, type = "dist.neighbours")
plot(fit, type = "codes")
```

The exact plotting functions depend on the package, but the interpretation is the same.

## Link to the other methods

| Method | Low-dimensional representation | Main organising principle |
|---|---|---|
| PCA | continuous component scores | preserve variance |
| MDS | continuous map coordinates | preserve dissimilarities |
| ISOMAP | continuous map coordinates | preserve graph-based geodesic distances |
| SOM | discrete grid coordinates | learn organised prototypes |
| K-means | cluster labels | minimise within-cluster prototype error |

SOMs sit naturally between dimensionality reduction and clustering.

They give a low-dimensional map, but the map is built from prototypes.

## Recap

A self-organising map learns prototypes attached to a fixed low-dimensional grid.

Each observation is assigned to its BMU.

The BMU and its neighbours move towards the observation.

This creates an organised map of high-dimensional data.

The k-means connection is:

$$
\text{SOM with no neighbourhood learning}
\approx
\text{online k-means}.
$$

The dimensionality-reduction connection is:

$$
\mathbf{x}_i\in\mathbb{R}^p
\quad\longmapsto\quad
\boldsymbol{\ell}_{c(i)}\in\mathbb{R}^2.
$$

## Conceptual checks

1. What is the difference between a prototype vector and a grid location?
2. Which distance is used to choose the BMU?
3. Which distance is used to decide how strongly neighbouring units are updated?
4. Why do we usually decrease the learning rate and neighbourhood radius during training?
5. In what sense is a SOM related to k-means?
6. Why can a SOM be useful for dimensionality reduction even if it is not a distance-preserving map?
7. What does a component plane show?
8. What is the difference between quantisation error and topographic error?
9. Why should visual clusters in a SOM be interpreted cautiously?
10. Why should variables often be scaled before fitting a SOM?

## Supplementary detail

### Batch SOM

The online algorithm updates prototypes one observation at a time.

A batch version updates each prototype using all observations assigned to relevant neighbouring units.

A simplified batch update has the form

$$
\mathbf{m}_j
=
\frac{
\sum_i w_{ij}\mathbf{x}_i
}{
\sum_i w_{ij}
},
$$

where $w_{ij}$ is larger when observation $i$ is assigned to a unit close to $j$ on the grid.

If the neighbourhood contains only the BMU, this resembles the usual k-means centroid update.

### Why the SOM constraint can be checked

Because SOM uses the same nearest-prototype representation as k-means, we can compare reconstruction errors using the same number of prototypes.

K-means should generally have smaller reconstruction error because its prototypes are unconstrained.

If the SOM error is much larger, the low-dimensional grid constraint may be unsuitable for the dataset.

If the SOM error is close, the organised grid may be a useful approximation.

