---
title: STA2005S Regression Question Bank
subtitle: Questions
date: "`r format(Sys.Date(), '%d %B %Y')`"
format:
  html:
    embed-resources: true
    number-sections: true
    toc: true
  pdf:
    embed-resources: true
    number-sections: true
    include-in-header:
      text: |
        \usepackage{graphicx} % Allows including images
        \usepackage{booktabs} 
        \usepackage{amsmath}
        \usepackage{url}
        \usepackage{subfigure}
        \usepackage{xcolor}
        \usepackage{amsthm,amsfonts,amssymb,amscd,amsxtra}
        \usepackage{pgfpages}
        \usepackage{animate}
        \usepackage[backend=bibtex,style=verbose-trad2]{biblatex}
        \bibliography{bibliography} 
        \usepackage{mathpazo}
        \usepackage{unicode-math}
---

```{r}
#| include: false
library(ggplot2)
library(tibble)
```

# Prerequisites

## Statistical distributions

## Linear algebra

### Basic calculations and properties

Suppose that we have the following $2\times 2$ matrix $\symbf{A}$:

$$
\begin{bmatrix}
1 & 2 \\
3 & 4
\end{bmatrix}
$$

a. What is $\symbf{a}_2$?
b. What is $\symbf{a}_{[2]}$?
c. What is $\mathrm{trace}(\symbf{A})$?
d. Is $\symbf{A}$ orthogonal?
e. Is $\symbf{A}$ full rank?
f. Is $\symbf{A}$ symmetric?
g. Based on the answer to the previous question - could $\symbf{A}$ be positive definite?
h. Is $\symbf{A}$ idempotent?

### Block matrices

#### Basics

Suppose $\symbf{X}:6\times 1=\begin{bmatrix} \symbf{X}^{(1)}:2\times 1 \\ \symbf{X}^{(2)}:m\times n \end{bmatrix}$.

a. What is $m$?
b. What is $n$?

Suppose $\symbf{A}:3\times 4 = \begin{bmatrix} \symbf{A}^{(1)}:3\times 2 & \symbf{A}^{(2)}:c\times c \end{bmatrix}$.

b. What is $c$?
c. What is $d$?

#### Calculation 

Let's say that $\symbf{A}:4\times 4=\begin{bmatrix} \symbf{I}_3 & \symbf{C} \\ \symbf{D} & \symbf{E} \end{bmatrix}$.

a. What is the dimension of $\symbf{C}$?
b. What is the dimension of $\symbf{D}$?
c. What is the dimension of $\symbf{E}$?

Suppose that $\symbf{B}:4\times 5=\begin{bmatrix} \symbf{F} & \symbf{G} \\ \symbf{H} & \symbf{I} \end{bmatrix}$.
Suppose that we can use these block-matrix representations of these matrices to calculate $\symbf{AB}$.

d. What is the dimension of $\symbf{AB}$?
e. What is $\symbf{AB}$?
f. What is the dimension of $\symbf{F}$?
g. What is the dimension of $\symbf{G}$?
h. What is the dimension of $\symbf{H}$?
i. What is the dimension of $\symbf{I}$?

#### Orthonormality

Suppose that $\symbf{P}:k\times k$ is orthonormal. Show that $\symbf{P}^{-1}=\symbf{P}'$.

### Special matrices

Suppose that $\symbf{A}:3\times 3$ is idempotent of rank $2$ with eigenvectors $\symbf{e}_1$, $\symbf{e}_2$ and $\symbf{e}_3$.

a. How many non-zero eigenvalues are there?
b. What is $\symbf{e}_1'\symbf{e}_1$?
c. What is $\symbf{e}_1'\symbf{e}_2$?
d. Factorise $\symbf{A}$ using one of the factorisations we've discussed in class.
e. Why are $\symbf{A}$'s eigenvalues only one or zero?

### Invertibility

a. In three lines (more or less), show that if  $\symbf{P}'\symbf{P}=\symbf{I}$ then $\symbf{P}^{-1}=\symbf{P}’$.

# General linear model formulation

In a general linear model, we have that $\symbf{Y}= \symbf{X}\beta + \symbf{\epsilon}$ where $\symbf{Y}$ is an $n \times 1$ vector of responses, $\symbf{X}$ is a fixed $n \times k$ matrix of predictors and $\symbf{\epsilon}$ is an $n \times 1$ vector of errors. In additiion, $\mathrm{E}[\symbf{\epsilon}] = \symbf{0}$ and $\mathrm{E}[\symbf{\epsilon}\symbf{\epsilon}'] = \sigma^2\symbf{I}$. Finally, suppose that $\symbf{X}$ is full column rank (i.e. no column is a linear combination of any other columns) and $n>k$ (i.e. more observations than parameters).

# Week Two

## Model formulation

### Simple linear regression

We model $y_i$ as

$$
y_i = \beta_0 + \beta_1 x_i + e_i,
$$

where $e_i \stackrel{i.i.d.}{\sim} N(0, \sigma^2)$.

a. Which quantities are **known** in this model?
b. Which quantities are **unknown**?
c. Which of the unknowns are parameters, and which are random variables?
d. Derive the mean of $y_i$.
e. Derive the variance of $y_i$.
f. Substituting in specific numbers for each of the unknown parameters, write the model in scalar notation.
g. Using your answer to question f, write the model in scalar notation for a specific value of $x_i$ (i.e. substitute in some number for $x_i$).
h. Work out the mean and variance of $y_i$ for the specific value of $x_i$ you used in g.
i. Add one unit to $x_i$ and work out the new mean and variance of $y_i$.
j. Comment on the effect of adding one unit to $x_i$ on the mean and variance of $y_i$.
k. Given that you have provided specific values for $x_i$ and $\beta_0$, $\beta_1$ and $\sigma^2$, what is the fitted value for $y_i$?
l. Suppose the error term is 2.5. What is the actual value of $y_i$?
m. Give plausible examples of variables $x_i$ and $y_i$, given the parameter values you used in 
n. Contrast the model you proposed in j with simply estimating the mean of $y$, regardless of $x_i$.

### Fitted values and residuals

Let

$$
\hat{y}_i=\sum_{j=0}^p x_{ij}\hat{\beta}_j,\qquad 
\hat{e}_i=y_i-\hat{y}_i,\qquad
e_i=y_i-\sum_{j=0}^p x_{ij}\beta_j,
$$

with $e_i\stackrel{iid}{\sim}N(0,\sigma^2)$ and $x_{i0}=1$.

a. Which quantities are known before fitting the model?
b. Which quantities become known after fitting the model?
c. Which quantities remain unknown even after fitting?

d. Suppose $p=2$. What is the **mean** of $y_i$ in terms of $\beta$ and $x_{i1},x_{i2}$?

e. Set $\beta_0=1,\ \beta_1=-2,\ \beta_2=3$. What is the **mean** of $y_i$?

f. Choose specific values for $x_{i1}$ and $x_{i2}$ and compute the **mean** of $y_i$.

g. Suppose $y_i=5$. Express $e_i$ in terms of $\beta$ and $x_{i1},x_{i2}$. (Keep it symbolic.)

h. We fit the model on all $n$ observations and obtain $\hat{\beta}_0=1,\ \hat{\beta}_1=-2,\ \hat{\beta}_2=3$.
What is the fitted value $\hat{y}_i$ (in terms of $x_{i1},x_{i2}$)?

i. Given $y_i=5$ and the estimates in (h), what is $\hat{e}_i$? 

j. Suppose $y_j=4$ for $x_{j1}=2$ and $x_{j2}=1$. With the estimates in (h), compute $\hat{y}_j$, $\hat{e}_j$.
Also write $e_j$ symbolically (in terms of the true $\beta$).

k. How does $\hat{y}_i$ relate to the mean $E[y_i]$?

n. Give plausible examples of variables $x_{i1}, x_{i2}, y_i$ consistent with the signs in (e).

### Model representation

Suppose that we observe the following values for our two explanatory variables for three observations:

- $x_{11} = 1$, $x_{12} = 2$,
- $x_{21} = 2$, $x_{22} = 3$,
- $x_{31} = 4$, $x_{32} = -1$.

a. Assuming we fit an intercept, how many regression parameters are there in this model?
b. What is $n$?
c. Write the model in scalar notation for each observation.
d. Write the model in vector notation.
e. Write the model in matrix notation.
f. What are the dimensions of the matrices in e?
g. What are the dimensions of $\mathbf{X}'\mathbf{X}$ and $\mathbf{X} \mathbf{X}'$?
h. Suppose we swop the order of the columns in $\mathbf{X}$, so that it is now $[x_{i1},\ 1,\ x_{i2}]$. Does this change the fitted values? Explain.
j. If we swop the order of the columns in $\mathbf{X}$, what part of the model formulation does change?
k. What notation denotes the vector of regression parameters?
l. What notation denotes the vector of estimated regression parameters?
m. Provide the matrix formula for the fitted values $\hat{\mathbf{y}}$.

Suppose that we observe $y_1 = 3, y_2 = 5, y_3 = 7$.

n. What are the fitted values $\hat{y}_i$ for each observation $i$? How can you work this out immediately?
o. What are the residuals $\hat{e}_i$ for each observation $i$? How can you work this out immediately?
p. What are the estimated regression coefficients?
q. What are the true regression coefficients?
r. Drop the second explanatory variable, so that we now have $x_{i1}$ and an intercept. Rewrite the model in scalar, vector and matrix notation.
s. What are the dimensions of the matrices in l?
t. Calculate $\hat{\mathbf{\beta}}$ for the new model.
u. Calculate the fitted values $\hat{\mathbf{y}}$ for the new model.
v. Calculate the residuals $\hat{\mathbf{e}}$ for the new model.
w. Compare your answers in u and v to the previous model (that used $x_{i2}$).


# Week three

## Confidence Intervals and Hypothesis Testing

### Individual regression coefficients

Suppose you have fitted a linear regression model to a dataset with 100 observations and 3 predictors (including the intercept). The estimated regression coefficient for the variable $X_1$ is $\hat{\beta}_1 = 2.5$ with an estimated standard error of $\hat{se}(\hat{\beta}_1) = 0.5$.

Here are the following critical values:

- $z^{0.025} \approx -1.96$
- $z^{0.05} \approx -1.645$
- $t^{0.025}_{3} \approx -3.18$
- $t^{0.05}_{3} \approx -2.35$
- $t^{0.025}_{97} \approx -1.9848$
- $t^{0.05}_{97} \approx -1.6607$
- $t^{0.025}_{100} \approx -1.984$
- $t^{0.05}_{100} \approx -1.6602$

a. Construct a 95% confidence interval for $\beta_1$.
b. Interpret this confidence interval.
c. Calculate the increase in the width of the confidence interval if there were only 6 observations (but, remarkably, the standard error remained the same).
d. Do lower sample sizes affect the expected value of $s^2$? Explain your answer.
e. Suppose that $s^2=0.05$, the sample size is $n$ and $k=1$ (i.e. an intercept-only model). Estimate the standard error of $\beta_0$.
f. Contrast the answer in $e$ to $d$ and explain the difference.
g. As $r$ increases, a $\chi^2_r$ random variable tends towards a normal random variable with mean $r$ and variance $2r$. Use this to show that the estimate of $s^2$ becomes more precise as $n$ increases, for sufficiently large $n$.
h. Based on g, what happens to the precision of the estimate of $s^2$ as: 1) $k$ increases, 2) $n-k$ increases, and 3) we vary $n$ and $k$ but keep $n-k$ constant?


Suppose now instead that 95% confidence interval for a regression coefficient $\beta_i$ is $1 \leq \beta_i \leq 20$, and that we again have 100 observations and 3 predictors.

i. What is $\hat{\beta}_i$ in this case?
j. What is the standard error of $\hat{\beta}_i$?
k. Would $\hat{se}(\hat{\beta}_i)=\hat{se}(\hat{\beta}_j)$ be the same for $i\neq j$? Explain your answer.

### Linear combinations of regression coefficients

Consider the following regression model:

$$
Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3 + \beta_4 X_4 + e,
$$

where the errors are iid $N(0, \sigma^2)$.

Suppose that $\hat{\symbf{\beta}}' = \begin{pmatrix} 2.5 & 1.8 & 0.7 & 0.3 & 1 \end{pmatrix}$ is the vector of estimated coefficients, and the estimated variance-covariance matrix of $\hat{\symbf{\beta}}$ is

$$
\begin{pmatrix}
0.25 & 0.12 & 0.05 & 0.02 & 0.01 \\
0.12 & 0.36 & 0.08 & 0.03 & 0.02 \\
0.05 & 0.08 & 0.16 & 0.04 & 0.03 \\
0.02 & 0.03 & 0.04 & 0.09 & 0.05 \\
0.01 & 0.02 & 0.03 & 0.05 & 0.04
\end{pmatrix}.
$$

a. Write the null and alternate hypotheses for testing that the effect of $X_1$ is the same as the effect of $X_2$.
b. Write the null and alternate hypothes for testing that the fitted value when $X_1=2$ and $X_2=3$ is the same as the fitted value when $X_1=3$ and $X_2=2$, for $X_3=X_4=0$ in both cases.
c. Write the $(1-\alpha)100\%$ confidence for $\beta_3 - \beta_1$ (note: different regression coefficients from above), for sample size $n$.
d. Suppose that $s^2=0.01$. What is $(\symbf{X}'\symbf{X})^{-1}_{5,5}$?
e. Test the null hypothesis that $\beta_4 = 0$ at the 5% level. Suppose that the sample size is $100$, and that $t_{95}^{0.95} \approx 1.66$ and that $t_{95}^{0.975} \approx 1.96$.
f. Provide the formula for the calculation of the p-value, in this case.
g. Provide the general formula for the calculation of the p-value.
h. Provide the formula for the calculation of the p-value, in this case, where the null hypothesis is $\beta_4 \leq 0$.
i. Provide the general formula for the calculation of the p-value, where the null hypothesis is $\beta_4 \leq 0$.
j. Provide the general formula for the calculation of the p-value, where the null hypothesis is $\beta_4 \geq 0$.

### Interpreting confidence intervals and P-values

Suppose the 95% confidence interval for a regression coefficient $\beta_i$ is $-0.2 \leq \beta_i \leq 0.5$.

a. What does this confidence interval imply about the statistical significance of the predictor $X_i$?

b. Can we conclude that $X_i$ has no effect on the response variable? Explain your reasoning.

c. Suppose we were further told that effects of absolute size 1 in magnitude are considered negligible (small enough to not be relevant). How would this information affect your interpretation of the confidence interval? If so, how?

d. Imagine instead that effects of size 0.1 in magnitude are considered large. How would this information affect your interpretation of the confidence interval? If so, how?

Now, consider a different regression coefficient.

e. Suppose that the p-value is large. What does that tell you?
f. Does the large p-value rule out an effect?
g. Imagine that the sample size were very large. What would be your likely explanation for why the p-value is large?
h. Imagine that the sample size were very small. How would that affect your above explanation?
i. Imagine that you were told that a drug to reduce blood pressure had a p-value of 0.0001. What would you do with that information?
j. Suppose that you were now told that the effect is negative (i.e. it reduces blood pressure). What would you do now?
k. Suppose that, in another study, a drug to reduce weight has confidence interval of $-7$ to $-0.5$ (kg). What would you do with that information?

### Difference between nominal and empirical coverage

a. Define nominal coverage and empirical coverage of a confidence interval.

b. Why is it important for the empirical coverage to match the nominal coverage in the context of decision-making?

c. Suppose you have a confidence interval for which the empirical coverage is known. Do you still need the nominal coverage?

d. Suppose you have a confidence interval for which the nominal coverage is known. Do you still need (at least some idea of) the empirical coverage?

e. For situations in which you only have the nominal coverage, what does the answer to d suggest?

### Using t-Distribution vs. Normal Distribution in Confidence Intervals

a. When constructing confidence intervals for regression coefficients, why do we use the t-distribution instead of the normal distribution?

b. What factor of your analysis makes it particularly important to use the t-distribution?

### Critiquing Interpretations of Regression Results

Suppose we fit a linear regression model of revenue against TV and radio advertising spending, obtaining $\hat{\beta}_{\text{TV}} = 2$, with a standard error of $1$, and a p-value of 0.1.

a. Critique the following statement: "We estimated $\beta_{\text{TV}} = 2$; therefore, TV advertising spend affects revenue."

b. If the 95% confidence interval for $\beta_{\text{TV}}$ is $0.0$ to $4.0$, what does this tell us about the effect of TV advertising?

\newpage

# Week Four

## ANOVA

### Categorical Variables

#### Questions Regarding a Particular Model

Suppose that we fit a model of blood pressure against age and sex, $\symbf{Y} = \symbf{1}\beta_0 + \symbf{X}_{\mathrm{Age}} \beta_{\mathrm{Age}} + \symbf{X}_{\mathrm{Sex}} \beta_{\mathrm{Sex}} + \symbf{\epsilon}$. We have that $\symbf{X}_{\mathrm{Sex}}$ has a value of 1 if the individual is male and 0 if they are female, and that $\symbf{X}_{\mathrm{Age}}$ is age in years since birth.

##### Initial Understanding

a. For which values of $X_{\mathrm{Age}}$ and $X_{\mathrm{Sex}}$ is $\beta_0$ the mean of $Y$? What are the characteristics of the individual for which $\beta_0$ is the mean? Finally, what is the interpretation of $\beta_{\mathrm{Sex}}$ and $\beta_{\mathrm{Age}}$?
b. Let's say that (purely for part b) instead, $\symbf{X}_{\mathrm{Sex}}$ has a value of 1 if the individual is female and 0 if male. What are the characteristics of the individual for which $\beta_0$ is the mean? In addition, what is the interpretation of $\beta_{\mathrm{Sex}}$? How does the sign of $\beta_{\mathrm{Sex}}$ change, relative to part a?
c. Suppose we subtract 20 from age (purely for part c).  What are the characteristics of the individual for which $\beta_0$ is the mean? In addition, what is the interpretation of $\beta_{\mathrm{Age}}$?

##### Alternative Coding

a. Suppose that instead of coding $\symbf{X}_{\mathrm{Sex}}$ using 0 and 1, we use -1 and 1. Let's say that we use -1 for females and 1 for males.
  - i. What is interpretation of $\beta_0$?
  - ii. What is the interpretation of $\beta_{\mathrm{Sex}}$?
b. Let's say that instead of coding $\symbf{X}_{\mathrm{Sex}}$ using 0 and 1, we use 0 and 0.5 where if an individual is female they have a value of 0 and if they are male they have a value of 0.5. 
  - i. What is the interpretation of $\beta_0$?
  - ii. What is the interpretation of $\beta_{\mathrm{Sex}}$?

#### Miscellaneous

a. Imagine we model some response against a three-level categorical variable. Suppose that we use the 0-1 encoding, as usual.  What is the correlation between $\hat{\beta}_1$ and $\hat{\beta}_2$, the coefficients for the effects of the second and third levels of the categorical variable relative to the first?
b. Suppose that we fit a model without an intercept and only with one categorical variable. Suppose also that we include one column per level of the categorical variable (so, if it has thre levels, then $\symbf{X}$ has three columns). Are the columns of $\symbf{X}$ linearly independent?

## Model checking

### Introduction

a. What is an outlier?
b. What is a high-leverage point?
c. In the plots below, add in a point and draw the corresponding fitted model line (approximately) such that the point would be i) a non-outlying, low-leverage point, ii) an outlying, low-leverage point, iii) an outlying, high-leverage point and iv) a non-outlying, high leverage point. Note that for iii, you will have to draw two points.

```{r}
#| include: false
set.seed(4)
plot_tbl <- tibble::tibble(x = seq(-20, 20, by = 4))
plot_tbl <- plot_tbl |>
  dplyr::mutate(y = rnorm(11, 0, 2))
p_points <- ggplot(plot_tbl) +
  geom_point(aes(x, y), alpha = 0.8) +
  cowplot::theme_cowplot() + 
  cowplot::background_grid(major = "xy") +
  lims(x = c(-77, 77), y = c(-30, 30)) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    x = "Explanatory varible",
    y = "Response variable"
  ) +
  theme(plot.title = element_text(size = 8)) +
  theme(axis.title = element_text(size = 7))

p_grid <- cowplot::plot_grid(
  p_points + ggtitle("i: Non-outlying, low leverage"),
  p_points + ggtitle("ii: Outlying, low leverage"),
  p_points + ggtitle("iii: Outlying, high leverage"),
  p_points + ggtitle("iv: Non-outlying, high leverage"),
  ncol = 2
)

cowplot::ggsave2(
  filename = projr::projr_path_get("cache", "p-grid-outliers.pdf"),
  plot = p_grid, width = 12, height = 12, units = "cm"
)

```

```{r}
#| results: asis
#| echo: false
pander::pandoc.image(
  projr::projr_path_get("cache", "p-grid-outliers.pdf")
)
```

### Leverage

a. How do we calculate the leverage for the $i$-th observation?
b. What is the maximum and minimum leverage for any point?
c. What is the average leverage for any point?
d. In a model with one predictor, what is the average leverage?
e. In a model with $n$ predictors, what is the average leverage?
f. Show that each fitted value is a linear combination of all response values.
g. How much does the $i$-th observation's own response ($Y_i$) contribute its fitted value ($\hat{Y}_i$)?
h. What is the fitted value if $h_i=1$?
i. A given observation has a response value of 10 when all other responses lie between 0 and 1. Does it have high leverage?





\newpage

# Week Five

## Prediction

### Prediction Intervals

a. What are the source(s) of variability when calculating the fitted values?
b. What are the source(s) of variability when sampling from $\symbf{Y}=\symbf{x}'\symbf{\beta} + \symbf{\epsilon}$? (Note: in this hypothetical case, you know $\symbf{\beta}$ and $\sigma^2$)
c. What are the sources of variability when calculating the prediction interval? 
d. When deriving the formula for the width of the prediction interval, we use the statistic $Z=Y_f-\hat{Y}_f$, where $Y_f$ is the future value of the response and $\hat{Y}_f$ is the fitted value for that future response. Why does it follow that $\mathrm{Var}[Z] = \mathrm{Var}[Y_f] + \mathrm{Var}[\hat{Y}_f]$?
e. Describe a scenario in which the prediction interval would be useful in decision-making.
f. For a given sample size, the prediction interval bands (i.e. the outer bounds) will be curved outwards. As the sample size increases, the bands will straighten. Why is this the case?
g. Show that the prediction interval for a model of only an intercept becomes narrower as the sample size increases.

### Variable Selection Algorithms

Which algorithm is more appropriate when you have a large number of predictors and a small sample size: forward selection or backward elimination?

### Variable Selection Criteria

For reference's sake, BIC = $n\mathrm{ln}(RSS) + k\mathrm{ln}(n) + n\ln(n)$ and AIC = $n\mathrm{ln}(SSE) + 2k$ and AIC = $n\mathrm{ln}(\mathrm{RSS}) + 2k+ n\ln(n)$. 

a. Assume $n=5$. Would BIC or AIC tend to include more variables?
b. Assume $n=1000$. Would BIC or AIC tend to include more variables?
c. During forward selection using the AIC, by how much does an additional variable have to reduce the RSS in order to merit inclusion?

### Hypothesis Testing

Suppose you use some selection procedure and criterion to select a model, where $n=1000$ and $k=200$. Let's say that the selection procedure chooses a final model with five variables, all of which have p-values in the range 0.04 to 0.05. How persuasive are the p-values that these variables actually have non-zero coefficients?

### Training/Test Split Size

Describe the trade-offs made when choosing the size of the split (e.g. 75/25, 80/20) when splitting the data into training and test sets. Consider how the size of the training data set affects the precision of the estimates of the regression coefficients (and consequently the fitted values), and how the size of the test data set affects the precision of the estimates of the test error. 

## Transformations

### Transforming Y or X

#### Transforming Y

a.
Suppose we apply a logarithmic transformation to the dependent variable $Y$ in the model $\log(Y) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \epsilon$. How do you interpret the coefficient $\beta_1$ in terms of the transformed $Y$ variable?

b.
Continuing with the logarithmic transformation of $Y$, how would you interpret $\beta_1$ in terms of the original $Y$ variable?

c. Suppose instead that we apply a square root transformation to the dependent variable $Y$ in the model $\sqrt(Y) = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \epsilon$. How do you interpret the coefficient $\beta_1$ in terms of the transformed $Y$ variable?

d. Continuing with the square root transformation of $Y$, how would you interpret $\beta_1$ in terms of the original $Y$ variable?

e. After applying a square root or log transformation to $Y$, what happens to the variance of the residuals at varying levels of the fitted values? Is this a good or bad thing?

#### Transforming X

a.
Suppose you apply a logarithmic transformation to the independent variable $X_1$ in the model $Y = \beta_0 + \beta_1 \log(X_1) + \beta_2 X_2 + \epsilon$. How do you interpret the coefficient $\beta_1$ in terms of $Y$?

b. If we apply the squared transformation to $X_1$, what relationship do we expect between $X_1$ and $Y$?

c.
After applying a squared transformation to $X_1$ (i.e., $X'_1 = X_1^2$) in the regression model $Y = \beta_0 + \beta_1 X'_1 + \beta_2 X_2 + \epsilon$, how would you interpret the coefficient $\beta_1$ in relation to $Y$?

d. If we apply the squared transformation to $X_1$, what relationship do we expect between $X_1$ and $Y$?

e. 
What effect does transforming only $X_1$ have on the variance of the residuals at varying levels of the fitted values?

### Interaction terms

#### Fitted Values and Interpretation

Suppose again that we again fit a model of blood pressure against age and sex, but this time we include an interaction term, i.e. $\symbf{Y} = \symbf{1}\beta_0 + \symbf{X}_{\mathrm{Age}} \beta_{\mathrm{Age}}  + \symbf{X}_{\mathrm{Sex}} \beta_{\mathrm{Sex}} + + \symbf{X}_{\mathrm{Sex}:\mathrm{Age}} \beta_{\mathrm{Sex}:\mathrm{Age}} + \symbf{\epsilon}$. We have that $\symbf{X}_{\mathrm{Sex}}$ has a value of 1 if the individual is male and 0 if they are female, and that $\symbf{X}_{\mathrm{Age}}$ is age in years since birth.

a. What is the expected response for a female aged $x$ years?
a. What is the expected response for a male aged $x$ years?
c. What is the interpretation of $\beta_{\mathrm{Sex}:\mathrm{Age}}$?
d. What is the interpreation of $\beta_{\mathrm{Age}}$?
e. What is the interpreation of $\beta_{\mathrm{Age}} + \beta_{\mathrm{Sex}:\mathrm{Age}}$?

#### Estimation, Interpretation and Hypothesis Testing

```{r}
#| include: false
set.seed(21)
n <- 30
x_vec_1 <- runif(n)
g_vec <- sample(c("A", "B", "C"), replace = TRUE, size = n)
y_vec <- 5 + 
  (g_vec == "A") * 3 +
  (g_vec == "C") * 2 +
  2 * x_vec_1 +
  (g_vec == "B") * x_vec_1 * 2 +
  (g_vec == "C") * x_vec_1 * 3 +
  rnorm(n)
data_tbl <- data.frame(
    X = x_vec_1,
    G = g_vec,
    y = y_vec
)
fit <- lm(y ~ G * X, data = data_tbl)
```

Suppose we fit the following model:

```{r}
#| eval: false
#| echo: true
fit <- lm(y ~ G * X, data = data_tbl)
```

and get the following estimated coefficients:

```{r}
#| echo: true
coef(fit) |> signif(2)
```

##### Model Structure

a. 
State the model formula for a regression model with a three-level categorical variable $G$ (levels A, B and C, with A as the reference level) interacting with a continuous variable $X$.

b.
What is the design matrix row for an observation with $G = A$ and $X = 0$?

c.
What is the design matrix row for an observation with $G = C$ and $X = 0$?

d. 
What is the design matrix row for an observation with $G = A$ and $X = 3$?

e. 
What is the design matrix row for an observation with $G = B$ and $X = 3$?

##### Fitted Values

a. 
What is the fitted value for each level of $G$ at $X = 0$?

b.  
What is the fitted value for each level of $G$ at $X = 1$?

##### Effects

a.  
State the estimated difference in the effect of $X$ between groups A and B.

b. 
State the estimated difference in the effect of $X$ between groups A and C.

c.
State the estimated difference in the effect of $X$ between groups B and C.

d. 
State the estimated effect of $X$ for Group A.

e.
State the estimated effect of $X$ for Group B.

f.
State the estimated effect of $X$ for Group C.

g.
State the estimated effect of belonging to Group B compared to Group A at $X = 0$.

h.
State the estimated effect of belonging to Group B compared to Group A at $X = 4$.


i.
At what value of $X$ is the fitted value the same for groups A and B?

j.
State the estimated effect of belonging to Group C compared to Group B at $X = 0$.

k.
State the estimated effect of belonging to Group C compared to Group B at $X = 4$.

##### Plot

Plot the fitted values for each group as a function of $X$, for $X$ ranging from 0 to 4.
(Note: I won't ask you to do this in a test context, but this rough (if by hand) drawing for practice is a good idea.)

##### Hypothesis Testing

a.  
State the null hypothesis for testing whether the effect of $X$ is the same between groups A and B.

b.
State the null hypothesis for testing whether the effect of $X$ is the same between groups A and C.

c.
State the null hypothesis for testing whether the effect of $X$ is the same between groups B and C.

d.
State the null hypothesis for testing whether the effect of $X$ is the same across all groups.

e. 
State the null hypothesis for testing whether $X$ has an effect for Group A.

f.
State the null hypothesis for testing whether $X$ has an effect for Group C.

g.
State the null hypothesis for testing whether $X$ has an effect for any group.

h.
State the null hypothesis for testing whether the intercept is zero for Group A.

i.
State the null hypothesis for testing whether the intercept is zero for Group C.

j.
State the null hypothesis for testing whether the intercept is zero for any group.

k.
State the null hypothesis for testing whether the effect of Group B at $X=0$ is zero.
Note: this is equivalent to testing whether group A and group B have the same effect at $X=0$.

l.
State the null hypothesis for testing whether the effect of Group B at $X=3$ is zero.
Note: this is equivalent to testing whether group A and group B have the same effect at $X=3$.

m.
State the null hypothesis for testing whether the effect of groups B or C is zero at $X=x$.
Note: this is equivalent to testing whether group A has the same effect as group B or C at $X=x$.

#### Calculations and Inference

Suppose we fit a model of crop yield against the amount of fertilizer used and the type of crop, incorporating an interaction term to assess whether the type of crop modifies the effect of fertilizer on yield.
We model this as

$$
\symbf{Y} = \symbf{1}\beta_0 + \symbf{X}_{\mathrm{Crop}} \beta_{\mathrm{Crop}} + \symbf{X}_{\mathrm{Fertilizer}} \beta_{\mathrm{Fertilizer}} + \symbf{X}_{\mathrm{Crop}:\mathrm{Fertilizer}} \beta_{\mathrm{Crop}:\mathrm{Fertilizer}} + \symbf{\epsilon}.
$$

Here, $\symbf{X}_{\mathrm{Crop}}$ takes a value of 0 if the crop is corn and 1 if it is wheat, and $\symbf{X}_{\mathrm{Fertilizer}}$ is the amount of fertilizer applied.

Here we generate some data, and fit the model.
Of course, IRL (in real life; get with it, Gen Z) we would not know the true model form, but we do here.

```{r}
set.seed(2106)
n_obs <- 60
crop_vec <- sample(0:1, size = n_obs, replace = TRUE)
fertilizer_vec <- rnorm(n_obs)
yield_vec <- 100 + 10 +
  fertilizer_vec * ifelse(crop_vec, 5, -5) +
  rnorm(n_obs, mean = 0, sd = 5)
yield_tbl <- tibble::tibble(
  yield = yield_vec, crop = crop_vec, fertilizer = fertilizer_vec
  )
fit_yield <- lm(yield ~ crop*fertilizer, data = yield_tbl)
```

Applied to the dataset, we have the following summary output:

```{r}
#| echo: false
summary(fit_yield)
```

Here is $(\symbf{X}'\symbf{X})^{-1}$:

```{r}
#| echo: false
X_mat <- model.matrix(fit_yield)
XtX_mat_inv <- solve(t(X_mat) %*% X_mat)
XtX_mat_inv |> signif(2)
```

Here is the variance-covariance matrix:

```{r}
#| echo: false
vcov(fit_yield) |> signif(2)
```


a. Use the estimated effects, $\hat{\symbf{\beta}}$, to conclude regarding the effect of fertilizer. Ignore inference (p-values and confidence intervals), and simply interpret the estimated effects.
b. Test the hypothesis that the effect of fertilizer is the same for both crops at the 5% level of significance.
c. Test the hypothesis that there is an effect of fertilizer at the 5% level of significance.
d. Calculate a 95% confidence interval for the effect an additional unit fertilizer for corn. Here are some (potentially) useful quantiles: $z^{0.95}=1.64$, $z^{0.975}=1.96$, $t^{0.95}_=1.67$ and $t^{0.975}=2$.
e. Calculate a confidence interval for the effect of an additional unit of fertilizer for wheat.
f. Motivate why $\beta_{\mathrm{Crop}:\mathrm{Fertilizer}}$ can be considered an effect modification term.
g. Test the hypothesis that there is an effect of fertilizer or crop type.
h. Suppose that crop were a three-level categorical variable, with additional crop type barley also included. How would the model form need to change? If so, how? Note that interactions are still included.

Continue with the three-level categorical variables for crop type.

i. How could we test whether additional fertilizer has the same effect on barley as corn?
j. How could we test whether additional fertilizer has the same effect on barley as wheat?

---

## Gauss-Markov theorem

### Basics

a. For the estimator $\hat{\symbf{\beta}} = (\symbf{X}'\symbf{X})^{-1}\symbf{X}\symbf{Y}$, in what sense is it:
  - i. Unbiased?
  - ii. Linear?
  - iii. Best, according to the Gauss-Markov theorem?

### Validity

a. Does the Gauss-Markov theorem hold when normally distributed?
b. Does the Gauss-Markov theorem hold only when the responses are normally distributed?
c. Does the Gauss-Markov theorem apply to models where the variance increases with the mean?

### Deeper Understanding

a. Suppose that we fit a model but exclude a relevant variable (i.e. one for which $\beta_i\neq 0$) that is also correlated with an explanatory variable (that is included). For example, let's say that the true model is $\symbf{Y}=\symbf{\beta}_0 + \symbf{X}_1\beta_1 + \symbf{X}_2\beta_2 + \symbf{\epsilon}$ for $\beta_i \neq 0 \; \forall \; i$ where $\symbf{X}_1$ and $\symbf{X}_2$ are correlated, but we fit $\symbf{Y}=\symbf{\beta}_0 + \symbf{X}_1\beta_1 + \symbf{\epsilon}$. Does the Gauss-Markov theorem apply?
b. Suppose we simply estimate $\symbf{\beta}$ by $\symbf{0}$ (i.e. the estimator is literally $\symbf{0}$, no matter what the data are). Suppose also that the true value of $\symbf{\beta}$ is $\begin{pmatrix} 1 \\ 1 \end{pmatrix}'$, and that $\mathrm{Var}[\hat{\symbf{\beta}}]= \begin{bmatrix} 5 & 0 \\ 0 & 5 \end{bmatrix}$$  In what sense is this better than $\hat{\symbf{\beta}} = (\symbf{X}'\symbf{X})^{-1}\symbf{X}\symbf{Y}$, and in what sense it is worse?
c. Suppose we model number of zebra against size of a game park. If we fit a linear regression model, would $\hat{\symbf{\beta}}$ be BLUE by the Gauss-Markov theorem?

\newpage

# Week Six

## Bootstrapping

### Sampling with Replacement

a. What is meant by "sampling with replacement" in the context of bootstrapping? 

b. Why is sampling with replacement essential for the bootstrapping process? 

### Sampling with replacement

a. What is meant by "sampling with replacement" in the context of bootstrapping? 
b. Why is sampling with replacement essential for the bootstrapping process? 

### Role of Bootstrapping in Inference

a. Explain the primary purpose of bootstrapping in statistical inference.
b. How does bootstrapping differ from traditional, more theoretical methods when performing inference? 
c. Explain why bootstrapping may be preferred to traditional methods when conducting inference (e.g. constructing confidence intervals)?

### Limitations of traditional methods

a. What assumptions does the traditional $t$-interval method make about the sampling distribution of the estimator? 

b. How does bootstrapping address the limitations of these assumptions? 

### Asymmetric bootstrap confidence intervals

a. Name and briefly describe two bootstrap methods for constructing confidence intervals that do not assume symmetry in the sampling distribution. 

b. In what scenarios might the bootstrap $t$-interval perform better than the percentile bootstrap method? 

### Non-parametric Nature of Bootstrapping

Explain why bootstrapping is considered a non-parametric method. 

---

## Principal Component Analysis

### Calculation and Application

We'll load and prepare a dataset for PCA.
You will be able to answer the questions without writing or running any code, but I'll provide the code to load the data and (in the solutions) to perform the calculations.

#### Preparation

##### Setup

First, we attach relevant packages and load the dataset:

```{r}
if (!requireNamespace("DataTidy23RodoHonsMult", quietly = TRUE)) {
  if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
  }
  remotes::install_github("MiguelRodo/DataTidy23RodoHonsMult@2025")
}

# Load libraries
library(ggplot2)
library(dplyr)
library(DataTidy23RodoHonsMult)
data("data_tidy_mali", package = "DataTidy23RodoHonsMult")
```

##### Data

Survey data to assess options for enhancing food security through sustainable use of natural resources in the Sikasso region of Mali is available in `data_tidy_mali` (instructions to get it below).

Variables include:​

- family (total number of individuals in household)​
- distrd (distance in kilometres to nearest passable road)​
- cotton (hectares of cotton planted in year 2000)​
- maize (hectares of maize planted in year 2000)​
- sorg (hectares of sorg planted in year 2000)​
- millet (hectares of millet planted in year 2000)​
- bull (total number of bullocks or draft animals)​
- cattle (total)​
- goats (total)

Here is the top of the dataset:

```{r}
head(data_tidy_mali |> signif(2))
```

Let's remove some outlying values:

```{r}
data_tidy_mali <- data_tidy_mali |>
  filter(dist_rd < 500, family < 100, cattle < 50)
```

We'll also select just five of the variables, to make the formulae easier to write:

```{r}
data_tidy_mali <- data_tidy_mali |>
  select(family, dist_rd, cotton, maize, cattle)
```

Let's scale the variables to have unit variance, as they're in different units:

```{r}
# Standardize the selected variables
data_tidy_mali <- scale(data_tidy_mali)

# Convert to data frame for easier handling
data_tidy_mali <- as.data.frame(data_tidy_mali)
```

##### Eigen-Decomposition of the Covariance Matrix

The principal component loadings (the coefficient vectors) and their variances are provided by the eigen-decomposition of the covariance matrix.

So, we calculate the covariance matrix and perform the eigen-decomposition:

```{r}
# Compute the covariance matrix
cov_matrix <- cov(data_tidy_mali)
cov_matrix |> signif(2)
```

```{r}
# Perform eigen-decomposition
eig_decomp <- eigen(cov_matrix)
# eigenvalues
eig_val_vec <- eig_decomp$values
eig_val_vec <- eig_val_vec |> setNames(paste0("EigenValue", 1:length(eig_val_vec)))
eig_val_vec |> signif(2)
# eigenvectors
eig_vec_mat <- eig_decomp$vectors
colnames(eig_vec_mat) <- paste0("EigenVector", 1:ncol(eig_vec_mat))
rownames(eig_vec_mat) <- colnames(cov_matrix)
eig_vec_mat |> signif(2)
```

Note that the eigenvectors are along the columns of the matrix `eig_vec_mat`, e.g. the first column is the first eigenvector.

#### Questions

a. Why does the covariance matrix have ones along the diagonal?
b. Calculate the first principal component score for an observation with values `family = 1`, `distrd = 0.5`, `cotton = 1`, `maize = 0`, and `cattle = 0`.
c. Calculate the third principal component score for the same observation.
d. What is the variance of the first principal component?
e. What is the variance of the fourth principal component?
f. What is the total variance of the original variables?
g. What is the total variance of the principal components?
h. What is the proportion of variance explained by the first principal component?
i. What is the proportion of variance explained by the fifth principal component?
j. What is the proportion of variance explained by the first two principal components?
k. What is the proportion of variance explained by all five principal components?
l. Construct a scree plot (roughly, if doing so by hand).
m. Using the scree plot, motivate the number of principal components to retain.
n. Using proportion of variation explained, motivate the number of principal components to retain.
o. Interpret the first principal component.
p. Interpret the second principal component.

### Deeper Understanding

a. Suppose that $\symbf{X}$ is not full column rank. What is the variance of the lowest-variance principal component?

b. Suppose that, instead of regressing $\symbf{Y}$ against $\symbf{X}$, you regress $\symbf{Y}$ against the principal components. In other words, instead of fitting $\symbf{Y} = \symbf{X}\hat{\symbf{\beta}}$, you fit $\symbf{Y} = \symbf{S}\hat{\symbf{\beta}}$ where $\symbf{S} = \symbf{X}\symbf{V}$ for $\symbf{V}$ the eigenvectors of $\symbf{X}'\symbf{X}$. What is the correlation between the regression coefficients?

c. Suppose that $\mathrm{Cov}[\symbf{X}] = \symbf{X}'\symbf{X} = \begin{bmatrix} 2 & 0 \\ 0 & 1 \end{bmatrix}$. What are the principal components? What are their variances?

d. Explain why providing a coefficient vector of length other than one is not correct in the context of PCA, and how to fix it.

e. Why do we choose the first $k$ principal components rather than some other subset (for example, the 1st, 5th, and 8th)? Why do we not select arbitrary subsets?

f. Why not just choose all the principal components if dimensionality reduction isn't a pressing concern?
- Discuss why the last components are typically "noise," as all the signal is captured in earlier components.
- Explain why they are probably useless.

g. If you were to plot two principal components to accurately represent the data, which would you choose, and why?

h. What is the advantage of plotting principal components over plotting all the original variables?

i. Describe the potential problem PCA faces when different variables are on different scales, and provide generic scenarios when this may arise.

j. Describe how to fix the above problem (standardize the variables).

