---
title: "ISOMAP in R"
subtitle: "STA5077Z coding walkthrough"
author: "Miguel Rodó"
format:
  html:
    toc: true
    number-sections: true
    code-fold: false
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This walkthrough applies ISOMAP to two very simple two-dimensional datasets.

In both examples, the data contain two multivariate normal clusters:

1. two roughly spherical clusters that are fairly well separated; and
2. two long clusters that lie close together.

We ask ISOMAP for two output dimensions so that we can compare the original data with the ISOMAP representation directly.

The cluster labels are used only to colour the plots. They are **not** supplied to ISOMAP.

## Setup

Install the required packages once.

```{r}
#| label: install-isomap-packages
#| eval: false
pkg_vec <- c("tidyverse", "vegan")
for (x in pkg_vec) {
  if (!requireNamespace(x, quietly = TRUE)) {
    install.packages(x)
  }
}
```

Load the packages.

```{r}
#| label: load-isomap-packages

library(tidyverse)
library(vegan)
```

The main ISOMAP code will be:

```r
fit <- vegan::isomap(
  dist(x),
  ndim = 2,
  k = 12
)
```

Here:

- `dist(x)` calculates the ordinary Euclidean distances;
- `k = 12` constructs a 12-nearest-neighbour graph;
- ISOMAP replaces the original distances with graph shortest-path distances; and
- `ndim = 2` asks classical MDS for two output dimensions.

# Scenario 1: spherical clusters

## Simulate the data

We first generate two spherical clusters that are separated, but not completely isolated from one another.

```{r}
#| label: simulate-spherical-clusters

set.seed(5077)

n_per_cluster <- 100

spherical_covariance <- diag(
  c(
    0.7^2,
    0.7^2
  )
)

spherical_cluster_1 <- MASS::mvrnorm(
  n = n_per_cluster,
  mu = c(-1.2, 0),
  Sigma = spherical_covariance
)

spherical_cluster_2 <- MASS::mvrnorm(
  n = n_per_cluster,
  mu = c(1.2, 0),
  Sigma = spherical_covariance
)

spherical_tbl <- bind_rows(
  as_tibble(
    spherical_cluster_1,
    .name_repair = ~ c("x1", "x2")
  ) |>
    mutate(cluster = "Cluster 1"),
  as_tibble(
    spherical_cluster_2,
    .name_repair = ~ c("x1", "x2")
  ) |>
    mutate(cluster = "Cluster 2")
)

spherical_tbl
```

## Plot the original data

```{r}
#| label: fig-spherical-original
#| fig-cap: "Two approximately spherical multivariate normal clusters."
#| fig-width: 7
#| fig-height: 5

ggplot(
  spherical_tbl,
  aes(
    x = x1,
    y = x2,
    colour = cluster
  )
) +
  geom_point(
    alpha = 0.75,
    size = 2
  ) +
  coord_equal() +
  labs(
    x = expression(x[1]),
    y = expression(x[2]),
    colour = NULL
  ) +
  theme_minimal()
```

## Fit ISOMAP

ISOMAP uses only the two numerical columns.

```{r}
#| label: fit-isomap-spherical

spherical_x <- spherical_tbl |>
  select(x1, x2) |>
  as.matrix()

spherical_isomap <- vegan::isomap(
  dist(spherical_x),
  ndim = 2,
  k = 12
)
```

The single call to `isomap()` performs the main steps:

1. construct the nearest-neighbour graph;
2. calculate shortest-path distances through the graph; and
3. apply classical MDS to those distances.

## Extract the ISOMAP coordinates

```{r}
#| label: extract-spherical-isomap-scores

spherical_scores <- vegan::scores(
  spherical_isomap
) |>
  as.data.frame()

names(spherical_scores) <- c(
  "ISOMAP1",
  "ISOMAP2"
)

spherical_embedding_tbl <- spherical_scores |>
  as_tibble() |>
  mutate(
    cluster = spherical_tbl$cluster
  )

spherical_embedding_tbl
```

## Plot the ISOMAP representation

```{r}
#| label: fig-spherical-isomap
#| fig-cap: "Two-dimensional ISOMAP representation of the spherical-cluster data."
#| fig-width: 7
#| fig-height: 5

ggplot(
  spherical_embedding_tbl,
  aes(
    x = ISOMAP1,
    y = ISOMAP2,
    colour = cluster
  )
) +
  geom_point(
    alpha = 0.75,
    size = 2
  ) +
  coord_equal() +
  labs(
    x = "ISOMAP axis 1",
    y = "ISOMAP axis 2",
    colour = NULL
  ) +
  theme_minimal()
```

The orientation of the ISOMAP plot is arbitrary. It may be rotated or reflected without changing the distances represented by the configuration.

# Scenario 2: close, elongated clusters

## Simulate the data

The next two clusters are much longer and lie close together.

The covariance matrices below simply create two elongated clusters with different orientations.

```{r}
#| label: simulate-elongated-clusters

set.seed(5077)

elongated_covariance_1 <- matrix(
  c(
    1.4,  1.1,
    1.1,  1.4
  ),
  nrow = 2
)

elongated_covariance_2 <- matrix(
  c(
     1.4, -1.1,
    -1.1,  1.4
  ),
  nrow = 2
)

elongated_cluster_1 <- MASS::mvrnorm(
  n = n_per_cluster,
  mu = c(-0.4, 0),
  Sigma = elongated_covariance_1
)

elongated_cluster_2 <- MASS::mvrnorm(
  n = n_per_cluster,
  mu = c(0.4, 0),
  Sigma = elongated_covariance_2
)

elongated_tbl <- bind_rows(
  as_tibble(
    elongated_cluster_1,
    .name_repair = ~ c("x1", "x2")
  ) |>
    mutate(cluster = "Cluster 1"),
  as_tibble(
    elongated_cluster_2,
    .name_repair = ~ c("x1", "x2")
  ) |>
    mutate(cluster = "Cluster 2")
)

elongated_tbl
```

## Plot the original data

```{r}
#| label: fig-elongated-original
#| fig-cap: "Two close, elongated multivariate normal clusters."
#| fig-width: 7
#| fig-height: 5

ggplot(
  elongated_tbl,
  aes(
    x = x1,
    y = x2,
    colour = cluster
  )
) +
  geom_point(
    alpha = 0.75,
    size = 2
  ) +
  coord_equal() +
  labs(
    x = expression(x[1]),
    y = expression(x[2]),
    colour = NULL
  ) +
  theme_minimal()
```

## Fit ISOMAP

The code is exactly the same as before.

```{r}
#| label: fit-isomap-elongated

elongated_x <- elongated_tbl |>
  select(x1, x2) |>
  as.matrix()

elongated_isomap <- vegan::isomap(
  dist(elongated_x),
  ndim = 2,
  k = 12
)
```

## Extract the ISOMAP coordinates

```{r}
#| label: extract-elongated-isomap-scores

elongated_scores <- vegan::scores(
  elongated_isomap
) |>
  as.data.frame()

names(elongated_scores) <- c(
  "ISOMAP1",
  "ISOMAP2"
)

elongated_embedding_tbl <- elongated_scores |>
  as_tibble() |>
  mutate(
    cluster = elongated_tbl$cluster
  )

elongated_embedding_tbl
```

## Plot the ISOMAP representation

```{r}
#| label: fig-elongated-isomap
#| fig-cap: "Two-dimensional ISOMAP representation of the close, elongated-cluster data."
#| fig-width: 7
#| fig-height: 5

ggplot(
  elongated_embedding_tbl,
  aes(
    x = ISOMAP1,
    y = ISOMAP2,
    colour = cluster
  )
) +
  geom_point(
    alpha = 0.75,
    size = 2
  ) +
  coord_equal() +
  labs(
    x = "ISOMAP axis 1",
    y = "ISOMAP axis 2",
    colour = NULL
  ) +
  theme_minimal()
```

# Compare the two examples

In the spherical example, the local neighbourhoods mostly remain within the two visible groups.

In the elongated example, observations from different clusters can be close in the original space. The nearest-neighbour graph may therefore contain many connections between the two groups.

ISOMAP is not a clustering algorithm. It does not use the known cluster labels, and it cannot be expected to separate groups when their local neighbourhoods are strongly mixed.

The graph is therefore critical:

> ISOMAP can preserve only the geometry represented by the neighbourhood graph that we construct.

## Try a different value of $k$

The same data can give a different representation when the neighbourhood size changes.

For example:

```{r}
#| label: refit-elongated-isomap-different-k

elongated_isomap_k6 <- vegan::isomap(
  dist(elongated_x),
  ndim = 2,
  k = 6
)

elongated_isomap_k20 <- vegan::isomap(
  dist(elongated_x),
  ndim = 2,
  k = 20
)
```

A very small value of $k$ can produce a fragmented graph.

A very large value of $k$ introduces more long-range connections and moves the graph distances closer to ordinary Euclidean distances.

# Minimal template

For a new numerical dataset called `my_data`, the complete workflow is:

```{r}
#| label: minimal-isomap-template
#| eval: false

# Keep only the numerical variables.
x <- my_data |>
  select(where(is.numeric)) |>
  as.matrix()

# Fit ISOMAP.
isomap_fit <- vegan::isomap(
  dist(x),
  ndim = 2,
  k = 12
)

# Extract the coordinates.
isomap_coordinates <- vegan::scores(
  isomap_fit
)

# Plot the first two ISOMAP axes.
plot(
  isomap_coordinates[, 1],
  isomap_coordinates[, 2],
  xlab = "ISOMAP axis 1",
  ylab = "ISOMAP axis 2",
  pch = 19
)
```

The only choices that must be changed for a new analysis are usually:

- the variables included in `x`;
- the number of output dimensions, `ndim`; and
- the neighbourhood size, `k`.
