---
title: "ISOMAP: worked coding demonstration"
subtitle: "STA5077Z coding practical"
author: "Miguel Rodó"
format:
  html:
    toc: true
    number-sections: true
    code-fold: false
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

This short demonstration shows how ISOMAP is implemented in R.

We will:

1. construct and plot a nearest-neighbour graph;
2. calculate shortest-path distances through the graph;
3. pass those distances to `cmdscale()`; and
4. obtain the same result directly using `vegan::isomap()`.

The example uses observations lying on a curved one-dimensional object in two-dimensional space. ISOMAP should replace the curve with a one-dimensional representation.

## Setup

The package installation only needs to be run once.

```{r}
#| label: install-isomap-demonstration-packages
#| eval: false

install.packages(c("FNN", "igraph", "vegan"))
```

Load the packages.

```{r}
#| label: load-isomap-demonstration-packages

library(FNN)
library(igraph)
library(vegan)
```

## Create a simple curved dataset

The variable `theta` records position along the curve. It is not supplied to ISOMAP, but will be used to colour the observations so that we can see whether their ordering is recovered.

```{r}
#| label: simulate-curved-data

n <- 60

theta <- seq(
  from = 0.2,
  to = 1.8 * pi,
  length.out = n
) + 0.008 * sin(seq_len(n))

curve_x <- cbind(
  x1 = cos(theta),
  x2 = sin(theta)
)

point_colour <- hcl.colors(n, palette = "Viridis")
```

Plot the observations in their original two-dimensional space.

```{r}
#| label: fig-curved-data
#| fig-cap: "Observations lying on a curved one-dimensional object. Colour records position along the curve."
#| fig-width: 6.2
#| fig-height: 5.2

plot(
  curve_x,
  pch = 19,
  col = point_colour,
  asp = 1,
  xlab = expression(x[1]),
  ylab = expression(x[2])
)
```

The observations require two coordinates to be displayed in the original space, but their position on the curve is described by only one quantity, `theta`.

# Construct ISOMAP step by step

## Step 1: Construct a nearest-neighbour graph

We use `FNN::get.knn()` to find the nearest neighbours and `igraph::graph_from_adjacency_matrix()` to construct the graph.

Set the number of nearest neighbours to retain.

```{r}
#| label: choose-neighbourhood-size

k <- 3
```

Calculate the ordinary Euclidean distances and find the three nearest neighbours of every observation.

```{r}
#| label: find-nearest-neighbours

euclidean_distance_mat <- as.matrix(dist(curve_x))

nearest_neighbours <- FNN::get.knn(
  curve_x,
  k = k
)$nn.index
```

Convert the neighbour indices into a symmetric adjacency matrix. Two observations are connected when either observation regards the other as one of its `k` nearest neighbours.

```{r}
#| label: construct-neighbourhood-adjacency

adjacent <- matrix(FALSE, nrow = n, ncol = n)

adjacent[cbind(
  rep(seq_len(n), each = k),
  as.vector(t(nearest_neighbours))
)] <- TRUE

adjacent <- adjacent | t(adjacent)
```

Construct a weighted graph. Each retained edge is weighted by the ordinary Euclidean distance between the two connected observations.

```{r}
#| label: construct-weighted-neighbourhood-graph

curve_graph <- igraph::graph_from_adjacency_matrix(
  adjacent * euclidean_distance_mat,
  mode = "undirected",
  weighted = TRUE,
  diag = FALSE
)

igraph::is_connected(curve_graph)
```

The graph must be connected if we want a finite shortest-path distance between every pair of observations.

## Step 2: Plot the graph

An `igraph` graph is usually plotted using an automatically generated layout. Here we instead use the original observations as the layout. This lets us see whether the graph edges follow the curve.

```{r}
#| label: fig-curved-neighbourhood-graph
#| fig-cap: "The three-nearest-neighbour graph plotted at the original observation coordinates."
#| fig-width: 6.2
#| fig-height: 5.2

plot(
  curve_graph,
  layout = curve_x,
  vertex.color = point_colour,
  vertex.size = 5,
  vertex.label = NA,
  edge.color = "grey70",
  asp = 1
)
```

Only local movements along the displayed edges are now permitted.

## Step 3: Calculate graph distances

Use `igraph::distances()` to calculate the shortest-path distance between every pair of observations.

```{r}
#| label: calculate-graph-distances

graph_distance_mat <- igraph::distances(
  curve_graph,
  weights = igraph::E(curve_graph)$weight
)
```

The matrix `graph_distance_mat` is the dissimilarity matrix used by ISOMAP. It contains distances through the graph, rather than direct distances through the surrounding two-dimensional space.

For example, compare the two types of distance between the first and last observations.

```{r}
#| label: compare-selected-distances

c(
  direct_euclidean_distance = euclidean_distance_mat[1, n],
  graph_distance = graph_distance_mat[1, n]
)
```

The endpoints are relatively close through the surrounding space, but far apart when movement is restricted to the curve.

## Step 4: Pass the graph distances to `cmdscale()`

ISOMAP now applies classical MDS to the graph-distance matrix.

```{r}
#| label: fit-isomap-with-cmdscale

isomap_step_by_step <- cmdscale(
  as.dist(graph_distance_mat),
  k = 1,
  eig = TRUE
)

step_by_step_coordinate <- isomap_step_by_step$points[, 1]
```

Plot the resulting one-dimensional configuration.

```{r}
#| label: fig-step-by-step-isomap
#| fig-cap: "One-dimensional ISOMAP configuration obtained from graph distances and cmdscale()."
#| fig-width: 7
#| fig-height: 2.5

plot(
  step_by_step_coordinate,
  rep(0, n),
  pch = 19,
  col = point_colour,
  yaxt = "n",
  xlab = "ISOMAP coordinate",
  ylab = ""
)
```

The smooth progression in colour shows that the one-dimensional coordinate has recovered position along the curve. The direction could be reversed without changing the result.

# Fit ISOMAP with one function

The `vegan` package performs the same three main operations inside a single function:

1. retain local distances;
2. replace the remaining distances with shortest-path distances; and
3. apply classical MDS.

```{r}
#| label: fit-isomap-with-vegan

isomap_vegan <- vegan::isomap(
  dist(curve_x),
  k = k,
  ndim = 1
)

vegan_coordinate <- isomap_vegan$points[, 1]
```

Plot the fitted coordinate.

```{r}
#| label: fig-vegan-isomap
#| fig-cap: "One-dimensional ISOMAP configuration obtained directly using vegan::isomap()."
#| fig-width: 7
#| fig-height: 2.5

plot(
  vegan_coordinate,
  rep(0, n),
  pch = 19,
  col = point_colour,
  yaxt = "n",
  xlab = "ISOMAP coordinate",
  ylab = ""
)
```

Finally, compare the two fitted coordinates.

```{r}
#| label: compare-isomap-implementations

abs(cor(
  step_by_step_coordinate,
  vegan_coordinate
))
```

We use the absolute correlation because an MDS axis may be reflected. A correlation close to one shows that the two approaches give the same one-dimensional representation, apart from arbitrary orientation and minor implementation details.

# What to notice

The complete calculation was

$$
\text{observations}
\longrightarrow
\text{nearest-neighbour graph}
\longrightarrow
\text{shortest-path distances}
\longrightarrow
\texttt{cmdscale()}.
$$

The last two steps can be seen directly in the code:

```{r}
#| label: show-essential-isomap-code
#| eval: false

graph_distance_mat <- igraph::distances(curve_graph)

isomap_fit <- cmdscale(
  as.dist(graph_distance_mat),
  k = 1
)
```

For routine use, the full calculation can instead be written as:

```{r}
#| label: show-one-function-isomap-code
#| eval: false

isomap_fit <- vegan::isomap(
  dist(curve_x),
  k = 3,
  ndim = 1
)
```

In the main practical, we will investigate what happens when the neighbourhood graph is too sparse, too dense or disconnected.
