---
title: "ISOMAP: Changing the Distances Before MDS"
author: Miguel Rodo
date: "2026-07-31"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
---

```{r}
#| label: setup
#| include: false

library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(cowplot)
library(ggrepel)
library(igraph)

knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "80%"
)

simulate_swiss_roll <- function(
  n = 500,
  seed = 5077,
  noise_sd = 0
) {
  set.seed(seed)

  t <- 1.5 * pi * (1 + 2 * runif(n))
  height <- runif(n, min = 0, max = 20)

  x1 <- t * cos(t) + rnorm(n, sd = noise_sd)
  x2 <- height + rnorm(n, sd = noise_sd)
  x3 <- t * sin(t) + rnorm(n, sd = noise_sd)

  # Arc length along the spiral:
  # integral sqrt(1 + t^2) dt
  arc_length <- 0.5 * (
    t * sqrt(1 + t^2) +
      asinh(t)
  )

  tibble(
    observation = seq_len(n),
    t = t,
    height = height,
    arc_length = arc_length,
    x1 = x1,
    x2 = x2,
    x3 = x3
  )
}

project_3d_to_2d <- function(
  x1,
  x2,
  x3,
  azimuth = -45,
  elevation = 25
) {
  theta <- azimuth * pi / 180
  phi <- elevation * pi / 180

  horizontal <- cos(theta) * x1 -
    sin(theta) * x3

  depth <- sin(theta) * x1 +
    cos(theta) * x3

  vertical <- -sin(phi) * depth +
    cos(phi) * x2

  tibble(
    display_x = horizontal,
    display_y = vertical
  )
}

build_knn_graph <- function(x, k) {
  x <- as.matrix(x)
  n <- nrow(x)

  if (k < 1 || k >= n) {
    stop("k must lie between 1 and n - 1.")
  }

  ambient_distance_mat <- as.matrix(
    dist(x, method = "euclidean")
  )

  search_distance_mat <- ambient_distance_mat
  diag(search_distance_mat) <- Inf

  neighbour_indicator <- matrix(
    FALSE,
    nrow = n,
    ncol = n
  )

  for (i in seq_len(n)) {
    nearest <- order(
      search_distance_mat[i, ],
      decreasing = FALSE
    )[seq_len(k)]

    neighbour_indicator[i, nearest] <- TRUE
  }

  # Symmetrise the k-nearest-neighbour relation:
  # connect i and j when either regards the other as a neighbour.
  adjacent <- neighbour_indicator |
    t(neighbour_indicator)

  edge_index <- which(
    adjacent & upper.tri(adjacent),
    arr.ind = TRUE
  )

  edge_tbl <- tibble(
    from = as.character(edge_index[, 1]),
    to = as.character(edge_index[, 2]),
    weight = ambient_distance_mat[edge_index]
  )

  graph <- igraph::graph_from_data_frame(
    edge_tbl,
    directed = FALSE,
    vertices = tibble(
      name = as.character(seq_len(n))
    )
  )

  list(
    graph = graph,
    edges = edge_tbl,
    ambient_distances = ambient_distance_mat,
    components = igraph::components(graph)
  )
}

fit_isomap <- function(x, k, ndim = 2) {
  graph_info <- build_knn_graph(x, k)

  if (graph_info$components$no != 1) {
    stop(
      paste0(
        "The neighbourhood graph has ",
        graph_info$components$no,
        " connected components."
      )
    )
  }

  graph_distance_mat <- igraph::distances(
    graph_info$graph,
    weights = igraph::E(
      graph_info$graph
    )$weight
  )

  mds_fit <- cmdscale(
    as.dist(graph_distance_mat),
    k = ndim,
    eig = TRUE
  )

  list(
    graph_info = graph_info,
    graph_distances = graph_distance_mat,
    mds = mds_fit,
    points = mds_fit$points
  )
}

align_configuration <- function(
  reference,
  target,
  allow_scaling = TRUE
) {
  reference <- as.matrix(reference)
  target <- as.matrix(target)

  reference_centre <- colMeans(reference)
  target_centre <- colMeans(target)

  reference_centred <- sweep(
    reference,
    MARGIN = 2,
    STATS = reference_centre,
    FUN = "-"
  )

  target_centred <- sweep(
    target,
    MARGIN = 2,
    STATS = target_centre,
    FUN = "-"
  )

  rotation_svd <- svd(
    crossprod(
      target_centred,
      reference_centred
    )
  )

  rotation <- rotation_svd$u %*%
    t(rotation_svd$v)

  target_aligned <- target_centred %*%
    rotation

  if (allow_scaling) {
    scale_factor <- sum(
      target_aligned *
        reference_centred
    ) /
      sum(target_aligned^2)

    target_aligned <- scale_factor *
      target_aligned
  }

  sweep(
    target_aligned,
    MARGIN = 2,
    STATS = reference_centre,
    FUN = "+"
  )
}

distance_preservation <- function(
  target_distances,
  coordinates
) {
  target <- as.vector(
    as.dist(target_distances)
  )

  fitted <- as.vector(
    dist(coordinates)
  )

  correlation <- cor(target, fitted)

  tibble(
    correlation = correlation,
    residual_variance = 1 -
      correlation^2
  )
}
```

<!--
Suggested pacing for a two-hour session:

00:00-00:12  Bridge from PCA and MDS
00:12-00:30  Curved manifolds and geodesic distance
00:30-00:48  Neighbourhood graphs and shortest paths
00:48-01:03  The ISOMAP algorithm
01:03-01:20  Swiss-roll example: PCA, classical MDS and ISOMAP
01:20-01:40  Choosing the neighbourhood size
01:40-01:52  Diagnostics and target dimension
01:52-02:00  Failure modes, workflow and recap

The more technical derivations and supplementary remarks are collapsed by default.
-->

## From PCA and MDS to ISOMAP

Principal component analysis (PCA) constructs new variables by taking linear combinations of the original variables, in order to represent a high-dimensional dataset in fewer dimensions.

Classical multidimensional scaling (MDS) differs from PCA in that it begins with pairwise dissimilarities rather than the original varibles.
But it is ultimately a linear method, and in fact is exactly equivalent to PCA when the dissimilarities are ordinary Euclidean distances.

$$
\text{PCA scores}
\quad\equiv\quad
\text{classical MDS coordinates}.
$$

So, if we have that the distance between observations is not accurately represented by the straight-line distance between them, the PCA and classical MDS will not recover the underlying structure of the data.

The key question is therefore:

> What should we do when straight-line Euclidean distances do not represent movement along the structure of the data?

Metric and non-metric MDS are non-linear extensions of MDS, and so can recover some non-linear structure.

ISOMAP is another alternative.
It relates closely to classical MDS, but addresses the issue by changing the dissimilarities before applying classical MDS.

$$
\boxed{
\text{data}
\longrightarrow
\text{neighbourhood graph}
\longrightarrow
\text{shortest-path distances}
\longrightarrow
\text{classical MDS}
}
$$

The final classical-scaling step is familiar.
The nonlinear behaviour enters through the graph-based distances supplied to it.

## Lower-dimensional objects embedded in higher-dimensional space

Ultimately, ISOMAP is going to to base the distance between two observations in the high-dimensional space on the effectively lower-dimensional objects that these observations are assumed to lie on.
We'll give a few examples to illustrate.

The first example is a plane in three-dimensional space.

```{r}
#| label: fig-plane-in-three-dimensions
#| fig-cap: "A two-dimensional plane embedded in three-dimensional space. The observations have three coordinates but vary in only two independent directions."
#| echo: false
#| fig-width: 8
#| fig-height: 6

set.seed(5077)

x1_grid <- seq(-2, 2, length.out = 30)
x2_grid <- seq(-2, 2, length.out = 30)

# The plane is x3 = 0.6 x1 - 0.4 x2.
x3_grid <- outer(
  x1_grid,
  x2_grid,
  function(x1, x2) {
    0.6 * x1 - 0.4 * x2
  }
)

plane_points <- tibble::tibble(
  x1 = runif(35, -1.8, 1.8),
  x2 = runif(35, -1.8, 1.8)
) |>
  dplyr::mutate(
    x3 = 0.6 * x1 - 0.4 * x2
  )

plane_projection <- persp(
  x = x1_grid,
  y = x2_grid,
  z = x3_grid,
  theta = 35,
  phi = 25,
  expand = 0.8,
  col = adjustcolor(
    "skyblue3",
    alpha.f = 0.45
  ),
  border = adjustcolor(
    "grey30",
    alpha.f = 0.35
  ),
  ticktype = "detailed",
  xlab = expression(x[1]),
  ylab = expression(x[2]),
  zlab = expression(x[3])
)

projected_points <- trans3d(
  x = plane_points$x1,
  y = plane_points$x2,
  z = plane_points$x3,
  pmat = plane_projection
)

points(
  projected_points,
  pch = 19,
  cex = 0.8,
  col = "dodgerblue4"
)
```

The plane is defined by

$$
x_3=0.6x_1-0.4x_2.
$$

Equivalently, its points can be written parametrically as

$$
\begin{bmatrix}
x_1 \\
x_2 \\
x_3
\end{bmatrix}
=

u
\begin{bmatrix}
1 \\
0 \\
0.6
\end{bmatrix}
+
v
\begin{bmatrix}
0 \\
1 \\
-0.4
\end{bmatrix},
\qquad u,v\in\mathbb R.
$$


There are three measured coordinates, but only two free quantities, (u) and (v).
The plane therefore is really only 2D, despite lying in 3D.

A simpler object is a line in 2D.
Suppose that the "real" distance between two points is the distance along the line.
This is the same as the straight-line distance between them, and so we can use ordinary Euclidean distance with no problem.

```{r}
#| label: prepare-straight-manifold
#| include: false

straight_tbl <- tibble::tibble(
  x = seq(-2, 2, length.out = 101)
)
straight_tbl <- straight_tbl |>
  dplyr::mutate(
    y = 2 * x + 1
  )

straight_endpoints <- tibble::tibble(
  label = c("A", "B"),
  x = c(-1.5, 1.5),
  y = 2 * c(-1.5, 1.5) + 1
)

straight_euclidean_distance <- sqrt(
  sum(
    (
      straight_endpoints[1, c("x", "y")] -
        straight_endpoints[2, c("x", "y")]
    )^2
  )
)

straight_euclidean_distance |> signif(2)
```

```{r}
#| label: fig-straight-manifold-distance
#| echo: false
#| fig-cap: "On a straight one-dimensional object, the Euclidean distance and geodesic distance are identical."

ggplot() +
  geom_path(
    data = straight_tbl,
    aes(x = x, y = y),
    linewidth = 1.4,
    colour = "grey25"
  ) +
  geom_segment(
    aes(
      x = straight_endpoints$x[1],
      y = straight_endpoints$y[1],
      xend = straight_endpoints$x[2],
      yend = straight_endpoints$y[2]
    ),
    colour = "red3",
    linewidth = 1.5,
    linetype = "dashed"
  ) +
  geom_point(
    data = straight_endpoints,
    aes(x = x, y = y),
    size = 4
  ) +
  ggrepel::geom_text_repel(
    data = straight_endpoints,
    aes(x = x, y = y, label = label),
    nudge_y = 0.16,
    fontface = "bold"
  ) +
  # annotate(
  #   "text",
  #   x = 0,
  #   y = -0.18,
  #   label = paste0(
  #     "Euclidean distance = geodesic distance = ",
  #     round(straight_euclidean_distance, 2)
  #   ),
  #   colour = "red3"
  # ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") + 
  theme(
    # axis.text = element_blank(),
    # axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

Suppose instead that the line was not straight, e.g.

$$
y=x^2.
$$

```{r}
#| label: prepare-curved-manifold-distances
#| include: false

curve_tbl <- tibble::tibble(
  x = seq(-1.5, 1.5, length.out = 501),
  y = x^2
)

curve_endpoints <- tibble::tibble(
  label = c("A", "B"),
  x = c(-1, 1),
  y = c(1, 1)
)

# Sampled observations on the part of the curve between A and B.
curve_sample_tbl <- tibble::tibble(
  x = seq(-1, 1, length.out = 13),
  y = x^2
)

curve_euclidean_distance <- sqrt(
  diff(curve_endpoints$x)^2 +
    diff(curve_endpoints$y)^2
)

# Exact arc length of y = x^2 between x = -1 and x = 1.
curve_geodesic_distance <- integrate(
  function(x) {
    sqrt(1 + (2 * x)^2)
  },
  lower = -1,
  upper = 1
)$value

# Approximation obtained by joining neighbouring sampled points.
curve_graph_distance <- sum(
  sqrt(
    diff(curve_sample_tbl$x)^2 +
      diff(curve_sample_tbl$y)^2
  )
)

tibble::tibble(
  distance = c(
    "Ambient Euclidean distance",
    "Exact geodesic distance",
    "Sample-based graph approximation"
  ),
  value = c(
    curve_euclidean_distance,
    curve_geodesic_distance,
    curve_graph_distance
  )
)
```

```{r}
#| label: fig-curved-manifold-distance
#| echo: false
#| fig-cap: "The straight-line Euclidean distance cuts through the space. The geodesic distance follows the curved object. The dashed polyline approximates the geodesic using sampled observations."

ggplot() +
  geom_path(
    data = curve_tbl,
    aes(x = x, y = y),
    linewidth = 1,
    colour = "grey40"
  ) +
  geom_path(
    data = dplyr::filter(
      curve_tbl,
      x >= -1,
      x <= 1
    ),
    aes(x = x, y = y),
    linewidth = 1.8,
    colour = "navy"
  ) +
  geom_path(
    data = curve_sample_tbl,
    aes(x = x, y = y),
    linewidth = 0.9,
    colour = "deepskyblue4",
    linetype = "dashed"
  ) +
  geom_point(
    data = curve_sample_tbl,
    aes(x = x, y = y),
    size = 2,
    colour = "deepskyblue4"
  ) +
  geom_segment(
    aes(
      x = curve_endpoints$x[1],
      y = curve_endpoints$y[1],
      xend = curve_endpoints$x[2],
      yend = curve_endpoints$y[2]
    ),
    colour = "red3",
    linewidth = 1.5
  ) +
  geom_point(
    data = curve_endpoints,
    aes(x = x, y = y),
    size = 4
  ) +
  geom_text(
    data = curve_endpoints,
    aes(x = x, y = y, label = label),
    nudge_y = 0.18,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = 0,
    y = 1.17,
    label = paste0(
      "Direct path (length: ",
      round(curve_euclidean_distance, 2),
      ")"
    ),
    colour = "red3"
  ) +
  annotate(
    "text",
    x = 0,
    y = 0.28,
    label = paste0(
      "Path along line (length: ",
      round(curve_geodesic_distance, 2),
      ")"
    ),
    colour = "navy"
  ) +
  coord_equal(
    xlim = c(-1.55, 1.55),
    ylim = c(-0.1, 2.35),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    # axis.text = element_blank(),
    # axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

## Calculating distances using the lower-dimensional spacial structure

The shortest distance along the curve between two points is termed the geodesic distance.

If we actually knsew the formula for the curve, we could calculate the lower-dimensional distance exactly.

$$
d_M(A,B)
= 

\int_{-1}^{1}
\sqrt{1+\left(\frac{dy}{dx}\right)^2},dx
= 

\int_{-1}^{1}\sqrt{1+4x^2},dx
\approx 2.96.
$$

If we did not actually know the full formula for the curve, but only knew its value at specific, short intervals, then we could approximate the geodesic distance by joining the points with short straight-line segments.

For example, suppose that we had sampled the curve at 10 evenly-spaced points joining A and B.
Then, the curve would look like this:

```{r}
#| label: fig-curved-manifold-distance-approx
#| echo: false
#| fig-cap: "The subset of sampled points along the curve. The dashed line approximates the geodesic using these sampled observations."

x_vec_approx <- seq(-1, 1, length.out = 5)
y_vec_approx <- approx(
  x = curve_tbl$x,
  y = curve_tbl$y,
  xout = x_vec_approx
)
curve_tbl_approx <- tibble::tibble(
  x = y_vec_approx$x,
  y = y_vec_approx$y
)
int_length_vec <- sqrt(
  diff(curve_tbl_approx$x)^2 +
    diff(curve_tbl_approx$y)^2
)
path_length <- sum(int_length_vec)

ggplot() +
  geom_path(
    data = curve_tbl,
    aes(x = x, y = y),
    linewidth = 1,
    colour = "grey40"
  ) +
  geom_path(
    data = curve_tbl_approx,
    aes(x = x, y = y),
    linewidth = 1.8,
    colour = "navy"
  ) +
  geom_path(
    data = curve_tbl_approx,
    aes(x = x, y = y),
    linewidth = 0.9,
    colour = "deepskyblue4",
    linetype = "dashed",
    alpha = 0.5
  ) +
  geom_point(
    data = curve_tbl_approx,
    aes(x = x, y = y),
    size = 2,
    colour = "deepskyblue4"
  ) +
  geom_segment(
    aes(
      x = curve_endpoints$x[1],
      y = curve_endpoints$y[1],
      xend = curve_endpoints$x[2],
      yend = curve_endpoints$y[2]
    ),
    colour = "red3",
    linewidth = 1.5
  ) +
  geom_point(
    data = curve_endpoints,
    aes(x = x, y = y),
    size = 4
  ) +
  geom_text(
    data = curve_endpoints,
    aes(x = x, y = y, label = label),
    nudge_y = 0.18,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = 0,
    y = 1.17,
    label = paste0(
      "Direct path (length: ",
      round(curve_euclidean_distance, 2),
      ")"
    ),
    colour = "red3"
  ) +
  annotate(
    "text",
    x = 0,
    y = 0.28,
    label = paste0(
      "Approximate path along curve (length: ",
      round(path_length, 2),
      ")"
    ),
    colour = "navy"
  ) +
  coord_equal(
    xlim = c(-1.55, 1.55),
    ylim = c(-0.1, 2.35),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    # axis.text = element_blank(),
    # axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

So, we can do a reasonable job of working out the distance along a shape even if we don't know the exact formula for it, as long as we have a sufficient number of points along the shape.

## Working out the geodesic distance when the shape is entirely unknown

In the previous example, we know that we needed to move from left to right along the x-axis to get from A to B, and so we could work out the geodesic distance by integrating along the curve.

However, in higher-dimensions, we may not know the shape of the object at all, and so we cannot pre-define the paths between points.

So, to approximate the geodesic distances between points, we're going to have to use the points themselves to define the paths.

### Graphs as discretised spaces

A **graph** is a collection of points and connections between points.
The points are called **vertices** or **nodes**.
The connections are called **edges**.

For our purposes, each vertex is an observation.
An edge means that two observations are treated as directly joined.
If an edge has a weight, the weight is the cost or distance of travelling along that edge.

Mathematically, a graph is usually written as

$$
G=(V,E),
$$

where $V$ is the set of vertices and $E$ is the set of edges.
For example,

$$
V=\{A,B,C,D,E\}
$$

and

$$
E=\{(A,B),(B,C),(C,D),(D,E)\}.
$$

The same information can be written as an adjacency list:

$$
\begin{aligned}
A&: B,\\
B&: A,C,\\
C&: B,D,\\
D&: C,E,\\
E&: D.
\end{aligned}
$$

This means that $A$ is connected to $B$, $B$ is connected to $C$, and so on.
It does **not** mean that $A$ is directly connected to $E$.
To travel from $A$ to $E$, we must move along the chain

$$
A \to B \to C \to D \to E.
$$

```{r}
#| label: prepare-simple-graph
#| include: false

simple_vertex_tbl <- tibble::tibble(
  label = c("A", "B", "C", "D", "E"),
  x = c(0.0, 1.0, 2.0, 3.0, 4.0),
  y = c(0.0, 0.4, 0.1, 0.7, 0.3)
)

simple_edge_tbl <- tibble::tribble(
  ~from, ~to,
  "A", "B",
  "B", "C",
  "C", "D",
  "D", "E"
)

make_edge_segment_tbl <- function(vertex_tbl, edge_tbl) {
  edge_tbl |>
    dplyr::left_join(
      vertex_tbl |>
        dplyr::rename(
          from = label,
          x_from = x,
          y_from = y
        ),
      by = "from"
    ) |>
    dplyr::left_join(
      vertex_tbl |>
        dplyr::rename(
          to = label,
          x_to = x,
          y_to = y
        ),
      by = "to"
    )
}

simple_edge_segment_tbl <- make_edge_segment_tbl(
  vertex_tbl = simple_vertex_tbl,
  edge_tbl = simple_edge_tbl
)
```

```{r}
#| label: fig-simple-graph
#| echo: false
#| fig-cap: "A simple graph. The only permitted moves are along the displayed edges."
#| fig-width: 8
#| fig-height: 4

ggplot() +
  geom_segment(
    data = simple_edge_segment_tbl,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 1.2,
    colour = "grey35"
  ) +
  geom_point(
    data = simple_vertex_tbl,
    aes(x = x, y = y),
    size = 4,
    colour = "deepskyblue4"
  ) +
  geom_text(
    data = simple_vertex_tbl,
    aes(x = x, y = y, label = label),
    nudge_y = 0.22,
    fontface = "bold"
  ) +
  coord_equal(
    xlim = c(-0.25, 4.25),
    ylim = c(-0.3, 1.15),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

In this graph, the distance from $A$ to $E$ is not the length of the straight line from $A$ to $E$.
There is no direct edge from $A$ to $E$.
The relevant distance is the shortest distance obtained by travelling through connected points.

If the edges have weights $w_{ij}$, the graph distance between vertices $i$ and $j$ is

$$
d_G(i,j)
=
\min_{\mathcal P:i\leadsto j}
\sum_{(r,s)\in\mathcal P} w_{rs},
$$

where $\mathcal P$ ranges over all paths from $i$ to $j$.

So, a graph turns the continuous space into a discretised space.
Instead of being able to move in any direction, we can move only by stepping from one sampled point to another sampled point.

### Complete graphs

A **complete graph** connects every point to every other point.
If there are $n$ observations, then every pair is joined by an edge.

For an undirected complete graph,

$$
E
=
\{(i,j):1\leq i<j\leq n\}.
$$

The number of edges is therefore

$$
\frac{n(n-1)}{2}.
$$

For example, with five observations there are

$$
\frac{5(5-1)}{2}=10
$$

edges.

```{r}
#| label: prepare-complete-graph-example
#| include: false

make_complete_edge_tbl <- function(vertex_tbl) {
  edge_mat <- t(
    utils::combn(
      vertex_tbl$label,
      2
    )
  )

  tibble::tibble(
    from = edge_mat[, 1],
    to = edge_mat[, 2]
  )
}

complete_edge_tbl <- make_complete_edge_tbl(
  simple_vertex_tbl
)

complete_edge_segment_tbl <- make_edge_segment_tbl(
  vertex_tbl = simple_vertex_tbl,
  edge_tbl = complete_edge_tbl
)
```

```{r}
#| label: fig-complete-graph
#| echo: false
#| fig-cap: "A complete graph on five observations. Every point is directly connected to every other point."
#| fig-width: 8
#| fig-height: 4

ggplot() +
  geom_segment(
    data = complete_edge_segment_tbl,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 0.8,
    alpha = 0.45,
    colour = "grey35"
  ) +
  geom_point(
    data = simple_vertex_tbl,
    aes(x = x, y = y),
    size = 4,
    colour = "deepskyblue4"
  ) +
  geom_text(
    data = simple_vertex_tbl,
    aes(x = x, y = y, label = label),
    nudge_y = 0.22,
    fontface = "bold"
  ) +
  coord_equal(
    xlim = c(-0.25, 4.25),
    ylim = c(-0.3, 1.15),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

A complete graph is the most permissive possible graph.
There are no restrictions on which pairs can be joined directly.

This is fine when the lower-dimensional object is a straight line.
Every direct connection between two sampled points still lies on the line, so the direct Euclidean distance agrees with the distance along the object.

```{r}
#| label: prepare-complete-graph-straight-and-curved
#| include: false

straight_complete_points <- tibble::tibble(
  label = paste0("P", seq_len(9)),
  x = seq(-1, 1, length.out = 9)
) |>
  dplyr::mutate(
    y = 0,
    display_label = dplyr::case_when(
      dplyr::row_number() == 1 ~ "A",
      dplyr::row_number() == dplyr::n() ~ "B",
      TRUE ~ ""
    )
  )

curved_complete_points <- tibble::tibble(
  label = paste0("P", seq_len(9)),
  x = seq(-1, 1, length.out = 9)
) |>
  dplyr::mutate(
    y = x^2,
    display_label = dplyr::case_when(
      dplyr::row_number() == 1 ~ "A",
      dplyr::row_number() == dplyr::n() ~ "B",
      TRUE ~ ""
    )
  )

straight_complete_edges <- make_complete_edge_tbl(
  straight_complete_points
)

curved_complete_edges <- make_complete_edge_tbl(
  curved_complete_points
)

straight_complete_segments <- make_edge_segment_tbl(
  vertex_tbl = straight_complete_points,
  edge_tbl = straight_complete_edges
)

curved_complete_segments <- make_edge_segment_tbl(
  vertex_tbl = curved_complete_points,
  edge_tbl = curved_complete_edges
)

straight_adjacent_edges <- tibble::tibble(
  from = straight_complete_points$label[-length(straight_complete_points$label)],
  to = straight_complete_points$label[-1]
)

curved_adjacent_edges <- tibble::tibble(
  from = curved_complete_points$label[-length(curved_complete_points$label)],
  to = curved_complete_points$label[-1]
)

straight_adjacent_segments <- make_edge_segment_tbl(
  vertex_tbl = straight_complete_points,
  edge_tbl = straight_adjacent_edges
)

curved_adjacent_segments <- make_edge_segment_tbl(
  vertex_tbl = curved_complete_points,
  edge_tbl = curved_adjacent_edges
)

straight_complete_endpoint_distance <- sqrt(
  diff(straight_complete_points$x[c(1, 9)])^2 +
    diff(straight_complete_points$y[c(1, 9)])^2
)

straight_adjacent_path_distance <- sum(
  sqrt(
    diff(straight_complete_points$x)^2 +
      diff(straight_complete_points$y)^2
  )
)

curved_complete_endpoint_distance <- sqrt(
  diff(curved_complete_points$x[c(1, 9)])^2 +
    diff(curved_complete_points$y[c(1, 9)])^2
)

curved_adjacent_path_distance <- sum(
  sqrt(
    diff(curved_complete_points$x)^2 +
      diff(curved_complete_points$y)^2
  )
)

complete_graph_distance_tbl <- tibble::tibble(
  object = c(
    "Straight line",
    "Straight line",
    "Curved line",
    "Curved line"
  ),
  graph_used = c(
    "Complete graph",
    "Adjacent-neighbour graph",
    "Complete graph",
    "Adjacent-neighbour graph"
  ),
  distance_from_A_to_B = c(
    straight_complete_endpoint_distance,
    straight_adjacent_path_distance,
    curved_complete_endpoint_distance,
    curved_adjacent_path_distance
  )
)
```

```{r}
#| label: fig-complete-graph-straight-line
#| echo: false
#| fig-cap: "A complete graph on sampled points from a straight line. Direct connections do not leave the lower-dimensional object."
#| fig-width: 8
#| fig-height: 3.5

ggplot() +
  geom_segment(
    data = straight_complete_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 0.8,
    alpha = 0.35,
    colour = "grey35"
  ) +
  geom_segment(
    data = straight_adjacent_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 1.8,
    colour = "navy"
  ) +
  geom_point(
    data = straight_complete_points,
    aes(x = x, y = y),
    size = 3,
    colour = "deepskyblue4"
  ) +
  geom_text(
    data = straight_complete_points |>
      dplyr::filter(display_label != ""),
    aes(x = x, y = y, label = display_label),
    nudge_y = 0.18,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = 0,
    y = -0.22,
    label = paste0(
      "Complete-graph distance = adjacent-path distance = ",
      round(straight_complete_endpoint_distance, 2)
    ),
    colour = "navy"
  ) +
  coord_equal(
    xlim = c(-1.15, 1.15),
    ylim = c(-0.35, 0.35),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

For the straight line, the complete graph does not introduce an artificial shortcut.
The direct edge from $A$ to $B$ lies on the same one-dimensional object as the sampled points.

The situation changes when the line is curved.
The complete graph connects points that are close in the ambient space, even if the direct edge between them cuts across the lower-dimensional object.

```{r}
#| label: fig-complete-graph-curved-line
#| echo: false
#| fig-cap: "A complete graph on sampled points from a curved line. Some direct connections cut across the curve, creating artificial shortcuts."
#| fig-width: 8
#| fig-height: 5

ggplot() +
  geom_path(
    data = curve_tbl,
    aes(x = x, y = y),
    linewidth = 1.0,
    colour = "grey45"
  ) +
  geom_segment(
    data = curved_complete_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 0.7,
    alpha = 0.25,
    colour = "grey35"
  ) +
  geom_segment(
    aes(
      x = curved_complete_points$x[1],
      y = curved_complete_points$y[1],
      xend = curved_complete_points$x[9],
      yend = curved_complete_points$y[9]
    ),
    linewidth = 1.6,
    colour = "red3"
  ) +
  geom_segment(
    data = curved_adjacent_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 1.8,
    colour = "navy"
  ) +
  geom_point(
    data = curved_complete_points,
    aes(x = x, y = y),
    size = 3,
    colour = "deepskyblue4"
  ) +
  geom_text(
    data = curved_complete_points |>
      dplyr::filter(display_label != ""),
    aes(x = x, y = y, label = display_label),
    nudge_y = 0.15,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = 0,
    y = 1.15,
    label = paste0(
      "Complete-graph shortcut = ",
      round(curved_complete_endpoint_distance, 2)
    ),
    colour = "red3"
  ) +
  annotate(
    "text",
    x = 0,
    y = 0.22,
    label = paste0(
      "Adjacent-neighbour path = ",
      round(curved_adjacent_path_distance, 2)
    ),
    colour = "navy"
  ) +
  coord_equal(
    xlim = c(-1.15, 1.15),
    ylim = c(-0.1, 1.35),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

```{r}
#| label: tbl-complete-graph-distances
#| echo: false

complete_graph_distance_tbl |>
  dplyr::mutate(
    distance_from_A_to_B = round(
      distance_from_A_to_B,
      3
    )
  )
```

The complete graph is therefore not what we want for ISOMAP.
If every pair of points is connected, then the shortest path between two observations is usually just the direct edge between them.
That brings us back to ordinary Euclidean distances.

For a flat or straight object, this is harmless.
For a curved, folded, or obstructed object, it is damaging.
The graph creates shortcuts through the surrounding space, rather than forcing paths to remain on the lower-dimensional structure.

So ISOMAP does **not** use a complete graph.
Instead, it tries to connect only points that are close enough for local Euclidean distance to be trustworthy.
That gives a sparse neighbourhood graph.
Long-range distances are then built up by travelling through chains of local connections.

### Neighbourhood graphs

This is where the neighbourhood graph comes in.

In a neighbourhood graph, each point is connected only to a subset of the other points.

Typically, there are two major approaches:

- k-nearest neighbours (kNN): each point is connected to its k closest points.
- Epsilon-radius ($\epsilon$-radius): each point is connected to all points within a distance $\epsilon$.

The $\epsilon$-radius approach is attractive in that points are consistently connected only if they are close enough to be trusted.

The difficulty is in defining what an appropriate $\epsilon$ is, which becomes increasingly challenging in high dimensions.
Moreover, if the density of points varies across the space, then a single $\epsilon$ may be too small in some regions and too large in others.

Therefore, a kNN approach is often preferred.
In a kNN graph, each point is connected to its k closest points.

### k-nearest-neighbour graphs

A **k-nearest-neighbour graph** connects each observation to its $k$ closest observations, measured using ordinary Euclidean distance in the input space.

The important point is that Euclidean distance is used only **locally**.
We trust the straight-line distance between very nearby observations, but we do not trust the straight-line distance between arbitrary distant observations.

For a collection of points $\mathbf x_1,\ldots,\mathbf x_n$, the basic construction is:

1. calculate all pairwise Euclidean distances;
2. for each point, find the $k$ nearest other points;
3. join those pairs by edges;
4. weight each edge by the corresponding Euclidean distance;
5. estimate long-range distances by shortest paths through the graph.

Now apply this to the curved one-dimensional example.
The true lower-dimensional distance from $A$ to $B$ is the arc length of the curve.
The graph distance estimates this by travelling through neighbouring sampled points.

```{r}
#| label: prepare-knn-curve-graph
#| include: false

construct_knn_graph_for_curve <- function(point_tbl, k) {
  coords <- as.matrix(
    point_tbl[, c("x", "y")]
  )

  n <- nrow(coords)
  distance_mat <- as.matrix(
    dist(coords)
  )

  search_mat <- distance_mat
  diag(search_mat) <- Inf

  adjacent <- matrix(
    FALSE,
    nrow = n,
    ncol = n
  )

  for (i in seq_len(n)) {
    neighbours <- order(search_mat[i, ])[seq_len(k)]
    adjacent[i, neighbours] <- TRUE
  }

  adjacent <- adjacent | t(adjacent)

  edge_index <- which(
    adjacent & upper.tri(adjacent),
    arr.ind = TRUE
  )

  edge_tbl <- tibble::tibble(
    from = as.character(edge_index[, 1]),
    to = as.character(edge_index[, 2]),
    weight = distance_mat[edge_index]
  )

  graph <- igraph::graph_from_data_frame(
    edge_tbl,
    directed = FALSE,
    vertices = tibble::tibble(
      name = as.character(seq_len(n))
    )
  )

  list(
    graph = graph,
    edges = edge_tbl,
    distance_mat = distance_mat
  )
}

make_weighted_edge_segment_tbl <- function(point_tbl, edge_tbl) {
  point_tbl <- point_tbl |>
    dplyr::mutate(
      name = as.character(dplyr::row_number())
    )

  edge_tbl |>
    dplyr::left_join(
      point_tbl |>
        dplyr::rename(
          from = name,
          x_from = x,
          y_from = y
        ),
      by = "from"
    ) |>
    dplyr::left_join(
      point_tbl |>
        dplyr::rename(
          to = name,
          x_to = x,
          y_to = y
        ),
      by = "to"
    )
}

curve_graph_points <- tibble::tibble(
  x = seq(-1, 1, length.out = 17)
) |>
  dplyr::mutate(
    y = x^2,
    label = dplyr::case_when(
      dplyr::row_number() == 1 ~ "A",
      dplyr::row_number() == dplyr::n() ~ "B",
      TRUE ~ ""
    )
  )

knn_k <- 2
knn_curve_graph <- construct_knn_graph_for_curve(
  point_tbl = curve_graph_points,
  k = knn_k
)

knn_curve_segments <- make_weighted_edge_segment_tbl(
  point_tbl = curve_graph_points,
  edge_tbl = knn_curve_graph$edges
)

knn_shortest_vertices <- igraph::shortest_paths(
  knn_curve_graph$graph,
  from = "1",
  to = as.character(nrow(curve_graph_points)),
  weights = igraph::E(knn_curve_graph$graph)$weight,
  output = "vpath"
)$vpath[[1]] |>
  as.character()

knn_shortest_edge_tbl <- tibble::tibble(
  from = knn_shortest_vertices[-length(knn_shortest_vertices)],
  to = knn_shortest_vertices[-1]
)

knn_shortest_segments <- make_weighted_edge_segment_tbl(
  point_tbl = curve_graph_points,
  edge_tbl = knn_shortest_edge_tbl
)

knn_graph_distance <- igraph::distances(
  knn_curve_graph$graph,
  v = "1",
  to = as.character(nrow(curve_graph_points)),
  weights = igraph::E(knn_curve_graph$graph)$weight
)[1, 1]
```

```{r}
#| label: fig-knn-curve-graph
#| echo: false
#| fig-cap: "A k-nearest-neighbour graph on sampled points from the curve. The shortest path through local edges estimates the geodesic distance."
#| fig-width: 8
#| fig-height: 5

ggplot() +
  geom_path(
    data = curve_tbl,
    aes(x = x, y = y),
    linewidth = 1,
    colour = "grey45"
  ) +
  geom_segment(
    data = knn_curve_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 0.8,
    alpha = 0.35,
    colour = "grey35"
  ) +
  geom_segment(
    data = knn_shortest_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 1.7,
    colour = "navy"
  ) +
  geom_segment(
    aes(
      x = curve_graph_points$x[1],
      y = curve_graph_points$y[1],
      xend = curve_graph_points$x[nrow(curve_graph_points)],
      yend = curve_graph_points$y[nrow(curve_graph_points)]
    ),
    linewidth = 1.3,
    colour = "red3",
    linetype = "dashed"
  ) +
  geom_point(
    data = curve_graph_points,
    aes(x = x, y = y),
    size = 2.6,
    colour = "deepskyblue4"
  ) +
  geom_text(
    data = curve_graph_points |>
      dplyr::filter(label != ""),
    aes(x = x, y = y, label = label),
    nudge_y = 0.13,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = 0,
    y = 1.17,
    label = paste0(
      "Straight-line distance = ",
      round(curve_euclidean_distance, 2)
    ),
    colour = "red3"
  ) +
  annotate(
    "text",
    x = 0,
    y = 0.23,
    label = paste0(
      "kNN graph distance = ",
      round(knn_graph_distance, 2),
      "\ntrue geodesic = ",
      round(curve_geodesic_distance, 2)
    ),
    colour = "navy"
  ) +
  coord_equal(
    xlim = c(-1.15, 1.15),
    ylim = c(-0.1, 1.35),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

The estimate improves when the graph is connected and local.
If $k$ is too small, the graph may break into components.
If $k$ is too large, the graph begins to add shortcuts across the curve.

```{r}
#| label: tbl-knn-geodesic-accuracy
#| echo: false

estimate_knn_distance <- function(k) {
  graph_info <- construct_knn_graph_for_curve(
    point_tbl = curve_graph_points,
    k = k
  )

  estimate <- igraph::distances(
    graph_info$graph,
    v = "1",
    to = as.character(nrow(curve_graph_points)),
    weights = igraph::E(graph_info$graph)$weight
  )[1, 1]

  tibble::tibble(
    graph_distance = estimate,
    absolute_error = abs(estimate - curve_geodesic_distance),
    connected = is.finite(estimate)
  )
}

tibble::tibble(
  k = c(1, 2, 3, 5, 8, 16)
) |>
  dplyr::mutate(
    result = purrr::map(k, estimate_knn_distance)
  ) |>
  tidyr::unnest(result) |>
  dplyr::mutate(
    graph_distance = round(graph_distance, 3),
    absolute_error = round(absolute_error, 3)
  )
```

The goal is not to make $k$ as large as possible.
The goal is to choose $k$ large enough to keep the graph connected, but small enough that edges remain local.

### Epsilon-neighbourhood graphs

An **epsilon-neighbourhood graph** uses a distance threshold instead of a fixed number of neighbours.
Two observations are joined if their Euclidean distance is at most $\epsilon$.

$$
(i,j)\in E
\quad\text{when}\quad
\|\mathbf x_i-\mathbf x_j\|\leq \epsilon.
$$


This is conceptually simple: every edge is local because every edge must have length at most $\epsilon$.
The difficulty is that a single value of $\epsilon$ may be too small in sparse regions and too large in dense regions.

```{r}
#| label: prepare-epsilon-curve-graph
#| include: false

construct_epsilon_graph_for_curve <- function(point_tbl, epsilon) {
  coords <- as.matrix(
    point_tbl[, c("x", "y")]
  )

  n <- nrow(coords)
  distance_mat <- as.matrix(
    dist(coords)
  )

  adjacent <- distance_mat <= epsilon
  diag(adjacent) <- FALSE

  edge_index <- which(
    adjacent & upper.tri(adjacent),
    arr.ind = TRUE
  )

  edge_tbl <- tibble::tibble(
    from = as.character(edge_index[, 1]),
    to = as.character(edge_index[, 2]),
    weight = distance_mat[edge_index]
  )

  graph <- igraph::graph_from_data_frame(
    edge_tbl,
    directed = FALSE,
    vertices = tibble::tibble(
      name = as.character(seq_len(n))
    )
  )

  list(
    graph = graph,
    edges = edge_tbl,
    distance_mat = distance_mat
  )
}

epsilon_value <- 0.27
epsilon_curve_graph <- construct_epsilon_graph_for_curve(
  point_tbl = curve_graph_points,
  epsilon = epsilon_value
)

epsilon_curve_segments <- make_weighted_edge_segment_tbl(
  point_tbl = curve_graph_points,
  edge_tbl = epsilon_curve_graph$edges
)

epsilon_shortest_vertices <- igraph::shortest_paths(
  epsilon_curve_graph$graph,
  from = "1",
  to = as.character(nrow(curve_graph_points)),
  weights = igraph::E(epsilon_curve_graph$graph)$weight,
  output = "vpath"
)$vpath[[1]] |>
  as.character()

epsilon_shortest_edge_tbl <- tibble::tibble(
  from = epsilon_shortest_vertices[-length(epsilon_shortest_vertices)],
  to = epsilon_shortest_vertices[-1]
)

epsilon_shortest_segments <- make_weighted_edge_segment_tbl(
  point_tbl = curve_graph_points,
  edge_tbl = epsilon_shortest_edge_tbl
)

epsilon_graph_distance <- igraph::distances(
  epsilon_curve_graph$graph,
  v = "1",
  to = as.character(nrow(curve_graph_points)),
  weights = igraph::E(epsilon_curve_graph$graph)$weight
)[1, 1]
```

```{r}
#| label: fig-epsilon-curve-graph
#| echo: false
#| fig-cap: "An epsilon-neighbourhood graph on sampled points from the curve. Edges are included only when the ambient Euclidean distance is below the threshold."
#| fig-width: 8
#| fig-height: 5

ggplot() +
  geom_path(
    data = curve_tbl,
    aes(x = x, y = y),
    linewidth = 1,
    colour = "grey45"
  ) +
  geom_segment(
    data = epsilon_curve_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 0.8,
    alpha = 0.35,
    colour = "grey35"
  ) +
  geom_segment(
    data = epsilon_shortest_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 1.7,
    colour = "navy"
  ) +
  geom_segment(
    aes(
      x = curve_graph_points$x[1],
      y = curve_graph_points$y[1],
      xend = curve_graph_points$x[nrow(curve_graph_points)],
      yend = curve_graph_points$y[nrow(curve_graph_points)]
    ),
    linewidth = 1.3,
    colour = "red3",
    linetype = "dashed"
  ) +
  geom_point(
    data = curve_graph_points,
    aes(x = x, y = y),
    size = 2.6,
    colour = "deepskyblue4"
  ) +
  geom_text(
    data = curve_graph_points |>
      dplyr::filter(label != ""),
    aes(x = x, y = y, label = label),
    nudge_y = 0.13,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = 0,
    y = 1.17,
    label = paste0(
      "Straight-line distance = ",
      round(curve_euclidean_distance, 2)
    ),
    colour = "red3"
  ) +
  annotate(
    "text",
    x = 0,
    y = 0.23,
    label = paste0(
      "epsilon graph distance = ",
      round(epsilon_graph_distance, 2),
      "\ntrue geodesic = ",
      round(curve_geodesic_distance, 2)
    ),
    colour = "navy"
  ) +
  coord_equal(
    xlim = c(-1.15, 1.15),
    ylim = c(-0.1, 1.35),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

As with $k$, the value of $\epsilon$ has to balance two errors.
If $\epsilon$ is too small, the graph may be disconnected.
If $\epsilon$ is too large, the graph may add shortcuts and move back toward ordinary Euclidean distances.

```{r}
#| label: tbl-epsilon-geodesic-accuracy
#| echo: false

estimate_epsilon_distance <- function(epsilon) {
  graph_info <- construct_epsilon_graph_for_curve(
    point_tbl = curve_graph_points,
    epsilon = epsilon
  )

  estimate <- igraph::distances(
    graph_info$graph,
    v = "1",
    to = as.character(nrow(curve_graph_points)),
    weights = igraph::E(graph_info$graph)$weight
  )[1, 1]

  tibble::tibble(
    graph_distance = estimate,
    absolute_error = abs(estimate - curve_geodesic_distance),
    connected = is.finite(estimate)
  )
}

tibble::tibble(
  epsilon = c(0.15, 0.27, 0.50, 2.10)
) |>
  dplyr::mutate(
    result = purrr::map(epsilon, estimate_epsilon_distance)
  ) |>
  tidyr::unnest(result) |>
  dplyr::mutate(
    graph_distance = round(graph_distance, 3),
    absolute_error = round(absolute_error, 3)
  )
```

### What both neighbourhood graphs are trying to do

Both constructions are based on the same idea.

Short straight-line distances are usually acceptable local approximations to movement along the underlying object.
Long straight-line distances are not.

So ISOMAP replaces one global distance by many small local distances:

$$
\|\mathbf x_i-\mathbf x_j\|
\quad\leadsto\quad
\min_{\mathcal P:i\leadsto j}
\sum_{(r,s)\in\mathcal P}
\|\mathbf x_r-\mathbf x_s\|.
$$

The left-hand side is the direct Euclidean distance through the input space.
The right-hand side is the graph shortest-path distance, obtained by travelling through local edges.

That shortest-path distance is not the true geodesic distance.
It is an estimate of the geodesic distance based on the sampled observations.
With enough data, and with a neighbourhood graph that is local but connected, this estimate can be much closer to the true lower-dimensional distance than the direct Euclidean distance.

## From one shortest path to all pairwise distances

The previous examples asked for one distance: how far is point A from point B?

ISOMAP needs more than this.
It needs an estimated distance between every pair of observations.

The same graph idea applies.
For each pair of points, we ask:

> What is the shortest route through the graph from one point to the other?

The edge lengths are still ordinary Euclidean distances.
The change is that we are not allowed to jump directly between any two points unless there is an edge between them.

```{r}
#| label: calculate-all-pair-curve-distance-matrices
#| include: false

curve_euclidean_mat <- as.matrix(
  dist(
    curve_graph_points[, c("x", "y")]
  )
)

curve_graph_distance_mat <- igraph::distances(
  knn_curve_graph$graph,
  weights = igraph::E(knn_curve_graph$graph)$weight
)

selected_curve_vertices <- c(1, 5, 9, 13, 17)
selected_curve_labels <- paste0("point ", selected_curve_vertices)

curve_euclidean_small <- round(
  curve_euclidean_mat[
    selected_curve_vertices,
    selected_curve_vertices
  ],
  2
)

curve_graph_small <- round(
  curve_graph_distance_mat[
    selected_curve_vertices,
    selected_curve_vertices
  ],
  2
)

rownames(curve_euclidean_small) <- selected_curve_labels
colnames(curve_euclidean_small) <- selected_curve_labels

rownames(curve_graph_small) <- selected_curve_labels
colnames(curve_graph_small) <- selected_curve_labels
```

### The original Euclidean distance matrix

The ordinary Euclidean distance matrix connects every pair of points directly.

```{r}
#| label: tbl-curve-euclidean-distance-matrix
#| echo: false

knitr::kable(
  curve_euclidean_small,
  caption = "Selected entries from the ordinary Euclidean distance matrix."
)
```

### The graph-distance matrix

The graph-distance matrix measures the shortest route through local graph edges.

```{r}
#| label: tbl-curve-graph-distance-matrix
#| echo: false

knitr::kable(
  curve_graph_small,
  caption = "Selected entries from the graph shortest-path distance matrix."
)
```

These two matrices answer different questions.

The Euclidean matrix asks:

> How far apart are two points if we draw a straight line between them?

The graph-distance matrix asks:

> How far apart are two points if we have to travel through the graph?

For neighbouring points, the answers are usually very similar.
For points far apart along the curve, the graph distance can be much closer to the distance along the underlying object.

```{r}
#| label: tbl-curve-distance-comparison
#| echo: false

curve_pair_comparison <- tibble::tibble(
  pair = c(
    "point 1 to point 2",
    "point 1 to point 9",
    "point 1 to point 17"
  ),
  euclidean_distance = c(
    curve_euclidean_mat[1, 2],
    curve_euclidean_mat[1, 9],
    curve_euclidean_mat[1, 17]
  ),
  graph_distance = c(
    curve_graph_distance_mat[1, 2],
    curve_graph_distance_mat[1, 9],
    curve_graph_distance_mat[1, 17]
  )
) |>
  dplyr::mutate(
    euclidean_distance = round(euclidean_distance, 2),
    graph_distance = round(graph_distance, 2)
  )

knitr::kable(
  curve_pair_comparison,
  caption = "Direct Euclidean distances compared with shortest-path distances through the graph."
)
```

The important point is simple:

> ISOMAP replaces the original Euclidean distance matrix with a graph shortest-path distance matrix.

## The ISOMAP algorithm

We can now state ISOMAP as a sequence of four steps.

### Step 1: calculate ordinary Euclidean distances

Start with the observed coordinates.
Calculate the ordinary Euclidean distance between every pair of observations.

```{r}
#| label: code-isomap-step-1
#| eval: false

x <- as.matrix(my_data)

euclidean_distance_mat <- as.matrix(
  dist(x)
)
```

### Step 2: construct a local neighbourhood graph

Keep only local edges.
This can be done using a $k$-nearest-neighbour graph or an $\epsilon$-neighbourhood graph.

```{r}
#| label: code-isomap-step-2
#| eval: false

neighbourhood <- build_knn_graph(
  x = x,
  k = 8
)

graph <- neighbourhood$graph
```

### Step 3: calculate shortest-path distances through the graph

The graph now tells us which moves are allowed.
The shortest-path calculation gives a new distance matrix.

```{r}
#| label: code-isomap-step-3
#| eval: false

graph_distance_mat <- igraph::distances(
  graph,
  weights = igraph::E(graph)$weight
)
```

This is the key ISOMAP move.
The graph distances are trying to estimate distances along the lower-dimensional object.

### Step 4: apply classical MDS

Finally, apply classical MDS to the graph-distance matrix.

```{r}
#| label: code-isomap-step-4
#| eval: false

isomap_fit <- cmdscale(
  as.dist(graph_distance_mat),
  k = 2,
  eig = TRUE
)

isomap_coordinates <- isomap_fit$points
```

So the overall method is:

$$
\boxed{
\text{data}
\to
\text{local graph}
\to
\text{shortest-path distances}
\to
\text{classical MDS}
}
$$

ISOMAP is therefore not a completely new version of MDS.
It is classical MDS applied after changing the distances.

As we will later see, the correct graph is *critical* for the final ISOMAP algorithm.

## A two-dimensional manifold in three-dimensional space

The one-dimensional curve examples show the idea, but ISOMAP is most useful when the lower-dimensional object is not so easy to see.

A standard example is the Swiss roll.
It is a two-dimensional surface embedded in three-dimensional space.

```{r}
#| label: simulate-swiss-roll-for-isomap

swiss_tbl <- simulate_swiss_roll(
  n = 700,
  seed = 5077
)

swiss_x <- swiss_tbl |>
  dplyr::select(x1, x2, x3) |>
  as.matrix()

swiss_display_tbl <- swiss_tbl |>
  dplyr::bind_cols(
    project_3d_to_2d(
      x1 = swiss_tbl$x1,
      x2 = swiss_tbl$x2,
      x3 = swiss_tbl$x3,
      azimuth = -45,
      elevation = 20
    )
  )
```

```{r}
#| label: fig-swiss-roll-view
#| echo: false
#| fig-cap: "A two-dimensional Swiss-roll surface embedded in three-dimensional space. Colour shows position along the roll."

plotly::plot_ly(
  data = swiss_tbl,
  x = ~x1,
  y = ~x3,
  z = ~x2,
  colour = ~t,
  colours = viridisLite::viridis(100),
  type = "scatter3d",
  mode = "markers",
  marker = list(
    size = 2,
    opacity = 0.8
  )
) |>
  plotly::layout(
    scene = list(
      xaxis = list(title = "x1"),
      yaxis = list(title = "x3"),
      zaxis = list(title = "Height"),
      aspectmode = "data",
      camera = list(
        eye = list(
          x = 1.6,
          y = 1.6,
          z = 0.8
        )
      )
    )
  )
```

Each observation has three measured coordinates.
But the surface itself has only two intrinsic directions:

1. position along the rolled sheet;
2. height along the sheet.

## Why Euclidean distance is a problem on the Swiss roll

Points on neighbouring layers of the roll can be close in ordinary three-dimensional distance.
But to move between them while staying on the surface, one may have to travel a long way around the sheet.

This is the same problem as the curved-line and river examples, but now in three dimensions.

```{r}
#| label: fit-swiss-roll-pca-and-euclidean-mds
#| include: false

swiss_pca <- prcomp(
  swiss_x,
  center = TRUE,
  scale. = FALSE
)

swiss_pca_scores <- swiss_pca$x[, 1:2]

swiss_cmds <- cmdscale(
  dist(swiss_x),
  k = 2,
  eig = TRUE
)

swiss_linear_tbl <- dplyr::bind_rows(
  tibble::tibble(
    dimension_1 = swiss_pca_scores[, 1],
    dimension_2 = swiss_pca_scores[, 2],
    t = swiss_tbl$t,
    method = "PCA"
  ),
  tibble::tibble(
    dimension_1 = swiss_cmds$points[, 1],
    dimension_2 = swiss_cmds$points[, 2],
    t = swiss_tbl$t,
    method = "Classical MDS\nwith Euclidean distance"
  )
) |>
  dplyr::mutate(
    method = factor(
      method,
      levels = c(
        "PCA",
        "Classical MDS\nwith Euclidean distance"
      )
    )
  )
```

```{r}
#| label: fig-swiss-roll-pca-and-euclidean-mds
#| echo: false
#| fig-cap: "PCA and Euclidean classical MDS preserve the rolled geometry rather than unfolding the surface."
#| out-width: "100%"

ggplot(
  swiss_linear_tbl,
  aes(
    x = dimension_1,
    y = dimension_2,
    colour = t
  )
) +
  geom_point(
    alpha = 0.85,
    size = 1.2
  ) +
  facet_wrap(~ method) +
  coord_equal() +
  scale_colour_viridis_c(
    name = "Position\nalong roll"
  ) +
  labs(
    x = "first dimension",
    y = "second dimension"
  ) +
  cowplot::theme_cowplot() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

The two plots need not have the same orientation.
Eigenvector signs and rotations are arbitrary.
The important point is that both methods are still working with ordinary Euclidean geometry.

## ISOMAP on the Swiss roll

Now use local graph distances instead of direct Euclidean distances.

```{r}
#| label: choose-connected-swiss-roll-graph
#| include: false

candidate_k <- c(6, 8, 10, 12, 15)

swiss_connectivity_tbl <- tibble::tibble(
  k = candidate_k
) |>
  dplyr::mutate(
    graph_info = purrr::map(
      k,
      ~ build_knn_graph(
        x = swiss_x,
        k = .x
      )
    ),
    components = purrr::map_int(
      graph_info,
      ~ .x$components$no
    )
  )

swiss_k <- swiss_connectivity_tbl |>
  dplyr::filter(components == 1) |>
  dplyr::slice(1) |>
  dplyr::pull(k)

if (length(swiss_k) == 0) {
  stop("No connected Swiss-roll graph was found among the candidate k values.")
}
```

```{r}
#| label: fit-swiss-roll-isomap-main
#| include: false

swiss_isomap <- fit_isomap(
  x = swiss_x,
  k = swiss_k,
  ndim = 2
)

swiss_isomap_tbl <- tibble::tibble(
  dimension_1 = swiss_isomap$points[, 1],
  dimension_2 = swiss_isomap$points[, 2],
  t = swiss_tbl$t,
  height = swiss_tbl$height
)
```

```{r}
#| label: fig-swiss-roll-isomap-main
#| echo: false
#| fig-cap: "ISOMAP applied to the Swiss roll. The method uses shortest-path distances through a local neighbourhood graph before applying classical MDS."
#| fig-width: 8
#| fig-height: 6

ggplot(
  swiss_isomap_tbl,
  aes(
    x = dimension_1,
    y = dimension_2,
    colour = t
  )
) +
  geom_point(
    alpha = 0.85,
    size = 1.3
  ) +
  coord_equal() +
  scale_colour_viridis_c(
    name = "Position\nalong roll"
  ) +
  labs(
    x = "first ISOMAP dimension",
    y = "second ISOMAP dimension"
  ) +
  cowplot::theme_cowplot() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

The colour progression is now much closer to an unrolled representation of the surface.

For this simulated example only, we know the true unrolled coordinates.
That lets us check whether the ISOMAP representation is recovering the intended structure.

```{r}
#| label: prepare-swiss-roll-intrinsic-comparison
#| include: false

swiss_intrinsic_tbl <- swiss_tbl |>
  dplyr::transmute(
    dimension_1 = arc_length,
    dimension_2 = height,
    t = t,
    method = "True unrolled\ncoordinates"
  )

aligned_isomap_coordinates <- align_configuration(
  reference = swiss_intrinsic_tbl |>
    dplyr::select(dimension_1, dimension_2),
  target = swiss_isomap$points
)

swiss_aligned_isomap_tbl <- tibble::tibble(
  dimension_1 = aligned_isomap_coordinates[, 1],
  dimension_2 = aligned_isomap_coordinates[, 2],
  t = swiss_tbl$t,
  method = "ISOMAP\nconfiguration"
)

swiss_compare_tbl <- dplyr::bind_rows(
  swiss_intrinsic_tbl,
  swiss_aligned_isomap_tbl
) |>
  dplyr::mutate(
    method = factor(
      method,
      levels = c(
        "True unrolled\ncoordinates",
        "ISOMAP\nconfiguration"
      )
    )
  )
```

```{r}
#| label: fig-swiss-roll-isomap-versus-intrinsic
#| echo: false
#| fig-cap: "The true unrolled coordinates compared with the ISOMAP configuration, after allowing for rotation, reflection and scaling."
#| out-width: "100%"

ggplot(
  swiss_compare_tbl,
  aes(
    x = dimension_1,
    y = dimension_2,
    colour = t
  )
) +
  geom_point(
    alpha = 0.85,
    size = 1.2
  ) +
  facet_wrap(~ method) +
  coord_equal() +
  scale_colour_viridis_c(
    name = "Position\nalong roll"
  ) +
  labs(
    x = "first dimension",
    y = "second dimension"
  ) +
  cowplot::theme_cowplot() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

The next practical question is how we choose the graph.
If the graph is too sparse, it may break apart.
If it is too dense, it may add shortcuts between different layers of the roll.
That is why the graph construction is not just a technical detail.


## Choosing the neighbourhood size

The choice of $k$ or $\varepsilon$ is central to ISOMAP.

A useful neighbourhood must be:

- large enough for the graph to be connected; but
- small enough for edges to remain genuinely local.

### If the neighbourhood is too small

- the graph may be disconnected;
- shortest paths between components do not exist;
- within a sparse graph, paths may be unnecessarily long and unstable.

### If the neighbourhood is too large

- edges may jump between nearby folds;
- these short-circuit edges create unrealistically short graph paths;
- graph distances move back towards ordinary ambient Euclidean distances;
- the nonlinear advantage of ISOMAP is lost.

### Sensitivity of the embedding to $k$

Compare a reasonable neighbourhood with a much larger one.

```{r}
#| label: fit-swiss-roll-isomap-over-k

k_embedding_values <- c(
  4,
  5,
  8,
  20
)

swiss_k_fit_tbl <- map_dfr(
  k_embedding_values,
  function(k_value) {
    graph_info <- build_knn_graph(
      swiss_x,
      k = k_value
    )

    if (graph_info$components$no != 1) {
      return(
        tibble(
          dimension_1 = numeric(),
          dimension_2 = numeric(),
          t = numeric(),
          k = integer()
        )
      )
    }

    graph_distance_mat <- igraph::distances(
      graph_info$graph,
      weights = igraph::E(
        graph_info$graph
      )$weight
    )

    coordinates <- cmdscale(
      as.dist(graph_distance_mat),
      k = 2
    )

    tibble(
      dimension_1 = coordinates[, 1],
      dimension_2 = coordinates[, 2],
      t = swiss_tbl$t,
      k = k_value
    )
  }
) |>
  mutate(
    k = factor(
      k,
      levels = k_embedding_values,
      labels = paste0(
        "k = ",
        k_embedding_values
      )
    )
  )
```

```{r}
#| label: fig-swiss-roll-isomap-over-k
#| echo: false
#| fig-cap: "ISOMAP embeddings of the Swiss roll under different nearest-neighbour choices."
#| out-width: "100%"

ggplot(
  swiss_k_fit_tbl,
  aes(
    x = dimension_1,
    y = dimension_2,
    colour = t
  )
) +
  geom_point(
    alpha = 0.75,
    size = 0.9
  ) +
  facet_wrap(
    ~ k
  ) +
  coord_equal() +
  scale_colour_viridis_c(
    name = "Position\nalong roll"
  ) +
  cowplot::theme_cowplot() +
  theme(
    panel.background = element_rect(
      fill = "white"
    ),
    plot.background = element_rect(
      fill = "white"
    )
  ) +
  labs(
    x = "First ISOMAP dimension",
    y = "Second ISOMAP dimension"
  )
```

A single visually pleasing result is not enough. A sensible interpretation should remain reasonably stable across nearby defensible values of $k$.

## Comparing dimensionality-reduction methods

| Method | Starting information | Main quantity represented | Where nonlinearity enters |
|:--|:--|:--|:--|
| PCA | Data matrix | Variation under linear projection | It does not |
| Classical MDS | Dissimilarity matrix | Geometry of the supplied dissimilarities | Only through the supplied dissimilarities |
| Non-metric MDS | Dissimilarity matrix | Ordering of supplied dissimilarities | Through a monotone transformation |
| ISOMAP | Data matrix and a neighbourhood rule | Estimated geodesic distances | Through graph construction and shortest paths |

Non-metric MDS and ISOMAP solve different problems.

- Non-metric MDS changes how the supplied dissimilarities are matched.
- ISOMAP changes the dissimilarities themselves before applying classical MDS.

## Failure modes

ISOMAP can work strikingly well on a clean, densely sampled Swiss roll. Its assumptions may be much less convincing in real datasets.

### Disconnected graphs

If the neighbourhood is too small, the graph may separate into components. No finite shortest path joins different components.

Possible responses include:

- increasing the neighbourhood size;
- analysing components separately; or
- reconsidering whether a single connected manifold is plausible.

### Short-circuit edges

If the neighbourhood is too large, or folds are close together, edges may cross gaps in the manifold. A single shortcut can distort many shortest paths.

### Uneven sampling density

A global $k$ or $\varepsilon$ may be too large in dense regions and too small in sparse regions.

### Noise

Noise can obscure local geometry, create false neighbours and make graph paths unstable.

### Outliers

An outlier may become isolated or act as an artificial bridge between otherwise separate regions.

### Multiple manifolds

Observations from genuinely distinct populations may not lie on one connected smooth manifold. One global ISOMAP map may then be inappropriate.

## What ISOMAP does not provide

ISOMAP does not automatically provide:

- loadings that relate its axes directly to the original variables;
- a universally correct value of $k$ or $\varepsilon$;
- robustness to noise and outliers;
- a guarantee that apparent groups are genuine clusters;
- a simple transformation for mapping new observations;
- recovery of every form of nonlinear structure.

The output axes are coordinates chosen to represent graph distances. Their signs and orientations are arbitrary, just as in MDS.