---
title: "ISOMAP: Changing the Distances Before MDS"
author: Miguel Rodo
date: "2026-07-31"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
---

```{r}
#| label: setup
#| include: false

library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(cowplot)
library(ggrepel)
library(igraph)

knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "80%"
)

simulate_swiss_roll <- function(
  n = 500,
  seed = 5077,
  noise_sd = 0
) {
  set.seed(seed)

  t <- 1.5 * pi * (1 + 2 * runif(n))
  height <- runif(n, min = 0, max = 20)

  x1 <- t * cos(t) + rnorm(n, sd = noise_sd)
  x2 <- height + rnorm(n, sd = noise_sd)
  x3 <- t * sin(t) + rnorm(n, sd = noise_sd)

  # Arc length along the spiral:
  # integral sqrt(1 + t^2) dt
  arc_length <- 0.5 * (
    t * sqrt(1 + t^2) +
      asinh(t)
  )

  tibble(
    observation = seq_len(n),
    t = t,
    height = height,
    arc_length = arc_length,
    x1 = x1,
    x2 = x2,
    x3 = x3
  )
}

project_3d_to_2d <- function(
  x1,
  x2,
  x3,
  azimuth = -45,
  elevation = 25
) {
  theta <- azimuth * pi / 180
  phi <- elevation * pi / 180

  horizontal <- cos(theta) * x1 -
    sin(theta) * x3

  depth <- sin(theta) * x1 +
    cos(theta) * x3

  vertical <- -sin(phi) * depth +
    cos(phi) * x2

  tibble(
    display_x = horizontal,
    display_y = vertical
  )
}

build_knn_graph <- function(x, k) {
  x <- as.matrix(x)
  n <- nrow(x)

  if (k < 1 || k >= n) {
    stop("k must lie between 1 and n - 1.")
  }

  ambient_distance_mat <- as.matrix(
    dist(x, method = "euclidean")
  )

  search_distance_mat <- ambient_distance_mat
  diag(search_distance_mat) <- Inf

  neighbour_indicator <- matrix(
    FALSE,
    nrow = n,
    ncol = n
  )

  for (i in seq_len(n)) {
    nearest <- order(
      search_distance_mat[i, ],
      decreasing = FALSE
    )[seq_len(k)]

    neighbour_indicator[i, nearest] <- TRUE
  }

  # Symmetrise the k-nearest-neighbour relation:
  # connect i and j when either regards the other as a neighbour.
  adjacent <- neighbour_indicator |
    t(neighbour_indicator)

  edge_index <- which(
    adjacent & upper.tri(adjacent),
    arr.ind = TRUE
  )

  edge_tbl <- tibble(
    from = as.character(edge_index[, 1]),
    to = as.character(edge_index[, 2]),
    weight = ambient_distance_mat[edge_index]
  )

  graph <- igraph::graph_from_data_frame(
    edge_tbl,
    directed = FALSE,
    vertices = tibble(
      name = as.character(seq_len(n))
    )
  )

  list(
    graph = graph,
    edges = edge_tbl,
    ambient_distances = ambient_distance_mat,
    components = igraph::components(graph)
  )
}

fit_isomap <- function(x, k, ndim = 2) {
  graph_info <- build_knn_graph(x, k)

  if (graph_info$components$no != 1) {
    stop(
      paste0(
        "The neighbourhood graph has ",
        graph_info$components$no,
        " connected components."
      )
    )
  }

  graph_distance_mat <- igraph::distances(
    graph_info$graph,
    weights = igraph::E(
      graph_info$graph
    )$weight
  )

  mds_fit <- cmdscale(
    as.dist(graph_distance_mat),
    k = ndim,
    eig = TRUE
  )

  list(
    graph_info = graph_info,
    graph_distances = graph_distance_mat,
    mds = mds_fit,
    points = mds_fit$points
  )
}

align_configuration <- function(
  reference,
  target,
  allow_scaling = TRUE
) {
  reference <- as.matrix(reference)
  target <- as.matrix(target)

  reference_centre <- colMeans(reference)
  target_centre <- colMeans(target)

  reference_centred <- sweep(
    reference,
    MARGIN = 2,
    STATS = reference_centre,
    FUN = "-"
  )

  target_centred <- sweep(
    target,
    MARGIN = 2,
    STATS = target_centre,
    FUN = "-"
  )

  rotation_svd <- svd(
    crossprod(
      target_centred,
      reference_centred
    )
  )

  rotation <- rotation_svd$u %*%
    t(rotation_svd$v)

  target_aligned <- target_centred %*%
    rotation

  if (allow_scaling) {
    scale_factor <- sum(
      target_aligned *
        reference_centred
    ) /
      sum(target_aligned^2)

    target_aligned <- scale_factor *
      target_aligned
  }

  sweep(
    target_aligned,
    MARGIN = 2,
    STATS = reference_centre,
    FUN = "+"
  )
}

distance_preservation <- function(
  target_distances,
  coordinates
) {
  target <- as.vector(
    as.dist(target_distances)
  )

  fitted <- as.vector(
    dist(coordinates)
  )

  correlation <- cor(target, fitted)

  tibble(
    correlation = correlation,
    residual_variance = 1 -
      correlation^2
  )
}
```

<!--
Suggested pacing for a two-hour session:

00:00-00:12  Bridge from PCA and MDS
00:12-00:30  Curved manifolds and geodesic distance
00:30-00:48  Neighbourhood graphs and shortest paths
00:48-01:03  The ISOMAP algorithm
01:03-01:20  Swiss-roll example: PCA, classical MDS and ISOMAP
01:20-01:40  Choosing the neighbourhood size
01:40-01:52  Diagnostics and target dimension
01:52-02:00  Failure modes, workflow and recap

The more technical derivations and supplementary remarks are collapsed by default.
-->

## From PCA and MDS to ISOMAP

Principal component analysis (PCA) constructs new variables by taking linear combinations of the original variables, in order to represent a high-dimensional dataset in fewer dimensions.

Classical multidimensional scaling (MDS) differs from PCA in that it begins with pairwise dissimilarities rather than the original varibles.
But it is ultimately a linear method, and in fact is exactly equivalent to PCA when the dissimilarities are ordinary Euclidean distances.

$$
\text{PCA scores}
\quad\equiv\quad
\text{classical MDS coordinates}.
$$

So, if we have that the distance between observations is not accurately represented by the straight-line distance between them, the PCA and classical MDS will not recover the underlying structure of the data.

The key question is therefore:

> What should we do when straight-line Euclidean distances do not represent movement along the structure of the data?

Metric and non-metric MDS are non-linear extensions of MDS, and so can recover some non-linear structure.

ISOMAP is another alternative.
It relates closely to classical MDS, but addresses the issue by changing the dissimilarities before applying classical MDS.

$$
\boxed{
\text{data}
\longrightarrow
\text{neighbourhood graph}
\longrightarrow
\text{shortest-path distances}
\longrightarrow
\text{classical MDS}
}
$$

The final classical-scaling step is familiar.
The nonlinear behaviour enters through the graph-based distances supplied to it.

## Lower-dimensional objects embedded in higher-dimensional space

Ultimately, ISOMAP is going to to base the distance between two observations in the high-dimensional space on the effectively lower-dimensional objects that these observations are assumed to lie on.
We'll give a few examples to illustrate.

The first example is a plane in three-dimensional space.

```{r}
#| label: fig-plane-in-three-dimensions
#| fig-cap: "A two-dimensional plane embedded in three-dimensional space. The observations have three coordinates but vary in only two independent directions."
#| fig-width: 8
#| fig-height: 6

set.seed(5077)

x1_grid <- seq(-2, 2, length.out = 30)
x2_grid <- seq(-2, 2, length.out = 30)

# The plane is x3 = 0.6 x1 - 0.4 x2.
x3_grid <- outer(
  x1_grid,
  x2_grid,
  function(x1, x2) {
    0.6 * x1 - 0.4 * x2
  }
)

plane_points <- tibble::tibble(
  x1 = runif(35, -1.8, 1.8),
  x2 = runif(35, -1.8, 1.8)
) |>
  dplyr::mutate(
    x3 = 0.6 * x1 - 0.4 * x2
  )

plane_projection <- persp(
  x = x1_grid,
  y = x2_grid,
  z = x3_grid,
  theta = 35,
  phi = 25,
  expand = 0.8,
  col = adjustcolor(
    "skyblue3",
    alpha.f = 0.45
  ),
  border = adjustcolor(
    "grey30",
    alpha.f = 0.35
  ),
  ticktype = "detailed",
  xlab = expression(x[1]),
  ylab = expression(x[2]),
  zlab = expression(x[3])
)

projected_points <- trans3d(
  x = plane_points$x1,
  y = plane_points$x2,
  z = plane_points$x3,
  pmat = plane_projection
)

points(
  projected_points,
  pch = 19,
  cex = 0.8,
  col = "dodgerblue4"
)
```

The plane is defined by

$$
x_3=0.6x_1-0.4x_2.
$$

Equivalently, its points can be written parametrically as

$$
\begin{bmatrix}
x_1 \\
x_2 \\
x_3
\end{bmatrix}
=

u
\begin{bmatrix}
1 \\
0 \\
0.6
\end{bmatrix}
+
v
\begin{bmatrix}
0 \\
1 \\
-0.4
\end{bmatrix},
\qquad u,v\in\mathbb R.
$$

There are three measured coordinates, but only two free quantities, (u) and (v).
The plane therefore is really only 2D, despite lying in 3D.

A simpler object is a line in 2D.
Suppose that the "real" distance between two points is the distance along the line.
This is the same as the straight-line distance between them, and so we can use ordinary Euclidean distance with no problem.

```{r}
#| label: prepare-straight-manifold
#| include: false

straight_tbl <- tibble::tibble(
  x = seq(-2, 2, length.out = 101)
)
straight_tbl <- straight_tbl |>
  dplyr::mutate(
    y = 2 * x + 1
  )

straight_endpoints <- tibble::tibble(
  label = c("A", "B"),
  x = c(-1.5, 1.5),
  y = 2 * c(-1.5, 1.5) + 1
)

straight_euclidean_distance <- sqrt(
  sum(
    (
      straight_endpoints[1, c("x", "y")] -
        straight_endpoints[2, c("x", "y")]
    )^2
  )
)

straight_euclidean_distance |> signif(2)
```

```{r}
#| label: fig-straight-manifold-distance
#| echo: false
#| fig-cap: "On a straight one-dimensional object, the Euclidean distance and geodesic distance are identical."

ggplot() +
  geom_path(
    data = straight_tbl,
    aes(x = x, y = y),
    linewidth = 1.4,
    colour = "grey25"
  ) +
  geom_segment(
    aes(
      x = straight_endpoints$x[1],
      y = straight_endpoints$y[1],
      xend = straight_endpoints$x[2],
      yend = straight_endpoints$y[2]
    ),
    colour = "red3",
    linewidth = 1.5,
    linetype = "dashed"
  ) +
  geom_point(
    data = straight_endpoints,
    aes(x = x, y = y),
    size = 4
  ) +
  ggrepel::geom_text_repel(
    data = straight_endpoints,
    aes(x = x, y = y, label = label),
    nudge_y = 0.16,
    fontface = "bold"
  ) +
  # annotate(
  #   "text",
  #   x = 0,
  #   y = -0.18,
  #   label = paste0(
  #     "Euclidean distance = geodesic distance = ",
  #     round(straight_euclidean_distance, 2)
  #   ),
  #   colour = "red3"
  # ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") + 
  theme(
    # axis.text = element_blank(),
    # axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

Suppose instead that the line was not straight, e.g.

$$
y=x^2.
$$

```{r}
#| label: prepare-curved-manifold-distances
#| include: false

curve_tbl <- tibble::tibble(
  x = seq(-1.5, 1.5, length.out = 501),
  y = x^2
)

curve_endpoints <- tibble::tibble(
  label = c("A", "B"),
  x = c(-1, 1),
  y = c(1, 1)
)

# Sampled observations on the part of the curve between A and B.
curve_sample_tbl <- tibble::tibble(
  x = seq(-1, 1, length.out = 13),
  y = x^2
)

curve_euclidean_distance <- sqrt(
  diff(curve_endpoints$x)^2 +
    diff(curve_endpoints$y)^2
)

# Exact arc length of y = x^2 between x = -1 and x = 1.
curve_geodesic_distance <- integrate(
  function(x) {
    sqrt(1 + (2 * x)^2)
  },
  lower = -1,
  upper = 1
)$value

# Approximation obtained by joining neighbouring sampled points.
curve_graph_distance <- sum(
  sqrt(
    diff(curve_sample_tbl$x)^2 +
      diff(curve_sample_tbl$y)^2
  )
)

tibble::tibble(
  distance = c(
    "Ambient Euclidean distance",
    "Exact geodesic distance",
    "Sample-based graph approximation"
  ),
  value = c(
    curve_euclidean_distance,
    curve_geodesic_distance,
    curve_graph_distance
  )
)
```

```{r}
#| label: fig-curved-manifold-distance
#| echo: false
#| fig-cap: "The straight-line Euclidean distance cuts through the space. The geodesic distance follows the curved object. The dashed polyline approximates the geodesic using sampled observations."

ggplot() +
  geom_path(
    data = curve_tbl,
    aes(x = x, y = y),
    linewidth = 1,
    colour = "grey40"
  ) +
  geom_path(
    data = dplyr::filter(
      curve_tbl,
      x >= -1,
      x <= 1
    ),
    aes(x = x, y = y),
    linewidth = 1.8,
    colour = "navy"
  ) +
  geom_path(
    data = curve_sample_tbl,
    aes(x = x, y = y),
    linewidth = 0.9,
    colour = "deepskyblue4",
    linetype = "dashed"
  ) +
  geom_point(
    data = curve_sample_tbl,
    aes(x = x, y = y),
    size = 2,
    colour = "deepskyblue4"
  ) +
  geom_segment(
    aes(
      x = curve_endpoints$x[1],
      y = curve_endpoints$y[1],
      xend = curve_endpoints$x[2],
      yend = curve_endpoints$y[2]
    ),
    colour = "red3",
    linewidth = 1.5
  ) +
  geom_point(
    data = curve_endpoints,
    aes(x = x, y = y),
    size = 4
  ) +
  geom_text(
    data = curve_endpoints,
    aes(x = x, y = y, label = label),
    nudge_y = 0.18,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = 0,
    y = 1.17,
    label = paste0(
      "Direct path (length: ",
      round(curve_euclidean_distance, 2),
      ")"
    ),
    colour = "red3"
  ) +
  annotate(
    "text",
    x = 0,
    y = 0.28,
    label = paste0(
      "Path along line (length: ",
      round(curve_geodesic_distance, 2),
      ")"
    ),
    colour = "navy"
  ) +
  coord_equal(
    xlim = c(-1.55, 1.55),
    ylim = c(-0.1, 2.35),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    # axis.text = element_blank(),
    # axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

## Calculating distances using the lower-dimensional spacial structure

The shortest distance along the curve between two points is termed the geodesic distance.

If we actually knew the formula for the curve, we could calculate the lower-dimensional distance exactly.

$$
d_M(A,B)
= 

\int_{-1}^{1}
\sqrt{1+\left(\frac{dy}{dx}\right)^2},dx
= 

\int_{-1}^{1}\sqrt{1+4x^2},dx
\approx 2.96.
$$

If we did not actually know the full formula for the curve, but only knew its value at specific, short intervals, then we could approximate the geodesic distance by joining the points with short straight-line segments.

For example, suppose that we had sampled the curve at 10 evenly-spaced points joining A and B.
Then, the curve would look like this:

```{r}
#| label: fig-curved-manifold-distance-approx
#| echo: false
#| fig-cap: "The subset of sampled points along the curve. The dashed line approximates the geodesic using these sampled observations."

x_vec_approx <- seq(-1, 1, length.out = 5)
y_vec_approx <- approx(
  x = curve_tbl$x,
  y = curve_tbl$y,
  xout = x_vec_approx
)
curve_tbl_approx <- tibble::tibble(
  x = y_vec_approx$x,
  y = y_vec_approx$y
)
int_length_vec <- sqrt(
  diff(curve_tbl_approx$x)^2 +
    diff(curve_tbl_approx$y)^2
)
path_length <- sum(int_length_vec)

ggplot() +
  geom_path(
    data = curve_tbl,
    aes(x = x, y = y),
    linewidth = 1,
    colour = "grey40"
  ) +
  geom_path(
    data = curve_tbl_approx,
    aes(x = x, y = y),
    linewidth = 1.8,
    colour = "navy"
  ) +
  geom_path(
    data = curve_tbl_approx,
    aes(x = x, y = y),
    linewidth = 0.9,
    colour = "deepskyblue4",
    linetype = "dashed",
    alpha = 0.5
  ) +
  geom_point(
    data = curve_tbl_approx,
    aes(x = x, y = y),
    size = 2,
    colour = "deepskyblue4"
  ) +
  geom_segment(
    aes(
      x = curve_endpoints$x[1],
      y = curve_endpoints$y[1],
      xend = curve_endpoints$x[2],
      yend = curve_endpoints$y[2]
    ),
    colour = "red3",
    linewidth = 1.5
  ) +
  geom_point(
    data = curve_endpoints,
    aes(x = x, y = y),
    size = 4
  ) +
  geom_text(
    data = curve_endpoints,
    aes(x = x, y = y, label = label),
    nudge_y = 0.18,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = 0,
    y = 1.17,
    label = paste0(
      "Direct path (length: ",
      round(curve_euclidean_distance, 2),
      ")"
    ),
    colour = "red3"
  ) +
  annotate(
    "text",
    x = 0,
    y = 0.28,
    label = paste0(
      "Approximate path along curve (length: ",
      round(path_length, 2),
      ")"
    ),
    colour = "navy"
  ) +
  coord_equal(
    xlim = c(-1.55, 1.55),
    ylim = c(-0.1, 2.35),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    # axis.text = element_blank(),
    # axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

So, we can do a reasonable job of working out the distance along a shape even if we don't know the exact formula for it, as long as we have a sufficient number of points along the shape.

## Working out the geodesic distance when the shape is entirely unknown

In the previous example, we know that we needed to move from left to right along the x-axis to get from A to B, and so we could work out the geodesic distance by integrating along the curve.

However, in higher-dimensions, we may not know the shape of the object at all, and so we cannot pre-define the paths between points.

So, to approximate the geodesic distances between points, we're going to have to use the points themselves to define the paths.

### Graphs as discretised spaces

A **graph** is a collection of points and connections between points.
The points are called **vertices** or **nodes**.
The connections are called **edges**.

For our purposes, each vertex is an observation.
An edge means that two observations are treated as directly joined.
If an edge has a weight, the weight is the cost or distance of travelling along that edge.

Mathematically, a graph is usually written as

$$
G=(V,E),
$$

where $V$ is the set of vertices and $E$ is the set of edges.
For example,

$$
V=\{A,B,C,D,E\}
$$

and

$$
E=\{(A,B),(B,C),(C,D),(D,E)\}.
$$

The same information can be written as an adjacency list:

$$
\begin{aligned}
A&: B,\\
B&: A,C,\\
C&: B,D,\\
D&: C,E,\\
E&: D.
\end{aligned}
$$

This means that $A$ is connected to $B$, $B$ is connected to $C$, and so on.
It does **not** mean that $A$ is directly connected to $E$.
To travel from $A$ to $E$, we must move along the chain

$$
A \to B \to C \to D \to E.
$$

```{r}
#| label: prepare-simple-graph
#| include: false

simple_vertex_tbl <- tibble::tibble(
  label = c("A", "B", "C", "D", "E"),
  x = c(0.0, 1.0, 2.0, 3.0, 4.0),
  y = c(0.0, 0.4, 0.1, 0.7, 0.3)
)

simple_edge_tbl <- tibble::tribble(
  ~from, ~to,
  "A", "B",
  "B", "C",
  "C", "D",
  "D", "E"
)

make_edge_segment_tbl <- function(vertex_tbl, edge_tbl) {
  edge_tbl |>
    dplyr::left_join(
      vertex_tbl |>
        dplyr::rename(
          from = label,
          x_from = x,
          y_from = y
        ),
      by = "from"
    ) |>
    dplyr::left_join(
      vertex_tbl |>
        dplyr::rename(
          to = label,
          x_to = x,
          y_to = y
        ),
      by = "to"
    )
}

simple_edge_segment_tbl <- make_edge_segment_tbl(
  vertex_tbl = simple_vertex_tbl,
  edge_tbl = simple_edge_tbl
)
```

```{r}
#| label: fig-simple-graph
#| echo: false
#| fig-cap: "A simple graph. The only permitted moves are along the displayed edges."
#| fig-width: 8
#| fig-height: 4

ggplot() +
  geom_segment(
    data = simple_edge_segment_tbl,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 1.2,
    colour = "grey35"
  ) +
  geom_point(
    data = simple_vertex_tbl,
    aes(x = x, y = y),
    size = 4,
    colour = "deepskyblue4"
  ) +
  geom_text(
    data = simple_vertex_tbl,
    aes(x = x, y = y, label = label),
    nudge_y = 0.22,
    fontface = "bold"
  ) +
  coord_equal(
    xlim = c(-0.25, 4.25),
    ylim = c(-0.3, 1.15),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

In this graph, the distance from $A$ to $E$ is not the length of the straight line from $A$ to $E$.
There is no direct edge from $A$ to $E$.
The relevant distance is the shortest distance obtained by travelling through connected points.

If the edges have weights $w_{ij}$, the graph distance between vertices $i$ and $j$ is

$$
d_G(i,j)
=
\min_{\mathcal P:i\leadsto j}
\sum_{(r,s)\in\mathcal P} w_{rs},
$$

where $\mathcal P$ ranges over all paths from $i$ to $j$.

So, a graph turns the continuous space into a discretised space.
Instead of being able to move in any direction, we can move only by stepping from one sampled point to another sampled point.

### Complete graphs

A **complete graph** connects every point to every other point.
If there are $n$ observations, then every pair is joined by an edge.

For an undirected complete graph,

$$
E
=
\{(i,j):1\leq i<j\leq n\}.
$$

The number of edges is therefore

$$
\frac{n(n-1)}{2}.
$$

For example, with five observations there are

$$
\frac{5(5-1)}{2}=10
$$

edges.

```{r}
#| label: prepare-complete-graph-example
#| include: false

make_complete_edge_tbl <- function(vertex_tbl) {
  edge_mat <- t(
    utils::combn(
      vertex_tbl$label,
      2
    )
  )

  tibble::tibble(
    from = edge_mat[, 1],
    to = edge_mat[, 2]
  )
}

complete_edge_tbl <- make_complete_edge_tbl(
  simple_vertex_tbl
)

complete_edge_segment_tbl <- make_edge_segment_tbl(
  vertex_tbl = simple_vertex_tbl,
  edge_tbl = complete_edge_tbl
)
```

```{r}
#| label: fig-complete-graph
#| echo: false
#| fig-cap: "A complete graph on five observations. Every point is directly connected to every other point."
#| fig-width: 8
#| fig-height: 4

ggplot() +
  geom_segment(
    data = complete_edge_segment_tbl,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 0.8,
    alpha = 0.45,
    colour = "grey35"
  ) +
  geom_point(
    data = simple_vertex_tbl,
    aes(x = x, y = y),
    size = 4,
    colour = "deepskyblue4"
  ) +
  geom_text(
    data = simple_vertex_tbl,
    aes(x = x, y = y, label = label),
    nudge_y = 0.22,
    fontface = "bold"
  ) +
  coord_equal(
    xlim = c(-0.25, 4.25),
    ylim = c(-0.3, 1.15),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

A complete graph is the most permissive possible graph.
There are no restrictions on which pairs can be joined directly.

This is fine when the lower-dimensional object is a straight line.
Every direct connection between two sampled points still lies on the line, so the direct Euclidean distance agrees with the distance along the object.

```{r}
#| label: prepare-complete-graph-straight-and-curved
#| include: false

straight_complete_points <- tibble::tibble(
  label = paste0("P", seq_len(9)),
  x = seq(-1, 1, length.out = 9)
) |>
  dplyr::mutate(
    y = 0,
    display_label = dplyr::case_when(
      dplyr::row_number() == 1 ~ "A",
      dplyr::row_number() == dplyr::n() ~ "B",
      TRUE ~ ""
    )
  )

curved_complete_points <- tibble::tibble(
  label = paste0("P", seq_len(9)),
  x = seq(-1, 1, length.out = 9)
) |>
  dplyr::mutate(
    y = x^2,
    display_label = dplyr::case_when(
      dplyr::row_number() == 1 ~ "A",
      dplyr::row_number() == dplyr::n() ~ "B",
      TRUE ~ ""
    )
  )

straight_complete_edges <- make_complete_edge_tbl(
  straight_complete_points
)

curved_complete_edges <- make_complete_edge_tbl(
  curved_complete_points
)

straight_complete_segments <- make_edge_segment_tbl(
  vertex_tbl = straight_complete_points,
  edge_tbl = straight_complete_edges
)

curved_complete_segments <- make_edge_segment_tbl(
  vertex_tbl = curved_complete_points,
  edge_tbl = curved_complete_edges
)

straight_adjacent_edges <- tibble::tibble(
  from = straight_complete_points$label[-length(straight_complete_points$label)],
  to = straight_complete_points$label[-1]
)

curved_adjacent_edges <- tibble::tibble(
  from = curved_complete_points$label[-length(curved_complete_points$label)],
  to = curved_complete_points$label[-1]
)

straight_adjacent_segments <- make_edge_segment_tbl(
  vertex_tbl = straight_complete_points,
  edge_tbl = straight_adjacent_edges
)

curved_adjacent_segments <- make_edge_segment_tbl(
  vertex_tbl = curved_complete_points,
  edge_tbl = curved_adjacent_edges
)

straight_complete_endpoint_distance <- sqrt(
  diff(straight_complete_points$x[c(1, 9)])^2 +
    diff(straight_complete_points$y[c(1, 9)])^2
)

straight_adjacent_path_distance <- sum(
  sqrt(
    diff(straight_complete_points$x)^2 +
      diff(straight_complete_points$y)^2
  )
)

curved_complete_endpoint_distance <- sqrt(
  diff(curved_complete_points$x[c(1, 9)])^2 +
    diff(curved_complete_points$y[c(1, 9)])^2
)

curved_adjacent_path_distance <- sum(
  sqrt(
    diff(curved_complete_points$x)^2 +
      diff(curved_complete_points$y)^2
  )
)

complete_graph_distance_tbl <- tibble::tibble(
  object = c(
    "Straight line",
    "Straight line",
    "Curved line",
    "Curved line"
  ),
  graph_used = c(
    "Complete graph",
    "Adjacent-neighbour graph",
    "Complete graph",
    "Adjacent-neighbour graph"
  ),
  distance_from_A_to_B = c(
    straight_complete_endpoint_distance,
    straight_adjacent_path_distance,
    curved_complete_endpoint_distance,
    curved_adjacent_path_distance
  )
)
```

```{r}
#| label: fig-complete-graph-straight-line
#| echo: false
#| fig-cap: "A complete graph on sampled points from a straight line. Direct connections do not leave the lower-dimensional object."
#| fig-width: 8
#| fig-height: 3.5

ggplot() +
  geom_segment(
    data = straight_complete_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 0.8,
    alpha = 0.35,
    colour = "grey35"
  ) +
  geom_segment(
    data = straight_adjacent_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 1.8,
    colour = "navy"
  ) +
  geom_point(
    data = straight_complete_points,
    aes(x = x, y = y),
    size = 3,
    colour = "deepskyblue4"
  ) +
  geom_text(
    data = straight_complete_points |>
      dplyr::filter(display_label != ""),
    aes(x = x, y = y, label = display_label),
    nudge_y = 0.18,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = 0,
    y = -0.22,
    label = paste0(
      "Complete-graph distance = adjacent-path distance = ",
      round(straight_complete_endpoint_distance, 2)
    ),
    colour = "navy"
  ) +
  coord_equal(
    xlim = c(-1.15, 1.15),
    ylim = c(-0.35, 0.35),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

For the straight line, the complete graph does not introduce an artificial shortcut.
The direct edge from $A$ to $B$ lies on the same one-dimensional object as the sampled points.

The situation changes when the line is curved.
The complete graph connects points that are close in the ambient space, even if the direct edge between them cuts across the lower-dimensional object.

```{r}
#| label: fig-complete-graph-curved-line
#| echo: false
#| fig-cap: "A complete graph on sampled points from a curved line. Some direct connections cut across the curve, creating artificial shortcuts."
#| fig-width: 8
#| fig-height: 5

ggplot() +
  geom_path(
    data = curve_tbl,
    aes(x = x, y = y),
    linewidth = 1.0,
    colour = "grey45"
  ) +
  geom_segment(
    data = curved_complete_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 0.7,
    alpha = 0.25,
    colour = "grey35"
  ) +
  geom_segment(
    aes(
      x = curved_complete_points$x[1],
      y = curved_complete_points$y[1],
      xend = curved_complete_points$x[9],
      yend = curved_complete_points$y[9]
    ),
    linewidth = 1.6,
    colour = "red3"
  ) +
  geom_segment(
    data = curved_adjacent_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 1.8,
    colour = "navy"
  ) +
  geom_point(
    data = curved_complete_points,
    aes(x = x, y = y),
    size = 3,
    colour = "deepskyblue4"
  ) +
  geom_text(
    data = curved_complete_points |>
      dplyr::filter(display_label != ""),
    aes(x = x, y = y, label = display_label),
    nudge_y = 0.15,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = 0,
    y = 1.15,
    label = paste0(
      "Complete-graph shortcut = ",
      round(curved_complete_endpoint_distance, 2)
    ),
    colour = "red3"
  ) +
  annotate(
    "text",
    x = 0,
    y = 0.22,
    label = paste0(
      "Adjacent-neighbour path = ",
      round(curved_adjacent_path_distance, 2)
    ),
    colour = "navy"
  ) +
  coord_equal(
    xlim = c(-1.15, 1.15),
    ylim = c(-0.1, 1.35),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

```{r}
#| label: tbl-complete-graph-distances
#| echo: false

complete_graph_distance_tbl |>
  dplyr::mutate(
    distance_from_A_to_B = round(
      distance_from_A_to_B,
      3
    )
  )
```

The complete graph is therefore not what we want for ISOMAP.
If every pair of points is connected, then the shortest path between two observations is usually just the direct edge between them.
That brings us back to ordinary Euclidean distances.

For a flat or straight object, this is harmless.
For a curved, folded, or obstructed object, it is damaging.
The graph creates shortcuts through the surrounding space, rather than forcing paths to remain on the lower-dimensional structure.

So ISOMAP does **not** use a complete graph.
Instead, it tries to connect only points that are close enough for local Euclidean distance to be trustworthy.
That gives a sparse neighbourhood graph.
Long-range distances are then built up by travelling through chains of local connections.

### Neighbourhood graphs

This is where the neighbourhood graph comes in.

In a neighbourhood graph, each point is connected only to a subset of the other points.

Typically, there are two major approaches:

- k-nearest neighbours (kNN): each point is connected to its k closest points.
- Epsilon-radius ($\epsilon$-radius): each point is connected to all points within a distance $\epsilon$.

The $\epsilon$-radius approach is attractive in that points are consistently connected only if they are close enough to be trusted.

The difficulty is in defining what an appropriate $\epsilon$ is, which becomes increasingly challenging in high dimensions.
Moreover, if the density of points varies across the space, then a single $\epsilon$ may be too small in some regions and too large in others.

Therefore, a kNN approach is often preferred.
In a kNN graph, each point is connected to its k closest points.

### k-nearest-neighbour graphs

A **k-nearest-neighbour graph** connects each observation to its $k$ closest observations, measured using ordinary Euclidean distance in the input space.

The important point is that Euclidean distance is used only **locally**.
We trust the straight-line distance between very nearby observations, but we do not trust the straight-line distance between arbitrary distant observations.

For a collection of points $\mathbf x_1,\ldots,\mathbf x_n$, the basic construction is:

1. calculate all pairwise Euclidean distances;
2. for each point, find the $k$ nearest other points;
3. join those pairs by edges;
4. weight each edge by the corresponding Euclidean distance;
5. estimate long-range distances by shortest paths through the graph.

Here is the construction directly in R.

```{r}
#| label: code-construct-knn-graph
#| eval: false

construct_knn_graph <- function(point_tbl, k) {
  coords <- as.matrix(
    point_tbl[, c("x", "y")]
  )

  n <- nrow(coords)
  distance_mat <- as.matrix(
    dist(coords)
  )

  search_mat <- distance_mat
  diag(search_mat) <- Inf

  adjacent <- matrix(
    FALSE,
    nrow = n,
    ncol = n
  )

  for (i in seq_len(n)) {
    neighbours <- order(search_mat[i, ])[seq_len(k)]
    adjacent[i, neighbours] <- TRUE
  }

  # Treat the graph as undirected.
  # This connects i and j if either one regards the other as a neighbour.
  adjacent <- adjacent | t(adjacent)

  edge_index <- which(
    adjacent & upper.tri(adjacent),
    arr.ind = TRUE
  )

  edge_tbl <- tibble::tibble(
    from = as.character(edge_index[, 1]),
    to = as.character(edge_index[, 2]),
    weight = distance_mat[edge_index]
  )

  igraph::graph_from_data_frame(
    edge_tbl,
    directed = FALSE,
    vertices = tibble::tibble(
      name = as.character(seq_len(n))
    )
  )
}
```

Now apply this to the curved one-dimensional example.
The true lower-dimensional distance from $A$ to $B$ is the arc length of the curve.
The graph distance estimates this by travelling through neighbouring sampled points.

```{r}
#| label: prepare-knn-curve-graph
#| include: false

construct_knn_graph_for_curve <- function(point_tbl, k) {
  coords <- as.matrix(
    point_tbl[, c("x", "y")]
  )

  n <- nrow(coords)
  distance_mat <- as.matrix(
    dist(coords)
  )

  search_mat <- distance_mat
  diag(search_mat) <- Inf

  adjacent <- matrix(
    FALSE,
    nrow = n,
    ncol = n
  )

  for (i in seq_len(n)) {
    neighbours <- order(search_mat[i, ])[seq_len(k)]
    adjacent[i, neighbours] <- TRUE
  }

  adjacent <- adjacent | t(adjacent)

  edge_index <- which(
    adjacent & upper.tri(adjacent),
    arr.ind = TRUE
  )

  edge_tbl <- tibble::tibble(
    from = as.character(edge_index[, 1]),
    to = as.character(edge_index[, 2]),
    weight = distance_mat[edge_index]
  )

  graph <- igraph::graph_from_data_frame(
    edge_tbl,
    directed = FALSE,
    vertices = tibble::tibble(
      name = as.character(seq_len(n))
    )
  )

  list(
    graph = graph,
    edges = edge_tbl,
    distance_mat = distance_mat
  )
}

make_weighted_edge_segment_tbl <- function(point_tbl, edge_tbl) {
  point_tbl <- point_tbl |>
    dplyr::mutate(
      name = as.character(dplyr::row_number())
    )

  edge_tbl |>
    dplyr::left_join(
      point_tbl |>
        dplyr::rename(
          from = name,
          x_from = x,
          y_from = y
        ),
      by = "from"
    ) |>
    dplyr::left_join(
      point_tbl |>
        dplyr::rename(
          to = name,
          x_to = x,
          y_to = y
        ),
      by = "to"
    )
}

curve_graph_points <- tibble::tibble(
  x = seq(-1, 1, length.out = 17)
) |>
  dplyr::mutate(
    y = x^2,
    label = dplyr::case_when(
      dplyr::row_number() == 1 ~ "A",
      dplyr::row_number() == dplyr::n() ~ "B",
      TRUE ~ ""
    )
  )

knn_k <- 2
knn_curve_graph <- construct_knn_graph_for_curve(
  point_tbl = curve_graph_points,
  k = knn_k
)

knn_curve_segments <- make_weighted_edge_segment_tbl(
  point_tbl = curve_graph_points,
  edge_tbl = knn_curve_graph$edges
)

knn_shortest_vertices <- igraph::shortest_paths(
  knn_curve_graph$graph,
  from = "1",
  to = as.character(nrow(curve_graph_points)),
  weights = igraph::E(knn_curve_graph$graph)$weight,
  output = "vpath"
)$vpath[[1]] |>
  as.character()

knn_shortest_edge_tbl <- tibble::tibble(
  from = knn_shortest_vertices[-length(knn_shortest_vertices)],
  to = knn_shortest_vertices[-1]
)

knn_shortest_segments <- make_weighted_edge_segment_tbl(
  point_tbl = curve_graph_points,
  edge_tbl = knn_shortest_edge_tbl
)

knn_graph_distance <- igraph::distances(
  knn_curve_graph$graph,
  v = "1",
  to = as.character(nrow(curve_graph_points)),
  weights = igraph::E(knn_curve_graph$graph)$weight
)[1, 1]
```

```{r}
#| label: fig-knn-curve-graph
#| echo: false
#| fig-cap: "A k-nearest-neighbour graph on sampled points from the curve. The shortest path through local edges estimates the geodesic distance."
#| fig-width: 8
#| fig-height: 5

ggplot() +
  geom_path(
    data = curve_tbl,
    aes(x = x, y = y),
    linewidth = 1,
    colour = "grey45"
  ) +
  geom_segment(
    data = knn_curve_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 0.8,
    alpha = 0.35,
    colour = "grey35"
  ) +
  geom_segment(
    data = knn_shortest_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 1.7,
    colour = "navy"
  ) +
  geom_segment(
    aes(
      x = curve_graph_points$x[1],
      y = curve_graph_points$y[1],
      xend = curve_graph_points$x[nrow(curve_graph_points)],
      yend = curve_graph_points$y[nrow(curve_graph_points)]
    ),
    linewidth = 1.3,
    colour = "red3",
    linetype = "dashed"
  ) +
  geom_point(
    data = curve_graph_points,
    aes(x = x, y = y),
    size = 2.6,
    colour = "deepskyblue4"
  ) +
  geom_text(
    data = curve_graph_points |>
      dplyr::filter(label != ""),
    aes(x = x, y = y, label = label),
    nudge_y = 0.13,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = 0,
    y = 1.17,
    label = paste0(
      "Straight-line distance = ",
      round(curve_euclidean_distance, 2)
    ),
    colour = "red3"
  ) +
  annotate(
    "text",
    x = 0,
    y = 0.23,
    label = paste0(
      "kNN graph distance = ",
      round(knn_graph_distance, 2),
      "\ntrue geodesic = ",
      round(curve_geodesic_distance, 2)
    ),
    colour = "navy"
  ) +
  coord_equal(
    xlim = c(-1.15, 1.15),
    ylim = c(-0.1, 1.35),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

The estimate improves when the graph is connected and local.
If $k$ is too small, the graph may break into components.
If $k$ is too large, the graph begins to add shortcuts across the curve.

```{r}
#| label: tbl-knn-geodesic-accuracy
#| echo: false

estimate_knn_distance <- function(k) {
  graph_info <- construct_knn_graph_for_curve(
    point_tbl = curve_graph_points,
    k = k
  )

  estimate <- igraph::distances(
    graph_info$graph,
    v = "1",
    to = as.character(nrow(curve_graph_points)),
    weights = igraph::E(graph_info$graph)$weight
  )[1, 1]

  tibble::tibble(
    graph_distance = estimate,
    absolute_error = abs(estimate - curve_geodesic_distance),
    connected = is.finite(estimate)
  )
}

tibble::tibble(
  k = c(1, 2, 3, 5, 8, 16)
) |>
  dplyr::mutate(
    result = purrr::map(k, estimate_knn_distance)
  ) |>
  tidyr::unnest(result) |>
  dplyr::mutate(
    graph_distance = round(graph_distance, 3),
    absolute_error = round(absolute_error, 3)
  )
```

The goal is not to make $k$ as large as possible.
The goal is to choose $k$ large enough to keep the graph connected, but small enough that edges remain local.

### Epsilon-neighbourhood graphs

An **epsilon-neighbourhood graph** uses a distance threshold instead of a fixed number of neighbours.
Two observations are joined if their Euclidean distance is at most $\epsilon$.

$$
(i,j)\in E
\quad\text{when}\quad
\|\mathbf x_i-\mathbf x_j\|\leq \epsilon.
$$

Here is the corresponding R construction.

```{r}
#| label: code-construct-epsilon-graph
#| eval: false

construct_epsilon_graph <- function(point_tbl, epsilon) {
  coords <- as.matrix(
    point_tbl[, c("x", "y")]
  )

  n <- nrow(coords)
  distance_mat <- as.matrix(
    dist(coords)
  )

  adjacent <- distance_mat <= epsilon
  diag(adjacent) <- FALSE

  edge_index <- which(
    adjacent & upper.tri(adjacent),
    arr.ind = TRUE
  )

  edge_tbl <- tibble::tibble(
    from = as.character(edge_index[, 1]),
    to = as.character(edge_index[, 2]),
    weight = distance_mat[edge_index]
  )

  igraph::graph_from_data_frame(
    edge_tbl,
    directed = FALSE,
    vertices = tibble::tibble(
      name = as.character(seq_len(n))
    )
  )
}
```

This is conceptually simple: every edge is local because every edge must have length at most $\epsilon$.
The difficulty is that a single value of $\epsilon$ may be too small in sparse regions and too large in dense regions.

```{r}
#| label: prepare-epsilon-curve-graph
#| include: false

construct_epsilon_graph_for_curve <- function(point_tbl, epsilon) {
  coords <- as.matrix(
    point_tbl[, c("x", "y")]
  )

  n <- nrow(coords)
  distance_mat <- as.matrix(
    dist(coords)
  )

  adjacent <- distance_mat <= epsilon
  diag(adjacent) <- FALSE

  edge_index <- which(
    adjacent & upper.tri(adjacent),
    arr.ind = TRUE
  )

  edge_tbl <- tibble::tibble(
    from = as.character(edge_index[, 1]),
    to = as.character(edge_index[, 2]),
    weight = distance_mat[edge_index]
  )

  graph <- igraph::graph_from_data_frame(
    edge_tbl,
    directed = FALSE,
    vertices = tibble::tibble(
      name = as.character(seq_len(n))
    )
  )

  list(
    graph = graph,
    edges = edge_tbl,
    distance_mat = distance_mat
  )
}

epsilon_value <- 0.27
epsilon_curve_graph <- construct_epsilon_graph_for_curve(
  point_tbl = curve_graph_points,
  epsilon = epsilon_value
)

epsilon_curve_segments <- make_weighted_edge_segment_tbl(
  point_tbl = curve_graph_points,
  edge_tbl = epsilon_curve_graph$edges
)

epsilon_shortest_vertices <- igraph::shortest_paths(
  epsilon_curve_graph$graph,
  from = "1",
  to = as.character(nrow(curve_graph_points)),
  weights = igraph::E(epsilon_curve_graph$graph)$weight,
  output = "vpath"
)$vpath[[1]] |>
  as.character()

epsilon_shortest_edge_tbl <- tibble::tibble(
  from = epsilon_shortest_vertices[-length(epsilon_shortest_vertices)],
  to = epsilon_shortest_vertices[-1]
)

epsilon_shortest_segments <- make_weighted_edge_segment_tbl(
  point_tbl = curve_graph_points,
  edge_tbl = epsilon_shortest_edge_tbl
)

epsilon_graph_distance <- igraph::distances(
  epsilon_curve_graph$graph,
  v = "1",
  to = as.character(nrow(curve_graph_points)),
  weights = igraph::E(epsilon_curve_graph$graph)$weight
)[1, 1]
```

```{r}
#| label: fig-epsilon-curve-graph
#| echo: false
#| fig-cap: "An epsilon-neighbourhood graph on sampled points from the curve. Edges are included only when the ambient Euclidean distance is below the threshold."
#| fig-width: 8
#| fig-height: 5

ggplot() +
  geom_path(
    data = curve_tbl,
    aes(x = x, y = y),
    linewidth = 1,
    colour = "grey45"
  ) +
  geom_segment(
    data = epsilon_curve_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 0.8,
    alpha = 0.35,
    colour = "grey35"
  ) +
  geom_segment(
    data = epsilon_shortest_segments,
    aes(
      x = x_from,
      y = y_from,
      xend = x_to,
      yend = y_to
    ),
    linewidth = 1.7,
    colour = "navy"
  ) +
  geom_segment(
    aes(
      x = curve_graph_points$x[1],
      y = curve_graph_points$y[1],
      xend = curve_graph_points$x[nrow(curve_graph_points)],
      yend = curve_graph_points$y[nrow(curve_graph_points)]
    ),
    linewidth = 1.3,
    colour = "red3",
    linetype = "dashed"
  ) +
  geom_point(
    data = curve_graph_points,
    aes(x = x, y = y),
    size = 2.6,
    colour = "deepskyblue4"
  ) +
  geom_text(
    data = curve_graph_points |>
      dplyr::filter(label != ""),
    aes(x = x, y = y, label = label),
    nudge_y = 0.13,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = 0,
    y = 1.17,
    label = paste0(
      "Straight-line distance = ",
      round(curve_euclidean_distance, 2)
    ),
    colour = "red3"
  ) +
  annotate(
    "text",
    x = 0,
    y = 0.23,
    label = paste0(
      "epsilon graph distance = ",
      round(epsilon_graph_distance, 2),
      "\ntrue geodesic = ",
      round(curve_geodesic_distance, 2)
    ),
    colour = "navy"
  ) +
  coord_equal(
    xlim = c(-1.15, 1.15),
    ylim = c(-0.1, 1.35),
    expand = FALSE
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white")
  )
```

As with $k$, the value of $\epsilon$ has to balance two errors.
If $\epsilon$ is too small, the graph may be disconnected.
If $\epsilon$ is too large, the graph may add shortcuts and move back toward ordinary Euclidean distances.

```{r}
#| label: tbl-epsilon-geodesic-accuracy
#| echo: false

estimate_epsilon_distance <- function(epsilon) {
  graph_info <- construct_epsilon_graph_for_curve(
    point_tbl = curve_graph_points,
    epsilon = epsilon
  )

  estimate <- igraph::distances(
    graph_info$graph,
    v = "1",
    to = as.character(nrow(curve_graph_points)),
    weights = igraph::E(graph_info$graph)$weight
  )[1, 1]

  tibble::tibble(
    graph_distance = estimate,
    absolute_error = abs(estimate - curve_geodesic_distance),
    connected = is.finite(estimate)
  )
}

tibble::tibble(
  epsilon = c(0.15, 0.27, 0.50, 2.10)
) |>
  dplyr::mutate(
    result = purrr::map(epsilon, estimate_epsilon_distance)
  ) |>
  tidyr::unnest(result) |>
  dplyr::mutate(
    graph_distance = round(graph_distance, 3),
    absolute_error = round(absolute_error, 3)
  )
```

### What both neighbourhood graphs are trying to do

Both constructions are based on the same idea.

Short straight-line distances are usually acceptable local approximations to movement along the underlying object.
Long straight-line distances are not.

So ISOMAP replaces one global distance by many small local distances:

$$
\|\mathbf x_i-\mathbf x_j\|
\quad\leadsto\quad
\min_{\mathcal P:i\leadsto j}
\sum_{(r,s)\in\mathcal P}
\|\mathbf x_r-\mathbf x_s\|.
$$

The left-hand side is the direct Euclidean distance through the input space.
The right-hand side is the graph shortest-path distance, obtained by travelling through local edges.

That shortest-path distance is not the true geodesic distance.
It is an estimate of the geodesic distance based on the sampled observations.
With enough data, and with a neighbourhood graph that is local but connected, this estimate can be much closer to the true lower-dimensional distance than the direct Euclidean distance.
