---
title: Tidy data for analysis for STA5069Z course
format:
  html:
    embed-resources: true
---

# Introduction

```{r}
#| include: false
library(tibble)
```


## MDS

### Morse code data

```{r}
data_raw_morse <- readr::read_csv(
  projr::projr_path_get("raw-data", "morse.code.csv")
)
data_tidy_morse <- data_raw_morse
projr::projr_use_data(data_tidy_morse, overwrite = TRUE)
```

### SA distance data

```{r}
data_raw_sa_distance <- openxlsx::read.xlsx(
  projr::projr_path_get("raw-data", "SAdistances.xlsx"),
  sheet = 2
) |>
  tibble::as_tibble()
data_tidy_sa_distance <- data_raw_sa_distance
colnames(data_tidy_sa_distance)[which(colnames(data_tidy_sa_distance) == "CapeTOwn")] <- "CapeTown"
projr::projr_use_data(data_tidy_sa_distance, overwrite = TRUE)
```

## Cluster analysis

### Food

```{r}
data_raw_food <- readr::read_table(
  projr::projr_path_get("raw-data", "food.txt")
)
data_tidy_food <- data_raw_food
ugly_to_display_vec <- ugly_to_display_vec <- c(
  "\"Fat.grams\"" = "fat_grams",
  "\"Food.energy.calories\"" = "food_energy_calories",
  "\"Carbohydrates.grams\"" = "carbohydrates_grams",
  "\"Protein.grams\"" = "protein_grams",
  "\"Cholesterol.mg\"" = "cholesterol_mg",
  "\"weight.grams\"" = "weight_grams",
  "\"Saturated.fat.grams\"" = "saturated_fat_grams",
  "\"cluster.id\"" = "cluster_id",
  "\"Fat2.grams\"" = "fat2_grams",
  "\"Food.energy2.calories\"" = "food_energy2_calories",
  "\"Carbohydrates2.grams\"" = "carbohydrates2_grams",
  "\"Protein2.grams\"" = "protein2_grams",
  "\"Cholesterol2.mg\"" = "cholesterol2_mg",
  "\"Saturated.fat2.grams\"" = "saturated_fat2_grams"
)
colnames(data_tidy_food) <- ugly_to_display_vec[colnames(data_tidy_food)]
projr::projr_use_data(data_tidy_food, overwrite = TRUE)
```

## Example 12.3.4

```{r}
data_raw_example_12_3_4 <- readr::read_csv(
  projr::projr_path_get("raw-data", "Ex12.3.4.csv")
)
data_tidy_example_12_3_4 <- data_raw_example_12_3_4
projr::projr_use_data(data_tidy_example_12_3_4, overwrite = TRUE)
```

## Primate scapulae

```{r}
data_raw_primate <- readr::read_csv(
  projr::projr_path_get("raw-data", "primate.scapulae.csv")
)
data_tidy_primate <- data_raw_primate |>
  dplyr::rename(
    genus = "genus",
    AD_BD = "AD.BD",
    AD_CD = "AD.CD",
    EA_CD = "EA.CD",
    Dx_CD = "Dx.CD",
    SH_ACR = "SH.ACR",
    EAD = "EAD",
    beta = "beta",
    gamma = "gamma",
    class = "class",
    classdigit = "classdigit"
  )
projr::projr_use_data(data_tidy_primate, overwrite = TRUE)
```

## Satellite images

```{r}
data_raw_satellite <- readr::read_table(
  projr::projr_path_get("raw-data", "satimage.txt")
)
data_tidy_satellite <- data_raw_satellite
colnames(data_tidy_satellite) <- gsub("\"", "", colnames(data_tidy_satellite))
projr::projr_use_data(data_tidy_satellite, overwrite = TRUE)
```

## T11-5

```{r}
data_raw_t11_5 <- readr::read_table(
  projr::projr_path_get("raw-data", "T11-5.DAT"),
  col_names = FALSE
)
data_tidy_t11_5 <- data_raw_t11_5
projr::projr_use_data(data_tidy_t11_5, overwrite = TRUE)
```

## Linear dimensionality reduction

### Financial data

Hirschey and Wichern (referenced in Johnson and Wichern) investigate the consistency, determinants and uses of accounting and market-value measures of profitability.

The correlation matrix of accounting historical, accounting replacement and market-value measures of profitability for a sample of firms operating in 1977 is as follows:

```{r}
#| results: asis
#| echo: false
var_vec <- c(
  "Historical return on assets (HRA)",
  "Historical return on equity (HRE)",
  "Historical return on sales (HRS)",
  "Replacement return on assets (RA)",
  "Replacement return on equity (RRE)",
  "Replacement return on sales (RRS)",
  "Market Q ratio (Q)",
  "Market relative excess value (REV)"
  )
hra_vec <- c(1, 0.738, 0.731, 0.828, 0.681, 0.712, 0.625, 0.604)
hre_vec <- c(NA_real_, 1, 0.52, 0.688, 0.831, 0.543, 0.322, 0.303)
hrs_vec <- c(NA_real_, NA_real_, 1, 0.652, 0.5113, 0.826, 0.579, 0.617)
rra_vec <- c(NA_real_, NA_real_, NA_real_, 1, 0.887, 0.867, 0.639, 0.563)
rre_vec <- c(NA_real_, NA_real_, NA_real_, NA_real_, 1, 0.692, 0.419, 0.352)
rrs_vec <- c(NA_real_, NA_real_, NA_real_, NA_real_, NA_real_, 1, 0.608, 0.610)
q_vec <- c(NA_real_, NA_real_, NA_real_, NA_real_, NA_real_, NA_real_, 1, 0.937)
rev_vec <- c(NA_real_, NA_real_, NA_real_, NA_real_, NA_real_, NA_real_, NA_real_, 1)
corr_tbl <- tibble::tibble(
  Variable = var_vec,
  HRA = hra_vec,
  HRE = hre_vec,
  HRS = hrs_vec,
  RRA = rra_vec,
  RRE = rre_vec,
  RRS = rrs_vec,
  Q = q_vec,
  REV = rev_vec
)
knitr::kable(corr_tbl)
```

```{r}
#| echo: false
corr_mat <- matrix(c(
  HRA = hra_vec,
  HRE = hre_vec,
  HRS = hrs_vec,
  RRA = rra_vec,
  RRE = rre_vec,
  RRS = rrs_vec,
  Q = q_vec,
  REV = rev_vec
), nrow = 8, byrow = FALSE)
for (i in seq_len(nrow(corr_mat))) {
  for (j in seq_len(nrow(corr_mat))) {
    if (is.na(corr_mat[i, j])) corr_mat[i, j] <- corr_mat[j, i]
  }
}
```

The correlation matrix is saved as `data_tidy_profit`.

```{r}
#| eval: true
#| echo: true
data_tidy_profit <- corr_mat
projr::projr_use_data(data_tidy_profit, overwrite = TRUE)
```

### Luminex data

```{r}
#| echo: true
#| results: hide
#| message: false
#| warning: false
lum_tbl <- openxlsx::read.xlsx(
  projr::projr_path_get("raw-data", "age_inflammation_data.xlsx"),
  sheet = "luminex"
) |>
  tibble::as_tibble()
lum_tbl <- lum_tbl |>
  dplyr::rename(
    pid = `Donor.ID`,
    stim = `Stim.condition`
  )
colnames(lum_tbl) <- gsub("\\(\\d+\\)", "", colnames(lum_tbl))
colnames(lum_tbl) <- gsub("\\.|\\-", "", colnames(lum_tbl))
lum_tbl <- lum_tbl |>
  dplyr::mutate(
    dplyr::across(IFNa2:DDimer, as.numeric)
  )
(lum_tbl[,1:2]) |> UtilsDataRSV::view_cols()
(lum_tbl) |> UtilsDataRSV::view_cols()
lum_tbl <- lum_tbl |>
  tidyr::pivot_longer(
    -c(pid, stim),
    names_to = "protein",
    values_to = "response"
  ) |>
  dplyr::mutate(
    response = log(response)
  ) |>
  dplyr::group_by(
    protein, pid
  ) |>
  dplyr::filter(
    stim %in% c("Unstim", "H37Rv")
  ) |>
  dplyr::summarise(
    response = response[stim == "H37Rv"] - response[stim == "Unstim"]
  ) |>
  dplyr::ungroup()
if (FALSE) {
  p <- ggplot(lum_tbl, aes(y = response, x = protein, fill = protein)) +
    geom_boxplot()
  ggplot2::ggsave(
    filename = projr_path_get("cache", "test", "p-lum.pdf")
  )
}
lum_tbl <- lum_tbl |>
  tidyr::pivot_wider(
    names_from = "protein",
    values_from = "response"
  )
meta_data_tbl <- openxlsx::read.xlsx(
  projr::projr_path_get("raw-data", "age_inflammation_data.xlsx"),
  sheet = "metadata"
) |>
  tibble::as_tibble()
meta_data_tbl <- meta_data_tbl |>
  dplyr::rename(
    pid = `Donor.ID`,
    infxn = `QFT.status`,
    age = Age,
    sex = Gender,
    ethnicity = Ethnicity
  )
# remove all observations from meta_data_tbl
# that are not in lum_tbl
meta_data_tbl <- meta_data_tbl |>
  dplyr::semi_join(lum_tbl, by = "pid")
projr::projr_use_data(lum_tbl, overwrite = TRUE)
projr::projr_use_data(meta_data_tbl, overwrite = TRUE)
```

## Mtb-specific T cell response data

```{r}
#| results: hide
#| message: false
#| warning: false
if (!requireNamespace("VaccCompData", quietly = TRUE)) {
  remotes::install_github("MiguelRodo/VaccCompData")
}
data("tmaxSub17ExcTbl", package = "VaccCompData")
data("bl17ExcTbl", package = "VaccCompData")

tcell_response_data_memory <- tmaxSub17ExcTbl |>
  tibble::as_tibble() |>
  dplyr::filter(cd == "4") |>
  dplyr::mutate(
    cytCombo = gsub("^CD4", "", cytCombo)
  ) |>
  dplyr::select(-c(cd, timePoint, prid)) |>
  dplyr::filter(vaccine != "8")

chr_to_vacc_vec <- c(
  "1" = "Aeras402",
  "2" = "H56:IC31",
  "3" = "M72:AS01E",
  "4" = "MVA85A",
  "5" = "H1:IC31",
  "6" = "ID93:GLA-SE",
  "7" = "BCG",
  "8" = "Megapool"
)

tcell_response_data_memory <- tcell_response_data_memory |>
  dplyr::mutate(
    vaccine = chr_to_vacc_vec[vaccine]
  )

# tcell_response_data_memory |>
#   dplyr::group_by(vaccine, ptid) |>
#   dplyr::slice(1) |>
#   dplyr::ungroup() |>
#   dplyr::group_by(vaccine) |>
#   dplyr::summarise(count = dplyr::n())

tcell_response_data_baseline <- bl17ExcTbl |>
  tibble::as_tibble() |>
  dplyr::filter(cd == "4") |>
  dplyr::mutate(
    cytCombo = gsub("^CD4", "", cytCombo)
  ) |>
  dplyr::select(-c(cd, timePoint, prid)) |>
  dplyr::filter(vaccine != "8")
tcell_response_data_baseline <- tcell_response_data_baseline |>
  dplyr::mutate(
    vaccine = chr_to_vacc_vec[vaccine]
  )

pid_vec <- intersect(
  tcell_response_data_baseline$ptid,
  tcell_response_data_memory$ptid
)
tcell_response_data_baseline <- tcell_response_data_baseline |>
  dplyr::filter(ptid %in% pid_vec)
tcell_response_data_memory <- tcell_response_data_memory |>
  dplyr::filter(ptid %in% pid_vec)
projr::projr_use_data(tcell_response_data_baseline, overwrite = TRUE)
projr::projr_use_data(tcell_response_data_memory, overwrite = TRUE)
```
