---
title: Lecture 3 - Clustering
subtitle: Graph-based clustering
bibliography: zotero.bib
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
---

```{r}
#| include: false
library(mclust)
library(ggplot2)
library(tibble)
library(igraph)
library(FNN)
library(cowplot)
library(RColorBrewer)
library(patchwork)
theme_bg <- function() {
  cowplot::theme_cowplot() +
    theme(
      plot.background = element_rect(fill = "white", colour = NA),
      panel.background = element_rect(fill = "white", colour = NA)
    )
}
# Display individually
path_tmp <- "_tmp/graph/"
if (dir.exists(path_tmp)) {
  unlink(path_tmp, recursive = TRUE)
}
dir.create(path_tmp, recursive = TRUE)
path_p_init <- file.path(path_tmp, "p-init.png")
```

# 1. Introduction: Graph Basics

A graph models pairwise relationships. It consists of nodes (data points) and edges (connections, indicating similarity or neighborhood).

A graph $G$ is formally defined as an ordered pair $G = (V, E)$:

* $V$ is a set of vertices (or nodes). In clustering, these are your data points. $V = \{v_1, v_2, \dots, v_n\}$.
* $E$ is a set of edges (or links). An edge connects two vertices. $E = \{(v_i, v_j) \mid v_i, v_j \in V\}$.

To use graphs in algorithms, we translate $V$ and $E$ into data structures:

* Adjacency Matrix ($A$): An $n \times n$ matrix where $A_{ij} = 1$ if there is an edge between $v_i$ and $v_j$, and $A_{ij} = 0$ otherwise.
* Adjacency List: A list of length $n$ where the $i$-th element contains an array of all vertices connected to $v_i$. This is highly efficient for sparse graphs (like a 3-NN graph) because you don't store thousands of zeros.

Let's simulate 15 points, connect them using a 3-Nearest Neighbour (3-NN) rule, and extract both the adjacency matrix and adjacency list for the first 5 points.

```{r}
# -------------------------------------------------------------------
# Simulate 15 points and build a 3-NN graph
# -------------------------------------------------------------------

set.seed(42)
n_pts <- 15

# Generate 15 random points
pts <- data.frame(
  id = 1:n_pts,
  x = runif(n_pts, 0, 10),
  y = runif(n_pts, 0, 10)
)

# Find 3 nearest neighbors for each point
knn_res <- FNN::get.knn(pts[, c("x", "y")], k = 3)

# Build an edge list (from point i to its 3 neighbors)
edges <- do.call(rbind, lapply(1:n_pts, function(i) {
  data.frame(from = i, to = knn_res$nn.index[i, ])
}))

# We use simplify to remove duplicate edges (e.g., if A is a 3-NN of B, and B is a 3-NN of A)
g_15 <- graph_from_data_frame(edges, directed = FALSE)
g_15 <- simplify(g_15)

# -------------------------------------------------------------------
# Plotting the 3-NN Graph
# -------------------------------------------------------------------

# Extract edge coordinates for ggplot
edge_df <- as_data_frame(g_15, what = "edges")
edge_df$from <- as.numeric(edge_df$from)
edge_df$to <- as.numeric(edge_df$to)

edge_df$x <- pts$x[edge_df$from]
edge_df$y <- pts$y[edge_df$from]
edge_df$xend <- pts$x[edge_df$to]
edge_df$yend <- pts$y[edge_df$to]

p_15nn <- ggplot() +
  geom_segment(data = edge_df, aes(x = x, y = y, xend = xend, yend = yend), color = "gray70") +
  geom_point(data = pts, aes(x = x, y = y), size = 8, color = "steelblue") +
  geom_text(data = pts, aes(x = x, y = y, label = id), color = "white", size = 3) +
  theme_bg() +
  labs(
    title = "15 Points: 3-Nearest Neighbor Graph",
    x = "X", y = "Y"
  )

# Display the plot
path_tmp <- "_tmp/graph/"
if (!dir.exists(path_tmp)) dir.create(path_tmp, recursive = TRUE)
path_p_15nn <- file.path(path_tmp, "p-15nn.png")
ggsave(path_p_15nn, p_15nn, units = "cm", height = 10, width = 12)

# -------------------------------------------------------------------
# Adjacency Matrix and List (First 5 points)
# -------------------------------------------------------------------

# 1. Adjacency Matrix
adj_matrix_15 <- as.matrix(as_adjacency_matrix(g_15))
print("Adjacency Matrix (Points 1 to 5):")
print(adj_matrix_15[1:5, 1:5])

# 2. Adjacency List
adj_list_15 <- as_adj_list(g_15)
print("Adjacency List (Points 1 to 5):")
print(adj_list_15[1:5])
```

::: {#fig-graph-example}
![](`r path_p_15nn`)

Example of a graph constructed from a distance matrix.s-
:::

# 2. Geodesic vs. global distance

As we have already seen, clusters are rarely perfect spheres; they often take on complex shapes, varying in both volume and orientation. Because of this, the global distance between two points (the straight-line, Euclidean distance) does not always reflect how far apart they "really" are within the context of the data.

If two points lie on the same curved structure but are separated by an empty gap in space, a straight line might suggest they are close. To capture their true relationship, we must rely on the distance along the underlying shape of the data itself.
This is captured by the **geodesic distance**.

Geodesic distance is the shortest possible path between two points while strictly remaining on the underlying surface (or manifold) of the space they inhabit. 

Unlike Euclidean distance, which measures the straight line between two points by cutting through empty space, geodesic distance respects the actual structural shape of the data. 

A classic analogy is traveling between two cities on Earth: the Euclidean distance is a straight line tunneling directly through the planet's core. The geodesic distance is the actual flight path across the curved surface of the globe. 

In graph-based clustering, we approximate this continuous geodesic distance by calculating the shortest path that hops from neighbor to neighbor along the discrete edges of our graph.

```{r}
#| echo: FALSE
#| warning: FALSE

k_val <- 5

# -------------------------------------------------------------------
# 1. Simulate and Plot Two 2D Normals
# -------------------------------------------------------------------

set.seed(123)
n_norm <- 400

# Cluster 1 centered at (0, 0)
norm1 <- data.frame(
  x = rnorm(n_norm, mean = 0, sd = 1),
  y = rnorm(n_norm, mean = 0, sd = 1),
  cluster = as.factor("A")
)

# Cluster 2 centered at (4, 4)
norm2 <- data.frame(
  x = rnorm(n_norm, mean = 4, sd = 1),
  y = rnorm(n_norm, mean = 4, sd = 1),
  cluster = as.factor("B")
)

normals_data <- rbind(norm1, norm2)

# Base Plot
p_normals <- ggplot(normals_data, aes(x = x, y = y, color = cluster)) +
  geom_point(alpha = 0.8, size = 1.5) +
  scale_color_brewer(palette = "Set2", name = "True Cluster") +
  theme_bg() +
  labs(
    title = "Two 2D Normals",
    x = "X coordinate",
    y = "Y coordinate"
  )

# k-NN Calculation
knn_norm <- FNN::get.knn(normals_data[, c("x", "y")], k = k_val)
edges_norm <- do.call(rbind, lapply(1:nrow(normals_data), function(i) {
  data.frame(from = i, to = knn_norm$nn.index[i, ])
}))
edges_norm_df <- data.frame(
  x = normals_data$x[edges_norm$from],
  y = normals_data$y[edges_norm$from],
  xend = normals_data$x[edges_norm$to],
  yend = normals_data$y[edges_norm$to]
)

# k-NN Plot
p_normals_knn <- ggplot() +
  geom_segment(data = edges_norm_df, aes(x = x, y = y, xend = xend, yend = yend), 
               color = "gray30", alpha = 0.8) +
  geom_point(data = normals_data, aes(x = x, y = y, color = cluster), 
             size = 1.5, alpha = 0.3) +
  scale_color_brewer(palette = "Set2", name = "True Cluster") +
  theme_bg() +
  labs(
    title = "Two 2D Normals (5-NN Graph)",
    x = "X coordinate",
    y = "Y coordinate"
  )

# -------------------------------------------------------------------
# 2. Simulate and Plot the Swiss Roll
# -------------------------------------------------------------------

n_swiss <- 1500

# The 't' variable represents the 1D position along the unrolled manifold
t <- runif(n_swiss, 1.5 * pi, 4.5 * pi)

# Generate 3D coordinates (x, y, z)
x_swiss <- t * cos(t) + rnorm(n_swiss, 0, 0.5)
y_swiss <- runif(n_swiss, 0, 20) 
z_swiss <- t * sin(t) + rnorm(n_swiss, 0, 0.5)

swiss_roll <- data.frame(x = x_swiss, y = y_swiss, z = z_swiss, t = t)

# Base Plot
p_swiss <- ggplot(swiss_roll, aes(x = x, y = z, color = t)) +
  geom_point(alpha = 0.8, size = 1.5) +
  scale_color_distiller(palette = "Spectral", name = "Manifold\nPosition") +
  theme_bg() +
  labs(
    title = "Swiss Roll (2D Projection)",
    x = "X coordinate",
    y = "Z coordinate"
  )

# k-NN Calculation (Computed in 3D, plotted in 2D)
knn_swiss <- FNN::get.knn(swiss_roll[, c("x", "y", "z")], k = k_val)
edges_swiss <- do.call(rbind, lapply(1:nrow(swiss_roll), function(i) {
  data.frame(from = i, to = knn_swiss$nn.index[i, ])
}))
edges_swiss_df <- data.frame(
  x = swiss_roll$x[edges_swiss$from],
  y = swiss_roll$z[edges_swiss$from], # Z on Y-axis
  xend = swiss_roll$x[edges_swiss$to],
  yend = swiss_roll$z[edges_swiss$to]
)

# k-NN Plot
p_swiss_knn <- ggplot() +
  geom_segment(data = edges_swiss_df, aes(x = x, y = y, xend = xend, yend = yend), 
               color = "gray30", alpha = 0.8) +
  geom_point(data = swiss_roll, aes(x = x, y = z, color = t), 
             size = 1.5, alpha = 0.3) +
  scale_color_distiller(palette = "Spectral", name = "Manifold\nPosition") +
  theme_bg() +
  labs(
    title = "Swiss Roll (3D 5-NN Graph)",
    x = "X coordinate",
    y = "Z coordinate"
  )

# -------------------------------------------------------------------
# 3. Simulate and Plot 3 Concentric Circles
# -------------------------------------------------------------------

n_circle <- 400
radii <- c(2, 6, 10) 

# Generate the circles
circles_list <- lapply(seq_along(radii), function(i) {
  r <- radii[i]
  theta <- runif(n_circle, 0, 2 * pi)
  x_circ <- r * cos(theta) + rnorm(n_circle, 0, 0.3)
  y_circ <- r * sin(theta) + rnorm(n_circle, 0, 0.3)
  data.frame(x = x_circ, y = y_circ, cluster = as.factor(i))
})
circles <- do.call(rbind, circles_list)

# Base Plot
p_circles <- ggplot(circles, aes(x = x, y = y, color = cluster)) +
  geom_point(alpha = 0.8, size = 1.5) +
  scale_color_brewer(palette = "Set1", name = "True Cluster") +
  theme_bg() +
  labs(
    title = "Three Concentric Circles",
    x = "X coordinate",
    y = "Y coordinate"
  )

# k-NN Calculation
knn_circ <- FNN::get.knn(circles[, c("x", "y")], k = k_val)
edges_circ <- do.call(rbind, lapply(1:nrow(circles), function(i) {
  data.frame(from = i, to = knn_circ$nn.index[i, ])
}))
edges_circ_df <- data.frame(
  x = circles$x[edges_circ$from],
  y = circles$y[edges_circ$from],
  xend = circles$x[edges_circ$to],
  yend = circles$y[edges_circ$to]
)

# k-NN Plot
p_circles_knn <- ggplot() +
  geom_segment(data = edges_circ_df, aes(x = x, y = y, xend = xend, yend = yend), 
               color = "gray30", alpha = 0.8) +
  geom_point(data = circles, aes(x = x, y = y, color = cluster), 
             size = 1.5, alpha = 0.3) +
  scale_color_brewer(palette = "Set1", name = "True Cluster") +
  theme_bg() +
  labs(
    title = "Three Concentric Circles (5-NN Graph)",
    x = "X coordinate",
    y = "Y coordinate"
  )

# -------------------------------------------------------------------
# 4. Display the Plots
# -------------------------------------------------------------------

path_tmp <- "_tmp/graph/"
if (!dir.exists(path_tmp)) {
  dir.create(path_tmp, recursive = TRUE)
}
path_p_init <- file.path(path_tmp, "p-init.png")

# Combine plots into a 3x2 grid
# Column 1: Base plots | Column 2: k-NN plots
p_grid <- cowplot::plot_grid(
  p_normals,
  p_swiss,
  p_circles,
  ncol = 1,
  nrow = 3
) +
  theme_bg()

# Increased width to accommodate 2 columns side-by-side
ggsave(
  path_p_init,
  p_grid,
  units = "cm",
  height = 25, 
  width = 15
)
```

::: {#fig-nonlinear}
![](`r path_p_init`)

Example spaces.
:::

# 3. Capturing geodesic distances in graphs

We can formalize our continuous spaces into discrete graphs by linking each point to its $k$ nearest neighbors.
Here we use $k = 5$ for all three datasets.
We then use an algorithm to place the points in two-dimensional space such that their geodesic distance is optimally represented.

```{r}
#| echo: FALSE
#| warning: FALSE

k_val <- 5

# -------------------------------------------------------------------
# 1. Simulate and Plot Two 2D Normals
# -------------------------------------------------------------------

set.seed(123)
n_norm <- 150 # Reduced for visual clarity

norm1 <- data.frame(x = rnorm(n_norm, 0, 1), y = rnorm(n_norm, 0, 1), cluster = as.factor("A"))
norm2 <- data.frame(x = rnorm(n_norm, 4, 1), y = rnorm(n_norm, 4, 1), cluster = as.factor("B"))
normals_data <- rbind(norm1, norm2)

knn_norm <- FNN::get.knn(normals_data[, c("x", "y")], k = k_val)
edges_norm <- do.call(rbind, lapply(1:nrow(normals_data), function(i) {
  data.frame(from = i, to = knn_norm$nn.index[i, ])
}))

edges_norm_df <- data.frame(
  x = normals_data$x[edges_norm$from], y = normals_data$y[edges_norm$from],
  xend = normals_data$x[edges_norm$to], yend = normals_data$y[edges_norm$to]
)

# LEFT: Spatial k-NN
p_normals_left <- ggplot() +
  geom_segment(data = edges_norm_df, aes(x = x, y = y, xend = xend, yend = yend), 
               color = "gray50", alpha = 0.3) +
  geom_point(data = normals_data, aes(x = x, y = y, color = cluster), size = 1.5) +
  scale_color_brewer(palette = "Set2") +
  theme_bg() + theme(legend.position = "none") +
  labs(title = "Two 2D Normals (Spatial)", x = "X", y = "Y")

# RIGHT: Topological Layout (Kamada-Kawai)
g_norm <- graph_from_data_frame(edges_norm, directed = FALSE)
g_norm <- simplify(g_norm)
lay_norm <- layout_components(g_norm, layout = layout_with_kk)
normals_data$lay_x <- lay_norm[, 1]
normals_data$lay_y <- lay_norm[, 2]

edges_norm_lay_df <- data.frame(
  x = normals_data$lay_x[edges_norm$from], y = normals_data$lay_y[edges_norm$from],
  xend = normals_data$lay_x[edges_norm$to], yend = normals_data$lay_y[edges_norm$to]
)

p_normals_right <- ggplot() +
  geom_segment(data = edges_norm_lay_df, aes(x = x, y = y, xend = xend, yend = yend), 
               color = "gray60", alpha = 0.3) +
  geom_point(data = normals_data, aes(x = lay_x, y = lay_y, color = cluster), size = 1.5) +
  scale_color_brewer(palette = "Set2") +
  theme_bg() + theme(legend.position = "none", axis.text = element_blank(), axis.ticks = element_blank()) +
  labs(title = "Topological Layout", x = "", y = "")


# -------------------------------------------------------------------
# 2. Simulate and Plot the Swiss Roll
# -------------------------------------------------------------------

n_swiss <- 600 # Reduced for visual clarity
t <- runif(n_swiss, 1.5 * pi, 4.5 * pi)
x_swiss <- t * cos(t) + rnorm(n_swiss, 0, 0.5)
y_swiss <- runif(n_swiss, 0, 3) 
z_swiss <- t * sin(t) + rnorm(n_swiss, 0, 0.5)
swiss_roll <- data.frame(x = x_swiss, y = y_swiss, z = z_swiss, t = t)

knn_swiss <- FNN::get.knn(swiss_roll[, c("x", "y", "z")], k = k_val)
edges_swiss <- do.call(rbind, lapply(1:nrow(swiss_roll), function(i) {
  data.frame(from = i, to = knn_swiss$nn.index[i, ])
}))

edges_swiss_df <- data.frame(
  x = swiss_roll$x[edges_swiss$from], y = swiss_roll$z[edges_swiss$from],
  xend = swiss_roll$x[edges_swiss$to], yend = swiss_roll$z[edges_swiss$to]
)

# LEFT: Spatial k-NN
p_swiss_left <- ggplot() +
  geom_segment(data = edges_swiss_df, aes(x = x, y = y, xend = xend, yend = yend), 
               color = "gray50", alpha = 0.3) +
  geom_point(data = swiss_roll, aes(x = x, y = z, color = t), size = 1.5) +
  scale_color_distiller(palette = "Spectral") +
  theme_bg() + theme(legend.position = "none") +
  labs(title = "Swiss Roll (Spatial 2D Projection)", x = "X", y = "Z")

# RIGHT: Topological Layout (Kamada-Kawai)
g_swiss <- graph_from_data_frame(edges_swiss, directed = FALSE)
g_swiss <- simplify(g_swiss)
lay_swiss <- layout_components(g_swiss, layout = layout_with_kk)
swiss_roll$lay_x <- lay_swiss[, 1]
swiss_roll$lay_y <- lay_swiss[, 2]

edges_swiss_lay_df <- data.frame(
  x = swiss_roll$lay_x[edges_swiss$from], y = swiss_roll$lay_y[edges_swiss$from],
  xend = swiss_roll$lay_x[edges_swiss$to], yend = swiss_roll$lay_y[edges_swiss$to]
)

p_swiss_right <- ggplot() +
  geom_segment(data = edges_swiss_lay_df, aes(x = x, y = y, xend = xend, yend = yend), 
               color = "gray60", alpha = 0.3) +
  geom_point(data = swiss_roll, aes(x = lay_x, y = lay_y, color = t), size = 1.5) +
  scale_color_distiller(palette = "Spectral") +
  theme_bg() + theme(legend.position = "none", axis.text = element_blank(), axis.ticks = element_blank()) +
  labs(title = "Topological Layout", x = "", y = "")


# -------------------------------------------------------------------
# 3. Simulate and Plot 3 Concentric Circles
# -------------------------------------------------------------------

n_circle <- 150 # Reduced for visual clarity
radii <- c(2, 6, 10) 
circles_list <- lapply(seq_along(radii), function(i) {
  r <- radii[i]
  theta <- runif(n_circle, 0, 2 * pi)
  data.frame(
    x = r * cos(theta) + rnorm(n_circle, 0, 0.1),
    y = r * sin(theta) + rnorm(n_circle, 0, 0.1),
    cluster = as.factor(i)
  )
})
circles <- do.call(rbind, circles_list)

knn_circ <- FNN::get.knn(circles[, c("x", "y")], k = k_val)
edges_circ <- do.call(rbind, lapply(1:nrow(circles), function(i) {
  data.frame(from = i, to = knn_circ$nn.index[i, ])
}))

edges_circ_df <- data.frame(
  x = circles$x[edges_circ$from], y = circles$y[edges_circ$from],
  xend = circles$x[edges_circ$to], yend = circles$y[edges_circ$to]
)

# LEFT: Spatial k-NN
p_circles_left <- ggplot() +
  geom_segment(data = edges_circ_df, aes(x = x, y = y, xend = xend, yend = yend), 
               color = "gray50", alpha = 0.3) +
  geom_point(data = circles, aes(x = x, y = y, color = cluster), size = 1.5) +
  scale_color_brewer(palette = "Set1") +
  theme_bg() + theme(legend.position = "none") +
  labs(title = "Concentric Circles (Spatial)", x = "X", y = "Y")

# RIGHT: Topological Layout (Kamada-Kawai)
g_circ <- graph_from_data_frame(edges_circ, directed = FALSE)
g_circ <- simplify(g_circ)
lay_circ <- layout_components(g_circ, layout = layout_with_kk)
circles$lay_x <- lay_circ[, 1]
circles$lay_y <- lay_circ[, 2]

edges_circ_lay_df <- data.frame(
  x = circles$lay_x[edges_circ$from], y = circles$lay_y[edges_circ$from],
  xend = circles$lay_x[edges_circ$to], yend = circles$lay_y[edges_circ$to]
)

p_circles_right <- ggplot() +
  geom_segment(data = edges_circ_lay_df, aes(x = x, y = y, xend = xend, yend = yend), 
               color = "gray60", alpha = 0.3) +
  geom_point(data = circles, aes(x = lay_x, y = lay_y, color = cluster), size = 1.5) +
  scale_color_brewer(palette = "Set1") +
  theme_bg() + theme(legend.position = "none", axis.text = element_blank(), axis.ticks = element_blank()) +
  labs(title = "Topological Layout", x = "", y = "")


# -------------------------------------------------------------------
# 4. Display the Plots
# -------------------------------------------------------------------

path_tmp <- "_tmp/graph/"
if (!dir.exists(path_tmp)) dir.create(path_tmp, recursive = TRUE)
path_p_init_graph_2 <- file.path(path_tmp, "p-init-graph-2.png")

p_grid <- cowplot::plot_grid(
  p_normals_left, p_normals_right,
  p_swiss_left,   p_swiss_right, 
  p_circles_left, p_circles_right,
  ncol = 2, nrow = 3
) + theme_bg()

ggsave(path_p_init_graph_2, p_grid, units = "cm", height = 25, width = 25)
```

::: {#fig-nonlinear-2}
![](`r path_p_init_graph_2`)

:::

# 4. A further problem The curse of dimensionality

In high-dimensional data, the distance to your nearest neighbor and farthest neighbor become almost identical - assuming that the additional dimensions do not discriminate (i.e. are "noise" more than "signal").
Global Euclidean distance loses contrast, meaning points appear to be roughly the same distance from each other.
Consequently, standard distance-based clustering methods fail.

# 5. A common graph: minimum spanning tree (MST)

We can also connect our space using a Minimum Spanning Tree.
An MST is a subgraph connecting all nodes with the minimum possible total edge weight and no cycles.

```{r}
# Generate simulated data
set.seed(123)
n_A <- 100
mean_A <- c(0, 0)
Sigma_A <- matrix(c(4, 0, 0, 1), nrow = 2)
data_A <- MASS::mvrnorm(n = n_A, mu = mean_A, Sigma = Sigma_A)

n_B <- 100
mean_B <- c(3, 3)
Sigma_B <- diag(2)
data_B <- MASS::mvrnorm(n = n_B, mu = mean_B, Sigma = Sigma_B)

data_sim <- rbind(data_A, data_B)

# Compute the Euclidean distance matrix
dist_matrix <- as.matrix(dist(data_sim))

# Define a connectivity threshold (e.g., 10th percentile of all distances)
threshold <- quantile(dist_matrix, 0.1)

# Create an adjacency matrix
adj_matrix <- (dist_matrix < threshold) * 1
diag(adj_matrix) <- 0 # Remove self-loops

# Build an undirected graph
g_distance <- graph_from_adjacency_matrix(
  adj_matrix, mode = "undirected", weighted = TRUE
)

# Visualize the graph
path_p_graph <- projr::projr_path_get(
  "cache", "notes-l3-cluster", "p-graph.png"
)
png(path_p_graph, units = "cm", width = 12, height = 12, res = 300)
plot(
  g_distance, vertex.size = 5, vertex.label = NA,
  main = "Graph Constructed from Distance Matrix"
)
dev.off()
```


::: {#fig-graph-construct}
![](`r path_p_graph`)

Example of a graph constructed from a distance matrix. Each point connects to neighbors within a threshold.
:::

```{r}
g_weighted <- graph_from_adjacency_matrix(
  dist_matrix, mode = "undirected", weighted = TRUE, diag = FALSE
)

mst_graph <- mst(g_weighted, weights = E(g_weighted)$weight)

# Visualize the MST
path_p_mst <- projr::projr_path_get("cache", "notes-l3-cluster", "p-mst.png")
png(path_p_mst, units = "cm", width = 12, height = 12, res = 300)
plot(
  mst_graph, vertex.size = 5, vertex.label = NA,
  main = "Minimum Spanning Tree from Distance Graph"
)
dev.off()
```

::: {#fig-mst-construct}
![](`r path_p_mst`)

Example of an MST constructed from the distance graph.
:::

# 5. The Objective: Finding Clusters (Community Detection)

Once we have mapped our data into a graph, clustering transforms into a community detection problem. Intuitively, a community (or cluster) is a group of nodes that are highly connected to each other, but only sparsely connected to the rest of the network.

But how do we mathematically define what makes a "good" community? If a graph is just densely connected everywhere, finding a clump of edges doesn't mean much. We need a baseline to compare against. 

This is where modularity comes in. Modularity acts as a quality score. It answers the question: *Are there more edges inside this group than we would expect by pure, random chance?*

### Deconstructing the Modularity Formula

To understand the formula, we compare reality to a random baseline (often called the configuration model). 

$$
H = \frac{1}{2m} \sum_{c} \left( e_c \;-\; \frac{K_c^2}{2m} \right)
$$

Let's break down the components:

* $m$ (Total edges): The total number of edges in the entire graph. Because every edge has two ends, there are $2m$ total "half-edges" or "stubs" available to connect.
* $e_c$ (Reality): The actual, counted number of edges that lie completely inside community $c$.
* $K_c$ (Capacity): The sum of the degrees of all nodes in community $c$. Think of this as the total number of "stubs" sticking out of the nodes in this community.
* $\frac{K_c^2}{2m}$ (The Random Expectation):  This is the core of the formula. If you took all $K_c$ stubs in this community and connected them completely randomly to any of the $2m$ available stubs in the entire network, this fraction is the exact number of edges you would expect to accidentally form *inside* community $c$.

### The Subtraction: Reality vs. Expectation

The magic happens in the subtraction: $( \text{Actual Edges} - \text{Expected Edges} )$.

* If the result is positive, it means nodes in this group are connecting to each other much more frequently than random chance dictates. This is a strong, meaningful cluster.
* If the result is zero or negative, it means the edges inside this group are no better than random wiring. It is not a real cluster.

Finally, we sum this difference over every community ($\sum_c$) and divide by $2m$ just to normalize the final score $H$ so it generally falls between -0.5 and 1.0. 

Therefore, when algorithms like Louvain or Leiden run, they are simply moving nodes between different groups, constantly recalculating this formula, and searching for the partition that gives the absolute maximum modularity score.

# 6. Algorithms for Allocating

## The Leiden Algorithm

The Leiden algorithm is a community detection method that maximizes modularity.
It improves on the Louvain algorithm by refining communities so each is guaranteed to be internally well-connected.

Leiden expects higher weights to indicate similarities, so we often convert our distances into a $k$-nearest neighbor graph before running it.

```{r}
# Build a k-nearest neighbor graph
k <- 5
knn_result <- get.knn(data_sim, k = k)

# Build an edge list
edge_list <- do.call(rbind, lapply(1:nrow(data_sim), function(i) {
  data.frame(from = i, to = knn_result$nn.index[i, ])
}))

# Create an undirected graph
g_knn <- graph_from_data_frame(edge_list, directed = FALSE)
g_knn <- simplify(g_knn)

# Run the Leiden Algorithm
leiden_clusters <- cluster_leiden(
  g_knn,
  objective_function = "modularity",
  resolution = 0.2
)

membership_vec <- membership(leiden_clusters)
table(membership_vec)
```

We can plot the raw coordinates and color them by their Leiden cluster membership.

```{r}
#| warning: false
df_plot <- as_tibble(data_sim)
df_plot$cluster <- as.character(membership_vec)
colnames(df_plot)[1:2] <- c("x", "y")

path_p_leiden_clust_gg <- projr::projr_path_get(
  "cache", "notes-l3-cluster", "p-leiden-clust-gg.png"
)
p <- ggplot(df_plot, aes(x = x, y = y, color = cluster)) +
  geom_point(size = 2) +
  cowplot::theme_cowplot() +
  cowplot::background_grid() +
  labs(x = "X", y = "Y") +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.position = "none"
  )
ggsave(
  path_p_leiden_clust_gg, p, width = 12, height = 12, units = "cm"
)
```

## The HDBSCAN Algorithm

HDBSCAN uses the MST to find clusters of varying densities. 

It constructs a graph using mutual reachability distance, gradually removes edges exceeding a threshold to form clusters from the remaining connected components, and scores them based on stability over the threshold range. Highly stable clusters are kept, and unassigned points are labeled as noise.

# 7. Advantages & Disadvantages

Advantages:

- Captures complex, non-linear cluster shapes.
- Handles higher dimensions well by utilizing local connectivity.
- Automatically determines the number of clusters.

Disadvantages:

- Building the initial graph is computationally expensive ($O(N^2)$) for large datasets.
- Results are highly sensitive to local construction parameters like $k$ or the distance threshold.
