---
title: Lecture 3 - Clustering
subtitle: DBSCAN
bibliography: zotero.bib
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
---

<!--
Useful conversiations:
- Cluster Analysis Questions
  - Lot of discussion, fleshes out deciding on # of clusters
- DBScan clustering overview:
  - Discusses HDBSCAN
!-->

```{r}
#| include: false
library(mclust)
library(ggplot2)
library(tibble)
library(dbscan)
```

# Fixed-shape and volume clustering can over-partition data

Below, the data consist of two clusters of varying size and shape:

- Cluster A (stretched along first dimension): $\mathbf{X} \sim \mathcal{N}\left(\begin{pmatrix} 0 \\ 0 \end{pmatrix}, \begin{pmatrix} 4 & 0 \\ 0 & 1 \end{pmatrix}\right)$.
- Cluster B (spherical): $\mathbf{X} \sim \mathcal{N}\left(\begin{pmatrix} 3 \\ 3 \end{pmatrix}, \begin{pmatrix} 2 & 0 \\ 0 & 2 \end{pmatrix}\right)$.

```{r}
#| include: false
set.seed(123)
n_A <- 100
mean_A <- c(0, 0)
# Covariance matrix for Cluster A: Var1 = 2, Var2 = 1
Sigma_A <- matrix(
   c(4, 0, 0, 1),
   nrow = 2
)
data_A <- MASS::mvrnorm(n = n_A, mu = mean_A, Sigma = Sigma_A)
n_B <- 100
mean_B <- c(3, 3)
# Covariance matrix for Cluster B: unit variance for both variables
Sigma_B <- diag(2)
data_B <- MASS::mvrnorm(n = n_B, mu = mean_B, Sigma = Sigma_B)
data_sim <- rbind(data_A, data_B)
true_labels <- factor(rep(c("A", "B"), each = 100))
# Fit overly simple model (EII)
settings_vec <- c("EII", "VVI", "VVV")
mclust_list <- lapply(settings_vec, function(x) {
   Mclust(data_sim, G = 1:4, modelNames = x)
}) |>
   setNames(c("simple", "appropriate", "flexible"))
path_vec <- NULL
for (i in seq_along(mclust_list)) {
   path_p <- projr::projr_path_get(
     "cache", "notes-l3-cluster",
     paste0("p-2b-", names(mclust_list)[i], ".png")
   )
   path_vec <- c(path_vec, path_p)
   png(path_p, units = "cm", width = 8, height = 8, res = 300)
   plot(mclust_list[[i]], what = "classification")
   dev.off()
}
```

::: {#fig-flexible}

![EII (simple)](`r path_vec[[1]]`)

![EEI (appropriate)](`r path_vec[[2]]`)

![VVV (flexible)](`r path_vec[[3]]`)

Classification results for each model.

:::

- K-means is a "hard" classification equivalent to EII (fixed shape, volume, and orientation), so the over-partitioning problem is not unique to Mclust.
- The VVV model is sufficiently flexible, but:
  - It requires up-front specification of the number of clusters.
    - One can fit multiple models and select the best one, but this is computationally expensive.
  - All points are clustered, even if they are noise.
    - In the example below, we add a single outlier point to the data.
    - All models assign this to a cluster, but only the simplest model regards it as a different cluster to cluster at (0, 0).

```{r}
#| include: false
data_sim_noise <- data_sim |>
  rbind(
    matrix(c(-10, -10), nrow = 1)
  )
set.seed(123)
mclust_list <- lapply(settings_vec, function(x) {
   Mclust(data_sim_noise, G = 1:4, modelNames = x)
}) |>
   setNames(c("simple", "appropriate", "flexible"))
path_vec <- NULL
for (i in seq_along(mclust_list)) {
   path_p <- projr::projr_path_get(
     "cache", "notes-l3-cluster",
     paste0("p-2b-", names(mclust_list)[i], "-noise.png")
   )
   path_vec <- c(path_vec, path_p)
   png(path_p, units = "cm", width = 8, height = 8, res = 300)
   plot(mclust_list[[i]], what = "classification")
   dev.off()
}
```

::: {#fig-flexible}

![EII (simple)](`r path_vec[[1]]`)

![EEI (appropriate)](`r path_vec[[2]]`)

![VVV (flexible)](`r path_vec[[3]]`)

Classification results for each model.

:::

# DBSCAN: Density-Based Spatial Clustering of Applications with Noise

## 1. Overview

- Definition:  
  DBSCAN is a density-based clustering algorithm that groups together points that are closely packed (i.e., high density) and marks points in low-density regions as outliers or noise.

- Key idea:  
  Instead of requiring a pre-specified number of clusters (like k-means), DBSCAN defines clusters based on the density of data points.

- Main reference:
  @schubert_etal17

## 2. Key concepts

```{r}
#| include: false
path_img_dbscan_intro <- projr::projr_path_get(
  "raw-img", "schubert_etal17-dbscan_points.png"
)
```

::: {#fig-dbscan-points}

![](`r path_img_dbscan_intro`)

DBSCAN clustering example (Schubert et al., 2017)
:::

### 2.1 Core, border and noise points

- Core point:  
  A point is considered a core point if it has at least `minPts` points (including itself) within its neighborhood of radius `eps`.
- Border point:  
  A point that has fewer than `minPts` neighbors but is within the `eps` neighborhood of a core point.
- Noise point:  
  Any point that is neither a core point nor a border point; considered as outlier.

### 2.2 Density reachability and connectivity

- Direct density-reachability:  
  A point $q$ is directly density-reachable from point $p$ if $q$ is within $eps$ of $p$ and $p$ is a core point.
- Density-reachability:  
  A point $q$ is density-reachable from $p$ if there is a chain of points $p, p_1, p_2, \dots, q$ where each point is directly density-reachable from the previous one.
- Density-connectivity:  
  Two points $p$ and $q$ are density-connected if there exists a point $o$ such that both $p$ and $q$ are density-reachable from $o$.

## 3. Applied to the original and noisy data

```{r}
#| include: false
path_p_knn_init <- projr::projr_path_get(
  "cache", "notes-l3-cluster",
  "p-dbscan-knn-init.png"
)
png(path_p_knn_init, units = "cm", width = 8, height = 8, res = 300)
kNNdistplot(data_sim, k = 10)
dev.off()
```

```{r}
dbscan_obj <- dbscan(data_sim, eps = 0.8, minPts = 10)
```

```{r}
#| include: false
path_p_dbscan_noiseless <- projr::projr_path_get(
  "cache", "notes-l3-cluster",
  "p-dbscan-noiseless.png"
)
plot_tbl <- data_sim |>
  as_tibble() 
colnames(plot_tbl) <- c("x", "y")
plot_tbl$cluster <- as.factor(dbscan_obj$cluster)
p <- ggplot(
  plot_tbl,
  aes(x = x, y = y, color = cluster)
) +
  geom_point() +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    plot.background = element_rect(fill = "white")
  ) +
  coord_equal() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  )
ggsave(
  path_p_dbscan_noiseless,
  plot = p,
  width = 8, height = 8, units = "cm", dpi = 300
)
```

::: {#fig-dbscan-noiseless}

![](`r path_p_dbscan_noiseless`)

DBSCAN clustering of noiseless data
:::

```{r}
#| include: false
path_p_knn_noisy <- projr::projr_path_get(
  "cache", "notes-l3-cluster",
  "p-dbscan-knn-noisy.png"
)
png(path_p_knn_noisy, units = "cm", width = 8, height = 8, res = 300)
kNNdistplot(data_sim_noise, k = 10)
dev.off()
```

```{r}
#| warning: false
dbscan_obj <- dbscan(data_sim_noise, eps = 0.8, minPts = 10)
```

```{r}
#| include: false
path_p_dbscan_noisy <- projr::projr_path_get(
  "cache", "notes-l3-cluster",
  "p-dbscan-noisy.png"
)
plot_tbl <- data_sim_noise |>
  as_tibble()
colnames(plot_tbl) <- c("x", "y")
plot_tbl$cluster <- as.factor(dbscan_obj$cluster)
p <- ggplot(
  plot_tbl,
  aes(x = x, y = y, color = cluster)
) +
  geom_point() +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    plot.background = element_rect(fill = "white")
  ) +
  coord_equal() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  )
ggsave(
  path_p_dbscan_noisy,
  plot = p,
  width = 8, height = 8, units = "cm", dpi = 300
)
```

::: {#fig-dbscan-noisy}

![](`r path_p_dbscan_noisy`)

DBSCAN clustering of noisy data
:::

## 4. The DBSCAN algorithm

### 4.1 Pseudocode outline [@schubert_etal17]

1. For each point $p$ in the dataset:
   - If $p$ is already assigned to a cluster, skip it.
   - Otherwise, get all points within distance $eps$ of $p$.
   - If the number of neighbors < minPts:  
     - Mark $p$ as noise, and continue.
   - Else:
     - Create a new cluster.
     - Mark $p$ as part of this cluster.
     - Recursively add all points that are density-reachable from $p$ to the cluster.
2. Repeat until all points are processed.

## 5. Parameters and their selection

### 5.1 minPts

- Definition:  
  The minimum number of points required to form a dense region (including the core point itself).
- Typical Values:
  Often set to $2p$, but can be larger fo4r datasets with more noise or higher-dimensional.
- Impact:
  Higher values require denser clusters to be recognized; lower values might cause noise to be included in clusters.

### 5.2 eps ($\epsilon$)

- Definition:  
  The maximum distance between two points for one to be considered as in the neighborhood of the other.
- Selection Strategy:  
  - k-NN distance plot:  
    Plot the distance to the k-th nearest neighbor for all points (where $k = minPts$ for 2D data and $k=minPts+1$ for higher-dimensional data).
    Look for an "elbow" in the sorted plot; set $eps$ just above this point.
  - Rules of thumb:  
    For low-dimensional data, a small percentage of the range of values might work. For higher dimensions, domain knowledge is crucial.
- Impact:  
  A small $eps$ may result in many points being labeled as noise; a large $eps$ might merge distinct clusters.

```{r}
#| eval: false
kNNdistplot(data_sim, k = 10)
```

::: {#fig-dbscan-knn}

![](`r path_p_knn_init`)

k-NN distance plot for noiseless data
:::


### 5.3 Red flags for degenerate clustering results

When evaluating DBSCAN outcomes, consider these two key red flag criteria:

- **Number of clustered points vs noise:**

  A good clustering result should cover a large part of the dataset while still detecting noise. Typically, the level of noise should be between 1% and 30% of the data. If too few points are labeled as noise, it may indicate that the $\epsilon$ value is set too high.

- **Size of the largest component:**

  If the largest cluster (component) contains more than 20% to 50% of the clustered points, it might be a sign that the clustering is degenerate. This can mean that $\epsilon$ was chosen too large, causing distinct clusters to merge. In such cases, either a hierarchical approach (like OPTICS or HDBSCAN*) should be considered, or DBSCAN should be rerun with a smaller $\epsilon$ value.

While these criteria do not precisely quantify clustering quality, they serve as useful guidelines for spotting degenerate solutions and adjusting parameters. Increasing $\epsilon$ generally increases both the number of clustered points and the size of the largest cluster, while decreasing $\epsilon$ tends to produce more noise and smaller clusters.

## 6. Advantages and limitations

### 6.1 Advantages

- No need to specify the number of clusters in advance.
- Ability to discover clusters of arbitrary shape.
- Effective handling of noise and outliers.
- Intuitive parameter interpretation (eps and minPts).

### 6.2 Limitations

- Parameter sensitivity:  
  The quality of clustering depends heavily on the choice of $eps$ and $minPts$.
  Results can be erratic.
- Varied densities:  
  DBSCAN struggles with datasets where clusters have significantly different densities.
- Scalability:  
  While efficient with spatial indexing (e.g., kd-trees or R-trees), performance can degrade in very high-dimensional spaces.

