---
title: Answer Key
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
---

# MDS

## Section 1: Measurement and Data Structures

* **Answer 1:**
  * Ratio: Time taken to complete a task.
  * Interval: Difference from an arbitrary standard (as we had with degrees Celcius), e.g. calendar years, latitude and longitude.
  * Ordinal: Ratings (cakes, movies, etc.).
  * Nominal: Car brands or people's names (no greater/less than relationship).

* **Answer 2:**
  * **Proof:** For a ratio scale to hold, the ratio of two values must be preserved. The ratio of the original values is $x/y$. After adding the constant, the new ratio is $(x+c)/(y+c)$, which is not equal to $x/y$ (unless $c=0$ or $x=y$). Thus, the ratio scale is lost, and $A$ and $B$ operate on an interval scale. However, the difference between the new values is $A - B = (x + c) - (y + c) = x - y$. The arbitrary constant $c$ cancels out, exactly recovering the difference between the original ratio-scaled values. Therefore, the differences operate on a ratio scale.
  * **Implication:** If a constant is added to all dissimilarities before mapping (shifting them to an interval scale), we cannot meaningfully compare the ratios of the resulting configuration distances (e.g., we cannot say Point 1 is "twice as far" from Point 2 as it is from Point 3). However, we *can* meaningfully compare the *differences* between distances (e.g., the difference in distance between Pair 1 and Pair 2 is twice as large as the difference between Pair 3 and Pair 4).

* **Answer 3:**
  We cannot attempt to get coordinates in Euclidean space such that points have a negative distance between them.

* **Answer 4:**
  Let three points in 1D space be $x=0$, $y=1$, and $z=2$. The squared Euclidean distances are: $d(x,y) = (0-1)^2 = 1$; $d(y,z) = (1-2)^2 = 1$; and $d(x,z) = (0-2)^2 = 4$. According to the triangle inequality, $d(x,z)$ must be $\le d(x,y) + d(y,z)$. Since $4 \not\le 1 + 1$, squared Euclidean distance is not a valid distance metric.
  More generally, one can show this by considering Pythagoras theorem and pulling the corner opposite the hypotenuse in (not needed for an answer).

## Section 2: Classical Scaling vs. Least Squares

* **Answer 1:**
  Classical scaling does not directly approximate the dissimilarities themselves; instead, it approximates the centered inner products of the data.

* **Answer 2:**
  There are two main reasons:
  1. We can simply ignore negative eigenvalues, or add positive constants to original dissimilarities.
  2. The main reason, however, is practical - regardless of the underlying dissimilarity measure, we interpret points visually best when they are in Euclidean space.

* **Answer 3:**
  Let $X = U D V^T$. Then $B = XX^T = (UDV^T)(VD U^T) = U D^2 U^T$. The eigenvectors of $B$ are $U$, and the eigenvalues are $\Lambda = D^2$. Classical scaling computes coordinates as $Y = U_k \Lambda_k^{1/2} = U_k D_k$. PCA computes coordinates as $XV = (UDV^T)V = UD$. Thus, the first $k$ dimensions are identical.

* **Answer 4:**
  Classical scaling, in being equivalent to PCA, only considers coordinates in the low-dimensional space that are linear combinations of values in the high(er)-dimensional space. Least-squares scaling, by contrast, has no such restriction, and simply places points where they fit best as per the numerical optimiser.

## Section 3: Least Squares Optimization & Transformations

* **Answer 1:**
  $$\mathcal{L}(y_1, ..., y_n) = \sum_{i<j}^n (d_{ij} - \delta_{ij})^2$$

* **Answer 2:**
  $$\mathcal{L}_f = \sum_{i<j}^n w_{ij} \left( \sqrt{\sum_{k=1}^t (y_{ik} - y_{jk})^2} - f(\delta_{ij}) \right)^2$$

* **Answer 3:**
  There are 2n parameters to select, which quickly becomes large. Moreover, the distance matrix has $n(n-1)/2$ elements, which becomes enormous to hold in memory as $n$ becomes very big.

* **Answer 4:**
  Logging pre-optimisation fixes the base without consideration of the data, whereas making the base a parameter to optimise means that it is selected based on the data.

* **Answer 5:**
  You should use Sammon mapping, which applies the weight $w_{ij} = 1 / (\delta_{ij} \sum_{i<j}\delta_{ij})$ to down-weight larger original distances.

## Section 4: Interpretation and Diagnostics

* **Answer 1:**
  The same-size difference along the y-axis will appear to be smaller/larger than the same-size difference along the x-axis. It is very important in MDS that we have equal visual scaling of the axes.

* **Answer 2:**
  Map B is the one where the constant was added. Adding a constant $\alpha > 0$ effectively "pushes apart" all the dissimilarities, which affects the small dissimilarites more than large dissimilarites and so results in the relevant points being relatively further apart.

* **Answer 3:**
  The denominator in the provided formula normalizes by the square of the disparities ($\sum_{i<j} (\hat{d}_{ij})^2$). For Kruskal's Stress-1, the denominator must normalize by the square of the configuration distances, making the correct denominator $\sum_{i<j} (d_{ij})^2$.

* **Answer 4:**
  Number of dots/dashes the most critical factor (top-right vs bottom-left).
  The original paper had another observation, but I can't remember what it was.

* **Answer 5:**
  * Start: $\{3, 1, 4, 2.5\}$
  * Rank 1: $3$.
  * Rank 2: $1$. Because $1 < 3$, we average them: $(3+1)/2 = 2$. Current state: $\{2, 2\}$.
  * Rank 3: $4$. Because $4 > 2$, it is valid. Current state: $\{2, 2, 4\}$.
  * Rank 4: $2.5$. Because $2.5 < 4$, we average the violating pair: $(4+2.5)/2 = 3.25$.
  * Final state: $\{2, 2, 3.25, 3.25\}$.

# Clustering I

## Section 1: Conceptual Foundations & Algorithm Choice

* **Answer 1:**
  Density-based clustering approaches do not explicitly define a probability distribution but rather group observations together that lie in regions that are sufficiently dense throughout.

* **Answer 2:**
  For $n$ observations to be allocated to $k$ clusters, the number of different ways of allocating observations to clusters is given by a very large combinatorial expression.

* **Answer 3:**
  A centroid is the mathematical mean vector of all points in a cluster (and rarely an actual data point itself). A medoid is a strictly actual, observed data point within the cluster that minimizes the average dissimilarity to all other points in that cluster.

* **Answer 4:**
  K-means is vastly computationally faster because updating a centroid requires only calculating a simple mean. K-medoids is highly robust to extreme outliers and flexible because it can handle any arbitrary pairwise dissimilarity matrix, not just squared Euclidean distance.

## Section 2: K-Means & Partitioning Mechanics

* **Answer 1:**
  Let the within-cluster sum of squares for a cluster $C$ with $n_c$ points and prototype $c$ be $J(c) = \sum_{x \in C} ||x - c||^2$. Taking the derivative with respect to $c$ and setting it to zero: $\frac{\partial J}{\partial c} = -2 \sum_{x \in C} (x - c) = 0$. This implies $\sum x - n_c c = 0$, leading to $c = \frac{1}{n_c} \sum x$. Thus, the mathematical mean minimizes the function.

* **Answer 2:**
  The length of a vector is the same as its distance from the origin. Its length squared is equal to the dot product of the vector with itself. So, if we think of the vector as a difference vector (e.g., $x-y$), then its squared length $||x-y||^2$ is equal to its dot product with itself. Therefore, we can write the inner summand for the K-means objective function as either the squared length of the difference vector or its dot product with itself.

* **Answer 3:**
  No, K-means only guarantees convergence to a *local* optimum. Because both the assignment step and the update step monotonically decrease the within-cluster sum of squares, the algorithm will stop when it hits a local valley, but it cannot jump out of that valley to find the global optimum.

* **Answer 4:**
  Mini-batch K-means can be used, which takes random subsamples of the data to update centroids instead of waiting to evaluate every single point, vastly speeding up iteration time on large datasets.

* **Answer 5:**
  In each iteration, PAM tests swapping every medoid ($k$) with every non-medoid ($n-k$). Therefore, it evaluates $k(n-k)$ swaps. This value is maximized (worst-case performance) when $k = n/2$, yielding $n^2 / 4$ evaluations.

## Section 3: Hierarchical Clustering (AGNES & DIANA)

* **Answer 1:**
  AGNES and DIANA never reassign points.

* **Answer 2:**
  For a 3x3 matrix, you compute the initial distances ($\binom{3}{2} = 3$). After one merge, you have 2 clusters, requiring 1 distance computation between them. Total = 4. For an $n \times n$ matrix, calculating the initial pairwise distances takes $\frac{n(n-1)}{2}$ computations.

* **Answer 3:**
  For single linkage, the height is the largest nearest-neighbour distance. For complete linkage, the height is the largest distance. Therefore, for single linkage the merge heights before the final merge are compared visually to a lower value (the nearest-neighbour distance) than for complete linkage (largest distance). As the agglomerative coefficient is dependent on this ratio, it is heavily dependent in absolute terms on the linkage chosen.

* **Answer 4:**
  * Single Linkage: Excels at finding elongated, non-spherical shapes (like concentric rings). Fails terribly when there is background noise, causing "chaining" where distinct clusters are merged by a single bridge of noise points.
  * Complete Linkage: Excels at finding compact, spherical clusters of roughly equal diameter. Fails when underlying clusters naturally have vastly different variances/diameters, as it will break up large clusters prematurely.

## Section 4: Clustering Diagnostics & Metrics

* **Answer 1:**
  In the elbow method, we choose the number of clusters such that after that point the within-cluster dispersion stops dropping rapidly. The gap statistic formalises this by calculating how much better than random the within-cluster dispersion is for each $k$, which will tend to select points near the elbow (for conservative gap-stat criteria for $k$) as after that point the increasingly-small reductions in within-cluster dispersion are unlikely to outweigh the increased flexibility from more clusters.

* **Answer 2:**
  The PCs will take into account correlation between original features, ensuring better alignment with the actual data.

* **Answer 3:**
  It captures the natural variability in the within-cluster dispersion under the null scenario of no clusters.

* **Answer 4:**
  Long clusters with a region of very low density between them.

## Section 4: Mixture models

### Part A: Optimization and Lagrange Multipliers

* **Answer 1:**
  The unconstrained function is $f(x, y) = x^2 + x + y$. Taking partial derivatives and setting to zero: $\frac{\partial f}{\partial x} = 2x + 1 = 0 \implies x = -0.5$. However, $\frac{\partial f}{\partial y} = 1 \neq 0$. There is no critical point where the derivative with respect to $y$ is zero. This function represents a surface that continues upward infinitely in the $y$ direction, meaning it has no global or local maximum.

* **Answer 2:**

The Lagrangian is $\mathcal{L}(x, y, \lambda) = x^2 + x + y - \lambda(x^2 + y^2 - 1)$.
Taking partial derivatives:
1. $\frac{\partial \mathcal{L}}{\partial x} = 2x + 1 - 2\lambda x = 0 \implies x(2 - 2\lambda) = -1$
2. $\frac{\partial \mathcal{L}}{\partial y} = 1 - 2\lambda y = 0 \implies y = \frac{1}{2\lambda}$
3. $\frac{\partial \mathcal{L}}{\partial \lambda} = -(x^2 + y^2 - 1) = 0 \implies x^2 + y^2 = 1$
By substituting equations 1 and 2 into equation 3, we can solve for $\lambda$, which yields the maximizing coordinates $(x,y)$.
I don't know how to do this analytically, so we can plug the resultant equation ($1/\lambda^2+1/(\lambda-1)^2-4=0$) into `optmimize`.
The function has asymptotes at 0 and 1, so we'll optimise within each section:

```{r}
root_fn <- function(lambda) 1/lambda^2 + 1/(lambda-1)^2 - 4
root_neg <- uniroot(root_fn, interval = c(-100, -0.01))$root
root_pos_01 <- try(uniroot(root_fn, interval = c(0.01, 0.99))$root)
root_pos_g1 <- uniroot(root_fn, interval = c(1.01, 100))$root
root_vec <- c("neg" = root_neg, "pos" = root_pos_g1)
```

There is no root in [0, 1], so it errors out there.

The roots were:

```{r}
root_vec
```


```{r}
x_fn <- function(lambda) 1 / 2 / (lambda - 1)
y_fn <- function(lambda) 1 / 2 / lambda
orig_fn <- function(x, y) x^2 + x + y
# for lambda in [-100, -0.01]:
x_neg <- x_fn(root_neg)
y_neg <- y_fn(root_neg)
x_neg^2 + y_neg^2
orig_fn(x_neg, y_neg)
# for lambda in [0.01, 100]:
x_pos <- x_fn(root_pos_g1)
y_pos <- y_fn(root_pos_g1)
x_pos^2 + y_pos^2
orig_fn(x_pos, y_pos)
```

We had a minimum for the point in the [-100, 0.01] interval and a maximum in the [0.01, 100] interval, so the maximising parameter values are `x=0.945` and `y=0.327` leading to a maximum of `2.16`.

*Note*: I wouldn't ask you to do this in an actual written exam, of course. This is too time-consuming. I just didn't pick functions that led to a nice solution for the multiplier, $\lambda$.

### Part B: Deriving the M-Step

* **Answer 1:**

The weighted log-likelihood term for component $k$ in the $Q$-function is:
$$
Q_k(\lambda_k) = \sum_{i=1}^n \gamma_{ik} (x_i \log(\lambda_k) - \lambda_k - \log(x_i!))
$$

Take the derivative with respect to $\lambda_k$ and set to zero:

$$
\frac{\partial Q}{\partial \lambda_k} = \sum_{i=1}^n \gamma_{ik} \left( \frac{x_i}{\lambda_k} - 1 \right) = 0
$$

$$
\sum_{i=1}^n \gamma_{ik} \frac{x_i}{\lambda_k} = \sum_{i=1}^n \gamma_{ik}
$$

$$
\lambda_k = \frac{\sum_{i=1}^n \gamma_{ik} x_i}{\sum_{i=1}^n \gamma_{ik}}
$$

The MLE is simply the responsibility-weighted sample mean.

* **Answer 2:**

If all components share a single variance $\sigma^2$, the term in the $Q$-function involves a sum over all components $k$:

$$Q(\sigma^2) = \sum_{i=1}^n \sum_{k=1}^K \gamma_{ik} \left( -\frac{1}{2}\log(2\pi\sigma^2) - \frac{(x_i - \mu_k)^2}{2\sigma^2} \right)$$

Taking the derivative with respect to $\sigma^2$ (let $v = \sigma^2$) and setting to zero:

$$
\frac{\partial Q}{\partial v} = \sum_{i=1}^n \sum_{k=1}^K \gamma_{ik} \left( -\frac{1}{2v} + \frac{(x_i - \mu_k)^2}{2v^2} \right) = 0
$$

Multiplying by $2v^2$ and rearranging:

$$
v \sum_{i=1}^n \sum_{k=1}^K \gamma_{ik} = \sum_{i=1}^n \sum_{k=1}^K \gamma_{ik} (x_i - \mu_k)^2
$$

Since $\sum_{k=1}^K \gamma_{ik} = 1$ for every observation, the sum on the left simplifies to $n$:

$$
\sigma^2 = \frac{1}{n} \sum_{i=1}^n \sum_{k=1}^K \gamma_{ik} (x_i - \mu_k)^2
$$

Unlike component-specific variances, this pools the squared deviations from *all* groups, weighted by their respective responsibilities, and divides by the total sample size $n$.

* **Answer 3:**
  * a. In the E-step, $\gamma_{i1}$ is calculated using the known, fixed $\mu_1$ inside the density function $f_1(x_i)$. The responsibilities $\gamma_{i2}$ and $\gamma_{i3}$ use the iteratively updated estimates $\hat{\mu}_2^{(t)}$ and $\hat{\mu}_3^{(t)}$.
  * b. The M-step formulas for $\hat{\mu}_2$ and $\hat{\mu}_3$ are completely unaffected; they remain the responsibility-weighted means for their respective groups. The parameter $\mu_1$ is not updated at all.
  * c. The common variance uses the fixed $\mu_1$ alongside the newly updated $\mu_2$ and $\mu_3$:
    $$\sigma^2 = \frac{1}{n} \left[ \sum_{i=1}^n \gamma_{i1}(x_i - \mu_1)^2 + \sum_{i=1}^n \gamma_{i2}(x_i - \hat{\mu}_2)^2 + \sum_{i=1}^n \gamma_{i3}(x_i - \hat{\mu}_3)^2 \right]$$

In effect, there is no real change, except that we don't need to work out the mean each time.

### Part C: Conceptual Foundations

* **Answer 1:**

  Because a sum rather than a product of the $K$ mixture components is inside the log ($\log(\sum \pi_k f_k(x)), we cannot apply the rule that the log of the product is the sum of the logs, which is how we normally separate factors of different parameters into sums of different parameters that make differentiation easier.

* **Answer 2:**
  $Z_{ik}$ is an unobserved indicator variable equal to 1 if observation $i$ truly belongs to component $k$, and 0 otherwise.
  Introducing it allows us to write the log-likelihood as if we knew group membership.
  Combined with the "trick" to place it as the exponent of each component's contribution to the PDF, we end up with a log of a product ($\prod f_k(x_i)^{Z_{ik}}$) and so can rewrite it as a sum of logs.

* **Answer 3:**
  $\gamma_{ik}$ is the posterior probability that observation $i$ belongs to group $k$, given the data and current parameter estimates. Bayes' theorem is what allows us to derive the formula. Mathematically, it is the expected value of the latent variable $Z_{ik}$.

* **Answer 4:**

The Lagrangian is:

$$
\mathcal{L}(\boldsymbol{\pi}, \lambda) = \sum_{i=1}^n \sum_{k=1}^K \gamma_{ik} \log(\pi_k) + \lambda \left( 1 - \sum_{k=1}^K \pi_k \right)
$$

Taking the partial derivative with respect to $\pi_k$ and setting to zero:

$$
\frac{\partial \mathcal{L}}{\partial \pi_k} = \sum_{i=1}^n \frac{\gamma_{ik}}{\pi_k} - \lambda = 0 \implies \sum_{i=1}^n \gamma_{ik} = \lambda \pi_k
$$

Summing both sides over $k$:

$$
\sum_{k=1}^K \sum_{i=1}^n \gamma_{ik} = \lambda \sum_{k=1}^K \pi_k
$$

Because $\sum_k \gamma_{ik} = 1$ for each of the $n$ observations, the left side equals $n$. Because of our constraint, $\sum_k \pi_k = 1$. Therefore, $\lambda = n$.
Substituting $\lambda = n$ back in gives $\pi_k = \frac{1}{n} \sum_{i=1}^n \gamma_{ik}$.

### Part D: Practical Implementations and Limits

* **Answer 1:**
  We can use numerical optimisation (e.g. Newton-Raphson) for that component.

* **Answer 2:**
  Initialization is crucial because the EM algorithm only guarantees convergence to a *local* maximum, not the global maximum.
  Poor initialization can lead to suboptimal clustering. Common strategies include:
  * Running K-means first and using its cluster centers and variances as the starting parameters.
  * Randomly assigning starting responsibilities to all data points and proceeding directly to the first M-step.

* **Answer 3:**
  The most common criterion is checking the change in the observed-data log-likelihood between iterations.
  Iterations stop when $\ell^{(t+1)} - \ell^{(t)} < \epsilon$, where $\epsilon$ is a predefined small tolerance value.

* **Answer 4:**

The responsibility for object $i$ with respect to group $1$ is given by

$$
\gamma_{i1} = \frac{\pi_1 \exp(-\frac{1}{2\sigma^2}(x_i-\mu_1)^2)}{
  \pi_1 \exp(-\frac{1}{2\sigma^2}(x_i-\mu_i)^2) + (1-\pi_1) \exp(-\frac{1}{2\sigma^2}(x_i-\mu_2)^2)
}
$$

The responsibility with respect to group $2$ is clearly $1$ minus the above.

As $\sigma^2 \to 0$, the component density $f_k(\mathbf{x}_i)$ for the closer mean becomes infinitely larger than the density for the other mean.
This arises because the (squared) differences to the mean are multiplied by an increasingly large factor $1/(2\sigma^2)$ and then exponentiated.
As a result, the responsibility for the compnent density of the closer mean (regardless, eventualy, of the prior weights) tends to 1.

By way of example, assuming that the squared difference between $x_i$ and the mean is 1 for component 1 and 1.5 for component 2, with equal prior variances.

Here is a plot of $\gamma_{i1}$ as  $\sigma^2\rightarrow 0$:

```{r}
library(ggplot2)
get_responsibility <- function(sigma_squared,
                                squared_diff_1 = 1,
                                squared_diff_2 = 1.5) {
  comp_1 <- exp(-1 / sigma_squared * squared_diff_1)
  comp_2 <- exp(-1 / sigma_squared * squared_diff_2)
  comp_1 / (comp_1 + comp_2)
}

exponent_vec <- c(-5, -4, -3, -2, -1, 0, 1, 2, 3, 4, 5, 6)
sigma_2_vec <- 2^(c(-5, -4, -3, -2, -1, 0, 1, 2, 3, 4, 5, 6))
gamma_ik_vec <- sapply(sigma_2_vec, get_responsibility)
plot_tbl <- tibble::tibble(
  exponent = exponent_vec,
  sigma_2 = sigma_2_vec,
  gamma_ik = gamma_ik_vec
)
p <- ggplot(
  plot_tbl,
  aes(x = exponent, y = gamma_ik)
) +
  geom_line() +
  geom_point() +
  scale_x_continuous(
    breaks = exponent_vec,
    labels = scales::math_format(2^.x)
  ) +
  labs(
    x = expression(sigma^2),
    y = expression(gamma[i1])
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy")
p
```

All probabilities are therefore driven to 0 or 1, which makes the algorithm akin to hard clustering algorithm, such as K-means, where points either belong to a cluster or they do not.

As another comment, observe that $\gamma_{i1}$ tends to 0.5 as the variance tends to infinity, which makes intuitive sense because the component densities become flatter and more similar to each other, so the prior weights (which are equal in this example) dominate the responsibilities.

### Part E: Geometric Covariance Decompositions (mclust)

* **Answer 1:**
  * a. $\lambda_k$ is a scalar controlling the overall **volume** of the cluster. $D_k$ is an orthogonal matrix of eigenvectors controlling the **orientation** of the cluster in space. $\Lambda_k$ is a diagonal matrix of eigenvalues (normalized so its determinant is 1) controlling the geometric **shape**.
  * b. We force the volume parameter to be identical across all groups ($\lambda_k = \lambda$) but allow the shape matrices ($\Lambda_k$) and/or the orientation matrices ($D_k$) to vary independently for each $k$.
  * c. No, it makes no difference. If the relative stretch along each axis is identical, it makes no difference what the orientation is.
  * d. We constrain all elements to be identical across components: $\lambda_k = \lambda$, $D_k = D$, and $\Lambda_k = \Lambda$ (this is the "EEE" model in `mclust`). To allow the volume to be something other than 1, we simply leave $\lambda$ as a free parameter to be estimated via Maximum Likelihood from the data, rather than fixing it to a constant scalar of 1.

# Clustering II

## Section 1: DBSCAN

