---
title: "Autoencoder Tutorial"
author: "Juwa"
date: "06/09/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(ggplot2)
library(keras3)
library(scatterplot3d)

set_random_seed(5077)
```

# PCA versus an autoencoder

This example compares PCA and an autoencoder for dimensionality reduction. We use the following libraries:

- `keras3`, for fitting autoencoders and loading the MNIST dataset;
- `ggplot2`, for two-dimensional plots; and
- `scatterplot3d`, for three-dimensional plots.

The `lle` package is no longer required.
It was loaded only for visualisation, while the document actually uses `scatterplot3d` for the three-dimensional plots. The `h2o` package was also loaded but not used.

Run the following once, outside the knitted document, if Keras has not yet been installed:

```{r install-keras, eval=FALSE}
install.packages(c("keras3", "ggplot2", "scatterplot3d"))
keras3::install_keras(backend = "tensorflow")
```



By visualising only two principal components, we do not see all of the variation in the data. Plotting the observations in three dimensions can give a better indication of their structure, although variation represented by the remaining components is still omitted.

```{r pca-3d}
scatterplot3d(
  x = pca$x[plot_index, 1],
  y = pca$x[plot_index, 2],
  z = pca$x[plot_index, 3],
  color = as.integer(factor(y_train[plot_index])),
  pch = 16,
  cex.symbols = 0.35,
  xlab = "PC1",
  ylab = "PC2",
  zlab = "PC3"
)
```

The third dimension can help separate some of the digits.



# Autoencoder in Keras

The installation commands should not be placed inside a chunk that runs whenever the document is knitted. The document therefore assumes that `keras3` and its backend have already been installed using the installation chunk above.





# Compile the model

```{r autoencoder-2d-compile}
model |>
  compile(
    loss = "mean_squared_error",
    optimizer = "adam"
  )
```

# Fit the model





# Summary of the model


# Compile the model

```{r autoencoder-3d-compile}
model3 |>
  compile(
    loss = "mean_squared_error",
    optimizer = "adam"
  )
```

# Fit the model



# Evaluate the model


# Extract the bottleneck layer





