---
title: "Autoencoders: Learning a Representation by Reconstruction"
author: Miguel Rodo
date: "2026-08-05"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
---

```{r}
#| label: setup
#| include: false

library(tibble)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)

knitr::opts_chunk$set(
  fig.align = "center",
  out.width = "80%"
)

plot_theme <- theme_minimal(base_size = 13) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

curve_data <- function(
  n = 250,
  seed = 5077,
  noise_sd = 0.08
) {
  set.seed(seed)
  t <- sort(runif(n, -2.8, 2.8))

  x1 <- t
  x2 <- 0.45 * t^2 + sin(1.5 * t)

  tibble(
    t = t,
    x1 = x1 + rnorm(n, sd = noise_sd),
    x2 = x2 + rnorm(n, sd = noise_sd),
    curve_x1 = x1,
    curve_x2 = x2
  )
}

reconstruct_pca <- function(x, k) {
  pca_fit <- prcomp(x, center = TRUE, scale. = FALSE)

  scores <- pca_fit$x[, seq_len(k), drop = FALSE]
  reconstruction <-
    scores %*% t(pca_fit$rotation[, seq_len(k), drop = FALSE])

  reconstruction <- sweep(
    reconstruction,
    MARGIN = 2,
    STATS = pca_fit$center,
    FUN = "+"
  )

  list(
    pca = pca_fit,
    reconstruction = reconstruction
  )
}

architecture_tbl <- tibble(
  layer = c(
    "Input",
    "Hidden",
    "Code",
    "Hidden",
    "Output"
  ),
  units = c("p", "h", "q", "h", "p"),
  x = seq_along(layer),
  y = 1
)

make_architecture_plot <- function() {
  edge_tbl <- tibble(
    x = architecture_tbl$x[-nrow(architecture_tbl)],
    xend = architecture_tbl$x[-1],
    y = 1,
    yend = 1
  )

  ggplot() +
    geom_segment(
      data = edge_tbl,
      aes(x = x, xend = xend, y = y, yend = yend),
      arrow = arrow(length = unit(0.18, "cm")),
      linewidth = 0.5
    ) +
    geom_point(
      data = architecture_tbl,
      aes(x = x, y = y, size = layer == "Code"),
      shape = 21,
      fill = "white",
      stroke = 1.1
    ) +
    geom_text(
      data = architecture_tbl,
      aes(x = x, y = y + 0.22, label = layer),
      size = 4
    ) +
    geom_text(
      data = architecture_tbl,
      aes(x = x, y = y - 0.22, label = units),
      size = 5
    ) +
    scale_size_manual(values = c(`TRUE` = 12, `FALSE` = 9), guide = "none") +
    coord_cartesian(xlim = c(0.6, 5.4), ylim = c(0.55, 1.45)) +
    theme_void()
}

make_denoising_plot <- function() {
  base_curve <- curve_data(n = 120, seed = 99, noise_sd = 0)
  corrupted <- base_curve |>
    mutate(
      noisy_x1 = x1 + rnorm(n(), sd = 0.25),
      noisy_x2 = x2 + rnorm(n(), sd = 0.25)
    ) |>
    slice(seq(1, n(), by = 8))

  ggplot() +
    geom_path(
      data = base_curve,
      aes(x = x1, y = x2),
      linewidth = 1
    ) +
    geom_segment(
      data = corrupted,
      aes(
        x = noisy_x1,
        y = noisy_x2,
        xend = x1,
        yend = x2
      ),
      arrow = arrow(length = unit(0.16, "cm")),
      alpha = 0.7
    ) +
    geom_point(
      data = corrupted,
      aes(x = noisy_x1, y = noisy_x2),
      size = 1.7,
      alpha = 0.65
    ) +
    labs(
      x = "Original variable 1",
      y = "Original variable 2"
    ) +
    plot_theme
}
```

## Context

We have already looked at several ways to represent multivariate observations in fewer dimensions.

$$
\begin{array}{lll}
\text{PCA}
&:&
\text{linear component scores},
\$$3pt]
\text{MDS}
&:&
\text{coordinates chosen to represent dissimilarities},
\$$3pt]
\text{ISOMAP}
&:&
\text{coordinates chosen to represent graph distances},
\$$3pt]
\text{SOM}
&:&
\text{assignment to a position on a prototype grid}.
\end{array}
$$

Autoencoders approach dimension reduction from a different angle.

Instead of beginning with a distance matrix or a variance criterion, they begin with a reconstruction task.

### The question

Can we learn a lower-dimensional representation that still contains enough information to reconstruct the original observation?

$$
\text{observation}
\longrightarrow
\text{low-dimensional representation}
\longrightarrow
\text{reconstruction}
$$

This makes autoencoders useful for dimension reduction, but also makes them different from methods that explicitly preserve distances, ranks or variance.

## Encoder, code and decoder

For an observation

$$
\mathbf x_i\in\mathbb R^p,
$$

an autoencoder learns two functions.

The **encoder** maps the observation into a code or latent representation:

$$
\mathbf z_i = f_{\boldsymbol\theta}(\mathbf x_i).
$$

The **decoder** maps the code back into the original data space:

$$
\widehat{\mathbf x}_i
=
g_{\boldsymbol\phi}(\mathbf z_i).
$$

Putting the two functions together,

$$
\widehat{\mathbf x}_i
=
g_{\boldsymbol\phi}
\left(
  f_{\boldsymbol\theta}(\mathbf x_i)
\right).
$$

### Basic architecture

```{r}
#| label: fig-basic-autoencoder-architecture
#| echo: false
#| fig-width: 8
#| fig-height: 3
#| out-width: "75%"
#| fig-cap: "An autoencoder maps an observation through an encoder, a low-dimensional code and a decoder."

make_architecture_plot()
```

The middle layer is often called the **bottleneck**.

A typical architecture might be written as

$$
p
\longrightarrow
h
\longrightarrow
q
\longrightarrow
h
\longrightarrow
p.
$$

For example, for MNIST images,

$$
784
\longrightarrow
256
\longrightarrow
3
\longrightarrow
256
\longrightarrow
784.
$$

The architecture is often drawn symmetrically, but symmetry is not required.

## Why this is dimension reduction

If

$$
\mathbf x_i\in\mathbb R^p
\qquad\text{and}\qquad
\mathbf z_i\in\mathbb R^q,
$$

with

$$
q<p,
$$

then the observation is being represented using fewer coordinates than the original data.

This is called an **undercomplete autoencoder**.

### Two spaces

It is important to keep two spaces separate.

| Space | Objects |
|---|---|
| Original data space | $\mathbf x_i$ and $\widehat{\mathbf x}_i$ |
| Latent space | $\mathbf z_i$ |

The encoder moves from the data space to the latent space.

The decoder moves from the latent space back to the data space.

## Reconstruction loss

The autoencoder is trained to make reconstructions close to the original observations.

For continuous observations, a common loss is squared reconstruction error.

For one observation,

$$
\ell_i
=
\|\mathbf x_i-\widehat{\mathbf x}_i\|^2.
$$

For the full dataset,

$$
L(\boldsymbol\theta,\boldsymbol\phi)
=
\frac1n
\sum_{i=1}^n
\left\|
\mathbf x_i
-
g_{\boldsymbol\phi}
\left(
  f_{\boldsymbol\theta}(\mathbf x_i)
\right)
\right\|^2.
$$

The parameters $\boldsymbol\theta$ and $\boldsymbol\phi$ are chosen to reduce this reconstruction loss.

### Input and target

There is no external response variable.

The input observation supplies its own training target:

$$
\text{input}=\mathbf x_i,
\qquad
\text{target}=\mathbf x_i.
$$

For this reason, autoencoders are usually treated as unsupervised methods, although the training task is sometimes described as self-supervised.

## Why reconstruction alone is not enough

A very flexible network may learn to copy the input:

$$
g(f(\mathbf x_i))\approx \mathbf x_i.
$$

That can give low reconstruction error without giving a useful representation.

So the key question is not just whether the network reconstructs well.

The key question is whether the reconstruction task forces the network to learn meaningful structure.

### Two ways to prevent trivial copying

There are two broad strategies.

1. **Restrict the architecture.**

   Use a bottleneck with $q<p$.

2. **Regularise the learnt mapping.**

   Add restrictions that discourage copying even when the code is not very small.

Examples include denoising, sparsity and penalising sensitivity to small perturbations.

### Important qualification

A narrow bottleneck does not automatically make the latent coordinates simple or interpretable.

The encoder and decoder can still be complicated functions.

The bottleneck restricts information flow, but it does not guarantee that the latent representation preserves distances, clusters or any particular scientific structure.

## A minimal neural-network layer

A neural-network layer has the form

$$
\mathbf h
=
\sigma(\mathbf W\mathbf x+\mathbf b),
$$

where:

- $\mathbf W$ contains weights;
- $\mathbf b$ contains bias terms;
- $\sigma$ is an activation function.

If $\sigma$ is nonlinear, stacking layers gives a flexible nonlinear function.

For example, an encoder could be written as

$$
\mathbf z
=
\sigma_2
\left[
  \mathbf W_2
  \sigma_1(\mathbf W_1\mathbf x+\mathbf b_1)
  +
  \mathbf b_2
\right].
$$

The decoder applies additional layers to map $\mathbf z$ back to the original data space.

### Common activation functions

| Activation | Typical role |
|---|---|
| Linear | Useful for unrestricted continuous output |
| ReLU | Common in hidden layers |
| Sigmoid | Useful for outputs constrained to $[0,1]$ |

Without nonlinear activation functions, a stack of linear layers is still just a linear transformation.

## How the network learns

Training usually follows the same basic loop.

$$
\boxed{
\text{forward pass}
\longrightarrow
\text{loss}
\longrightarrow
\text{gradients}
\longrightarrow
\text{parameter update}
}
$$

In words:

1. Pass a batch of observations through the encoder and decoder.
2. Calculate the reconstruction loss.
3. Use backpropagation to calculate how the loss changes with each parameter.
4. Use an optimiser, such as Adam, to update the parameters.
5. Repeat for several epochs.

The aim here is not to derive backpropagation, but to understand what the optimisation is trying to achieve.

## Training, validation and test data

The data are usually divided into three roles.

| Data | Role |
|---|---|
| Training data | Used to estimate the network weights |
| Validation data | Used to monitor performance during training |
| Test data | Used for final assessment |

A **batch** is a subset of observations used for one parameter update.

An **epoch** is one pass through the training data.

Validation loss is useful because a flexible autoencoder can overfit.

$$
\text{low training loss}
\not\Rightarrow
\text{good reconstruction of new observations}.
$$

### Early stopping

One simple safeguard is early stopping.

The model is trained until validation loss stops improving.

This avoids continuing to reduce training loss after the model has stopped improving on held-out data.

## Linear autoencoders and PCA

The connection with PCA is useful because PCA is already familiar.

Suppose the data are centred and the encoder and decoder are linear.

Let

$$
\mathbf z_i
=
\mathbf W_E\mathbf x_i
$$

and

$$
\widehat{\mathbf x}_i
=
\mathbf W_D\mathbf z_i
=
\mathbf W_D\mathbf W_E\mathbf x_i.
$$

The autoencoder minimises

$$
\sum_{i=1}^n
\left\|
\mathbf x_i
-
\mathbf W_D\mathbf W_E\mathbf x_i
\right\|^2.
$$

Since $\mathbf W_D\mathbf W_E$ has rank at most $q$, the model is looking for a rank-$q$ reconstruction.

### PCA connection

A linear undercomplete autoencoder trained with squared reconstruction error learns the same optimal reconstruction subspace as PCA.

So PCA can be seen as a special linear reconstruction method.

A nonlinear autoencoder replaces the linear encoder and decoder with flexible nonlinear functions.

$$
\boxed{
\text{linear autoencoder}
\approx
\text{PCA reconstruction subspace}
}
$$

$$
\boxed{
\text{nonlinear autoencoder}
=
\text{learnt nonlinear reconstruction}
}
$$

### What this does not mean

The latent coordinates need not be exactly the PCA scores.

If $\mathbf A$ is an invertible $q\times q$ matrix, then

$$
\mathbf z_i^*=
\mathbf A\mathbf z_i
$$

can be reversed by changing the decoder.

For example, the decoder weights can be transformed using $\mathbf A^{-1}$.

The reconstruction can remain the same even though the latent axes have been rotated or rescaled.

Thus:

- PCA gives ordered orthogonal axes;
- an unconstrained linear autoencoder gives the same principal subspace;
- the autoencoder code may have a different orientation and scale.

## A curved one-dimensional example

PCA is linear.

This means that a one-dimensional PCA reconstruction lies on a straight line.

But many datasets have nonlinear structure.

```{r}
#| label: fig-curved-observations
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.5
#| out-width: "65%"
#| fig-cap: "A simple dataset concentrated near a curved one-dimensional structure."

curve_tbl <- curve_data()

curve_tbl |>
  ggplot(aes(x = x1, y = x2)) +
  geom_point(alpha = 0.65, size = 1.4) +
  labs(
    x = "Original variable 1",
    y = "Original variable 2"
  ) +
  plot_theme
```

The observations are recorded in two variables, but most variation follows a curved one-dimensional path.

## PCA reconstruction of the curve

With one component, PCA must reconstruct the data onto a straight line.

```{r}
#| label: fig-pca-curve-reconstruction
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.5
#| out-width: "65%"
#| fig-cap: "One-dimensional PCA reconstruction is restricted to a straight line."

curve_x <- as.matrix(curve_tbl[, c("x1", "x2")])
curve_pca <- reconstruct_pca(curve_x, k = 1)

pca_recon_tbl <- bind_cols(
  curve_tbl,
  tibble(
    recon_x1 = curve_pca$reconstruction[, 1],
    recon_x2 = curve_pca$reconstruction[, 2]
  )
)

pca_recon_tbl |>
  ggplot(aes(x = x1, y = x2)) +
  geom_segment(
    aes(xend = recon_x1, yend = recon_x2),
    alpha = 0.15
  ) +
  geom_point(alpha = 0.55, size = 1.2) +
  geom_point(
    aes(x = recon_x1, y = recon_x2),
    size = 0.7,
    alpha = 0.7
  ) +
  labs(
    x = "Original variable 1",
    y = "Original variable 2"
  ) +
  plot_theme
```

This is the best one-dimensional linear reconstruction under squared error.

But it cannot bend with the data.

## Nonlinear reconstruction idea

A nonlinear autoencoder with a one-dimensional code can learn

$$
\mathbb R^2
\longrightarrow
\mathbb R
\longrightarrow
\mathbb R^2.
$$

The decoder maps a one-dimensional latent coordinate back into the original two-dimensional space.

As the latent coordinate varies, the decoder can trace out a curved path.

```{r}
#| label: fig-autoencoder-curve-idea
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.5
#| out-width: "65%"
#| fig-cap: "A nonlinear decoder can represent a curved reconstruction path in the original data space."

curve_line <- curve_tbl |>
  arrange(t) |>
  transmute(x1 = curve_x1, x2 = curve_x2)

curve_tbl |>
  ggplot(aes(x = x1, y = x2)) +
  geom_point(alpha = 0.45, size = 1.2) +
  geom_path(
    data = curve_line,
    aes(x = x1, y = x2),
    linewidth = 1
  ) +
  labs(
    x = "Original variable 1",
    y = "Original variable 2"
  ) +
  plot_theme
```

The key point is not that the curve above has been fitted by an autoencoder.

The point is geometric.

A nonlinear decoder is not restricted to straight reconstruction lines or planes.

## MNIST example

We now consider handwritten digit images.

Each image has size

$$
28\times 28.
$$

After flattening the image, each observation is a vector in

$$
\mathbb R^{784}.
$$

The labels are not used to train the autoencoder.

They are useful afterwards, because they help us interpret the learnt representation.

### Data preparation

The usual preparation is:

1. load the MNIST images;
2. flatten each image into a 784-dimensional vector;
3. scale pixel intensities to lie between 0 and 1;
4. split data into training, validation and test sets.

```{r}
#| label: mnist-preparation-template
#| eval: false

mnist <- keras3::dataset_mnist()

x_train <- mnist$train$x
x_test <- mnist$test$x

y_train <- mnist$train$y
y_test <- mnist$test$y

x_train <- keras3::array_reshape(
  x_train,
  c(nrow(x_train), 784)
)

x_test <- keras3::array_reshape(
  x_test,
  c(nrow(x_test), 784)
)

x_train <- x_train / 255
x_test <- x_test / 255
```

## PCA benchmark on MNIST

PCA gives a familiar baseline.

Questions to ask:

- How much variance is explained by the first few components?
- Which digits separate clearly in the first two or three components?
- Which digits overlap?
- How well can images be reconstructed using $k$ components?

```{r}
#| label: pca-mnist-template
#| eval: false

pca <- prcomp(x_train)

cumulative_variance <-
  cumsum(pca$sdev^2) / sum(pca$sdev^2)

plot(
  cumulative_variance,
  type = "l",
  xlab = "Number of principal components",
  ylab = "Cumulative proportion of variance explained"
)
```

This reminds us that PCA can be viewed as a reconstruction method.

With $k$ components, PCA reconstructs each observation using a $k$-dimensional linear subspace.

## Autoencoder architecture for MNIST

A simple architecture is

$$
784
\longrightarrow
256
\longrightarrow
q
\longrightarrow
256
\longrightarrow
784.
$$

The bottleneck dimension $q$ controls the size of the latent representation.

For a visualisation, one might use $q=2$ or $q=3$.

For a reconstruction task, one might compare several values of $q$.

```{r}
#| label: mnist-autoencoder-template
#| eval: false

model <- keras_model_sequential(
  input_shape = c(ncol(x_train))
) |>
  layer_dense(units = 256, activation = "relu") |>
  layer_dense(
    units = q,
    activation = "relu",
    name = "bottleneck"
  ) |>
  layer_dense(units = 256, activation = "relu") |>
  layer_dense(units = ncol(x_train))

model |>
  compile(
    loss = "mean_squared_error",
    optimizer = "adam"
  )
```

### Fitting the model

The target is the same image supplied as input.

$$
\mathbf x_i
\longrightarrow
\mathbf z_i
\longrightarrow
\widehat{\mathbf x}_i.
$$

```{r}
#| label: mnist-autoencoder-fit-template
#| eval: false

history <- model |>
  fit(
    x = x_train,
    y = x_train,
    epochs = 25,
    batch_size = 128,
    validation_split = 0.2,
    callbacks = list(
      callback_early_stopping(
        monitor = "val_loss",
        patience = 3,
        restore_best_weights = TRUE
      )
    )
  )
```

The training output should be interpreted using both training loss and validation loss.

The final assessment should use test-set reconstruction error.

## Extracting the latent representation

After fitting, the encoder can be extracted from the full autoencoder.

```{r}
#| label: mnist-encoder-template
#| eval: false

encoder <- keras_model(
  inputs = model$inputs,
  outputs = get_layer(model, "bottleneck")$output
)

latent <- encoder |>
  predict(x_test)
```

If $q=2$, the latent coordinates can be plotted directly.

If $q=3$, they can be plotted in three dimensions.

Digit labels can be used to colour the points, but they were not used during training.

### Why can digit groups appear?

The autoencoder is not told the digit labels.

However, digit identity affects the pixel patterns needed for reconstruction.

So some digit structure may appear in the latent representation because it helps to reconstruct the images.

But this is not guaranteed.

The loss function asks for good reconstruction, not class separation.

## Reconstruction comparison

A useful display uses the same test images in three rows.

1. Original image.
2. PCA reconstruction.
3. Autoencoder reconstruction.

Questions to ask:

- Which reconstruction is sharper?
- Which digit features are retained?
- Which digits are confused?
- Does lower reconstruction error imply better class separation?

### Reconstruction error

For each test observation,

$$
\ell_i
=
\|\mathbf x_i-\widehat{\mathbf x}_i\|^2.
$$

The test reconstruction MSE is the average over test observations.

$$
\text{test MSE}
=
\frac1{n_{\text{test}}}
\sum_{i=1}^{n_{\text{test}}}
\|\mathbf x_i-\widehat{\mathbf x}_i\|^2.
$$

## Choosing the latent dimension

There is no direct autoencoder equivalent of the PCA proportion-of-variance-explained curve.

A practical approach is to fit several bottleneck dimensions and compare reconstruction error.

$$
q
\quad\text{against}\quad
\text{test reconstruction MSE}.
$$

```{r}
#| label: fig-latent-dimension-reconstruction-illustration
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
#| out-width: "70%"
#| fig-cap: "Illustrative reconstruction-error curves. Actual values depend on the data, architecture and training procedure."

illustrative_df <- tibble(
  q = rep(1:10, 2),
  method = rep(c("PCA", "Autoencoder"), each = 10),
  mse = c(
    0.17, 0.13, 0.105, 0.087, 0.075,
    0.066, 0.059, 0.053, 0.048, 0.044,
    0.15, 0.105, 0.080, 0.064, 0.054,
    0.047, 0.042, 0.039, 0.036, 0.034
  )
)

illustrative_df |>
  ggplot(aes(x = q, y = mse, linetype = method)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    x = "Reduced dimension",
    y = "Test reconstruction MSE",
    linetype = "Method"
  ) +
  plot_theme
```

As $q$ increases, reconstruction error usually decreases.

But increasing $q$ also weakens the compression.

The choice of $q$ is therefore a trade-off between compression, reconstruction accuracy and interpretability.

## Interpreting the latent space

The latent space is learnt through reconstruction.

It is not automatically a faithful map of the original distances.

A good slide to remember is:

$$
\boxed{
\text{reconstruction objective}
\neq
\text{distance-preservation objective}
}
$$

Low reconstruction error does not guarantee that:

- Euclidean distances in latent space are meaningful;
- nearest neighbours are preserved;
- clusters are separated;
- axes are ordered by importance;
- axes are uniquely identified;
- two training runs have the same orientation.

### What the encoder gives us

A major practical advantage is that the encoder is a fitted function:

$$
f_{\boldsymbol\theta}:
\mathbb R^p\to\mathbb R^q.
$$

This gives a direct representation for a new observation.

Many embedding methods, such as classical MDS or ordinary ISOMAP, do not naturally provide such a direct encoder.

### What the decoder gives us

The decoder is also a fitted function:

$$
g_{\boldsymbol\phi}:
\mathbb R^q\to\mathbb R^p.
$$

This allows us to map latent coordinates back into the original data space.

For images, this means that we can inspect what reconstructed image corresponds to a given latent coordinate.

## Denoising autoencoders

A denoising autoencoder is trained using corrupted inputs.

Let

$$
\widetilde{\mathbf x}
\sim
C(\widetilde{\mathbf x}\mid\mathbf x)
$$

be a corrupted version of $\mathbf x$.

The network receives $\widetilde{\mathbf x}$, but is trained to reconstruct $\mathbf x$.

$$
\widetilde{\mathbf x}
\longrightarrow
f(\widetilde{\mathbf x})
\longrightarrow
\widehat{\mathbf x}
\approx
\mathbf x.
$$

So:

$$
\text{input}=\widetilde{\mathbf x},
\qquad
\text{target}=\mathbf x.
$$

### Denoising loss

For squared error,

$$
L
=
\frac1n
\sum_{i=1}^n
\left\|
\mathbf x_i
-
g
\left(
  f(\widetilde{\mathbf x}_i)
\right)
\right\|^2.
$$

The model cannot solve the task by simply copying its input.

It must learn regularities that allow it to infer plausible clean observations.

## Geometric interpretation of denoising

If observations concentrate near a low-dimensional structure, corruption can move them away from that structure.

A denoising autoencoder learns to map corrupted observations back towards plausible observations.

```{r}
#| label: fig-denoising-manifold-idea
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.5
#| out-width: "65%"
#| fig-cap: "Denoising can be interpreted as moving corrupted observations back towards the learnt data structure."

make_denoising_plot()
```

This links autoencoders back to manifold learning.

ISOMAP uses a neighbourhood graph and shortest-path distances to infer manifold geometry.

A denoising autoencoder learns a parametric mapping that removes variation that looks like corruption.

## Other regularised autoencoders

### Sparse autoencoder

A sparse autoencoder adds a penalty that encourages most latent activations to be near zero.

This allows the code to be larger while encouraging each observation to use only a small number of active features.

### Contractive autoencoder

A contractive autoencoder penalises excessive sensitivity of the encoder to small changes in the input.

Informally, nearby observations should often have similar codes unless the change corresponds to real variation in the data.

The formal penalty uses derivatives of the encoder.

### Variational autoencoder

A variational autoencoder is a probabilistic extension.

It learns a structured latent distribution and can be used for generating new observations.

We will not cover the variational autoencoder objective here.

## Practical decisions

### Scaling and loss

Squared reconstruction error is scale-dependent.

Variables on larger scales can dominate the loss.

Therefore, continuous variables are often standardised before training.

For MNIST, pixel intensities are scaled to lie in $[0,1]$.

### Match the loss to the data

| Data type | Common output choice | Possible loss |
|---|---|---|
| Continuous unrestricted | Linear | Squared error |
| Binary | Sigmoid | Binary cross-entropy |
| Values in $[0,1]$ | Linear or sigmoid | Depends on modelling choice |

Autoencoders do not have to use squared error.

The output layer and reconstruction loss should reflect the kind of data being reconstructed.

### Capacity and computation

Representation quality depends on:

- bottleneck dimension;
- number of hidden layers;
- width of hidden layers;
- activation functions;
- regularisation;
- training duration;
- random initialisation.

Autoencoders are much more flexible than PCA, but also more computationally demanding and less uniquely defined.

## Comparison with previous methods

| Method | Representation | Main objective | New observations | Reconstruction |
|---|---|---|---|---|
| PCA | Linear scores | Linear reconstruction or variance | Direct linear map | Direct |
| MDS | Coordinates | Represent dissimilarities | Usually indirect | No natural decoder |
| ISOMAP | Coordinates | Represent graph distances | Usually indirect | No natural decoder |
| SOM | Grid positions | Learn organised prototypes | Assign nearest prototype | Prototype approximation |
| Autoencoder | Learnt code | Reconstruction | Direct encoder | Direct decoder |

The main contrast is:

$$
\boxed{
\text{ISOMAP preserves a chosen geometry}
}
$$

while

$$
\boxed{
\text{autoencoders preserve information needed for reconstruction}.
}
$$

Those are related ideas, but they are not the same objective.

## Key ideas

1. An autoencoder learns an encoder and a decoder.
2. The code is a reduced representation when its dimension is smaller than the input dimension.
3. The model is trained by minimising reconstruction loss.
4. Reconstruction alone does not guarantee a useful representation.
5. Undercomplete and regularised autoencoders discourage trivial copying in different ways.
6. A linear undercomplete autoencoder trained by squared reconstruction error learns the PCA principal subspace.
7. Nonlinear autoencoders can learn curved reconstruction structures.
8. The latent coordinates are not ordered or uniquely identified like PCA components.
9. Good reconstruction does not guarantee preservation of distances, neighbours or clusters.
10. Denoising autoencoders reconstruct clean observations from corrupted inputs.
11. Autoencoders give a direct encoder for new observations.
12. Autoencoders also give a decoder back to the original data space.

## Conceptual checks

1. What are the roles of the encoder, code and decoder?
2. Why can an autoencoder be treated as an unsupervised method?
3. What is meant by reconstruction loss?
4. Why might a very flexible autoencoder learn an unhelpful identity mapping?
5. How does an undercomplete autoencoder restrict the model?
6. In what sense is a linear autoencoder related to PCA?
7. Why need the latent coordinates of a linear autoencoder not equal the PCA scores?
8. Why does nonlinear activation matter?
9. Does lower reconstruction error imply better preservation of pairwise distances?
10. Why can digit groups appear in an MNIST latent space even if labels were not used in training?
11. Why should both training and validation loss be examined?
12. Why can latent representations differ across training runs?
13. What is the input and what is the target in a denoising autoencoder?
14. Why does denoising discourage trivial copying?
15. What advantage does an autoencoder have over ISOMAP when a new observation arrives?
16. Why might clustering latent coordinates be misleading?

## References

Goodfellow, I., Bengio, Y. and Courville, A. (2016). *Deep Learning*. MIT Press. Chapter 14: Autoencoders.

Hastie, T., Tibshirani, R. and Friedman, J. (2009). *The Elements of Statistical Learning*. Springer. Chapters 11 and 14 provide useful background on neural networks and unsupervised learning, although autoencoders are not a central topic there.
