---
title: "Confidence Intervals, Contrasts, and Global Tests in Multiple Regression"
subtitle: "Third R Practical — CIs for coefficients, contrasts, nested-model ANOVA, and if–else coverage checks"
format:
  html:
    self-contained: true
    code-fold: false
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

# Main practical

## Part A: Confidence intervals and nested-model tests

We work with the multiple regression model

$$
y_i \;=\; \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \varepsilon_i, 
\quad \varepsilon_i \overset{\text{iid}}{\sim} \mathcal N(0, \sigma^2),\quad i=1,\dots,n.
$$

### 1. Construct the data

```{r}
set.seed(2025)

# Baseline parameters
n       <- 100
beta_0  <- 2
beta_1  <- +0.8
beta_2  <- -0.8
sigma   <- 1.5

# Simulate correlated predictors with Corr(x1, x2) ~ 0.4 using a simple construction
rho     <- 0.4
x1      <- rnorm(n, mean = 0, sd = 1)
eta     <- rnorm(n, mean = 0, sd = 1)
x2      <- rho * x1 + sqrt(1 - rho^2) * eta

# Response
y <- beta_0 + beta_1 * x1 + beta_2 * x2 + rnorm(n, 0, sigma)

dat <- data.frame(y, x1, x2)
head(dat, 5)
```

### 2. Fit the full model and inspect the summary

```{r}
fit_full <- lm(y ~ x1 + x2, data = dat)
summary(fit_full)
```

### 3. Inference for a single regression coefficient $\beta_1$

#### Calculating required quantities

a. Construct the design matrix $\mathbf{X}$, save it as `X`. Compare it to `dat` as well as the output of `model.matrix()`.

```{r}
#| include: false
X <- cbind(1, dat[, c("x1", "x2")]) |> as.matrix()  # add intercept column
colnames(X) <- c("(Intercept)", "x1", "x2")
```

b. Calculate the estimated regression coefficients $\hat{\mathbf{\beta}}$. Save the coefficient for $\beta_1$ as `b1_hat`.

As from last week, this is given by

```{r}
y_mat <- matrix(dat$y, ncol = 1)  # response vector as a matrix
beta_vec <- solve(t(X) %*% X) %*% t(X) %*% y_mat
beta_vec |> signif(3)
```

We can double check this against the output of `coef(fit_full)`:

```{r}
coef(fit_full) |> signif(3)
```

Saving the coefficient for $\beta_1$ as `b1_hat`:

```{r}
b1_hat <- beta_vec[2]
b1_hat |> signif(3)
```

c. Calculate the unbiased estimator of $\sigma^2$ using the residuals from `fit_full`. Compare it to the output of `summary(fit_full)$sigma`.

The unbiased estimator of $\sigma^2$ is given by

$$
s^2 = \frac{\sum_{i=1}^n\hat{e}_i^2}{n - k},
$$

Let's first obtain these quantities and then combine them.

We know from the simulation that $n$ is 100 and $k$ is 3 (intercept, $x_1$, $x_2$).
However, it is less error-prone to calculate these values automatically rather than hard-coding them.

One way to set `n` and `k` is to use the dimensions of the design matrix `X`:

```{r}
n <- nrow(X)  # number of observations
k <- ncol(X)  # number of parameters (including intercept)
```

Now, we need the residuals from the fitted model `fit_full`:

```{r}
resid_vec <- residuals(fit_full)  # residuals from the full model
```

**Extra info**: If you run `?residuals.lm`, you'll see that there are various definitions of residuals, but the default `"working"` is the difference between the observed and fitted values, which is what we want here.

Now, we can compute the unbiased estimator of $\sigma^2$:

```{r}
s2 <- sum(resid_vec^2) / (n - k)
s2 |> signif(3)
```

Examining the output of `summary(fit_full)$sigma`:

```{r}
summary(fit_full)$sigma |> signif(3)
```

They are not the same - the `summary()` output gives the **standard deviation** of the residuals, while we computed the **variance**. To get the standard deviation, we take the square root:

```{r}
sqrt(s2) |> signif(3)
```

d. Compute the estimated covariance matrix of $\hat{\mathbf{\beta}}$. Compare it to the output of `vcov(fit_full)`.

The formula for the covariance matrix of the estimated coefficients $\hat{\beta}$ is given by

$$
\widehat{\mathrm{Var}}(\hat\beta)\;=\;s^2\,(X^\top X)^{-1},
$$

```{r}
cov_manual <- s2 * solve(t(X) %*% X)
cov_manual |> signif(3)
```

Compare this to the output of `vcov(fit_full)`:

```{r}
cov_builtin <- vcov(fit_full)
cov_builtin |> signif(3)
```

We can check if they are equal:

```{r}
all.equal(cov_manual, cov_builtin)
```

**Extra info**: We use `all.equal()` instead of `identical` because there may be small numerical differences due to the way arithmetic works in R for real numbers (so-called "floating-point" numbers). The `tolerance` argument allows for a small margin of error. We can see that `identical` regards these covariance matrices are different:

```{r}
identical(cov_manual, cov_builtin)  # should be FALSE due to numerical precision
```

If we sum over the squared differences, we can see how close they are:

```{r}
sum((cov_manual - cov_builtin)^2)
```

The above is very small but non-zero, hence the different results between `all.equal()` and `identical()`.

e. Calculate the standard error of $\hat{\beta}_1$.

We just extract the diagonal element corresponding to $\hat{\beta}_1$ from the covariance matrix:

```{r}
se_b1 <- sqrt(cov_manual["x1", "x1"])
se_b1 |> signif(3)
```

**Important**:
Do *not* round (`round`) or truncate (`signif`) the standard error before using it in further calculations, as this can lead to incorrect results.
Same goes for any other intermediate quantities.
Always use the full precision of the number.
We just use `signif()` here for display purposes.

For example, do *not* truncate before saving the value to `se_b1_signif` (as below), and then use `se_b1_signif` in further calculations:

```{r}
se_b1_signif <- sqrt(cov_manual["x1", "x1"]) |>
  signif(3)  # this is wrong for further calculations!
```

#### Confidence interval 

f. Calculate the confidence interval for $\beta_1$.

The formula for the confidence interval is

$$
\hat\beta_1 \pm t_{1-\alpha/2,\,n-k}\,\widehat{\mathrm{SE}}(\hat\beta_1),
$$

Setting $\alpha = 0.05$ (for a 95% CI), we need the t-critical value for $n-k$ degrees of freedom, which we can obtain using the `qt()` function:

```{r}
t_crit <- qt(1 - 0.05 / 2, df = n - k)
t_crit |> signif(3)
```

**Tip**: If we had not automatically calculated $n$ and $k$ but hardcoded them, if the data changed we would have had to spot all these small places in which $n$ and $k$ are used. By using the dimensions of the design matrix (or whatever other way), we avoid this problem.

We can calculate the half-width ($t_{1-\alpha/2,\,n-k} \times \mathrm{SE}(\hat\beta_1)$) of the confidence interval as follows:

```{r}
half_width <- t_crit * se_b1
half_width |> signif(3)
```

The confidence interval is just the point estimate $\hat\beta_1$ plus/minus the half-width:

```{r}
lb <- b1_hat - half_width
ub <- b1_hat + half_width
ci_beta1 <- c(lb, ub)
ci_beta1 |> signif(3)
```

We can compare this against the output of `confint(fit_full)`:

```{r}
confint(fit_full) |> signif(3)
```

Amazing.

#### P-value

g. Calculate the p-value for the hypothesis test $H_0: \beta_1 = 0$. Compare the answer to `summary(fit_full)`.

The t-statistic for $\hat\beta_1$ is given by

$$
t = \frac{\hat\beta_1}{\mathrm{SE}(\hat\beta_1)}.
$$

To repeat what we mentioned in class, this is based on 

$$
\frac{\hat{\beta_1} - \beta_1}{\widehat{\mathrm{SE}}(\hat{\beta}_1)} \sim t_{n-k},
$$

but setting the null hypothesis $\beta_1 = 0$.

Calculating the t-statistic:

```{r}
t_stat_b1 <- b1_hat / se_b1
t_stat_b1 |> signif(3)
```

Now we can calculate the p-value using the t-distribution:

```{r}
p_value_b1 <- 2 * pt(abs(t_stat_b1), df = n - k, lower.tail = FALSE)
p_value_b1 |> signif(3)
```

We can compare this to the output of `summary(fit_full)`:

```{r}
summary(fit_full)
```

The $t$-statistic and p-value for $\hat\beta_1$ match.

#### Inference for $\beta_2$

h. Calculate the confidence interval and p-value for $\beta_2$.
i. **Extra**: If you would like to practice for loops, use a for loop to calculate the confidence intervals for $\beta_0$, $\beta_1$, and $\beta_2$. But I would suggest doing this once you've completed the remainder of the practical, which covers multiple hypothesis testing and `if`/`else` statements.

### 4. Simultaneous hypothesis testing for $\beta_1$ and $\beta_2$

Suppose we want to test the hypothesis $H_0: \beta_1 = 0, \beta_2 = 0$, where the alternative is $H_1: \text{At least one of }\beta_1\text{ and }\beta_2 \text{ not equal to }0$.

a. State the null hypothesis in the form $\symbf{L\beta}=\symbf{L\beta}^{(0)}$.
b. Construct the matrix $\symbf{L}$ for the hypothesis test, and save it as `L`.

We can construct $\symbf{L}$ as follows:

```{r}
L <- matrix(
  c(0, 1, 0,  # first row for beta_1
    0, 0, 1),   # second row for beta_2
  byrow = TRUE,
  ncol = 3
)
L
```

The $F$-statistic for the hypothesis test is given by

$$
F = \frac{(\symbf{L}\hat{\symbf{\beta}})^{\top}(\symbf{LC}\symbf{L}^{\top})^{-1}(\symbf{L}\hat{\symbf{\beta}})}{qs^2}
$$

where $q$ is the number of restrictions (here, $q=2$).

In general, there are the following quantities involved in the $F$-statistic:

- $\symbf{L}$, 
- $\hat{\symbf{\beta}}$,
- $\symbf{C}=(\symbf{X}^{\top}\symbf{X})^{-1}$,
- $s^2$, and
- $q$.

We already have $\symbf{L}$, $\hat{\symbf{\beta}}$ and $s^2$ from the previous steps.

b. Calculate $\symbf{C}$.

```{r}
C <- solve(t(X) %*% X)  # (X'X)^{-1}
C |> signif(3)
```

c. Calculate $q$.

$q$ is simply the number of restrictions, which is the number of rows in $\symbf{L}$:

```{r}
q_val <- nrow(L)
```

I saved it as `q_val` as there is a built-in function `q()` in R, so we avoid confusion.

d. Calculate the $F$-statistic for the hypothesis test.

To make the final calculation more readable, let us first calculate $\symbf{L}\hat{\symbf{\beta}}$ and then $(\symbf{LC}_{22}\symbf{L}^{\top})^{-1}$:

```{r}
L_beta <- L %*% beta_vec  # L * beta_hat
L_C_Lt_inv <- solve(L %*% C %*% t(L))  # (L * C * L')^{-1}  
```

We can first compute the numerator and denominator separately:

```{r}
numerator   <- t(L_beta) %*% L_C_Lt_inv %*% L_beta
denominator <- q_val * s2
```

```{r}
f_stat_b1_b2 <- numerator / denominator
f_stat_b1_b2 |> signif(3)
```

d. Calculate the p-value for the $F$-test.

The p-value for the $F$-test can be calculated using the `pf()` function, which gives the cumulative distribution function of the $F$-distribution:

```{r}
p_value_f_test <- pf(f_stat_b1_b2, df1 = q_val, df2 = n - k, lower.tail = FALSE)  
p_value_f_test
```

As compared to the $t-stat$ p-value, we do not multiply by 2, as the $F$-test is inherently a two-tailed test (it tests whether the linear combination of coefficients is significantly different from zero in either direction).

e. Compare the p-value to the output of `summary(fit_full)`.

The F-statistic at the bottom of the summary output tests the null hypothesis that all non-intercept coefficients are zero, which is equivalent to our hypothesis $H_0: \beta_1 = 0, \beta_2 = 0$.

```{r}
summary(fit_full)
```

Again, they match.

f. Calculate the $F$-statistic for the hypothesis test $H_0: \beta_0 = 0, \beta_1 = 0, \beta_2 = 0$.

## Part B: Programming — `if`, `else`, `ifelse`, `&&`, `||`, `&`, `|` and coverage

### Conditions

#### Individual conditions

Often we need to check whether something is true or false.

For this, we use **conditions**. A condition is an expression that evaluates to either `TRUE` or `FALSE`.

For example, this is a condition:

```{r}
3 < 5
```

This evaluates to `TRUE`, because 3 is indeed less than 5.

There are various conditional operators in `R`, such as:

- `<` (less than)
- `<=` (less than or equal to)
- `>` (greater than)
- `>=` (greater than or equal to)
- `==` (equal to)
- `!=` (not equal to)

We can use these conditions involving things other than numbers, such as strings:

```{r}
"hello" != "goodbye"
```

Various other functions return `TRUE` or `FALSE`, often starting with `is.` like `is.numeric()` but often not, like `grepl`:

```{r}
is.numeric("abc")  # checks if "abc" is numeric
grepl("hello", "he")  # checks if "hello" contains "he"
```

a. Write a condition that checks whether the string "hello" is equal to "world".
b. Write a condition that evaluates to `TRUE` when comparing "hello" and "world".
c. Write a condition that checks whether the number 10 is greater than 3.

#### Combining conditions

We can combine conditions using logical operators:

- `&` (element-wise AND):

Every condition needs to be `TRUE` for the whole expression to be `TRUE`.
This evaluates to `TRUE` as both conditions are `TRUE`:

```{r}
TRUE & TRUE
is.numeric(3) & is.numeric(5)
3 < 6 & 5 > 2
```

It is FALSE if at least one condition is `FALSE`:

```{r}
TRUE & FALSE
is.numeric("abc") & is.numeric(5)
3 < 6 & 5 < 2
```

- `|` (element-wise OR):

At least one condition needs to be `TRUE` for the whole expression to be `TRUE`.
This evaluates to `TRUE` as at least one condition is `TRUE`:

```{r}
TRUE | FALSE
is.numeric(3) | is.numeric("abc")
3 < 6 | 5 < 2
```

a. Write a condition that checks whether the number 3 is less than 5 **and** the number 10 is less than 8.
b. Write a condition that checks whether the number 3 is less than 5 **or** the number 10 is less than 8.

### `if`/`else` statements

Conditions are useful for controlling which code is run based on whether a condition is `TRUE` or `FALSE`.

To do this, we use `if` and `else` statements.

For example, suppose we wish to generate a message as to whether a p-value was statistically significant or not.
Suppose the p-value observed is 0.03:

```{r}
p_val <- 0.03
```

Then, we run a different message depending on whether the p-value is less than 0.05:

```{r}
if (p_val < 0.05) {
  "Significant, you may publish your paper"
} else {
  "Not significant, hack it further"
}
```

`if` and `else` are used to control what code is run based on a condition.
The syntax is as follows:

if (condition) {

  # code to execute if condition is TRUE

} else {

  # code to execute if condition is FALSE

}

If the condition is `TRUE`, only the code inside the first block is executed; if it is `FALSE`, only the code inside the `else` block is executed.

Note that `condition` must evaluate to a *single* `TRUE` or `FALSE` value.

For example, this throws an error, as the condition is not of type `logical`:

```{r}
#| error: TRUE
if ("abc") {
  "This will not run"
} else {
  "This will also not run"
}
```

This, too, will not run as the condition is not a single `TRUE` or `FALSE`:

```{r}
#| error: true
if (c(TRUE, TRUE)) {
  "This will not run"
} else {
  "This will also not run"
}
```

a. Write an `if`/`else` block that checks whether the number `2` is contained in the interval `c(1, 3)`. If it is, print "2 is in the interval"; otherwise, print "2 is not in the interval".
b. Write an `if`/`else` block that checks whether the string "hello" is equal to "world" OR if `3` is a number. If it is, print "Hello world!"; otherwise, print "Goodbye world!".
c. Generate an error by using an inappropriate condition in an `if`/`else`

### `ifelse` function

For conditions that evaluate to a vector of `TRUE` and `FALSE`, we can use the `ifelse()` function, which is vectorised:

```{r}
ifelse(c(TRUE, FALSE), "This is TRUE", "This is FALSE")
```

An `ifelse()` statement has the following syntax:

`ifelse(condition_vector, yes_vector, no_vector)`

where `condition_vector` is a vector of `TRUE` and `FALSE` values, `yes_vector` is the value to return when the condition is `TRUE`, and `no_vector` is the value to return when the condition is `FALSE`.

a. Use `ifelse()` to create a vector that contains "Yes" if the number is greater than 5 and "No" otherwise, for the numbers `1`, `3`, `5`, `7`, and `9`.

```{r}
numbers <- c(1, 3, 5, 7, 9)
ifelse(numbers > 5, "Yes", "No")
```

b. Use `ifelse()` to create a vector that contains "Even" if the number is even and "Odd" if the number is odd, for the numbers `1`, `2`, `3`, `4`, and `5`. Hint: use the check if the number divided by 2 has a remainder of 0, i.e. `<number> %% 2 == 0`, e.g. `4 %% 2 == 0` is `TRUE` as 4 is even and so `4 %% 2` returns `0` but `5 %% 2 == 0` is `FALSE` as 5 is odd and so `5 %% 2` returns `1`.
c. Generate an error by using an inappropriate condition in `ifelse()`. Something of a trick question.

# Extension material

None this week, sorry folks.
The prac is perhaps rather long already.