---
title: "The Distribution of MLEs for Linear Regression"
subtitle: "Second R Practical — Plotting densities and using for loops"
format:
  html:
    self-contained: true
    code-fold: false
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

# Main practical

## Part A: Plotting univariate densities

Suppose that $\hat{\mathbf{\beta}} \sim \mathcal N_p(\mathbf{\beta}, \sigma^2(\mathbf{X}^\top\mathbf{X})^{-1})$.

Let

$$
\mathbf{\beta} =
\begin{pmatrix}
\beta_0 \\
\beta_1 \\
\beta_2
\end{pmatrix} = 
\begin{pmatrix}
2 \\
-1 \\
2.5
\end{pmatrix},
$$

$$
\sigma^2 = 1.5,
$$

$$
\mathbf{X^\top}\mathbf{X} = \begin{pmatrix}
1 & 0 & 0 \\
0 & 1 & 0.6 \\
0 & 0.6 & 1
\end{pmatrix}.
$$

You can load $\mathbf{X}^\top\mathbf{X}$ as follows in `R`:

```{r}
xtx_mat <- matrix(
  c(1, 0, 0,
    0, 1, 0.6, 
    0, 0.6, 1),
  nrow = 3, byrow = TRUE)
```

a. Compute the covariance matrix of $\hat{\mathbf{\beta}}$. Name it `cov_mat`. Note that we can use the `solve` function in `R` to compute the inverse of a matrix, e.g. `example_mat <- matrix(1:4, nrow = 2); solve(example_mat)`.

```{r}
#| include: false
sigma2 <- 1.5
cov_mat <- sigma2 * solve(xtx_mat)
cov_mat
```

b. What is the variance of $\hat{\beta}_1$? What is the covariance between $\hat{\beta}_1$ and $\hat{\beta}_2$?
c. We wish to plot the univariate pdf for $\hat{\beta}_1$. To do this, we need a sequence of appropriate $x$ values, and the associated pdf values.

The sequence of $x$ values should be such that the pdf is not negligible.
Plotting three standard deviations on either side of the mean should be sufficient.
We note that the mean of $\hat{\beta}_1$ is $-1$ and its standard deviation is approximately 1.5.
Let's create a sequence of length 1e4 so that the pdf is smooth.

```{r}
# 1e4 means 10^4, i.e. 10 000. It's less error-prone to
# write 1e4 than 10000.
x_vec_b1 <- seq(-1 - 3 * 1.5, -1 + 3 * 1.5, length.out = 1e4) 
x_vec_b1[1:5]  # show first 5 values
x_vec_b1[9996:1e4]  # show last 5 values
```

Now, we need to evaluate the pdf of $\hat{\beta}_1$ at these $x$ values.

We first extract the precise variance of $\hat{\beta}_1$ from the covariance matrix.
Matrices have two dimensions, so we'll need to specify both the row and column indices:

```{r}
var_b1 <- cov_mat[2, 2]
var_b1
```

Now we can use the `dnorm` function to compute the pdf values for $\hat{\beta}_1$ at each element of `x_vec_b1`:

```{r}
# note that the dnorm function takes
# the standard deviation, rather than the variance
y_vec_b1 <- dnorm(x_vec_b1, mean = -1, sd = sqrt(var_b1))
y_vec_b1[1:5]  # show first 5 values
```

We can use standard `R` plotting functions to plot the pdf of $\hat{\beta}_1$.
Note that we set `type` to `"l"` to plot lines, and `lwd` to set the line width.

```{r}
plot(x_vec_b1, y_vec_b1, type = "l", lwd = 2)
```

d. Plot the univariate pdf for $\hat{\beta}_2$. Use `par(mfrow = <integer_vector>)` to set up a multi-panel figure with two plots, one for $\hat{\beta}_1$ and one for $\hat{\beta}_2$.
e. Reset the plotting area to a single panel using `par(mfrow = c(1, 1))`.

## Part B: For loops

Now we consider an important programming construct: the `for` loop.

Without further ado, here is a simple example of a `for` loop in `R`:

```{r}
for (i in 1:5) {
  print(i^2)
}
```

This loop will print the numbers 1 to 5 squared, one at a time.

It is equivalent to the following code, which does not use a loop:

```{r}
print(1^2)
print(2^2)
print(3^2)
print(4^2)
print(5^2)
```

Clearly, it isn't very difficult to simply write out five print statements, but `for` loops become very useful when we need to perform the same operation on a large number of elements, or when the number of elements is not known in advance.

The syntax for a `for` loop is as follows:

for (<variable> in <sequence>) {
  # code to execute for each element in the sequence
}

In the example above, `i` is the variable that takes on each value in the sequence `1:5`, which is equivalent to `c(1, 2, 3, 4, 5)`.
Each iteration of the loop executes the code inside the curly braces `{}` with `i` taking on the current value from the sequence.

`i` is in fact a regular `R` variable.
After running the loop, `i` will have the value of the last element in the sequence, which is 5 in this case.
You can check:

```{r}
i
```

Note that `i` is not a special variable; you can use any valid variable name in its place, e.g. `j`, `k`, or `index`.
Also note that the sequence can be any vector or even a list.

For example, we can loop over a character vector:

```{r}
vec <- c("a", "b", "c")
for (x in vec) {
  print(x)
}
```

Typically, I use `i` when looping over indices (e.g. looping over `1:length(vec)`), and `x` when looping over something else.

a. Write a `for` loop that computes the pdf of $\hat{\beta}_1$ at each element of `x_vec_b1`, and stores the results in a vector called `y_vec_b1_loop`. Use the `dnorm` function inside the loop.

Loops do not automatically create an object (such as a vector) to store results.
So, we need to initialize the vector `y_vec_b1_loop` before the loop starts.
Here, we create a numeric vector of the same length as `x_vec_b1` to store the results:

```{r}
y_vec_b1_loop <- rep(NA, length(x_vec_b1))
y_vec_b1_loop[1:5]  # show first 5 values
```

The above creates a vector of `NA` values, which we will fill in during the loop.
`NA` values are used to indicate missing or undefined values in `R`.
Pre-filling with NAs helps because it ensures that the vector has the correct length, and we can easily identify after running the loop which entries the loop failed to compute.

Now, we can write the loop:

```{r}
for (i in 1:length(x_vec_b1)) {
  y_vec_b1_loop[i] <- dnorm(x_vec_b1[i], mean = -1, sd = sqrt(var_b1))
}
```

We can check whether individual elements are equal to the corresponding elements in `y_vec_b1`, using the `==` operator:

```{r}
match_vec <- y_vec_b1 == y_vec_b1_loop
match_vec[1:5]  # show first 5 values
```

We can see that the first five elements are `TRUE`, indicating that the first five elements of `y_vec_b1` and `y_vec_b1_loop` are equal.

We can check whether all elements are equal using the `all` function:

```{r}
all(match_vec)  # should return TRUE
```

This can be run in one step, of course:

```{r}
all(y_vec_b1 == y_vec_b1_loop)
```

Of course, a loop was unnecessary in this example, as the `dnorm` function accepts a *vector* of $x$ values and returns a vector of pdf values.
In other words, `dnorm` could accept multiple $x$ values at once, and compute the pdf for all of them in one go.
However, we at least could check here that the loop works correctly.

b. Write a `for` loop that plots the PDFs for $\hat{\beta}_0$, $\hat{\beta}_1$, and $\hat{\beta}_2$ in a single figure with three panels. Hint: your loop will have three iterations, one for each $\hat{\beta}_i$. Loop from `i = 1` to `i = 3`.

# Extension material

This is extension material in the sense that it is not required to complete this to keep up with the `R` coding expectations.
Later practicals will not depend on this material, but it is useful to know.

### Part C: Bivariate pdfs

a. What is the correlation between $\hat{\beta}_1$ and $\hat{\beta}_2$? Use the covariance matrix you computed in part a to answer this question.
b. The plots are univariate and cannot show any covariance between $\hat{\beta}_1$ and $\hat{\beta}_2$. To visualize the covariance, we can plot the bivariate pdf for $(\hat{\beta}_1, \hat{\beta}_2)$.

As before, we need to create a sequence of $x$ values for both $\hat{\beta}_1$ and $\hat{\beta}_2$:

First, we extract the means and standard deviations from the covariance matrix:
```{r}
# Mean vector and 2x2 covariance block for (β̂1, β̂2)
mu_vec_12    <- c(beta1 = -1, beta2 = 2.5)
cov_mat_12 <- cov_mat[2:3, 2:3]
sd1 <- sqrt(cov_mat_12[1,1])
sd2 <- sqrt(cov_mat_12[2,2])
```

The sequence of $x$ values used for parts $c$ and $d$ can be reused here.
For completeness, we redefine them using the means and standard deviations we just computed:

```{r}
x_vec_b1 <- seq(
  mu_vec_12[1] - 3 * sd1, mu_vec_12[1] + 3 * sd1,
  length.out = 200
)
x_vec_b2 <- seq(
  mu_vec_12[2] - 3 * sd2, mu_vec_12[2] + 3 * sd2,
  length.out = 200
)
```

Now, we need to evaluate the bivariate pdf for $(\hat{\beta}_1, \hat{\beta}_2)$ at each *combination* of $x$ values.
We can do so by creating a grid of $x$ values for $\hat{\beta}_1$ and $\hat{\beta}_2$, with the `expand.grid` function.

For example, applying `expand.grid` to `1:3` and `4:6` would yield a data frame with all combinations of the two vectors:

```{r}
grid_example <- expand.grid(v1 = 1:3, v2 = 4:6)
grid_example
```

We now apply it to the two vectors `x_vec_b1` and `x_vec_b2`:

```{r}
grid_df_12 <- expand.grid(b1 = x_vec_b1, b2 = x_vec_b2)
```

We then evaluate the bivariate pdf for each combination of $x$ values, using the MVN density as provided by the mvtnorm package:

```{r}
if (!requireNamespace("mvtnorm", quietly = TRUE)) {
  install.packages("mvtnorm")
}
library(mvtnorm)
dens_vec_12 <- mvtnorm::dmvnorm(
  as.matrix(grid_df_12),
  mean = mu_vec_12, sigma = cov_mat_12
)
dens_vec_12[1:5]
```

Now, we can reshape the density vector into a matrix, which can be used for plotting:

```{r}
dens_mat_12 <- matrix(
  dens_vec_12,
  nrow = length(x_vec_b1),
  ncol = length(x_vec_b2),
  # FALSE is the default, but we specify it for clarity
  byrow = FALSE
)
```

Now, we can plot the bivariate pdf using the `contour` function, which creates contour lines (lines of equal density) for the density values:

```{r}
#| include: false
path_dir_p <- "_tmp/prac3"
if (dir.exists(path_dir_p)) {
  unlink(path_dir_p, recursive = TRUE)
}
dir.create(path_dir_p, recursive = TRUE)

png(file.path(path_dir_p, "bivariate_pdf.png"), width = 800, height = 600)
```

```{r}
contour(x_vec_b1, x_vec_b2, dens_mat_12)
points(mu_vec_12[1], mu_vec_12[2], pch = 19)  # mean

```

```{r}
#| include: false
dev.off()
```

c. Plot the bivariate PDF for $(\hat{\beta}_0, \hat{\beta}_1)$ using the covariance matrix you computed in part a.
d. Compare the bivariate pdf you plotted in part c with the one you plotted in part b. Do they look similar? Why or why not?

### Part D: Simulations

