---
title: STA2005S Regression Tutorial Four
subtitle: Questions
format:
  pdf:
    embed-resources: true
    include-in-header:
      text: |
        \usepackage{graphicx} % Allows including images
        \usepackage{booktabs} 
        \usepackage{amsmath}
        \usepackage{url}
        \usepackage{subfigure}
        \usepackage{xcolor}
        \usepackage{amsthm,amsfonts,amssymb,amscd,amsxtra}
        \usepackage{pgfpages}
        \usepackage{animate}
        \usepackage[backend=bibtex,style=verbose-trad2]{biblatex}
        \bibliography{bibliography} 
        \usepackage{mathpazo}
        \usepackage{unicode-math}
---

# Introduction

We are studying how **crop yield** (kg per hectare) depends on annual **rainfall** (mm), **fertiliser type** (three categories), and **region** (two categories).

We simulate some data:

```{r}
#| include: true
set.seed(7)
n <- 20
rainfall <- rnorm(n, mean = 500, sd = 80)
fertiliser <- sample(c("none", "organic", "chemical"), n, replace = TRUE)
region <- sample(c("north", "south"), n, replace = TRUE)
yield <- 20 + 0.08 * rainfall +
  5 * (fertiliser == "organic") + 10 * (fertiliser == "chemical") +
  -3 * (region == "south") +
  rnorm(n, 0, 8)

data_tbl <- tibble::tibble(
  yield = yield,
  rainfall = rainfall,
  fertiliser = fertiliser,
  region = region
)
```

Here are a few rows of the data:

```{r}
data_tbl[c(2, 5, 17), ]
```

Here are the summary statistics for the explanatory variables:

```{r}
data_tbl$rainfall |> summary() |> signif(2)
```

Here are the counts for the categorical variables:

```{r}
table(data_tbl$fertiliser)
table(data_tbl$region)
```

We fit the model:

```{r}
fit_tbl <- data_tbl |>
  dplyr::mutate(
    fertiliser = factor(fertiliser, levels = c("none", "organic", "chemical")),
    region = factor(region, levels = c("north", "south"))
  )
fit <- lm(yield ~ rainfall + fertiliser + region, data = fit_tbl)
summary(fit)
```

# Questions

## Interpreting the model

**a.** Write down the model’s mean structure (e.g., $E[Y] = \beta_0 + \beta_{\text{Rainfall}} X_{\text{Rainfall}} + \ldots$).

**b.** For fertiliser and region, what are the reference levels?

**c.** Write down the row of the design matrix for a farm with rainfall of 500 mm, fertiliser type “none,” and located in the “south” region.

**d.** Suppose there is an effect for region, but no effect for fertiliser. Sketch (roughly) the fitted values against rainfall for the two regions.

**e.** If there were an effect for fertiliser, how many separate lines would you expect to see in the plot from part (d) if you plotted a line for each combination of categorical variable levels? Briefly explain your answer.

**f.** Write down the null hypotheses for the effect of fertiliser tested in the `summary(lm(...))` output above.

**g.** Using the output, does the model suggest there is an effect of region on yield? Has it demonstrated conclusively that there is a large effect?

**h.** Provide the p-value for testing the joint null hypothesis that rainfall, fertiliser, and region have no effect on yield.

**i.** Comment on whether the null hypothesis that $\beta_{\text{rainfall}} = 0$ is likely of practical interest. Propose alternative statistics with practical relevance.

**j.** Compare the model estimates to the true values used to simulate the data. How close are they? How could this be improved?

**k.** Comment on why the estimated intercept is so far from the true value (10.1 vs 20).

**l.** Provide the full and reduced models for testing the following null hypotheses (can use words, or symbols or even R code - whatever is quickest and sufficiently clear):k

- *1.* Fertiliser has no effect on yield.
- *2.* Neither fertiliser nor region have an effect on yield.
- *3.* Organic fertiliser has no effect on yield, compared to no fertiliser.
- *4.* Organic fertiliser has no effect on yield, compared to chemical fertiliser.
- *5.* Fertiliser has no effect on yield, controlling only for rainfall.
- *6.* Fertiliser has no effect on yield, controlling neither for rainfall nor region.

## Model validation

**a.** Comment on how well the model captures the variability in the response.

**b.** What is the correlation between the fitted values and the response?

**c.** Suppose that the true model were

$$
\mathbb E(\text{yield}\mid X)=
\beta_0+\beta_1\,\text{rainfall}.
$$

Have we violated any model assumptions with our model mean structure?

**d.** Suppose that the true model were

$$
\mathbb E(\text{yield}\mid X)=
\beta_0+\beta_1\,\text{rainfall}^2.
$$

Have we violated any model assumptions with our model mean structure?

**e.** Image that we increased the range of observed rainfall to between 0 and 1000 mm. Comment on the plausibility of the linear relationship between yield and rainfall.

**f.** Comment on the residuals vs fitted values plot (@fig-resid-f), explaining whether any model assumptions appear to be violated. Why does it appear to fan out in the middle?

```{r}
#| include: false
path_p <- projr::projr_path_get("cache", "tut_reg_w5_ano_model_checking", "p-val-f.pdf")
pdf(path_p)
plot(resid(fit), fitted(fit))
dev.off()
plot_tbl <- tibble::tibble(
  fitted = fitted(fit),
  resid = resid(fit)
)
library(ggplot2)
p <- ggplot(data = plot_tbl, aes(x = fitted, y = resid)) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  labs(x = "Fitted values", y = "Residuals") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray20") +
  geom_point()
ggplot2::ggsave(filename = path_p, plot = p, width = 6, height = 4)
```

![Residuals versus fitted values](`r path_p`){#fig-resid-f}