---
title: "Principal component analysis"
subtitle: "STA5077Z coding practical"
author: "Miguel Rodó"
format:
  html:
    toc: true
    number-sections: true
    code-fold: false
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

This practical uses survey data collected to assess options for improving food security through the sustainable use of natural resources in the Sikasso region of Mali.

The aim is to practise the main steps of a principal component analysis:

1. exploring the original variables;
2. deciding whether the variables should be standardised;
3. fitting PCA with `prcomp()`;
4. examining the proportion of variation explained;
5. interpreting loadings and scores; and
6. producing and interpreting a biplot.

Use comments in your code and short written explanations below each result. Do not report numerical output without explaining what it means in the context of the data.

## Variables

The variables used in this practical are:

| Variable | Description |
|:--|:--|
| `family` | Total number of individuals in the household |
| `dist_rd` | Distance in kilometres to the nearest passable road |
| `cotton` | Hectares of cotton planted in 2000 |
| `maize` | Hectares of maize planted in 2000 |
| `sorg` | Hectares of sorghum planted in 2000 |
| `millet` | Hectares of millet planted in 2000 |
| `bull` | Total number of bullocks or draft animals |
| `cattle` | Total number of cattle |
| `goats` | Total number of goats |

## Setup

The package installation only needs to be run once.

```{r}
#| label: install-mali-data-package
#| eval: false

if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

remotes::install_github(
  "MiguelRodo/DataTidy23RodoHonsMult@2024"
)
```

Load the packages and data.

```{r}
#| label: load-pca-practical-data

library(tidyverse)

data(
  "data_tidy_mali",
  package = "DataTidy23RodoHonsMult"
)

pca_variables <- c(
  "family",
  "dist_rd",
  "cotton",
  "maize",
  "sorg",
  "millet",
  "bull",
  "cattle",
  "goats"
)

mali_pca_tbl <- data_tidy_mali |>
  dplyr::select(dplyr::all_of(pca_variables))
```

# Core practical

## Question 1: Inspect the data

Inspect the first few rows and calculate suitable numerical summaries for each variable.

Your answer should address:

- the number of observations and variables;
- whether any values are missing;
- the centre and spread of each variable; and
- whether the variables appear to have similar numerical scales.

```{r}
#| label: inspect-mali-pca-data

# Your code here
```

**Interpretation:**  
Write two or three sentences summarising what you found.

## Question 2: Explore pairwise relationships

Construct scatterplots of:

1. household size (`family`) against distance to the nearest passable road (`dist_rd`); and
2. distance to the nearest passable road (`dist_rd`) against the number of cattle (`cattle`).

```{r}
#| label: fig-family-distance
#| fig-cap: "Household size plotted against distance to the nearest passable road."

# Your code here
```

```{r}
#| label: fig-distance-cattle
#| fig-cap: "Number of cattle plotted against distance to the nearest passable road."

# Your code here
```

Identify any observations that appear unusually far from the main body of the data.

Do not remove an observation merely because it is large.
Remove an observation only when you can explain why it is likely to distort the analysis or reflect a data problem.
Record any exclusions clearly and create a dataset called `mali_pca_clean`.

```{r}
#| label: prepare-clean-mali-pca-data


```

**Interpretation:**  
Describe any potentially unusual observations and justify your decision to retain or exclude them.

## Question 3: Decide whether to standardise

PCA can be performed on:

- the centred variables in their original units; or
- standardised variables, for which each variable has variance one.

Use the variable descriptions and the numerical summaries from Question 1 to decide which approach is more appropriate here.

```{r}
#| label: summarise-variable-scales

```

**Interpretation:**  
Explain whether you will use `scale. = TRUE` or `scale. = FALSE` in `prcomp()`. Your explanation should refer to units, numerical scales and the substantive meaning of the variables.

## Question 4: Fit the PCA using `prcomp`

Fit a PCA using `prcomp()`. Centre the variables and apply the scaling decision made in Question 3.

Store the fitted object as `mali_pca_fit`.

```{r}
#| label: fit-mali-pca


```

The main parts of the fitted object are:

- `mali_pca_fit$rotation`: the loading vectors;
- `mali_pca_fit$x`: the principal component scores; and
- `mali_pca_fit$sdev^2`: the variance of each principal component.

Confirm that the score matrix has one row per observation and one column per principal component.

```{r}
#| label: inspect-mali-pca-object


```

## Question 5: Fit the PCA using `eigen`

Fit the PCA using the eigen decomposition of the covariance/correlation matrix (depending on the decision you made earlier regarding standardisation).
Confirm that the results are equivalent to those from `prcomp()`.

## Question 6: Examine the variation explained

Calculate the proportion and cumulative proportion of variation explained by each principal component.
Use both the output from `prcomp` and the eigen decomposition to confirm that the results are equivalent.

Store the results in a table called `mali_pve_tbl`.

```{r}
#| label: calculate-mali-pve


```

Construct a scree plot showing the proportion of variation explained by each component.

```{r}
#| label: fig-mali-pca-scree
#| fig-cap: "Proportion of variation explained by each principal component in the Mali survey data."


```

How many components would you retain for a low-dimensional summary of these data?


## Question 7: Interpret the first two loading vectors

Display the loadings for the first two principal components.

```{r}
#| label: inspect-mali-pca-loadings


```

Provide interpretations for the first two principal components.



## Question 8: Examine the principal component scores

Create a score table containing PC1 and PC2.

```{r}
#| label: prepare-mali-pca-scores


```

Plot the observations using their first two principal component scores.

```{r}
#| label: fig-mali-pca-scores
#| fig-cap: "Mali survey observations represented by their scores on the first two principal components."

```

Comment on the plot.

## Question 8: Construct and interpret a biplot

Construct a biplot using the first two principal components.

```{r}
#| label: fig-mali-pca-biplot
#| fig-cap: "Biplot of the first two principal components for the standardised Mali survey variables."

```

