---
title: "Regression Week VI: Resampling"
subtitle: "Slide deck I: Bootstrapping"
author: "Yovna Junglee, edited by Miguel Rodo"
institute: University of Cape Town
date: today
bibliography: zotero.bib
nocite: |
  @hesterberg14
format:
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: ../preamble.tex
    keep-tex: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(ggplot2)
library(tibble)
library(dplyr)
library(tidyr)
library(purrr)

set.seed(6061)

alpha <- 0.05
B <- 10000
plot_theme <- theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

theme_set(plot_theme)

boot_mean <- function(x, B = 10000) {
  replicate(B, mean(sample(x, size = length(x), replace = TRUE)))
}

boot_t_mean <- function(x, B = 10000) {
  xbar <- mean(x)
  replicate(B, {
    x_star <- sample(x, size = length(x), replace = TRUE)
    se_star <- sd(x_star) / sqrt(length(x_star))
    if (is.na(se_star) || se_star == 0) {
      return(NA_real_)
    }
    (mean(x_star) - xbar) / se_star
  }) |>
    stats::na.omit()
}

ci_t_mean <- function(x, alpha = 0.05) {
  mean(x) + c(-1, 1) * qt(1 - alpha / 2, df = length(x) - 1) *
    sd(x) / sqrt(length(x))
}

ci_percentile_mean <- function(x, B = 10000, alpha = 0.05) {
  quantile(boot_mean(x, B), probs = c(alpha / 2, 1 - alpha / 2), names = FALSE)
}

ci_bootse_mean <- function(x, B = 10000, alpha = 0.05) {
  boot_stats <- boot_mean(x, B)
  mean(x) + c(-1, 1) * qt(1 - alpha / 2, df = length(x) - 1) * sd(boot_stats)
}

ci_boott_mean <- function(x, B = 10000, alpha = 0.05) {
  t_star <- boot_t_mean(x, B)
  q <- quantile(t_star, probs = c(alpha / 2, 1 - alpha / 2), names = FALSE)
  se_hat <- sd(x) / sqrt(length(x))
  c(mean(x) - q[2] * se_hat, mean(x) - q[1] * se_hat)
}
```

# Why bootstrapping belongs in this course

- Earlier inference used exact finite-sample results under normal-error assumptions.

\pause

- Those results gave familiar intervals and tests for regression coefficients, fitted means and predictions.

\pause

- Bootstrapping gives a simulation-based way to approximate the sampling distribution of a statistic.

\pause

- It is especially useful when formulas are difficult, or when we want to see how much an answer depends on the sample.

# Lecture outline

By the end of this deck, you should be able to:

- Explain the plug-in idea behind the bootstrap
- Distinguish the sampling distribution from the bootstrap distribution
- Construct percentile, bootstrap-standard-error and bootstrap-$t$ intervals
- Explain why skewness and small samples can cause poor coverage
- Bootstrap regression quantities using pairs or residual resampling
- State what the bootstrap can and cannot fix

# What exact regression inference used

For the usual linear model,

$$
\mathbf Y=\mathbf X\boldsymbol\beta+\boldsymbol\varepsilon,
\qquad
\boldsymbol\varepsilon\sim N(\mathbf 0,\sigma^2\mathbf I_n),
$$

we obtained

$$
\frac{\hat\beta_j-\beta_j}{s\sqrt{c_{jj}}}
\sim t_{n-p},
$$

where $c_{jj}$ is the $j$th diagonal element of $(\mathbf X^T\mathbf X)^{-1}$.

\pause

This depends on strong structure:

- a correctly specified model
- independent errors with common variance
- normal errors for exact $t$ and $F$ distributions

# The first warning: skewness can survive averaging

```{r sampling-distributions-skewed}
#| fig-width: 8.4
#| fig-height: 4.0
#| out-width: "92%"
set.seed(6062)
sampling_tbl <- map_dfr(c(5, 20, 100), function(n) {
  means <- replicate(5000, mean(rexp(n, rate = 1)))
  tibble(
    n = paste0("n = ", n),
    z = sqrt(n) * (means - 1)
  )
})

normal_curve_tbl <- tibble(
  z = seq(-3.5, 4.5, length.out = 600),
  density = dnorm(z)
)

ggplot(sampling_tbl, aes(x = z)) +
  geom_density(fill = "#647C9C", alpha = 0.35, colour = "#003366", linewidth = 0.7) +
  geom_line(
    data = normal_curve_tbl,
    aes(x = z, y = density),
    inherit.aes = FALSE,
    linewidth = 0.9,
    linetype = 2,
    colour = "grey20"
  ) +
  facet_wrap(~ n, nrow = 1) +
  labs(
    x = expression(sqrt(n)(bar(X) - mu) / sigma),
    y = "Density",
    title = "Standardised sampling distributions for exponential data"
  )
```

The normal approximation improves as $n$ grows, but the improvement can be slow for skewed data.

# Exact normal-theory intervals

If $X_1,\ldots,X_n$ are a normal sample, then

$$
\frac{\bar X-\mu}{S/\sqrt n}
\sim t_{n-1}.
$$

So an exact $100(1-\alpha)\%$ confidence interval for $\mu$ is

$$
\bar x \pm t_{1-\alpha/2,n-1}\frac{s}{\sqrt n}.
$$

\pause

Two important facts are hidden inside this result:

- the sampling distribution is symmetric
- for a normal sample, $\bar X$ and $S$ are independent

# What changes outside the normal case?

For skewed data:

- the sampling distribution of $\bar X$ may still be skewed
- $\bar X$ and $S$ are generally correlated
- the statistic

$$
T
=
\frac{\bar X-\mu}{S/\sqrt n}
$$

need not have a $t$ distribution

\pause

Bootstrapping is one way to estimate the relevant distribution from the data rather than pretending it is exactly $t$.

# The ideal world

::::{.columns}

:::{.column width=50%}


a population distribution $F$

$$
F
$$

\vspace{0.3cm}

many repeated samples

$$
\mathbf X^{(1)},\mathbf X^{(2)},\ldots
$$

\vspace{0.3cm}

many statistics

$$
\hat\theta^{(1)},\hat\theta^{(2)},\ldots
$$

:::

:::{.column width=50%}


the sampling distribution

$$
\mathcal L_F(\hat\theta)
$$

\vspace{0.4cm}

This is what we want for inference.

\pause

But usually we only have one sample.

:::

::::

# The bootstrap world

::::{.columns}

:::{.column width=50%}

replace the unknown $F$ with the empirical distribution

$$
\widehat F_n
$$

\vspace{0.3cm}

many bootstrap samples

$$
\mathbf X^{*(1)},\mathbf X^{*(2)},\ldots
$$

\vspace{0.3cm}

many bootstrap statistics

$$
\hat\theta^{*(1)},\hat\theta^{*(2)},\ldots
$$

:::

:::{.column width=50%}


the bootstrap distribution

$$
\mathcal L_{\widehat F_n}(\hat\theta^*)
$$

\vspace{0.4cm}

This estimates the spread and shape of the sampling distribution.

\pause

It is centred near $\hat\theta$, not near the unknown $\theta$.

:::

::::

# The plug-in principle

The bootstrap is based on a familiar statistical idea:

$$
\text{unknown quantity}
\quad\leadsto\quad
\text{estimate of that quantity}.
$$

\pause

For the usual standard error of $\bar X$, we replace $\sigma$ with $s$:

$$
\operatorname{SE}(\bar X)=\frac{\sigma}{\sqrt n}
\quad\leadsto\quad
\widehat{\operatorname{SE}}(\bar X)=\frac{s}{\sqrt n}.
$$

\pause

The bootstrap takes this one step further:

$$
F
\quad\leadsto\quad
\widehat F_n.
$$

Then we repeatedly sample from $\widehat F_n$.

# A bootstrap sample by hand

Suppose the observed data are

$$
4,\,7,\,9,\,12.
$$

A bootstrap sample has the same size, but is drawn **with replacement**:

```{r hand-bootstrap-table}
#| echo: false
hand_boot_tbl <- tibble(
  sample = c("original data", "bootstrap sample 1", "bootstrap sample 2", "bootstrap sample 3"),
  values = c("4, 7, 9, 12", "4, 4, 9, 12", "7, 9, 9, 9", "4, 7, 7, 12"),
  mean = c(mean(c(4, 7, 9, 12)), mean(c(4, 4, 9, 12)), mean(c(7, 9, 9, 9)), mean(c(4, 7, 7, 12)))
)
knitr::kable(hand_boot_tbl, digits = 2)
```

\pause

Duplicates and omissions are not a bug. They are how the empirical distribution is resampled.

# The ordinary nonparametric bootstrap

For a statistic $\hat\theta = T(X_1,\ldots,X_n)$:

1. Draw $X_1^*,\ldots,X_n^*$ with replacement from the observed data.
2. Compute $\hat\theta^*=T(X_1^*,\ldots,X_n^*)$.
3. Repeat steps 1 and 2 many times.
4. Use the distribution of $\hat\theta^*$ to estimate uncertainty in $\hat\theta$.

\pause

In this course, take $B=10{,}000$ as a sensible routine default, and increase it when tail accuracy matters [@hesterberg14].

# Bootstrap distribution for the mean

```{r bootstrap-mean-distribution}
#| fig-width: 7.2
#| fig-height: 4.1
#| out-width: "82%"
set.seed(6063)
x_exp <- rexp(30, rate = 1)
xbar_exp <- mean(x_exp)
boot_xbar <- boot_mean(x_exp, B = B)

boot_mean_tbl <- tibble(boot_mean = boot_xbar)
ci_perc_exp <- quantile(boot_xbar, c(0.025, 0.975), names = FALSE)

boot_density <- ggplot(boot_mean_tbl, aes(x = boot_mean)) +
  geom_density(fill = "#647C9C", alpha = 0.35, colour = "#003366", linewidth = 0.8) +
  geom_vline(xintercept = xbar_exp, colour = "#7A1F1F", linewidth = 0.9) +
  geom_vline(xintercept = ci_perc_exp, linetype = 2, colour = "grey25") +
  labs(
    x = expression(bar(X)^"*"),
    y = "Density",
    title = "Bootstrap distribution of the sample mean"
  )
boot_density
```

The vertical solid line is the observed mean. The dashed lines mark the middle 95% of the bootstrap means.

# Method I: percentile interval

Let $q_{\alpha/2}^*$ and $q_{1-\alpha/2}^*$ be quantiles of the bootstrap statistics.

The percentile interval is

$$
\left(q_{\alpha/2}^*,\ q_{1-\alpha/2}^*\right).
$$

\pause

For the sample mean:

1. Generate $B$ bootstrap samples.
2. Calculate $\bar x_b^*$ for each bootstrap sample.
3. Use the $2.5\%$ and $97.5\%$ quantiles of the $\bar x_b^*$ values.

\pause

This is intuitive, but it can be too narrow in small samples.

# Method II: bootstrap standard error

The bootstrap standard error is the standard deviation of the bootstrap statistics:

$$
\widehat{\operatorname{SE}}_{\mathrm{boot}}(\hat\theta)
=
\sqrt{
\frac{1}{B-1}
\sum_{b=1}^{B}
\left(\hat\theta_b^*-\overline{\hat\theta^*}\right)^2
}.
$$

\pause

A simple interval is then

$$
\hat\theta
\pm
 t_{1-\alpha/2,\nu}\widehat{\operatorname{SE}}_{\mathrm{boot}}(\hat\theta).
$$

For a one-sample mean, $\nu=n-1$.

# Method III: bootstrap-$t$ interval

For the mean, define the bootstrap analogue of the $t$ statistic:

$$
T_b^*
=
\frac{\bar x_b^*-\bar x}{s_b^*/\sqrt n}.
$$

Let $q_{\alpha/2}^*$ and $q_{1-\alpha/2}^*$ be quantiles of the $T_b^*$ values.

Then the bootstrap-$t$ interval is

$$
\left(
\bar x - q_{1-\alpha/2}^*\frac{s}{\sqrt n},
\quad
\bar x - q_{\alpha/2}^*\frac{s}{\sqrt n}
\right).
$$

\pause

The upper bootstrap-$t$ quantile is used for the lower endpoint, and vice versa.

# Example: three intervals from the same data

```{r interval-comparison-plot}
#| fig-width: 8.0
#| fig-height: 3.6
#| out-width: "88%"
set.seed(6064)
x_example <- rexp(25, rate = 1)
mu_true <- 1
ci_tbl <- tibble(
  method = c("ordinary t", "percentile", "bootstrap SE", "bootstrap t"),
  lower = c(
    ci_t_mean(x_example)[1],
    ci_percentile_mean(x_example, B = B)[1],
    ci_bootse_mean(x_example, B = B)[1],
    ci_boott_mean(x_example, B = B)[1]
  ),
  upper = c(
    ci_t_mean(x_example)[2],
    ci_percentile_mean(x_example, B = B)[2],
    ci_bootse_mean(x_example, B = B)[2],
    ci_boott_mean(x_example, B = B)[2]
  )
) |>
  mutate(method = factor(method, levels = rev(method)))

ggplot(ci_tbl, aes(y = method)) +
  geom_vline(xintercept = mu_true, linetype = 3, linewidth = 0.8) +
  geom_segment(aes(x = lower, xend = upper, yend = method), linewidth = 1.3, colour = "#003366") +
  geom_point(aes(x = lower), size = 2.1, colour = "#003366") +
  geom_point(aes(x = upper), size = 2.1, colour = "#003366") +
  labs(
    x = expression(mu),
    y = NULL,
    title = "Intervals for the mean of one skewed sample"
  )
```

The dotted line is the true mean in this simulation. In real data, that line is not available.

# Why the intervals differ

The four methods use different approximations:

```{r interval-comparison-table}
#| echo: false
method_tbl <- tibble(
  Method = c("ordinary $t$", "percentile", "bootstrap SE", "bootstrap-$t$"),
  Uses = c(
    "$t_{n-1}$ distribution",
    "quantiles of $\\hat\\theta^*$",
    "standard deviation of $\\hat\\theta^*$",
    "quantiles of $T^*$"
  ),
  Main_strength = c(
    "exact for normal one-sample mean",
    "very intuitive",
    "simple bridge to formula intervals",
    "adapts to skewed $t$ statistics"
  ),
  Main_warning = c(
    "can be poor under skewness",
    "often too narrow for small $n$",
    "still symmetric around $\\hat\\theta$",
    "needs a standard error inside each bootstrap sample"
  )
)
knitr::kable(method_tbl, escape = FALSE)
```

# Small samples are still small samples

The bootstrap distribution reflects the original sample.

\pause

If the original sample is unrepresentative, the bootstrap distribution is also based on that unrepresentative sample.

\pause

Using more bootstrap samples reduces Monte Carlo noise, but it does not change the basic information in the original data.

\pause

This is why percentile bootstrap intervals can under-cover in small samples, even though the method feels data-driven [@hesterberg14].

# Regression target: bootstrap what you estimate

In regression, the statistic might be

$$
\hat\beta_j,
\qquad
\widehat{E}(Y\mid \mathbf x_0)=\mathbf x_0^T\hat{\boldsymbol\beta},
\qquad
\widehat{Y}_{\mathrm{new}}.
$$

\pause

The bootstrap recipe stays the same:

1. Generate a bootstrap version of the data.
2. Refit the regression model.
3. Save the statistic of interest.
4. Use the saved values to study uncertainty.

# Two common regression bootstraps

::::{.columns}

:::{.column width=50%}

\begin{block}{Pairs bootstrap}
Resample complete rows:

$$
(x_i,y_i)^*.
$$

Then refit the model.

\vspace{0.3cm}

Use this when the rows are regarded as a random sample from a population.
\end{block}

:::

:::{.column width=50%}

\begin{block}{Residual bootstrap}
Keep the same $x_i$ values and resample centred residuals:

$$
y_i^*=\hat y_i+e_i^*.
$$

Then refit the model.

\vspace{0.3cm}

Use this when conditioning on the observed design is appropriate.
\end{block}

:::

::::

# Residual bootstrap: a regression-line picture

```{r residual-bootstrap-lines}
#| fig-width: 7.2
#| fig-height: 4.2
#| out-width: "82%"
set.seed(6065)
n_reg <- 80
x_reg <- runif(n_reg, 0, 6)
y_reg <- 2 + 1.4 * x_reg + rnorm(n_reg, sd = 1.2)
reg_tbl <- tibble(x = x_reg, y = y_reg)
fit_reg <- lm(y ~ x, data = reg_tbl)
resid_centred <- residuals(fit_reg) - mean(residuals(fit_reg))
new_x <- seq(min(x_reg), max(x_reg), length.out = 80)

line_tbl <- map_dfr(1:200, function(b) {
  y_star <- fitted(fit_reg) + sample(resid_centred, n_reg, replace = TRUE)
  fit_star <- lm(y_star ~ x_reg)
  tibble(
    b = b,
    x = new_x,
    y = coef(fit_star)[1] + coef(fit_star)[2] * new_x
  )
})

orig_line_tbl <- tibble(
  x = new_x,
  y = predict(fit_reg, newdata = tibble(x = new_x))
)

ggplot(reg_tbl, aes(x = x, y = y)) +
  geom_line(data = line_tbl, aes(group = b), alpha = 0.08, colour = "grey25") +
  geom_point(alpha = 0.75, colour = "#003366") +
  geom_line(data = orig_line_tbl, linewidth = 1.2, colour = "#7A1F1F") +
  labs(
    x = "x",
    y = "y",
    title = "Bootstrap regression lines from residual resampling"
  )
```

The cloud of lines shows uncertainty in the fitted regression function.

# Regression coefficient bootstrap distribution

```{r slope-bootstrap-distribution}
#| fig-width: 7.2
#| fig-height: 4.0
#| out-width: "82%"
set.seed(6066)
slope_star <- replicate(B, {
  y_star <- fitted(fit_reg) + sample(resid_centred, n_reg, replace = TRUE)
  coef(lm(y_star ~ x_reg))[2]
})

slope_ci <- quantile(slope_star, c(0.025, 0.975), names = FALSE)
slope_hat <- coef(fit_reg)[2]

ggplot(tibble(slope = slope_star), aes(x = slope)) +
  geom_density(fill = "#647C9C", alpha = 0.35, colour = "#003366", linewidth = 0.8) +
  geom_vline(xintercept = slope_hat, colour = "#7A1F1F", linewidth = 0.9) +
  geom_vline(xintercept = slope_ci, linetype = 2, colour = "grey25") +
  labs(
    x = expression(hat(beta)[1]^"*"),
    y = "Density",
    title = "Bootstrap distribution of the slope"
  )
```

The same idea applies to an intercept, a fitted mean, or a prediction target.

# When should pairs and residual bootstraps differ?

They may differ when:

- the $x$ values are highly uneven
- there are rare factor levels or sparse interaction combinations
- residual variance changes with $x$
- the regression model fits poorly

\pause

A residual bootstrap conditions on the observed design matrix.

\pause

A pairs bootstrap also resamples the design, so some bootstrap datasets may contain awkward or unstable combinations of predictors.

# What the bootstrap cannot fix

Bootstrapping is not a remedy for every problem.

It does not fix:

- biased sampling or missing parts of the population
- dependence that is ignored during resampling
- a badly misspecified regression model
- severe outliers without modelling or robust methods
- very small samples that poorly represent the population

\pause

The resampling scheme must match the way the data were generated.

# Practical recommendations

For this course:

- Use at least $B=10{,}000$ bootstrap samples for final answers.
- Start with the percentile interval for intuition.
- Use a bootstrap standard error when comparing with formula-based intervals.
- Use bootstrap-$t$ when skewness matters and a standard error can be computed for each bootstrap sample.
- For regression, be explicit about whether you resample rows or residuals.

# Take-home messages

- The bootstrap replaces the unknown population distribution $F$ with the empirical distribution $\widehat F_n$.

- The bootstrap distribution is used for uncertainty, not to improve the original point estimate.

- Percentile intervals are intuitive, but can be too narrow in small samples.

- Bootstrap-$t$ intervals often behave better under skewness, but require more work.

- In regression, the resampling scheme is part of the statistical model.

# References
