---
title: "Regression Week I: Foundations for linear models"
subtitle: "Slide deck I: Course introduction and motivation"
author: Miguel Rodo
institute: University of Cape Town
date: today
format:
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: ../preamble.tex
    keep-tex: true
---

```{r}
#| include: false
library(ggplot2)
```

# What is the purpose of the regression section?

<!-- \pause !-->

- To make you a useful statistician!

<!-- \pause !-->

- Well, to really get you on your way
  - First year (STA1006S) is useful, but easily replicable by non-statisticians
  - STA2004F is (despite all the effort you put in) not immediately of great practical use
  - The practical skills and mathematical understanding you'll gain here are not easily replicable and can actually be used to assist decision-making

<!-- \pause !-->

## Key outcomes

- Mathematical understanding of simplest regression model
- Ability to code up an end-to-end modelling analysis
  - Reading in data
  - Analysing data
  - Displaying and interpreting results

# Lecture schedule

```{r }
#| echo: false
#| results: asis
#| label: tab-schedule
schedule_tbl <- tibble::tibble(
  Weeks = c("1-6", "Vac", "7-12"),
  `Start date` = c("27 Jul", "7 Sep", "14 Sep"),
  `End date` = c("4 Sep", "11 Sep", "23 Oct"),
  Section = c("Regression",  "", "Experimental design")
)
```

```{r}
#| echo: false
#| results: asis
#| label: tab-schedule-latex
#| eval: true
# Create and display the table
knitr::kable(schedule_tbl, format = "latex", booktabs = TRUE) |>
  kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
```

# Assessment contributions

```{r}
#| include: false

# Set the displayed row order here, from top to bottom
final_mark_order <- c("Class mark", "Exam mark")
class_mark_order <- c("Test", "Assignment", "Prac test")

assessment_colours <- c(
  "Exam mark"  = "#66c2a5",
  "Class mark" = "#fc8d62",
  "Regression" = "#8da0cb",
  "ED"         = "#e78ac3",
  "Prac test"  = "#a6d854"
)

final_mark_tbl <- tibble::tibble(
  assessment = final_mark_order,
  percentage = c(30, 70)
)

class_mark_tbl <- tibble::tribble(
  ~assessment,  ~section,     ~percentage, ~date,
  "Test",       "ED",         25,          "12 October",
  "Test",       "Regression", 25,          "1 September",
  "Assignment", "ED",         20,          "19 October",
  "Assignment", "Regression", 20,          "9 September",
  "Prac test",  "Prac test",  10,          "TBC"
) |>
  dplyr::mutate(
    assessment = factor(
      assessment,
      levels = rev(class_mark_order)
    )
  )

class_mark_totals <- class_mark_tbl |>
  dplyr::group_by(assessment) |>
  dplyr::summarise(
    total = sum(percentage),
    .groups = "drop"
  )

p_final_mark <- final_mark_tbl |>
  dplyr::mutate(
    assessment = factor(
      assessment,
      levels = rev(final_mark_order)
    )
  ) |>
  ggplot(
    aes(
      x = percentage,
      y = assessment,
      fill = assessment
    )
  ) +
  geom_col(width = 0.8) +
  geom_text(
    aes(
      x = percentage / 2,
      label = paste0(percentage, "%")
    ),
    size = 5
  ) +
  scale_fill_manual(values = assessment_colours) +
  scale_x_continuous(
    labels = scales::percent_format(scale = 1),
    limits = c(0, 75),
    breaks = seq(0, 60, by = 20),
    expand = c(0, 0)
  ) +
  cowplot::theme_cowplot() +
  labs(
    title = "Final mark",
    x = NULL,
    y = NULL
  ) +
  theme(
    legend.position = "none",
    plot.background = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "white", colour = NA),
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 10)
  )

p_class_mark <- ggplot(
  class_mark_tbl,
  aes(
    x = percentage,
    y = assessment,
    fill = section
  )
) +
  geom_col(
    width = 0.8,
    position = position_stack(reverse = FALSE)
  ) +
  geom_text(
    aes(label = date),
    position = position_stack(vjust = 0.5, reverse = FALSE),
    size = 4.2
  ) +
  geom_text(
    data = class_mark_totals,
    aes(
      x = total + 2.5,
      y = assessment,
      label = paste0(total, "%")
    ),
    inherit.aes = FALSE,
    hjust = 0,
    size = 4.5
  ) +
  scale_fill_manual(values = assessment_colours) +
  scale_x_continuous(
    labels = scales::percent_format(scale = 1),
    limits = c(0, 65),
    breaks = seq(0, 60, by = 20),
    expand = c(0, 0)
  ) +
  cowplot::theme_cowplot() +
  labs(
    title = "Class mark",
    x = NULL,
    y = NULL
  ) +
  theme(
    legend.position = "none",
    plot.background = element_rect(fill = "white", colour = NA),
    panel.background = element_rect(fill = "white", colour = NA),
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(size = 10)
  )

p <- cowplot::plot_grid(
  p_final_mark,
  p_class_mark,
  nrow = 1,
  rel_widths = c(0.8, 1.2),
  align = "h",
  axis = "tb"
)

path_p <- projr::projr_path_get(
  "cache",
  "fig",
  "w1s1",
  "p-assessment.png"
)

cowplot::ggsave2(
  path_p,
  p,
  width = 19,
  height = 8,
  units = "cm"
)
```

```{r}
#| results: asis
#| label: p-assessment
knitr::include_graphics(path_p)
```

# Relationship between assignment and test marks

```{r}
#| include: false
#| echo: false
path_p <- projr::projr_path_get("raw-data","img", "corr_reg.png")
```

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{`r path_p`}
\end{figure}

# Learning resources: content

- Regression notes:
  - Location: `Module 1: Regression/Notes/NotesReg.pdf`
  - Printed copies available at stats reception
- Slides:
  - Location: in `Slides` folder under each week's tab
- Lecture recordings:
  - Automatically uploaded to `Lecture Videos/2026 Lecture Videos` tab

# Learning resources: testing I

::::{.columns}

:::{.column width=50%}

- Q&A docs:
  - Comprehensive set of short questions on each slide deck's content
  - A review of all the facts, essentially
  - Will be in the `Slides` folder for each week's content
    - For example, `QARegNotesW1S0Prereq.pdf` is in `Introduction/Slides` folder alongside `NotesRegW1S0Prereq.html`
  - *Note*: these are not exam-style questions, but rather a way to a) remember the content, and b) check your understanding

:::

\pause

:::{.column width=50%}

- Tutorials
  - On Amathuba (`Module 1: Regression/Practice material/Tutorials`)

\pause

- Question bank:
  - Appropriate questions over and above the tutorials
  - Aim to progress from simpler to harder questions within each section
  - On Amathuba (`Module 1: Regression/Practice material/Question bank`)
  - Will be updated throughout the term

:::

::::

# Advised learning/revision route for exam and test

- Attend lectures
- After each lecture, do relevant Q&A content
- Do tutorials (and attend them)
- Do question bank questions

# Advice from past tutors

- Learning to code up analyses is perhaps the most useful part of the course.
- Don't underestimate the course - passing STA2004F is great, but this is a lot of content. Marks are often terrible. 
- Be wary of the assignments - they count as much as the tests, and take a long time. 
- The exam counts 70%, so don't presume that assignments and tests will carry you through. Focus on understanding the content.
- Questions to rehearse proofs are free marks. 
- Experimental design in term four tends to be more difficult, so bank regression marks.  

# Introduction to `R`

- Online lecture:
  - Day: Wednesday, 29 July
  - Time: 2pm
  - Link: Pending

\pause

- Practicals:
  - Days: Wednesday, Thursday and Friday (29-31 July)
  - Times: 2-4pm
  - Category: `R Intro Workshop`

# Practicals

- In a computer lab, with a tutor, who will walk through coding up part of a regression analysis
- For example:
  - Fitting a regression model
  - Validating a regression model

# Sign up for tutorials and practicals

- Frequency: weekly
- Start: week 2
- Signup:
  - `Amathuba` >> `Course info` >> `Groups`
  - Categories:
    - Practical: `Computer Lab Session`
    - Tutorial: `STA2005S Tutorials`

# Office hours

- Room: 5.52 PD Hahn (fifth floor, down the corridor from stats reception)
- Time: TBC once class rep is appointed
- Day: TBC
- Purpose: any content-related questions
- May also email for other times, if need be

\pause

- Email:
  - Address: `miguel.rodo@uct.ac.za`
  - Purpose: any admin-related issues specific to the regression section

# What is regression? {.smaller}

::::{.columns}

:::{.column width=50%}

- $Y\sim f(X,\epsilon)$:
  - $Y$ is the response variable (dependent variable)
  - $X$ contains the explanatory variable(s) (or independent variables) 
  - $\epsilon$ is the random variable
  - The function `f` describes how they relate
  - $Y$ cannot be categorical

- A regression model describes how the distribution of a response $Y$
  depends on one or more explanatory variables $X  .
- For the linear regression models considered in this section, $Y$ is
  continuous.
- Random variation is represented through an error term $\epsilon$.
:::

:::{.column width=50%}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../_raw_data/img/reg_img.png}
\end{figure}
:::

::::

# Regression versus the correlation coefficient: round one

- Imagine we work out what the correlation is between advertising spending and revenue.
- In what way might the correlation coefficient be deficient?

# Regression versus the correlation coefficient round two

- Now, suppose a doctor wants to decide whether a six-month old baby has a healthy weight.
- Why would the correlation coefficient not be useful?

# Regression versus the $t$-test: round one

- Let's say you want to understand the relationship between blood pressure (the dependent variable) and taking a drug (the independent variable).
- Also assert that more males took the drug than females.
- Suppose we perform a $t$-test. How might someone object?

# Regression versus the $t$-test: round two

- Suppose that we want to model blood pressure against BMI.
- What's the problem with the $t$-test in this case?

# Toolbox

\centering

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{../_raw_data/img/toolbox-wide-edit-removebg.png}
\end{figure}

<!--
- Points: simply adding to your statistical toolbox
- Have to deal with real life, so need multiple approaches with their own pros and cons
!-->

# Assumptions

::::{.columns}
:::{.column width=70%}
```{r}
# ==========================
# distribution
# ==========================
# density
set.seed(3)
# generate data from a normal distribution from a range of x values
p_normal <- ggplot(
  tibble::tibble(x = rnorm(1000, 200, 5)),
  aes(x = x)
) +
  geom_density(fill = "#a6cee3", adjust = 3, col = "gray65") +
  cowplot::theme_cowplot() +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  ) +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )
p_normal_non <- ggplot(
  tibble::tibble(x = rpois(1000, 200)^6),
  aes(x = x)
) +
  geom_density(fill = "#b2df8a", adjust = 3, col = "gray65") +
  cowplot::theme_cowplot() +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  ) +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )

# ==========================
# relationship
# ==========================

set.seed(5)
n_elem <- 1e2
plot_tbl <- tibble::tibble(
  x = c(1, runif(n_elem, 1, 10), 10)
) |>
  dplyr::mutate(
    y = 2 + 3 * x + rnorm(n_elem + 2, 0, 2),
    y2 = 2 + 3 * x^3 + rnorm(n_elem + 2, 0, 300),
    fit = 2 + 3 * x,
    fit2 = 2 + 3 * x^3
  )

p_rel_linear <- ggplot(plot_tbl, aes(x = x, y = y)) +
  geom_point(col = "#a6cee3") +
  geom_line(aes(y = fit), color = "gray15", size = 1) +
  cowplot::theme_cowplot() +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  ) +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )
p_rel_quad <- ggplot(plot_tbl, aes(x = x, y = y2)) +
  geom_point(col = "#b2df8a") +
  geom_line(aes(y = fit2), color = "gray15", size = 1) +
  cowplot::theme_cowplot() +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  ) +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )

# ==========================
# variance
# ==========================

set.seed(6)
n_elem <- 3e2
plot_tbl <- tibble::tibble(
  x = c(1, runif(n_elem, 1, 10), 10)
) |>
  dplyr::mutate(
    y = 1 * x + rnorm(n_elem + 2, 0, 2),
    y2 = (1 * x + rpois(n_elem + 2, x)^2)^0.9,
    y2 = y2 + rnorm(n_elem + 2, 0, 2)
  )

p_var_constant <- ggplot(plot_tbl, aes(x = x, y = y)) +
  geom_point(col = "#a6cee3") +
  geom_smooth(
    method = "lm", se = FALSE, col = "gray15", size = 1,
    formula = y ~ x
  ) +
  cowplot::theme_cowplot() +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  ) +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )
p_var_increasing <- ggplot(plot_tbl, aes(x = x, y = y2)) +
  geom_point(col = "#b2df8a") +
 geom_smooth(
    method = "lm", se = FALSE, col = "gray15", size = 1,
    formula = y ~ x
  ) +
  cowplot::theme_cowplot() +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  ) +
  theme(
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  )


p_grid <- cowplot::plot_grid(
  p_normal, p_rel_linear, p_var_constant,
  p_normal_non, p_rel_quad, p_var_increasing,
  ncol = 3,
  labels = c("Distribution", "Relationship", "Variance"),
  label_size = 18
)
path_p_grid <- projr::projr_path_get(
  "cache", "fig", "w1s1", "p-assumptions.png"
)
cowplot::ggsave2(
  path_p_grid, p_grid,
  width = 19, height = 14, units = "cm"
)
```

```{r}
#| results: asis
knitr::include_graphics(path_p_grid)
```
:::

:::{.column width=30%}

\vskip 1cm

- In our course, we consider **linear regression** only:
  - Normally distributed response
  - Linear relationship with response variables
  - Constant variance

:::

::::

# Regression becomes highly flexible

- Non-constant variance: mean-variance relationship, dependent on predictor variables, etc.
- Correlated errors:
  - within subsets of observations, e.g. measurements from the same individual, or
  - between "nearby" observations, e.g. time-series or spatial data
- Non-normally distributed data, e.g. count data
- Categorical dependent variables, e.g. disease status

# Mathematical foundations to linear regression

::::{.columns}

:::{.column width=50%}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../_raw_data/img/iceberg.png}
\end{figure}

:::

:::{.column width=50%}

\begin{figure}
\centering
\includegraphics[width=0.95\textwidth]{../_raw_data/img/lm_output.png}
\end{figure}

:::

::::

# Regression section outline

- **Week one**: statistical and mathematical prerequisites (statistical distributions and linear algebra), and useful proofs regarding the multivariate normal distribution
- **Week two**: General linear model introduction (formulation, estimation and parameter distributions)
- **Week three**: General linear model inference (confidence intervals and hypothesis testing)
- **Week four**: ANOVA and model checking
- **Week five**: Variable transformations, variable selection and Gauss-Markov theorem
- **Week six**: PCA and bootstrapping

# Prerequisites: Linear algebra and statistical theory I

::::{.columns}

:::{.column width=50%}

**Content**, under `Module 1: Regression/Introduction/Prerequisites`:

- Linear algebra:
  - Slides: `SlidesRegW1S0PrereqLA.pdf`
  - Video: `VideoRegW1S0PrereqLA.mp4`
  - Basics in `R`: `DemRegW1S0LinearAlgebra.pdf`
- Linear algebra + statistical distributions:
  - Linear algebra content from slides above, plus information on statistical distributions: `NotesRegW1S0Prereq.html`
  - Q&A on notes: `QARegNotesW1S0PrereqLA.pdf`
:::

:::{.column width=50%}

**Practice material**, under `Module 1: Regression/Practice material`:

  - `Tutorial/TutReg1IntroMVNQuestions.pdf`
    - The first few questions in this tutorial are about linear algebra

:::  

::::