---
title: "Regression Week I: Foundations for linear models"
subtitle: "Slide deck III: The normal distribution and related theorems"
author: "Yovna Junglee, edited by Miguel Rodo"
institute: University of Cape Town
date: today
format:
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: ../preamble.tex
    keep-tex: true
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

# Why the normal distribution matters

- In the linear regression model, we usually assume that the **errors** are normally distributed.

\pause

- This assumption is not needed for ordinary least squares point estimation.

\pause

- It is central for the exact finite-sample distributions used in inference:
  - normal distributions for linear combinations of normal random variables
  - chi-square distributions for residual sums of squares
  - $t$ and $F$ distributions built from those results

\pause

- Under normal errors, the least-squares estimates are also maximum likelihood estimates.

# Lecture outline

By the end of this deck, you should be able to:

- State the PDFs of the univariate and multivariate normal distributions
- State the MGF of the multivariate normal distribution
- Use the MGF to prove that a linear transformation of an MVN random vector is MVN
- Use selection matrices to obtain distributions of components and subvectors
- Prove the relationship between a standard MVN quadratic form and a $\chi^2$ random variable

# Exclusions compared to previous years' courses {.smaller}

The following appear in previous years' slides, tests and some notes, but will not be examined here:

- Deriving the MVN density
- Deriving the MVN MGF
- Determining the mean and covariance matrix of an MVN distribution from its exponent

\pause

The reasons are:

- Later results can be established without those derivations
- Learning the derivations well does not substantially help with other course content
- There is overlap with STA2004F
- They add a heavier calculus component than we need here

\pause

You still need to be able to **use** the MGF and the MVN results.

# The univariate normal distribution

::::{.columns}

:::{.column width=58%}

\begin{block}{Definition}
If $X \sim N(\mu, \sigma^2)$, then its probability density function is
$$
f(x)
=
\frac{1}{\sqrt{2\pi\sigma^2}}
\exp\left\{-\frac{(x-\mu)^2}{2\sigma^2}\right\},
\qquad x \in \mathbb R.
$$
Moreover,
$$
E(X)=\mu,
\qquad
\operatorname{Var}(X)=\sigma^2.
$$
\end{block}

:::

:::{.column width=42%}

```{r}
#| fig-width: 4.6
#| fig-height: 3.6
#| out-width: "100%"
x <- seq(-4, 4, length.out = 600)
y <- dnorm(x)
old_par <- par(
  mar = c(4.0, 4.2, 2.2, 0.6),
  mgp = c(2.4, 0.7, 0),
  tcl = -0.25
)
plot(
  x, y,
  type = "n",
  xlim = c(-4, 4),
  ylim = c(0, 0.42),
  xaxs = "i",
  yaxs = "i",
  xlab = expression(x),
  ylab = "Density",
  main = expression(X %~% N(0, 1)),
  las = 1
)
polygon(
  c(x, rev(x)),
  c(y, rep(0, length(y))),
  col = grDevices::adjustcolor("#647C9C", alpha.f = 0.22),
  border = NA
)
lines(x, y, lwd = 2.2, col = "#003366")
abline(v = 0, lty = 3, col = "grey45")
box()
par(old_par)
```

:::

::::

# The multivariate normal distribution {.smaller}

\begin{block}{Definition}
Let $\mathbf X=(X_1,\ldots,X_p)^T$. If
$\mathbf X \sim N_p(\boldsymbol\mu,\boldsymbol\Sigma)$ and
$\boldsymbol\Sigma$ is positive definite, then
$$
f(\mathbf x)
=
\frac{1}{(2\pi)^{p/2}|\boldsymbol\Sigma|^{1/2}}
\exp\left\{-\frac{1}{2}(\mathbf x-\boldsymbol\mu)^T
\boldsymbol\Sigma^{-1}(\mathbf x-\boldsymbol\mu)\right\},
\qquad \mathbf x \in \mathbb R^p.
$$
Moreover,
$$
E(\mathbf X)=\boldsymbol\mu,
\qquad
\operatorname{Cov}(\mathbf X)
=
E\{(\mathbf X-\boldsymbol\mu)(\mathbf X-\boldsymbol\mu)^T\}
=\boldsymbol\Sigma.
$$
\end{block}

- $\boldsymbol\Sigma$ is a covariance **matrix**, not a scalar.
- Its $(i,j)$ entry is $\operatorname{Cov}(X_i,X_j)$.

# Visualising a bivariate normal distribution

Let
$$
\mathbf X \sim N_2\left(
\begin{pmatrix}0\\0\end{pmatrix},
\begin{pmatrix}1&0\\0&1\end{pmatrix}
\right).
$$

```{r}
#| fig-width: 8.4
#| fig-height: 3.4
#| out-width: "96%"
x <- seq(-4, 4, length.out = 80)
y <- seq(-4, 4, length.out = 80)
z <- outer(x, y, function(a, b) dnorm(a) * dnorm(b))
old_par <- par(
  mfrow = c(1, 2),
  mar = c(3.2, 3.3, 2.3, 1.0),
  mgp = c(2.0, 0.6, 0),
  tcl = -0.25
)
persp(
  x, y, z,
  theta = 35,
  phi = 25,
  expand = 0.62,
  ticktype = "detailed",
  col = "#D9E2EC",
  border = "#647C9C",
  xlab = expression(X[1]),
  ylab = expression(X[2]),
  zlab = "Density",
  main = "Density surface",
  cex.axis = 0.75,
  cex.lab = 0.9,
  cex.main = 0.95
)
contour(
  x, y, z,
  xlim = c(-4, 4),
  ylim = c(-4, 4),
  levels = seq(0.02, 0.14, by = 0.02),
  col = "#647C9C",
  lwd = 1.4,
  xlab = expression(X[1]),
  ylab = expression(X[2]),
  main = "Contour plot",
  asp = 1,
  las = 1,
  cex.axis = 0.8,
  cex.lab = 0.9,
  cex.main = 0.95
)
par(old_par)
```

# Moment-generating functions {.smaller}

\begin{block}{Definition}
For a random vector $\mathbf X$, the moment-generating function is
$$
M_{\mathbf X}(\mathbf t)
=
E\left\{\exp(\mathbf t^T\mathbf X)\right\},
$$
where $\mathbf t$ is a deterministic vector conformable with $\mathbf X$.
\end{block}

- The MGF is said to exist if the expectation is finite in an open neighbourhood of $\mathbf 0$.
- For a univariate random variable, derivatives of the MGF at $0$ give moments.
- For a random vector, partial derivatives give marginal and mixed moments.

\pause

In this course, the main use of MGFs is not to calculate moments. It is to identify distributions, because an existing MGF uniquely determines the distribution.

# The MGF of the multivariate normal distribution

\begin{block}{Result}
If $\mathbf X \sim N_p(\boldsymbol\mu,\boldsymbol\Sigma)$, then
$$
M_{\mathbf X}(\mathbf t)
=
\exp\left\{
\mathbf t^T\boldsymbol\mu
+
\frac{1}{2}\mathbf t^T\boldsymbol\Sigma\mathbf t
\right\}.
$$
\end{block}

\pause

- We will use this result without deriving it.
- The important point is that the exponent contains:
  - a linear term, $\mathbf t^T\boldsymbol\mu$
  - a quadratic term, $\frac{1}{2}\mathbf t^T\boldsymbol\Sigma\mathbf t$

# Linear transformation of an MVN random variable: theorem {.smaller}

\begin{block}{Theorem}
Let $\mathbf X \sim N_p(\boldsymbol\mu,\boldsymbol\Sigma)$.
If $\mathbf C$ is a fixed $q \times p$ matrix, then
$$
\mathbf Y
=
\mathbf C\mathbf X
\sim
N_q\left(
\mathbf C\boldsymbol\mu,
\mathbf C\boldsymbol\Sigma\mathbf C^T
\right).
$$
\end{block}

\pause

- This result is essential because regression estimators are linear transformations of a normally distributed response vector.
- If $\mathbf C\boldsymbol\Sigma\mathbf C^T$ is singular, the resulting normal distribution is degenerate. The formula for the mean and covariance matrix still holds.

# Linear transformation of an MVN random variable: proof {.smaller}

Let $\mathbf Y=\mathbf C\mathbf X$. Then
$$
\begin{aligned}
M_{\mathbf Y}(\mathbf t)
&= E\left\{\exp(\mathbf t^T\mathbf Y)\right\} \\
&= E\left\{\exp(\mathbf t^T\mathbf C\mathbf X)\right\} \\
&= E\left\{\exp\left((\mathbf C^T\mathbf t)^T\mathbf X\right)\right\} \\
&= M_{\mathbf X}(\mathbf C^T\mathbf t).
\end{aligned}
$$

\pause

Using the MGF of $\mathbf X$,
$$
\begin{aligned}
M_{\mathbf Y}(\mathbf t)
&= \exp\left\{(\mathbf C^T\mathbf t)^T\boldsymbol\mu
+ \frac{1}{2}(\mathbf C^T\mathbf t)^T\boldsymbol\Sigma(\mathbf C^T\mathbf t)\right\} \\
&= \exp\left\{\mathbf t^T\mathbf C\boldsymbol\mu
+ \frac{1}{2}\mathbf t^T\mathbf C\boldsymbol\Sigma\mathbf C^T\mathbf t\right\}.
\end{aligned}
$$

# Linear transformation proof: conclusion

The final expression is the MGF of a multivariate normal random vector with

\begin{block}{Identified mean and covariance matrix}
$$
E(\mathbf Y)=\mathbf C\boldsymbol\mu,
\qquad
\operatorname{Cov}(\mathbf Y)
=
\mathbf C\boldsymbol\Sigma\mathbf C^T.
$$
\end{block}

\pause

Therefore
$$
\boxed{
\mathbf Y=\mathbf C\mathbf X
\sim
N_q(\mathbf C\boldsymbol\mu,
\mathbf C\boldsymbol\Sigma\mathbf C^T)
}.
$$

# Why selection matrices are useful

- A selection matrix is a matrix that extracts components of a vector.
- This lets us use the linear-transformation theorem to obtain distributions of:
  - a single component, such as $X_1$
  - a subvector, such as $(X_1,X_2)^T$
  - all components except one, such as $(X_2,\ldots,X_p)^T$
  - a sum, such as $X_1+\cdots+X_p$

\pause

# Example: identity matrix

Suppose
$$
\mathbf X \sim N_p(\boldsymbol\mu,\boldsymbol\Sigma),
\qquad
\mathbf Y=\mathbf C\mathbf X.
$$

If $\mathbf C=\mathbf I_p$, then
$$
\mathbf Y=\mathbf I_p\mathbf X=\mathbf X.
$$

Therefore
$$
E(\mathbf Y)=\mathbf I_p\boldsymbol\mu=\boldsymbol\mu,
\qquad
\operatorname{Cov}(\mathbf Y)
=
\mathbf I_p\boldsymbol\Sigma\mathbf I_p^T
=\boldsymbol\Sigma.
$$

# Example: one individual element {.smaller}

Suppose
$$
\mathbf X=(X_1,\ldots,X_p)^T
\sim
N_p(\boldsymbol\mu,\boldsymbol\Sigma).
$$

To extract $X_i$, use the row vector
$$
\mathbf e_i^T
=
\begin{pmatrix}0&\cdots&0&1&0&\cdots&0\end{pmatrix},
$$
where the $1$ is in position $i$.

\pause

Then
$$
Y=\mathbf e_i^T\mathbf X=X_i,
$$
so
$$
E(Y)=\mathbf e_i^T\boldsymbol\mu=\mu_i,
\qquad
\operatorname{Var}(Y)
=\mathbf e_i^T\boldsymbol\Sigma\mathbf e_i
=\sigma_{ii}.
$$

Therefore
$$
X_i \sim N(\mu_i,\sigma_{ii}).
$$

# Example: the first two elements {.smaller}

Let
$$
\mathbf C
=
\begin{pmatrix}
1&0&0&\cdots&0\\
0&1&0&\cdots&0
\end{pmatrix}.
$$

Then
$$
\mathbf Y=\mathbf C\mathbf X
=
\begin{pmatrix}X_1\\X_2\end{pmatrix}.
$$

\pause

By the linear-transformation theorem,
$$
\begin{pmatrix}X_1\\X_2\end{pmatrix}
\sim
N_2\left(
\begin{pmatrix}\mu_1\\\mu_2\end{pmatrix},
\begin{pmatrix}
\sigma_{11} & \sigma_{12}\\
\sigma_{21} & \sigma_{22}
\end{pmatrix}
\right).
$$

# Example: all elements except the first {.smaller}

Let
$$
\mathbf C
=
\begin{pmatrix}
0&1&0&\cdots&0\\
0&0&1&\cdots&0\\
\vdots&\vdots&\vdots&\ddots&\vdots\\
0&0&0&\cdots&1
\end{pmatrix}
=\begin{pmatrix}\mathbf 0_{(p-1)\times 1}&\mathbf I_{p-1}\end{pmatrix}.
$$

Then
$$
\mathbf Y=\mathbf C\mathbf X
=
\begin{pmatrix}X_2\\X_3\\\vdots\\X_p\end{pmatrix}.
$$

\pause

Therefore
$$
\begin{pmatrix}X_2\\X_3\\\vdots\\X_p\end{pmatrix}
\sim
N_{p-1}\left(
\begin{pmatrix}\mu_2\\\mu_3\\\vdots\\\mu_p\end{pmatrix},
\boldsymbol\Sigma_{2:p,\,2:p}
\right).
$$

# Example: the sum of all elements

Let
$$
\mathbf 1_p
=
\begin{pmatrix}1\\\vdots\\1\end{pmatrix},
\qquad
Y=\mathbf 1_p^T\mathbf X.
$$

Then
$$
Y=X_1+X_2+\cdots+X_p.
$$

\pause

Using the linear-transformation theorem,
$$
Y \sim N\left(
\mathbf 1_p^T\boldsymbol\mu,
\mathbf 1_p^T\boldsymbol\Sigma\mathbf 1_p
\right).
$$

Equivalently,
$$
Y \sim N\left(
\sum_{i=1}^p\mu_i,
\sum_{i=1}^p\sum_{j=1}^p \sigma_{ij}
\right).
$$

# MVN partitions are MVN: theorem {.smaller}

\begin{block}{Theorem}
Let
$$
\mathbf X
=
\begin{pmatrix}\mathbf X^{(1)}\\\mathbf X^{(2)}\end{pmatrix}
\sim
N_p\left(
\begin{pmatrix}\boldsymbol\mu^{(1)}\\\boldsymbol\mu^{(2)}\end{pmatrix},
\begin{pmatrix}
\boldsymbol\Sigma_{11}&\boldsymbol\Sigma_{12}\\
\boldsymbol\Sigma_{21}&\boldsymbol\Sigma_{22}
\end{pmatrix}
\right),
$$
where $\mathbf X^{(1)}$ is $q\times 1$ and $\mathbf X^{(2)}$ is $r\times 1$.
Then
$$
\mathbf X^{(1)}\sim N_q(\boldsymbol\mu^{(1)},\boldsymbol\Sigma_{11}),
\qquad
\mathbf X^{(2)}\sim N_r(\boldsymbol\mu^{(2)},\boldsymbol\Sigma_{22}).
$$
\end{block}

# MVN partitions are MVN: proof {.smaller}

To extract $\mathbf X^{(1)}$, define
$$
\mathbf C_1=\begin{pmatrix}\mathbf I_q&\mathbf 0_{q\times r}\end{pmatrix}.
$$
Then
$$
\mathbf C_1\mathbf X
=
\begin{pmatrix}\mathbf I_q&\mathbf 0\end{pmatrix}
\begin{pmatrix}\mathbf X^{(1)}\\\mathbf X^{(2)}\end{pmatrix}
=
\mathbf X^{(1)}.
$$

\pause

The mean is
$$
\mathbf C_1\boldsymbol\mu
=
\begin{pmatrix}\mathbf I_q&\mathbf 0\end{pmatrix}
\begin{pmatrix}\boldsymbol\mu^{(1)}\\\boldsymbol\mu^{(2)}\end{pmatrix}
=
\boldsymbol\mu^{(1)}.
$$

# MVN partitions are MVN: proof continued {.smaller}

The covariance matrix is
$$
\begin{aligned}
\mathbf C_1\boldsymbol\Sigma\mathbf C_1^T
&=
\begin{pmatrix}\mathbf I_q&\mathbf 0\end{pmatrix}
\begin{pmatrix}
\boldsymbol\Sigma_{11}&\boldsymbol\Sigma_{12}\\
\boldsymbol\Sigma_{21}&\boldsymbol\Sigma_{22}
\end{pmatrix}
\begin{pmatrix}\mathbf I_q\\\mathbf 0\end{pmatrix} \\
&=\boldsymbol\Sigma_{11}.
\end{aligned}
$$

Therefore
$$
\mathbf X^{(1)}\sim N_q(\boldsymbol\mu^{(1)},\boldsymbol\Sigma_{11}).
$$

\pause

The proof for $\mathbf X^{(2)}$ is identical, using
$$
\mathbf C_2=\begin{pmatrix}\mathbf 0_{r\times q}&\mathbf I_r\end{pmatrix}.
$$

# Example: bivariate MVN marginals {.smaller}

Suppose
$$
\begin{pmatrix}X_1\\X_2\end{pmatrix}
\sim
N_2\left(
\begin{pmatrix}\mu_1\\\mu_2\end{pmatrix},
\begin{pmatrix}
\sigma_1^2 & \sigma_{12}\\
\sigma_{21} & \sigma_2^2
\end{pmatrix}
\right).
$$

Then, by the partition theorem,
$$
X_1\sim N(\mu_1,\sigma_1^2),
\qquad
X_2\sim N(\mu_2,\sigma_2^2).
$$

\pause

- A jointly normal vector has normal marginal distributions.
- The converse is not true in general: normal marginal distributions do not by themselves guarantee joint normality.

# Quadratic forms and the chi-square distribution {.smaller}

\begin{block}{Theorem}
Let
$$
\mathbf Y \sim N_n(\mathbf 0,\mathbf I_n).
$$
If $\mathbf A$ is symmetric and idempotent, with rank $k$, then
$$
\mathbf Y^T\mathbf A\mathbf Y \sim \chi^2_k.
$$
\end{block}

\pause

- Idempotent means $\mathbf A^2=\mathbf A$.
- Symmetry is needed for the orthogonal diagonalisation used in the proof.
- The rank $k$ determines the degrees of freedom.

# Quadratic forms and chi-square: proof {.smaller}

Since $\mathbf A$ is symmetric and idempotent, there is an orthonormal matrix $\mathbf P$ such that
$$
\mathbf P^T\mathbf A\mathbf P
=
\begin{pmatrix}
\mathbf I_k&\mathbf 0\\
\mathbf 0&\mathbf 0
\end{pmatrix}.
$$

Let
$$
\mathbf Z=\mathbf P^T\mathbf Y.
$$

\pause

By the linear-transformation theorem,
$$
\mathbf Z\sim N_n(\mathbf P^T\mathbf 0,\mathbf P^T\mathbf I_n\mathbf P)
=N_n(\mathbf 0,\mathbf I_n),
$$
because $\mathbf P^T\mathbf P=\mathbf I_n$.

# Quadratic forms and chi-square: proof continued {.smaller}

::::{.columns}

:::{.column width=58%}

Since $\mathbf Y=\mathbf P\mathbf Z$,
$$
\begin{aligned}
\mathbf Y^T\mathbf A\mathbf Y
&=(\mathbf P\mathbf Z)^T\mathbf A(\mathbf P\mathbf Z)\\
&=\mathbf Z^T\mathbf P^T\mathbf A\mathbf P\mathbf Z\\
&=\mathbf Z^T
\begin{pmatrix}
\mathbf I_k&\mathbf 0\\
\mathbf 0&\mathbf 0
\end{pmatrix}
\mathbf Z.
\end{aligned}
$$

:::

:::{.column width=42%}

Partition $\mathbf Z=(Z_1,\ldots,Z_n)^T$. Then
$$
\mathbf Y^T\mathbf A\mathbf Y
=
\sum_{i=1}^k Z_i^2.
$$

Because $\mathbf Z\sim N_n(\mathbf 0,\mathbf I_n)$, the $Z_i$ are independent standard normal random variables. Hence
$$
\boxed{\sum_{i=1}^k Z_i^2\sim \chi^2_k}.
$$

:::

::::

# Why this theorem matters for regression

- In the general linear model, residual sums of squares can be written as quadratic forms.
- Projection matrices used in regression are symmetric and idempotent.
- Their ranks determine the relevant degrees of freedom.

\pause

That is why this theorem is one of the foundations for:

- residual variance estimation
- $t$ tests for individual coefficients
- $F$ tests comparing linear models

# Summary

In this deck, we used three main ideas:

- The multivariate normal distribution is stable under linear transformations:
$$
\mathbf C\mathbf X
\sim
N(\mathbf C\boldsymbol\mu,\mathbf C\boldsymbol\Sigma\mathbf C^T).
$$

- Components and partitions of an MVN random vector are themselves normally distributed.

- A symmetric idempotent quadratic form in a standard MVN random vector is chi-square:
$$
\mathbf Y^T\mathbf A\mathbf Y\sim\chi^2_{\operatorname{rank}(\mathbf A)}.
$$
