---
title: "Multidimensional scaling"
subtitle: "STA5077Z coding practical"
author: "Miguel Rodó"
format:
  html:
    toc: true
    number-sections: true
    code-fold: false
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

Across 1,000 cells sampled from human blood, measurements were made of six cell-surface markers.
These markers help distinguish different cell lineages.
For the purpose of this practical, what this means is that cells of different types should separate in the space defined by the marker measurements, while cells of the same type should cluster together.
This is similar to, say, tortoises of different species separating in the space defined by their shell measurements, while tortoises of the same species cluster together.

The observational units are **cells**, and the variables are the marker measurements:

- `cd3`
- `cd4`
- `cd8`
- `cd16`
- `cd19`
- `cd33`

The aim of this practical is to use multidimensional scaling (MDS) to construct low-dimensional representations of the cells and to interpret those representations biologically.

You are **not** expected to programme the full MDS algorithms from first principles.
A first-principles implementation of classical MDS is included as an optional extension.

## Setup

```{r}
#| label: install-packages
pkg_vec <- c("remotes", "tidyverse", "MASS", "cowplot")
for (x in pkg_vec) {
  if (!requireNamespace(x, quietly = TRUE)) {
    install.packages(x)
  }
  library(x, character.only = TRUE)
}
if (!requireNamespace("DataTidy24STA5069Z", quietly = TRUE)) {
  remotes::install_github(
    "MiguelRodo/DataTidy24STA5069Z@2025"
  )
}
```

Load the required packages and data.

```{r}
#| label: load-mds-practical-data

path_cell_lineage <- system.file(
  "extdata",
  "cytof",
  "data",
  "data_tidy_cell_lineage.csv",
  package = "DataTidy24STA5069Z"
)

if (!file.exists(path_cell_lineage)) {
  stop(
    paste0(
      "Cannot find file: ",
      path_cell_lineage,
      ". Is DataTidy24STA5069Z installed and up to date?"
    )
  )
}

data_tidy_cell_lineage <- readr::read_csv(
  path_cell_lineage,
  show_col_types = FALSE
)
```

Prepare the marker data.

```{r}
#| label: prepare-cell-marker-data

marker_names <- c(
  "cd3",
  "cd4",
  "cd8",
  "cd16",
  "cd19",
  "cd33"
)

cell_marker_tbl <- data_tidy_cell_lineage |>
  dplyr::select(
    cell_id,
    dplyr::all_of(marker_names)
  )

cell_marker_mat <- cell_marker_tbl |>
  dplyr::select(
    dplyr::all_of(marker_names)
  ) |>
  as.matrix()

rownames(cell_marker_mat) <- cell_marker_tbl$cell_id
```

Here is a density plot of the marker measurements.

```{r}
#| label: fig-cell-marker-densities
#| fig-cap: "Density plots of the six lineage-marker measurements across the sampled cells."

cell_marker_tbl |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(marker_names),
    names_to = "marker",
    values_to = "expression"
  ) |>
  ggplot(
    aes(
      x = expression,
      fill = marker
    )
  ) +
  geom_density(
    alpha = 0.65,
    colour = "black"
  ) +
  facet_wrap(
    ~ marker,
    scales = "free"
  ) +
  guides(fill = "none") +
  labs(
    x = "Marker expression",
    y = "Density"
  ) +
  theme_cowplot() +
  background_grid(major = "xy")
```

# Core practical



## Question 3: Define the dissimilarities

For the main analysis, use Euclidean distance between cells:

$$
\delta_{ij}
=
\sqrt{
\sum_{m=1}^{6}
\left(
x_{im}-x_{jm}
\right)^2
}.
$$

```{r}
#| label: calculate-cell-euclidean-dissimilarities

cell_delta <- dist(
  cell_marker_mat,
  method = "euclidean"
)

cell_delta
```

Explain in words what a small and a large Euclidean dissimilarity mean in this setting.

Also answer:

1. Why is the dissimilarity matrix symmetric?
2. Why is its diagonal zero?
3. How many distinct pairwise dissimilarities are there for 1,000 cells?
4. How would your interpretation change if maximum distance were used instead?

## Question 4: Fit non-metric MDS

Fit a two-dimensional non-metric MDS configuration.

```{r}
#| label: fit-cell-nonmetric-mds

cell_nmds_fit <- MASS::isoMDS(
  cell_delta,
  k = 2,
  trace = FALSE
)

cell_nmds_fit$stress
```

Store the fitted coordinates in a tibble.

```{r}
#| label: prepare-cell-nmds-coordinates

cell_nmds_tbl <- tibble::as_tibble(
  cell_nmds_fit$points,
  .name_repair = ~ c(
    "dimension_1",
    "dimension_2"
  )
) |>
  dplyr::mutate(
    cell_id = cell_marker_tbl$cell_id,
    .before = 1
  )
```

Answer the following:

1. How many rows and columns does the coordinate matrix contain?
2. What does each row represent?
3. What does the reported stress measure?
4. Why should stress be interpreted together with graphical diagnostics rather than by applying a single universal cut-off?

## A helper for plotting marker-coloured configurations

The following function handles the reshaping and plotting required to colour an MDS configuration by each marker.

```{r}
#| label: define-marker-configuration-plotter

plot_marker_configuration <- function(
  coordinates,
  marker_data,
  plot_title = NULL
) {
  coordinate_tbl <- tibble::as_tibble(
    coordinates,
    .name_repair = ~ c(
      "dimension_1",
      "dimension_2"
    )
  )

  plot_tbl <- marker_data |>
    dplyr::select(
      cell_id,
      dplyr::all_of(marker_names)
    ) |>
    dplyr::bind_cols(coordinate_tbl) |>
    tidyr::pivot_longer(
      cols = dplyr::all_of(marker_names),
      names_to = "marker",
      values_to = "expression"
    )

  ggplot(
    plot_tbl,
    aes(
      x = dimension_1,
      y = dimension_2,
      colour = expression
    )
  ) +
    geom_point(
      alpha = 0.70,
      size = 0.75
    ) +
    facet_wrap(~ marker) +
    coord_equal() +
    scale_colour_viridis_c() +
    labs(
      x = "First MDS dimension",
      y = "Second MDS dimension",
      colour = "Expression",
      title = plot_title
    ) +
    theme_minimal()
}
```

## Question 5: Interpret the non-metric MDS configuration

Plot the non-metric MDS configuration, colouring the cells by the expression of each marker.

```{r}
#| label: fig-cell-nmds-markers
#| fig-cap: "Two-dimensional non-metric MDS configuration of the cells, coloured separately by each lineage-marker measurement."

plot_marker_configuration(
  coordinates = cell_nmds_fit$points,
  marker_data = cell_marker_tbl,
  plot_title = "Non-metric multidimensional scaling"
)
```

Interpret the plot.

Your discussion should address:

- which markers form the clearest gradients across the map;
- whether high-CD4 and high-CD8 cells occupy different regions;
- whether CD16, CD19 or CD33 identify distinct regions;
- whether the map suggests several cell populations;
- how those apparent populations could be described using the marker measurements; and
- why visual separation in an MDS plot does not, by itself, establish formal clusters.

Remember that the absolute directions of the axes do not have inherent meanings. The configuration could be translated, rotated or reflected without changing the represented pairwise distances.

## Question 6: Assess the non-metric MDS fit

Construct a Shepard diagram.

```{r}
#| label: prepare-cell-nmds-shepard

cell_nmds_shepard <- MASS::Shepard(
  cell_delta,
  cell_nmds_fit$points
)

cell_nmds_shepard_tbl <- tibble::tibble(
  dissimilarity = cell_nmds_shepard$x,
  fitted_distance = cell_nmds_shepard$y,
  fitted_disparity = cell_nmds_shepard$yf
) |>
  dplyr::arrange(dissimilarity)
```

```{r}
#| label: fig-cell-nmds-shepard
#| fig-cap: "Shepard diagram comparing the original cell dissimilarities with distances in the two-dimensional non-metric MDS configuration."

ggplot(
  cell_nmds_shepard_tbl,
  aes(
    x = dissimilarity,
    y = fitted_distance
  )
) +
  geom_point(
    alpha = 0.25,
    size = 0.65
  ) +
  geom_step(
    aes(y = fitted_disparity),
    linewidth = 0.8
  ) +
  labs(
    x = "Original dissimilarity",
    y = "Distance in fitted configuration"
  ) +
  theme_minimal()
```

Comment on:

1. whether the fitted distances tend to increase with the original dissimilarities;
2. whether there is substantial scatter around the fitted monotone relationship; and
3. whether the two-dimensional configuration appears to give a useful representation.

## Question 7: Fit classical MDS

Fit classical MDS to the **same Euclidean dissimilarity matrix**.

```{r}
#| label: fit-cell-classical-mds

cell_cmds_fit <- cmdscale(
  cell_delta,
  k = 2,
  eig = TRUE
)

cell_cmds_fit
```

Extract the two-dimensional coordinates.

```{r}
#| label: prepare-cell-cmds-coordinates

cell_cmds_coordinates <- cell_cmds_fit$points

cell_cmds_tbl <- tibble::as_tibble(
  cell_cmds_coordinates,
  .name_repair = ~ c(
    "dimension_1",
    "dimension_2"
  )
) |>
  dplyr::mutate(
    cell_id = cell_marker_tbl$cell_id,
    .before = 1
  )
```

Inspect the eigenvalues.

```{r}
#| label: inspect-cell-cmds-eigenvalues

tibble::tibble(
  dimension = seq_along(cell_cmds_fit$eig),
  eigenvalue = cell_cmds_fit$eig
) |>
  dplyr::slice_head(n = 10)
```

Answer the following:

1. What do the leading eigenvalues indicate?
2. Are there any important negative eigenvalues?
3. Why can negative eigenvalues occur when a dissimilarity matrix is not exactly representable as Euclidean distances?
4. In this particular analysis, what would you expect given that the input dissimilarities were calculated using ordinary Euclidean distance?

## Question 8: Interpret the classical MDS configuration

Plot the classical MDS configuration using the supplied helper function.

```{r}
#| label: fig-cell-cmds-markers
#| fig-cap: "Two-dimensional classical MDS configuration of the cells, coloured separately by each lineage-marker measurement."

plot_marker_configuration(
  coordinates = cell_cmds_coordinates,
  marker_data = cell_marker_tbl,
  plot_title = "Classical multidimensional scaling"
)
```

Interpret the plot using the same questions considered for non-metric MDS.

Do not compare the raw signs of the axes between the two methods. A reflected or rotated configuration may represent essentially the same pairwise relationships.

## Question 9: Compare classical and non-metric MDS

Compare the two configurations.

Your answer should address:

- whether the same broad marker-defined regions appear;
- whether one method separates any cell populations more clearly;
- the distinction between preserving dissimilarity magnitudes and preserving their ordering;
- the fit diagnostics available for each method; and
- why the coordinates should not be compared axis by axis without first allowing for rotation and reflection.

Write a concise paragraph explaining which configuration you find more useful for these data and why.

# Optional extensions

::: {.callout-note collapse="true" title="Optional extension 1: maximum distance"}

Maximum distance between cells is defined by the marker on which they differ most:

$$
\delta_{ij}^{\max}
=
\max_m
\left|
x_{im}-x_{jm}
\right|.
$$

Calculate maximum-distance dissimilarities and fit non-metric MDS.

```{r}
#| label: optional-fit-maximum-distance-nmds

cell_delta_maximum <- dist(
  cell_marker_mat,
  method = "maximum"
)

cell_nmds_maximum_fit <- MASS::isoMDS(
  cell_delta_maximum,
  k = 2,
  trace = FALSE
)
```

```{r}
#| label: fig-cell-maximum-distance-nmds
#| fig-cap: "Two-dimensional non-metric MDS configuration based on maximum-distance dissimilarities, coloured by the lineage-marker measurements."

plot_marker_configuration(
  coordinates = cell_nmds_maximum_fit$points,
  marker_data = cell_marker_tbl,
  plot_title = "Non-metric MDS using maximum distance"
)
```

Explain how the meaning of proximity has changed.

In particular, consider whether it is scientifically appropriate for two cells to be regarded as highly dissimilar whenever they differ strongly on even one marker.

:::

::: {.callout-note collapse="true" title="Optional extension 2: classical MDS from first principles"}

Classical MDS begins from the squared dissimilarity matrix.

```{r}
#| label: optional-form-classical-mds-matrices

cell_distance_mat <- as.matrix(cell_delta)
n_cells <- nrow(cell_distance_mat)

centering_mat <- diag(n_cells) -
  matrix(
    1 / n_cells,
    nrow = n_cells,
    ncol = n_cells
  )

inner_product_mat <- -0.5 *
  centering_mat %*%
  (cell_distance_mat^2) %*%
  centering_mat
```

Eigendecompose the double-centred matrix.

```{r}
#| label: optional-eigendecompose-classical-mds-matrix

inner_product_eigen <- eigen(
  inner_product_mat,
  symmetric = TRUE
)

leading_eigenvalues <- inner_product_eigen$values[1:2]
leading_eigenvectors <- inner_product_eigen$vectors[, 1:2, drop = FALSE]
```

Construct the two-dimensional principal coordinates.

```{r}
#| label: optional-construct-manual-cmds-coordinates

cell_cmds_manual <- leading_eigenvectors %*%
  diag(
    sqrt(leading_eigenvalues),
    nrow = 2
  )
```

Compare the pairwise distances from the manual configuration with those returned by `cmdscale()`.

```{r}
#| label: optional-compare-manual-and-cmdscale-distances

comparison_tbl <- tibble::tibble(
  manual = as.vector(dist(cell_cmds_manual)),
  cmdscale = as.vector(dist(cell_cmds_coordinates))
)

cor(
  comparison_tbl$manual,
  comparison_tbl$cmdscale
)
```

A correlation extremely close to one confirms that the configurations preserve the same pairwise relationships, even if their coordinate axes differ in sign or orientation.

:::

# Final reflection

In no more than 200 words, summarise what the MDS analyses reveal about the cell-lineage data.

Your summary should state:

- how dissimilarity between cells was defined;
- what the major marker gradients or apparent populations were;
- whether the two-dimensional non-metric representation appeared adequate;
- how the classical and non-metric configurations compared; and
- one limitation of interpreting a two-dimensional MDS map.
