---
title: "Multidimensional scaling: solutions"
subtitle: "STA5077Z coding practical solutions"
author: "Miguel Rodó"
format:
  html:
    toc: true
    number-sections: true
    code-fold: false
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

This practical uses simulated dissimilarities to examine classical, least-squares and non-metric multidimensional scaling.
You will repair an invalid dissimilarity matrix, assess a classical MDS representation, investigate different definitions of dissimilarity, and compare what the three methods preserve.

## Setup

Load the required packages.

```{r}
#| label: load-mds-practical-packages
library(MASS)
library(tibble)
library(dplyr)
library(ggplot2)

dir.create(
  file.path("_tmp", "l2"),
  recursive = TRUE,
  showWarnings = FALSE
)
```

The following code generates three groups of observations and constructs a dissimilarity matrix (`diss_mat_orig`) containing several deliberate errors.

```{r}
#| label: create-mds-practical-data
#| include: false

set.seed(5077)

pos_def <- function(n, covEvMin = 0.25, covEvMax = 0.75) {
    # Generates a positive definite matrix of dimension n
    # Upper and lower bounds control the range of the variance eigenvalues.

    # Generate the bounded eigenvalues before matrix decomposition
    ev <- runif(n, min = covEvMin, max = covEvMax)

    if (n == 1) {
        return(matrix(ev, 1, 1))
    }

    z <- matrix(ncol = n, rnorm(n^2))
    decomp <- qr(z)
    q <- qr.Q(decomp)
    r <- qr.R(decomp)
    d <- diag(r)
    ph <- d / abs(d)

    o <- q %*% diag(ph, nrow = n)
    z <- t(o) %*% diag(ev, nrow = n) %*% o
    return(z)
}

simulate_group <- function(group, mean, n = 8) {
  simulated_values <- MASS::mvrnorm(
    n = n,
    mu = mean,
    Sigma = pos_def(length(mean))
  )
  colnames(simulated_values) <- paste0("v", seq_len(length(mean)))

  tibble(
    label = paste0(group, seq_len(n)),
    group = group
  ) |>
    dplyr::bind_cols(
      as_tibble(simulated_values)
    )
}

mds_observations_tbl <- bind_rows(
  simulate_group(
    group = "A",
    mean = c(-2.2, -0.6, -1, -4)
  ),
  simulate_group(
    group = "B",
    mean = c(0.0, 1.8, 1, 1.5)
  ),
  simulate_group(
    group = "C",
    mean = c(2.1, -0.7, 1, 3)
  )
)

mds_feature_mat <- mds_observations_tbl |>
  dplyr::select(v1, v2, v3, v4) |>
  as.matrix()

rownames(mds_feature_mat) <- mds_observations_tbl$label

original_euclidean_mat <- as.matrix(
  dist(mds_feature_mat, method = "euclidean")
)

diss_mat_orig <- original_euclidean_mat

diag(diss_mat_orig)[c(3, 11, 20)] <- c(
  0.45,
  0.80,
  0.60
)

diss_mat_orig[2, 15] <-
  1.15 * diss_mat_orig[2, 15]

diss_mat_orig[6, 21] <-
  diss_mat_orig[6, 21] + 0.80

diss_mat_orig[10, 18] <-
  0.75 * diss_mat_orig[10, 18]

diss_mat_orig[13, 24] <-
  diss_mat_orig[13, 24] + 0.50

mds_group_key_tbl <- mds_observations_tbl |>
  dplyr::select(label, group)
```

# Core practical

## Question 1: Repair the dissimilarity matrix

Identify and fix the errors in `diss_mat_orig`.
Hint: if need be, off-diagonal elements can be averaged using 

$$
\delta_{ij}^{*}
=
\frac{\delta_{ij}+\delta_{ji}}{2}.
$$

```{r}
#| label: identify-invalid-dissimilarities

diss_mat <- diss_mat_orig

for (i in seq_len(nrow(diss_mat))) {
  for (j in seq_len(nrow(diss_mat))) {
    if (i == j) {
      diss_mat[i, i] <- 0
    } else {
      diss_mat[i, j] <- mean(
        c(diss_mat_orig[i, j], diss_mat_orig[j, i])
      )
    }
  }
}

all(diag(diss_mat) == 0)
isSymmetric(diss_mat)

# A faster equivalent is:
# diag(diss_mat) <- 0
# diss_mat <- (diss_mat + t(diss_mat)) / 2
```

## Question 2: Apply and interpret classical MDS

Fit a two-dimensional classical MDS representation using `cmdscale()` to your repaired distance matrix.

```{r}
#| label: fit-classical-mds
cmdscale_obj <- cmdscale(
  diss_mat,
  k = 2,
  eig = FALSE,
  add = FALSE
)
```

Plot the coordinates, indicating group membership, which are provided by the object `mds_group_key_tbl` in the `group` column.
The ordering of the `diss_mat_orig` and `mds_group_key_tbl` are the same, i.e. the $i$-th row of `mds_group_key_tbl` matches the $i$-th row/column of `diss_mat_orig`.
Use equal scaling on the two axes and colour the observations by group.

```{r}
#| label: fig-classical-mds-configuration
plot_tbl <- tibble(
  group = mds_group_key_tbl$group,
  MDS1 = cmdscale_obj[, 1],
  MDS2 = cmdscale_obj[, 2]
)

p <- ggplot(
  plot_tbl,
  aes(
    x = MDS1,
    y = MDS2,
    colour = group
  )
) +
  geom_point(size = 2.5) +
  coord_equal() +
  labs(
    x = "MDS coordinate 1",
    y = "MDS coordinate 2",
    colour = "Group"
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  )

path_p <- file.path("_tmp", "l2", "p-q1.png")

ggsave(
  filename = path_p,
  plot = p,
  units = "cm",
  height = 10,
  width = 10
)
```

![](`r path_p`)

The observations form three visible groups. Observations within a group are generally closer together than observations belonging to different groups. Groups A and C lie furthest apart, with group B between them.

## Question 3: Choose the number of dimensions

Fit classical MDS representations using one to five dimensions and calculate the relative distance error for each representation.
Comment on the results.

$$
\text{relative distance error}
=
\left[
\frac{
\sum_{i<j}(d_{ij}-\delta_{ij})^2
}{
\sum_{i<j}\delta_{ij}^2
}
\right]^{1/2}.
$$

```{r}
#| label: calculate-dimension-fit
err_vec <- purrr::map_dbl(1:5, function(n_dim) {
  cmdscale_obj <- cmdscale(
    diss_mat,
    k = n_dim,
    eig = FALSE,
    add = FALSE
  )

  diss_vec <- diss_mat[lower.tri(diss_mat)]
  denom <- sum(diss_vec^2)

  dist_coord <- dist(
    cmdscale_obj,
    upper = TRUE,
    diag = TRUE
  ) |>
    as.matrix()

  numerator <- 0

  for (i in seq_len(ncol(diss_mat) - 1)) {
    for (j in seq(i + 1, ncol(diss_mat))) {
      numerator <- numerator +
        (diss_mat[i, j] - dist_coord[i, j])^2
    }
  }

  sqrt(numerator / denom)
})

err_vec |>
  stats::setNames(paste0("n_dim_", 1:5)) |>
  signif(2)
```

The error decreases as dimensions are added. The observations were generated in four dimensions, so the substantial improvement up to four dimensions is expected. The repaired matrix is not exactly the original Euclidean distance matrix, so the error need not become exactly zero.

## Question 4: Compare fitted and input dissimilarities

Plot the interpoint distances from the two-dimensional MDS fit on the y-axis against the original dissimilarities on the x-axis.

```{r}
#| label: fig-classical-mds-distance-comparison
cmdscale_obj <- cmdscale(
  diss_mat,
  k = 2,
  eig = FALSE,
  add = FALSE
)

dist_coord <- dist(
  cmdscale_obj,
  upper = TRUE,
  diag = TRUE
) |>
  as.matrix()

plot_tbl <- tibble(
  ind = seq_len((24 * 23) / 2),
  orig = rep(NA_real_, (24 * 23) / 2),
  fit = rep(NA_real_, (24 * 23) / 2)
)

k <- 0

for (i in seq_len(ncol(diss_mat) - 1)) {
  for (j in seq(i + 1, ncol(diss_mat))) {
    k <- k + 1
    plot_tbl[k, "orig"] <- diss_mat[i, j]
    plot_tbl[k, "fit"] <- dist_coord[i, j]
  }
}

p <- ggplot(plot_tbl, aes(x = orig, y = fit)) +
  geom_abline(slope = 1, intercept = 0) +
  geom_point(alpha = 0.5, size = 0.5) +
  coord_equal() +
  labs(
    x = "Input dissimilarity",
    y = "Coordinate distance"
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  )

path_p <- file.path("_tmp", "l2", "p-q3.png")

ggsave(
  filename = path_p,
  plot = p,
  units = "cm",
  height = 10,
  width = 10
)
```

![](`r path_p`)

The two-dimensional solution retains the broad distance structure, but does not reproduce the dissimilarities perfectly because the observations were generated in four dimensions.

## Question 5: Experiment with the dissimilarities

### Part a

Cube each of the distances in your dissimilarity matrix.

```{r}
#| label: cube-dissimilarities

diss_mat_cubed <- diss_mat^3
```

Fit the following two-dimensional representations:

1. classical MDS using `cmdscale()`;
2. non-metric MDS using `MASS::isoMDS()`; and
3. least-squares MDS minimising the following objective function:

$$
\sum_{i<j}(d_{ij}-\delta_{ij})^2/\delta_{ij}^2
$$

As some help to do the least-squares MDS:

- note that your goal is to obtain $n$ pairs of 2-dimensional coordinate positions, i.e. you are trying to obtain an $n\times 2$ matrix, where each row corresponds to a single observation and the columns correspond to the x- and y-axis coordinate positions (or, equivalently, two variable values).
- You can use the `optim` function.
- Use the `cmdscale` coordinates as the starting values.
- The function that you are minimising must treat the coordinates as the parameters with respect to which the objective is minimised, and should return the value of the objective function as defined above. It does not return the coordinate values themselves.
  - You will need to use the coordinate values internally to that function to calculate the MDS interpoint distances.
  - These need to be compared to the input dissimilarities.
- As a check on whether you are correct, the `MASS::sammon` function optimises a somewhat similar objective function, with denominator $\delta_{ij}$ rather than $\delta_{ij}^2$.

```{r}
#| label: fit-cubed-dissimilarity-mds

n_obs <- nrow(diss_mat_cubed)
diss_cubed_obj <- as.dist(diss_mat_cubed)
diss_cubed_vec <- as.vector(diss_cubed_obj)

classical_points <- cmdscale(
  diss_cubed_obj,
  k = 2,
  eig = FALSE,
  add = FALSE
)

nonmetric_fit <- MASS::isoMDS(
  diss_cubed_obj,
  y = classical_points,
  k = 2,
  trace = FALSE
)

weighted_ls_objective <- function(
  par,
  input_dissimilarities,
  n_obs
) {
  coordinates <- matrix(
    par,
    nrow = n_obs,
    ncol = 2
  )

  coordinate_distances <- as.vector(
    dist(coordinates)
  )

  sum(
    (coordinate_distances - input_dissimilarities)^2 /
      input_dissimilarities^2
  )
}

weighted_ls_optim <- optim(
  par = as.vector(classical_points),
  fn = weighted_ls_objective,
  input_dissimilarities = diss_cubed_vec,
  n_obs = n_obs,
  method = "BFGS",
  control = list(
    maxit = 5000,
    reltol = 1e-10
  )
)

weighted_ls_points <- matrix(
  weighted_ls_optim$par,
  nrow = n_obs,
  ncol = 2
)

weighted_ls_optim$convergence
weighted_ls_optim$value

sammon_check <- MASS::sammon(
  diss_cubed_obj,
  y = classical_points,
  k = 2,
  trace = FALSE
)

weighted_ls_objective(
  as.vector(sammon_check$points),
  input_dissimilarities = diss_cubed_vec,
  n_obs = n_obs
)
```

The final line evaluates the weighted least-squares objective at the Sammon coordinates. The two fits need not agree because `MASS::sammon()` uses $1/\delta_{ij}$ weights rather than $1/\delta_{ij}^2$ weights.

### Part b

Assess each approach, using the following plots:

1. Plot the MDS coordinates.
2. Plot the interpoint distances against the original dissimilarities.
3. Plot the interpoint distances against the ranks of the original dissimilarities.

```{r}
#| label: prepare-mds-comparisons

coordinate_tbl <- bind_rows(
  tibble(
    method = "Classical MDS",
    group = mds_group_key_tbl$group,
    coordinate_1 = classical_points[, 1],
    coordinate_2 = classical_points[, 2]
  ),
  tibble(
    method = "Non-metric MDS",
    group = mds_group_key_tbl$group,
    coordinate_1 = nonmetric_fit$points[, 1],
    coordinate_2 = nonmetric_fit$points[, 2]
  ),
  tibble(
    method = "Weighted least-squares MDS",
    group = mds_group_key_tbl$group,
    coordinate_1 = weighted_ls_points[, 1],
    coordinate_2 = weighted_ls_points[, 2]
  )
)

make_distance_comparison <- function(points, method_name) {
  tibble(
    method = method_name,
    input_dissimilarity = diss_cubed_vec,
    input_rank = rank(diss_cubed_vec),
    coordinate_distance = as.vector(dist(points))
  )
}

distance_comparison_tbl <- bind_rows(
  make_distance_comparison(
    classical_points,
    "Classical MDS"
  ),
  make_distance_comparison(
    nonmetric_fit$points,
    "Non-metric MDS"
  ),
  make_distance_comparison(
    weighted_ls_points,
    "Weighted least-squares MDS"
  )
)
```

```{r}
#| label: fig-compare-mds-configurations

plot_configuration <- function(method_name) {
  coordinate_tbl |>
    filter(method == method_name) |>
    ggplot(
      aes(
        x = coordinate_1,
        y = coordinate_2,
        colour = group
      )
    ) +
    geom_point(size = 2.5) +
    coord_equal() +
    labs(
      title = method_name,
      x = "Coordinate 1",
      y = "Coordinate 2",
      colour = "Group"
    ) +
    cowplot::theme_cowplot() +
    cowplot::background_grid(major = "xy") +
    theme(
      plot.background = element_rect(fill = "white"),
      panel.background = element_rect(fill = "white")
    )
}

p <- cowplot::plot_grid(
  plot_configuration("Classical MDS"),
  plot_configuration("Non-metric MDS"),
  plot_configuration("Weighted least-squares MDS"),
  nrow = 1
)

path_p <- file.path(
  "_tmp", "l2", "p-q5-configurations.png"
)

ggsave(
  filename = path_p,
  plot = p,
  units = "cm",
  height = 10,
  width = 22
)
```

![](`r path_p`)

```{r}
#| label: fig-compare-mds-distances

p <- ggplot(
  distance_comparison_tbl,
  aes(
    x = input_dissimilarity,
    y = coordinate_distance
  )
) +
  geom_abline(slope = 1, intercept = 0) +
  geom_point(alpha = 0.5, size = 0.7) +
  facet_wrap(~method, scales = "free") +
  labs(
    x = "Input dissimilarity",
    y = "Coordinate distance"
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  )

path_p <- file.path(
  "_tmp", "l2", "p-q5-distances.png"
)

ggsave(
  filename = path_p,
  plot = p,
  units = "cm",
  height = 10,
  width = 22
)
```

![](`r path_p`)

```{r}
#| label: fig-compare-mds-ranks

p <- ggplot(
  distance_comparison_tbl,
  aes(
    x = input_rank,
    y = coordinate_distance
  )
) +
  geom_line(alpha = 0.5) +
  geom_point(alpha = 0.5, size = 0.7) +
  facet_wrap(~method, scales = "free_y") +
  labs(
    x = "Rank of input dissimilarity",
    y = "Coordinate distance"
  ) +
  cowplot::theme_cowplot() +
  cowplot::background_grid(major = "xy") +
  theme(
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white")
  )

path_p <- file.path(
  "_tmp", "l2", "p-q5-ranks.png"
)

ggsave(
  filename = path_p,
  plot = p,
  units = "cm",
  height = 10,
  width = 22
)
```

![](`r path_p`)

```{r}
#| label: summarise-mds-comparison

comparison_summary_tbl <- distance_comparison_tbl |>
  group_by(method) |>
  summarise(
    relative_distance_error = sqrt(
      sum(
        (coordinate_distance - input_dissimilarity)^2
      ) /
        sum(input_dissimilarity^2)
    ),
    weighted_relative_error = sum(
      (coordinate_distance - input_dissimilarity)^2 /
        input_dissimilarity^2
    ),
    rank_correlation = cor(
      input_rank,
      coordinate_distance,
      method = "spearman"
    ),
    .groups = "drop"
  )

comparison_summary_tbl
```

Classical MDS targets the inner-product structure implied by the dissimilarities, rather than either of the two plotted criteria directly. Non-metric MDS is intended to preserve the ordering of the dissimilarities, so it should perform best in the rank comparison. The weighted least-squares fit directly minimises the squared relative discrepancies and therefore gives much less weight to errors associated with large dissimilarities.
