---
title: "Regression Week V: Matching the model to the data"
subtitle: "Slide deck I: Transformations and additional model terms"
author: "Miguel Rodo"
institute: University of Cape Town
date: today
bibliography: zotero.bib
format:
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: ../preamble.tex
    keep-tex: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

format_number <- function(x, digits = 3) {
  format(signif(x, digits), trim = TRUE, scientific = FALSE)
}

make_plot_window <- function() {
  par(
    mar = c(4.0, 4.2, 2.2, 0.8),
    mgp = c(2.3, 0.7, 0),
    tcl = -0.25,
    las = 1
  )
}

plot_residuals <- function(fit, x, xlab) {
  plot(
    x,
    resid(fit),
    pch = 19,
    cex = 0.75,
    col = "grey35",
    xlab = xlab,
    ylab = "residual",
    main = "Residual plot"
  )
  abline(h = 0, lwd = 2, col = "grey30")
  lines(stats::lowess(x, resid(fit)), lwd = 2.2, col = "#003366")
  box()
}

set.seed(501)
n_curve <- 120
curve_tbl <- data.frame(
  x = runif(n_curve, 0, 10)
)
curve_tbl$y <- 4 + 1.5 * curve_tbl$x + 0.35 * (curve_tbl$x - 5)^2 +
  rnorm(n_curve, sd = 1.4)

fit_curve_linear <- lm(y ~ x, data = curve_tbl)
fit_curve_quad <- lm(y ~ x + I((x - mean(x))^2), data = curve_tbl)
curve_grid <- data.frame(
  x = seq(min(curve_tbl$x), max(curve_tbl$x), length.out = 200)
)
curve_grid$linear <- predict(fit_curve_linear, newdata = curve_grid)
curve_grid$quadratic <- predict(fit_curve_quad, newdata = curve_grid)

set.seed(502)
n_log_y <- 160
log_y_tbl <- data.frame(
  x = runif(n_log_y, 0, 8)
)
log_y_tbl$y <- exp(1.0 + 0.22 * log_y_tbl$x + rnorm(n_log_y, sd = 0.28))
fit_log_y_raw <- lm(y ~ x, data = log_y_tbl)
fit_log_y <- lm(log(y) ~ x, data = log_y_tbl)
log_y_grid <- data.frame(x = seq(min(log_y_tbl$x), max(log_y_tbl$x), length.out = 200))
log_y_grid$raw_fit <- predict(fit_log_y_raw, newdata = log_y_grid)
log_y_grid$log_fit_original_scale <- exp(predict(fit_log_y, newdata = log_y_grid))

set.seed(503)
n_x <- 160
x_trans_tbl <- data.frame(
  x1 = runif(n_x, 0, 8),
  x2 = runif(n_x, 1, 80)
)
x_trans_tbl$y <- 3 + 1.2 * x_trans_tbl$x1 + 4.5 * log(x_trans_tbl$x2) +
  rnorm(n_x, sd = 1.4)
fit_x_raw <- lm(y ~ x1 + x2, data = x_trans_tbl)
fit_x_log <- lm(y ~ x1 + log(x2), data = x_trans_tbl)

set.seed(504)
n_disc <- 100
disc_tbl <- data.frame(x = runif(n_disc, 0, 10))
disc_tbl$y <- 2 + 0.45 * disc_tbl$x + 0.25 * (disc_tbl$x - 5)^2 +
  rnorm(n_disc, sd = 1.1)
disc_tbl$x_cut <- cut(disc_tbl$x, breaks = 4)
fit_disc_linear <- lm(y ~ x, data = disc_tbl)
fit_disc_cut <- lm(y ~ x_cut, data = disc_tbl)
```

# Why we are changing the model

- Week IV used residual plots and related diagnostics to ask whether a fitted model is credible.

\pause

- Sometimes the problem is not the method of estimation.

\pause

- The problem is that the model is trying to represent the data with the wrong shape.

\pause

- This deck considers targeted changes to the model, while retaining the regression framework developed so far.

# Lecture outline

By the end of this deck, you should be able to:

- Explain why a fitted linear model may not match the data
- Decide whether to transform the response, transform an explanatory variable, or add model terms
- Interpret coefficients after transforming the response or an explanatory variable
- Construct and interpret polynomial models
- Explain why discretising a numerical explanatory variable is usually crude
- Refit and re-check a model after modifying its structure

# From diagnosis to model modification

A useful workflow is:

$$
\boxed{
\text{fit model}
\longrightarrow
\text{check diagnostics}
\longrightarrow
\text{modify model}
\longrightarrow
\text{re-check diagnostics}
}
$$

\pause

Typical signs from diagnostics:

- Curved residual pattern: the conditional mean may be misspecified
- Residual spread changes with fitted values: error variance may not be constant
- A few extreme residuals dominate the fit: the scale of the response may be problematic
- Different groups show different trends: the model may need an interaction term

# What can go wrong?

The ordinary general linear model assumes, among other things, that

$$
Y=X\boldsymbol\beta+\mathbf e,
\qquad
E(\mathbf e\mid X)=\mathbf 0,
\qquad
\operatorname{Var}(\mathbf e\mid X)=\sigma^2I_n.
$$

\pause

The issue is not that the marginal distribution of $Y$ must be normal.

\pause

The issue is whether the model gives a reasonable description of the conditional mean and the errors.

# A motivating example: a straight line is too simple

```{r}
#| fig-width: 7.2
#| fig-height: 4.2
#| out-width: "82%"
old_par <- par(no.readonly = TRUE)
make_plot_window()
plot(
  curve_tbl$x,
  curve_tbl$y,
  pch = 19,
  cex = 0.75,
  col = "grey35",
  xlab = "x",
  ylab = "y",
  main = "A linear fit to a curved relationship"
)
abline(fit_curve_linear, lwd = 2.4, col = "#003366")
box()
par(old_par)
```

# The residuals show the problem

```{r}
#| fig-width: 7.2
#| fig-height: 4.2
#| out-width: "82%"
old_par <- par(no.readonly = TRUE)
make_plot_window()
plot_residuals(fit_curve_linear, curve_tbl$x, "x")
par(old_par)
```

A systematic residual pattern is evidence that the fitted conditional mean is missing structure.

# One targeted response: add a polynomial term

Instead of forcing a straight line, fit

$$
Y_i
=
\beta_0+
\beta_1x_i+
\beta_2(x_i-\bar{x})^2+e_i.
$$

\pause

This is still a linear regression model because the model is linear in the unknown coefficients.

\pause

The design matrix now has the columns

$$
\begin{bmatrix}
1 & x_i & (x_i-\bar{x})^2
\end{bmatrix}.
$$

# The modified model has the right shape

```{r}
#| fig-width: 7.2
#| fig-height: 4.2
#| out-width: "82%"
old_par <- par(no.readonly = TRUE)
make_plot_window()
plot(
  curve_tbl$x,
  curve_tbl$y,
  pch = 19,
  cex = 0.75,
  col = "grey35",
  xlab = "x",
  ylab = "y",
  main = "Linear and polynomial fits"
)
lines(curve_grid$x, curve_grid$linear, lwd = 2.2, col = "#003366")
lines(curve_grid$x, curve_grid$quadratic, lwd = 2.2, col = "#990000")
legend(
  "topright",
  legend = c("linear", "polynomial"),
  lwd = 2.2,
  col = c("#003366", "#990000"),
  bty = "n"
)
box()
par(old_par)
```

# Re-check after modifying the model

```{r}
#| fig-width: 7.2
#| fig-height: 4.2
#| out-width: "82%"
old_par <- par(no.readonly = TRUE)
make_plot_window()
plot_residuals(fit_curve_quad, curve_tbl$x, "x")
par(old_par)
```

A model modification is not complete until the new model has been checked.

# Three different tactics

The main tactics in this deck are:

\pause

1. Transform the response:

$$
\log(Y_i)=\beta_0+\beta_1x_i+e_i.
$$

\pause

2. Transform an explanatory variable:

$$
Y_i=\beta_0+\beta_1\log(x_i)+e_i.
$$

\pause

3. Add model terms:

$$
Y_i=\beta_0+\beta_1x_i+\beta_2x_i^2+e_i.
$$

# Transforming the response

Transforming $Y$ changes the scale on which the whole regression model is fitted.

\pause

This can be useful when:

- the response grows multiplicatively rather than additively;
- residual spread increases with the fitted mean;
- large response values dominate the fit;
- the transformed-scale errors are more plausibly symmetric and homoscedastic.

\pause

It also changes every relationship with every explanatory variable.

# Example: a log response model

```{r}
#| fig-width: 7.4
#| fig-height: 4.2
#| out-width: "82%"
old_par <- par(no.readonly = TRUE)
make_plot_window()
plot(
  log_y_tbl$x,
  log_y_tbl$y,
  pch = 19,
  cex = 0.7,
  col = "grey35",
  xlab = "x",
  ylab = "y",
  main = "Fitting on the original scale"
)
lines(log_y_grid$x, log_y_grid$raw_fit, lwd = 2.2, col = "#003366")
lines(log_y_grid$x, log_y_grid$log_fit_original_scale, lwd = 2.2, col = "#990000")
legend(
  "topleft",
  legend = c("linear fit to y", "log-response fit, back-transformed"),
  lwd = 2.2,
  col = c("#003366", "#990000"),
  bty = "n"
)
box()
par(old_par)
```

# The log-response model

The model is

$$
\log(Y_i)=\beta_0+\beta_1x_i+e_i.
$$

\pause

Equivalently,

$$
Y_i=\exp(\beta_0+\beta_1x_i+e_i).
$$

\pause

So the model is additive on the log scale, but multiplicative on the original scale.

# Interpreting a log-response coefficient

For

$$
\log(Y_i)=\beta_0+\beta_1x_i+e_i,
$$

a one-unit increase in $x$ changes the fitted value on the log scale by $\beta_1$.

\pause

On the original scale, it multiplies the fitted median by

$$
\exp(\beta_1).
$$

\pause

The corresponding percentage change is

$$
100\{\exp(\beta_1)-1\}\%.
$$

# Square-root response transformations

A square-root transformation is often considered for non-negative responses, especially counts.

\pause

For example,

$$
\sqrt{Y_i}=\beta_0+\beta_1x_i+e_i.
$$

\pause

Here $\beta_1$ is interpreted on the square-root scale.

\pause

There is usually no simple additive interpretation for the mean of $Y$ on the original scale.

# Be careful with automatic transformations

Transformations are modelling choices, not recipes.

\pause

In particular:

- logging $Y$ is not allowed when $Y\leq0$;
- using $\log(Y+1)$ changes the scale and depends on the units of measurement;
- back-transforming fitted values can change the estimand being described;
- transforming the response may help one problem while creating another.

# Transforming an explanatory variable

Sometimes only one relationship is curved.

\pause

Then transforming that explanatory variable may be more targeted than transforming the response.

\pause

For example,

$$
Y_i=\beta_0+\beta_1X_{1i}+\beta_2\log(X_{2i})+e_i.
$$

\pause

This changes the column in the design matrix, not the scale of the response.

# Example: one explanatory variable needs transformation

```{r}
#| fig-width: 9.2
#| fig-height: 4.1
#| out-width: "92%"
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
make_plot_window()
plot_residuals(fit_x_raw, x_trans_tbl$x2, expression(x[2]))
title("Raw x2", line = 0.3)
make_plot_window()
plot_residuals(fit_x_log, log(x_trans_tbl$x2), expression(log(x[2])))
title("Transformed x2", line = 0.3)
par(old_par)
```

Transforming $x_2$ directly targets the curved relationship involving $x_2$.

# Comparing the two fitted models {.smaller}

```{r}
#| results: asis
comparison_tbl <- data.frame(
  Model = c("Original explanatory variable", "Transformed explanatory variable"),
  Formula = c("y ~ x1 + x2", "y ~ x1 + log(x2)"),
  `Residual standard error` = c(
    summary(fit_x_raw)$sigma,
    summary(fit_x_log)$sigma
  ),
  `Adjusted R-squared` = c(
    summary(fit_x_raw)$adj.r.squared,
    summary(fit_x_log)$adj.r.squared
  ),
  check.names = FALSE
)
knitr::kable(comparison_tbl, format = "latex", digits = 3)
```

\pause

These summaries support the graphical diagnosis, but they do not replace model checking.

# Interpreting a log explanatory variable

For

$$
E(Y\mid X=x)=\beta_0+\beta_1\log(x),
$$

$\beta_1$ is the change in the mean response for a one-unit increase in $\log(x)$.

\pause

Equivalently, multiplying $x$ by $c$ changes the mean response by

$$
\beta_1\log(c).
$$

\pause

So doubling $x$ changes the mean response by

$$
\beta_1\log(2).
$$

# Common explanatory-variable transformations {.smaller}

```{r}
#| results: asis
trans_tbl <- data.frame(
  Term = c("x", "log(x)", "sqrt(x)", "x^2", "exp(x)"),
  `When it may help` = c(
    "approximately straight relationship",
    "effect decreases as x grows",
    "effect decreases, but less sharply than log(x)",
    "simple curvature",
    "response grows roughly exponentially in x"
  ),
  `Interpretation of coefficient` = c(
    "change per one-unit increase in x",
    "change when x is multiplied",
    "change per one-unit increase in sqrt(x)",
    "not usually interpreted alone if x is also included",
    "change per one-unit increase in exp(x)"
  ),
  check.names = FALSE
)
knitr::kable(trans_tbl, format = "latex", booktabs = TRUE)
```

# Polynomial terms

A polynomial model adds powers of the same explanatory variable:

$$
Y_i
=
\beta_0+
\beta_1x_i+
\beta_2x_i^2+
\cdots+
\beta_px_i^p+e_i.
$$

\pause

This is still a linear regression model because the coefficients enter linearly.

\pause

Usually, if $x^2$ is included, $x$ should also be included.

# Why include the lower-order term?

Compare the two model forms:

$$
Y_i=\beta_0+\beta_2x_i^2+e_i
$$

and

$$
Y_i=\beta_0+\beta_1x_i+\beta_2x_i^2+e_i.
$$

\pause

The first forces a very specific shape.

\pause

The second allows both a linear trend and curvature.

# Centring before polynomial terms

It is often useful to use

$$
z_i=x_i-\bar{x}
$$

and then fit

$$
Y_i=\beta_0+\beta_1z_i+\beta_2z_i^2+e_i.
$$

\pause

Centring can make the intercept and lower-order terms easier to interpret.

\pause

It can also reduce unnecessary correlation between the columns $z$ and $z^2$.

# Discretising a numerical explanatory variable

Another possibility is to split $x$ into intervals.

\pause

For four intervals, a model might have design-matrix row

$$
\begin{bmatrix}
1 & I(x\in I_2) & I(x\in I_3) & I(x\in I_4)
\end{bmatrix}.
$$

\pause

This estimates a separate mean for each interval, relative to the reference interval.

# Discretisation is usually crude

```{r}
#| fig-width: 7.4
#| fig-height: 4.2
#| out-width: "82%"
old_par <- par(no.readonly = TRUE)
make_plot_window()
plot(
  disc_tbl$x,
  disc_tbl$y,
  pch = 19,
  cex = 0.75,
  col = "grey35",
  xlab = "x",
  ylab = "y",
  main = "Discretising a continuous variable"
)
abline(fit_disc_linear, lwd = 2.2, col = "#003366")
ord <- order(disc_tbl$x)
stepfit <- fitted(fit_disc_cut)[ord]
lines(disc_tbl$x[ord], stepfit, lwd = 2.2, col = "#990000", type = "s")
legend(
  "topright",
  legend = c("straight line", "interval means"),
  lwd = 2.2,
  col = c("#003366", "#990000"),
  bty = "n"
)
box()
par(old_par)
```

# Problems with discretisation

Discretising a numerical explanatory variable:

- throws away the order and distance information inside each interval;
- creates artificial jumps at arbitrary cut-points;
- can reduce power;
- can make the fitted model depend strongly on chosen breaks.

\pause

It can be useful for communication or specific scientific thresholds, but it should not be the default solution to smooth non-linearity.

# How to choose a modification {.smaller}

\begin{center}
\begin{tabular}{ll}
\hline
Diagnostic pattern & More targeted response \\
\hline
Curvature involving one explanatory variable & Transform that variable or add polynomial terms \\
Variance increases with fitted mean & Consider transforming the response \\
Multiplicative response relationship & Consider a log response model \\
One group has a different slope & Add an interaction term \\
Several isolated thresholds are scientifically meaningful & Consider a discretised variable \\
\hline
\end{tabular}
\end{center}

# Practical workflow

After changing the model:

1. State the new model clearly.
2. Explain what changed in the design matrix.
3. Refit the model.
4. Re-check residuals and other diagnostics.
5. Interpret the coefficients on the correct scale.
6. Avoid presenting transformed-scale results as if they were original-scale results.

# Summary

- A transformation or additional term should answer a specific diagnostic problem.
- Transforming $Y$ changes the scale of the whole model.
- Transforming an explanatory variable changes one design-matrix column.
- Polynomial terms add columns while keeping the model linear in the coefficients.
- Discretisation is possible, but often discards useful information.
- Every modified model must be re-checked before it is trusted.
