---
title: "Regression Week V: Prediction"
subtitle: "Slide deck II: Prediction intervals, model selection and out-of-sample performance"
author: "Miguel Rodo"
institute: University of Cape Town
date: today
format:
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: preamble.tex
    keep-tex: true
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 4.5,
  out.width = "0.88\\linewidth"
)

set.seed(502)

if (requireNamespace("cowplot", quietly = TRUE)) {
  theme_reg <- function(font_size = 14) {
    cowplot::theme_cowplot(font_size = font_size) +
      ggplot2::theme(
        plot.background = ggplot2::element_rect(fill = "white", colour = NA),
        panel.background = ggplot2::element_rect(fill = "white", colour = NA)
      )
  }
} else {
  theme_reg <- function(font_size = 14) {
    ggplot2::theme_minimal(base_size = font_size) +
      ggplot2::theme(
        plot.background = ggplot2::element_rect(fill = "white", colour = NA),
        panel.background = ggplot2::element_rect(fill = "white", colour = NA)
      )
  }
}

format_number <- function(x, digits = 3) {
  format(signif(x, digits), trim = TRUE, scientific = FALSE)
}

make_interval_plot <- function(n, seed = 1) {
  set.seed(seed)
  x_vec <- runif(n, 0, 10)
  y_vec <- 4 + 2 * x_vec + rnorm(n, 0, 2)
  example_tbl <- tibble::tibble(x = x_vec, y = y_vec)
  fit <- lm(y ~ x, data = example_tbl)
  x_seq <- seq(min(x_vec), max(x_vec), length.out = 500)
  new_tbl <- tibble::tibble(x = x_seq)
  pred_int <- predict(fit, newdata = new_tbl, interval = "prediction")
  mean_int <- predict(fit, newdata = new_tbl, interval = "confidence")
  plot_tbl <- dplyr::bind_rows(
    tibble::tibble(x = x_seq, lower = mean_int[, "lwr"], upper = mean_int[, "upr"], type = "Mean response"),
    tibble::tibble(x = x_seq, lower = pred_int[, "lwr"], upper = pred_int[, "upr"], type = "New observation")
  )
  list(data = example_tbl, fit = fit, plot_tbl = plot_tbl)
}

small_interval <- make_interval_plot(10, 10)
large_interval <- make_interval_plot(200, 10)

set.seed(510)
n_leverage <- 60
x_lev <- sort(runif(n_leverage, 0, 10))
y_lev <- 5 + 1.5 * x_lev + rnorm(n_leverage, sd = 2)
lev_tbl <- tibble::tibble(x = x_lev, y = y_lev)
lev_fit <- lm(y ~ x, data = lev_tbl)
lev_grid <- tibble::tibble(x = seq(-3, 14, length.out = 400))
lev_pred <- predict(lev_fit, newdata = lev_grid, interval = "prediction")
lev_conf <- predict(lev_fit, newdata = lev_grid, interval = "confidence")
lev_grid <- lev_grid |>
  dplyr::mutate(
    pred_lower = lev_pred[, "lwr"],
    pred_upper = lev_pred[, "upr"],
    mean_lower = lev_conf[, "lwr"],
    mean_upper = lev_conf[, "upr"]
  )

set.seed(520)
n_select <- 160
p_select <- 20
rho <- 0.45
latent <- rnorm(n_select)
X_select <- sapply(seq_len(p_select), function(j) {
  rho * latent + sqrt(1 - rho^2) * rnorm(n_select)
})
colnames(X_select) <- paste0("X", seq_len(p_select))
beta_select <- c(1.2, -1.1, 0.9, 0.7, rep(0, p_select - 4))
y_select <- as.vector(X_select %*% beta_select + rnorm(n_select, sd = 2.2))
selection_tbl <- tibble::as_tibble(data.frame(y = y_select, X_select))

fit_null <- lm(y ~ 1, data = selection_tbl)
fit_full <- lm(y ~ ., data = selection_tbl)
fit_forward <- stats::step(
  fit_null,
  scope = list(lower = fit_null, upper = fit_full),
  direction = "forward",
  trace = 0
)
fit_backward <- stats::step(
  fit_full,
  direction = "backward",
  trace = 0
)
fit_both <- stats::step(
  fit_null,
  scope = list(lower = fit_null, upper = fit_full),
  direction = "both",
  trace = 0
)

true_terms <- paste0("X", 1:4)
selection_summary <- function(fit, label) {
  terms <- setdiff(names(coef(fit)), "(Intercept)")
  tibble::tibble(
    Method = label,
    Included = length(terms),
    `Correctly included` = sum(terms %in% true_terms),
    `Noise variables included` = sum(!(terms %in% true_terms)),
    AIC = AIC(fit),
    BIC = BIC(fit),
    `Training MSE` = mean(residuals(fit)^2)
  )
}
selection_results <- dplyr::bind_rows(
  selection_summary(fit_full, "Full"),
  selection_summary(fit_forward, "Forward"),
  selection_summary(fit_backward, "Backward"),
  selection_summary(fit_both, "Stepwise")
) |>
  dplyr::mutate(
    dplyr::across(tidyselect::where(is.numeric), ~ signif(.x, 3))
  )

set.seed(530)
n_null <- 150
p_null <- 80
X_null <- matrix(rnorm(n_null * p_null), nrow = n_null)
colnames(X_null) <- paste0("X", seq_len(p_null))
y_null <- rnorm(n_null)
null_tbl <- tibble::as_tibble(data.frame(y = y_null, X_null))
null_fit <- lm(y ~ 1, data = null_tbl)
null_full <- lm(y ~ ., data = null_tbl)
null_forward <- stats::step(
  null_fit,
  scope = list(lower = null_fit, upper = null_full),
  direction = "forward",
  trace = 0
)
null_selected <- length(setdiff(names(coef(null_forward)), "(Intercept)"))
null_train_mse <- mean(residuals(null_forward)^2)

set.seed(531)
train_index <- sample(seq_len(n_null), size = floor(0.75 * n_null))
train_tbl <- null_tbl[train_index, ]
test_tbl <- null_tbl[-train_index, ]
train_null <- lm(y ~ 1, data = train_tbl)
train_full <- lm(y ~ ., data = train_tbl)
train_forward <- stats::step(
  train_null,
  scope = list(lower = train_null, upper = train_full),
  direction = "forward",
  trace = 0
)
train_mse <- mean(residuals(train_forward)^2)
test_pred <- predict(train_forward, newdata = test_tbl)
test_mse <- mean((test_tbl$y - test_pred)^2)
selected_terms_train <- length(setdiff(names(coef(train_forward)), "(Intercept)"))

set.seed(532)
n_reps <- 100
split_results <- replicate(n_reps, {
  idx <- sample(seq_len(n_null), size = floor(0.75 * n_null))
  tr <- null_tbl[idx, ]
  te <- null_tbl[-idx, ]
  mod <- stats::step(
    lm(y ~ 1, data = tr),
    scope = list(lower = ~1, upper = ~.),
    direction = "forward",
    trace = 0
  )
  c(
    selected = length(setdiff(names(coef(mod)), "(Intercept)")),
    train_mse = mean(residuals(mod)^2),
    test_mse = mean((te$y - predict(mod, newdata = te))^2)
  )
})
split_results_tbl <- tibble::as_tibble(t(split_results)) |>
  dplyr::mutate(split = dplyr::row_number())
split_summary_tbl <- split_results_tbl |>
  dplyr::summarise(
    `Average selected variables` = mean(selected),
    `Average training MSE` = mean(train_mse),
    `Average test MSE` = mean(test_mse)
  ) |>
  dplyr::mutate(dplyr::across(tidyselect::where(is.numeric), ~ signif(.x, 3)))
```

# From inference to prediction

- In Week III, we used the fitted model to quantify uncertainty about unknown regression quantities.

\pause

- In Week IV, we asked whether the fitted model is believable enough to use.

\pause

- We now ask a practical question:

\pause

> If a new case arrives, what range of responses should we expect?

\pause

- That question leads naturally from prediction intervals to model selection and out-of-sample performance.

# Lecture outline

By the end of this deck, you should be able to:

- Distinguish uncertainty in a mean response from uncertainty in a new observation
- Derive and interpret a prediction interval in the general linear model
- Explain how prediction uncertainty depends on predictor location
- Explain why training error is usually too optimistic
- Describe forward, backward and stepwise variable selection
- Compare models using out-of-sample prediction error

# Set-up and notation {.smaller}

We continue to use the normal general linear model
$$
Y=X\boldsymbol\beta+\mathbf e,
\qquad
\mathbf e\sim N_n(\mathbf 0,\sigma^2I_n),
$$
where $X$ is an $n\times k$ design matrix with full column rank.

\pause

Let
$$
C=(X^TX)^{-1},
\qquad
s^2=\frac{RSS}{n-k}.
$$

\pause

For a new case, write its predictor vector as $\mathbf x_0$ and its response as
$$
Y_0=\mathbf x_0^T\boldsymbol\beta+e_0,
\qquad
 e_0\sim N(0,\sigma^2).
$$

\pause

We assume that $e_0$ is independent of the errors in the training data.

# Two different prediction questions

At the same predictor vector $\mathbf x_0$, there are two different targets.

\pause

1. The **mean response**:
$$
\mu_0=E(Y_0\mid \mathbf x_0)=\mathbf x_0^T\boldsymbol\beta.
$$

\pause

2. A **new individual response**:
$$
Y_0=\mathbf x_0^T\boldsymbol\beta+e_0.
$$

\pause

The second target is more variable because it includes the new random error $e_0$.

# Week III recap: interval for the mean response

The fitted mean response at $\mathbf x_0$ is
$$
\hat\mu_0=\mathbf x_0^T\hat{\boldsymbol\beta}.
$$

\pause

Because
$$
\hat{\boldsymbol\beta}\sim
N_k(\boldsymbol\beta,\sigma^2C),
$$
we have
$$
\hat\mu_0
\sim
N\left(\mu_0,\sigma^2\mathbf x_0^TC\mathbf x_0\right).
$$

\pause

A $100(1-\alpha)\%$ confidence interval for the mean response is therefore
$$
\boxed{
\hat\mu_0
\pm
t_{1-\alpha/2,n-k}
s\sqrt{\mathbf x_0^TC\mathbf x_0}
}.
$$

# Prediction error for a new observation

For a new response, the prediction error is
$$
Y_0-\hat\mu_0.
$$

\pause

Using
$$
Y_0=\mathbf x_0^T\boldsymbol\beta+e_0,
\qquad
\hat\mu_0=\mathbf x_0^T\hat{\boldsymbol\beta},
$$
we get
$$
Y_0-\hat\mu_0
=
e_0-\mathbf x_0^T(\hat{\boldsymbol\beta}-\boldsymbol\beta).
$$

\pause

So there are two sources of prediction error:

- the new random error $e_0$;
- uncertainty in estimating $\boldsymbol\beta$.

# Variance of the prediction error

The new error $e_0$ is independent of the training sample.

\pause

Therefore it is independent of $\hat{\boldsymbol\beta}$, and
$$
\operatorname{Var}(Y_0-\hat\mu_0)
=
\operatorname{Var}(e_0)
+
\operatorname{Var}\{\mathbf x_0^T(\hat{\boldsymbol\beta}-\boldsymbol\beta)\}.
$$

\pause

That gives
$$
\operatorname{Var}(Y_0-\hat\mu_0)
=
\sigma^2+
\sigma^2\mathbf x_0^TC\mathbf x_0
=
\sigma^2\left(1+\mathbf x_0^TC\mathbf x_0\right).
$$

\pause

The extra $1$ is the irreducible variation in the future observation itself.

# Prediction interval

Since the model is normal,
$$
Y_0-\hat\mu_0
\sim
N\left(0,\sigma^2\{1+\mathbf x_0^TC\mathbf x_0\}\right).
$$

\pause

Replacing $\sigma$ by $s$ gives the pivotal quantity
$$
\frac{Y_0-\hat\mu_0}{s\sqrt{1+\mathbf x_0^TC\mathbf x_0}}
\sim
t_{n-k}.
$$

\pause

A $100(1-\alpha)\%$ prediction interval is
$$
\boxed{
\hat\mu_0
\pm
t_{1-\alpha/2,n-k}
s\sqrt{1+\mathbf x_0^TC\mathbf x_0}
}.
$$

# Mean-response interval versus prediction interval

Compare the two standard errors:
$$
\text{Mean response:}
\qquad
s\sqrt{\mathbf x_0^TC\mathbf x_0},
$$

$$
\text{New observation:}
\qquad
s\sqrt{1+\mathbf x_0^TC\mathbf x_0}.
$$

\pause

The prediction interval is wider because we are trying to predict an individual future value, not just the mean of many such future values.

\pause

Even with a very large sample, a future observation still has random error.

# Example with ten observations

```{r}
plot_tbl <- small_interval$plot_tbl
example_tbl <- small_interval$data

ggplot2::ggplot(example_tbl, ggplot2::aes(x = x, y = y)) +
  theme_reg() +
  ggplot2::geom_point(colour = "dodgerblue", alpha = 0.85, size = 2.2) +
  ggplot2::geom_smooth(method = "lm", se = FALSE, colour = "orange", linewidth = 1.1) +
  ggplot2::geom_ribbon(
    data = dplyr::filter(plot_tbl, type == "Mean response"),
    ggplot2::aes(x = x, ymin = lower, ymax = upper),
    inherit.aes = FALSE,
    fill = "#66c2a5",
    alpha = 0.45
  ) +
  ggplot2::geom_line(
    data = dplyr::filter(plot_tbl, type == "New observation"),
    ggplot2::aes(x = x, y = lower),
    inherit.aes = FALSE,
    colour = "#e78ac3",
    linetype = "11",
    linewidth = 1
  ) +
  ggplot2::geom_line(
    data = dplyr::filter(plot_tbl, type == "New observation"),
    ggplot2::aes(x = x, y = upper),
    inherit.aes = FALSE,
    colour = "#e78ac3",
    linetype = "11",
    linewidth = 1
  ) +
  ggplot2::labs(x = "Predictor", y = "Response")
```

# Example with 200 observations

```{r}
plot_tbl <- large_interval$plot_tbl
example_tbl <- large_interval$data

ggplot2::ggplot(example_tbl, ggplot2::aes(x = x, y = y)) +
  theme_reg() +
  ggplot2::geom_point(colour = "dodgerblue", alpha = 0.25, size = 1.5) +
  ggplot2::geom_smooth(method = "lm", se = FALSE, colour = "orange", linewidth = 1.1) +
  ggplot2::geom_ribbon(
    data = dplyr::filter(plot_tbl, type == "Mean response"),
    ggplot2::aes(x = x, ymin = lower, ymax = upper),
    inherit.aes = FALSE,
    fill = "#66c2a5",
    alpha = 0.45
  ) +
  ggplot2::geom_line(
    data = dplyr::filter(plot_tbl, type == "New observation"),
    ggplot2::aes(x = x, y = lower),
    inherit.aes = FALSE,
    colour = "#e78ac3",
    linetype = "11",
    linewidth = 1
  ) +
  ggplot2::geom_line(
    data = dplyr::filter(plot_tbl, type == "New observation"),
    ggplot2::aes(x = x, y = upper),
    inherit.aes = FALSE,
    colour = "#e78ac3",
    linetype = "11",
    linewidth = 1
  ) +
  ggplot2::labs(x = "Predictor", y = "Response")
```

# What changed when $n$ increased?

- The confidence interval for the mean response became much narrower.

\pause

- The prediction interval narrowed much less.

\pause

- With increasing information at a well-supported predictor value,
$$
\mathbf x_0^TC\mathbf x_0
$$
tends to get smaller.

\pause

- But the prediction interval still contains
$$
1+\mathbf x_0^TC\mathbf x_0,
$$
so it cannot eliminate the new response's own random error.

# Prediction uncertainty depends on where we predict

Write
$$
h_0=\mathbf x_0^TC\mathbf x_0.
$$

\pause

Then
$$
\text{SE for mean response}=s\sqrt{h_0},
\qquad
\text{SE for new response}=s\sqrt{1+h_0}.
$$

\pause

The value $h_0$ is larger when $\mathbf x_0$ is unusual relative to the observed design matrix.

\pause

This is the same idea as leverage from model checking.

# Simple-regression intuition

In simple linear regression, with an intercept,
$$
h_0
=
\frac{1}{n}
+
\frac{(x_0-\bar{x})^2}{\sum_{i=1}^n(x_i-\bar{x})^2}.
$$

\pause

So predictions are most precise near $\bar{x}$.

\pause

Predictions become less precise as $x_0$ moves away from the centre of the observed predictor values.

\pause

This is why extrapolation is dangerous even when the fitted line looks reasonable inside the observed range.

# Interpolation and extrapolation

```{r}
ggplot2::ggplot(lev_tbl, ggplot2::aes(x = x, y = y)) +
  theme_reg() +
  ggplot2::geom_rect(
    ggplot2::aes(xmin = min(lev_tbl$x), xmax = max(lev_tbl$x), ymin = -Inf, ymax = Inf),
    inherit.aes = FALSE,
    fill = "grey90",
    alpha = 0.45
  ) +
  ggplot2::geom_point(colour = "dodgerblue", alpha = 0.8, size = 1.9) +
  ggplot2::geom_smooth(method = "lm", se = FALSE, colour = "orange", linewidth = 1) +
  ggplot2::geom_ribbon(
    data = lev_grid,
    ggplot2::aes(x = x, ymin = pred_lower, ymax = pred_upper),
    inherit.aes = FALSE,
    fill = "#e78ac3",
    alpha = 0.25
  ) +
  ggplot2::geom_ribbon(
    data = lev_grid,
    ggplot2::aes(x = x, ymin = mean_lower, ymax = mean_upper),
    inherit.aes = FALSE,
    fill = "#66c2a5",
    alpha = 0.45
  ) +
  ggplot2::labs(x = "Predictor", y = "Response")
```

# Prediction intervals in `R`

Once the model is fitted, `R` can produce either kind of interval.

```{r}
#| echo: true
fitted_model <- lm(y ~ x, data = lev_tbl)
new_observations <- data.frame(x = c(2, 8, 12))

predict(
  fitted_model,
  newdata = new_observations,
  interval = "prediction"
)
```

\pause

Use `interval = "confidence"` for a mean-response interval and `interval = "prediction"` for a new-observation interval.

# How to report a prediction interval

A suitable sentence is:

\pause

> For a case with predictor value $x_0=8$, the predicted response is 19.0, with a 95% prediction interval from 14.6 to 23.4.

\pause

That means:

- the fitted model predicts a value near 19.0;
- a future individual response could plausibly be anywhere in the interval;
- the interpretation relies on the regression model being appropriate.

\pause

It is not a statement that the unknown fixed mean response has 95% probability of lying in the interval.

# Prediction intervals depend on model assumptions

Prediction intervals inherit all the usual regression assumptions:

- linear mean structure;
- independent errors;
- constant error variance;
- normality if we want exact finite-sample intervals;
- no major extrapolation or influential model failures.

\pause

Model checking is therefore not optional before prediction.

\pause

A precise interval from a badly specified model can be precisely wrong.

# Why select variables?

Prediction usually starts with a set of possible predictors.

\pause

We may want fewer predictors because:

- a smaller model can be easier to interpret;
- some predictors may be expensive or difficult to measure;
- irrelevant predictors can increase prediction variance;
- a simpler model can sometimes predict better on new data.

\pause

The goal is not merely to fit the current data well.

The goal is to predict future data well.

# The candidate-model problem

With $p$ possible predictors, there are
$$
2^p
$$
possible subsets.

\pause

For example,
$$
2^{20}=1\,048\,576,
\qquad
2^{100}\approx 1.27\times 10^{30}.
$$

\pause

Exhaustive best-subset selection can therefore become computationally difficult.

\pause

Greedy procedures search only part of the model space.

# Best-subset search versus stepwise search

**Best-subset selection** tries to find the best model among all candidate subsets under a chosen criterion.

\pause

**Stepwise procedures** follow a sequence of local moves:

- add one variable;
- remove one variable;
- or alternate between adding and removing.

\pause

They are computationally convenient, but they are not guaranteed to find the globally best subset.

# Criteria for comparing models {.smaller}

Common criteria include:

- $R^2$: increases whenever a predictor is added, so it is not useful for selecting model size;
- adjusted $R^2$: penalises unnecessary predictors and should be maximised;
- AIC: balances fit and complexity, usually with a lighter penalty;
- BIC: balances fit and complexity, usually with a stronger penalty.

\pause

For comparable Gaussian linear models, up to an additive constant,
$$
\operatorname{AIC}
=
n\log(RSS/n)+2q,
$$
$$
\operatorname{BIC}
=
n\log(RSS/n)+q\log n,
$$
where $q$ is the number of estimated model parameters.

\pause

AIC and BIC are minimised.

# Forward selection

Forward selection starts with the intercept-only model.

\pause

At each step:

1. Try adding each variable not yet in the model.
2. Choose the addition that improves the criterion most.
3. Stop when no allowed addition improves the criterion.

\pause

It is useful when the full model is large, but it can miss variables that only help after other variables have entered.

# Backward and stepwise selection

**Backward elimination** starts with the full model.

\pause

At each step, it removes the variable whose removal gives the best criterion value.

\pause

This requires the full model to be estimable, so it can fail if there are too many predictors or exact collinearity.

\pause

**Bidirectional stepwise selection** allows both additions and removals.

\pause

It remains a greedy search, but permits earlier decisions to be partially reversed.

# A moderate simulation

We simulate 20 predictors.

\pause

Only the first four predictors truly affect the response.

\pause

The predictors are moderately correlated, and the noise level is large enough that selection is not completely trivial.

\pause

The question is:

> Which methods keep the signal variables without adding too many noise variables?

# Selection results

```{r}
knitr::kable(selection_results)
```

# What the table shows

- The full model includes all variables, including many noise variables.

\pause

- A stepwise method may remove many irrelevant variables.

\pause

- But the result depends on the data, the starting model, and the selection criterion.

\pause

- The selected model should not be treated as if it had been fixed before seeing the data.

# A warning about post-selection inference

The inference decks gave confidence intervals and hypothesis tests for a fixed model.

\pause

After data-driven model selection, the model is no longer fixed in the same sense.

\pause

Consequently:

- ordinary $p$-values after selection can be too optimistic;
- ordinary confidence intervals can have worse coverage than advertised;
- prediction intervals condition on the selected model and may not include all selection uncertainty.

\pause

This does not make model selection useless, but it changes how cautiously we should interpret the output.

# Training error is optimistic

Suppose the response is pure noise:
$$
Y_i\sim N(0,1),
$$
and none of the predictors truly matters.

\pause

If we search over many predictors, some will look useful by chance.

\pause

In one simulation, forward selection included `r null_selected` variables.

\pause

The training MSE was
$$
`r format_number(null_train_mse)`.
$$

\pause

That is much smaller than the irreducible variance of 1, but it is an artefact of using the same data to select and evaluate the model.

# Training and test data

To estimate predictive performance, separate the roles of the data.

\pause

- **Training data:** fit the model and perform variable selection.

\pause

- **Test data:** evaluate the already chosen modelling procedure on data not used for fitting.

\pause

The test data must remain untouched until the final evaluation.

\pause

Otherwise the test error also becomes optimistic.

# One training-test split

```{r}
tibble::tibble(
  `Selected variables` = selected_terms_train,
  `Training MSE` = signif(train_mse, 3),
  `Test MSE` = signif(test_mse, 3)
) |>
  knitr::kable()
```

\pause

The test error is much closer to the real prediction error than the training error.

\pause

It can also be greater than 1 because the selected model adds estimation and selection noise.

# Repeated splitting

One split can be noisy.

\pause

Repeating the split gives a more stable picture of the same optimism.

```{r}
knitr::kable(split_summary_tbl)
```

\pause

The average test MSE is much larger than the average training MSE.

# Visualising optimism

```{r}
plot_split_tbl <- split_results_tbl |>
  dplyr::select(split, `Training MSE` = train_mse, `Test MSE` = test_mse) |>
  tidyr::pivot_longer(-split, names_to = "Error type", values_to = "MSE")

ggplot2::ggplot(plot_split_tbl, ggplot2::aes(x = `Error type`, y = MSE, fill = `Error type`)) +
  theme_reg() +
  ggplot2::geom_boxplot(alpha = 0.75, outlier.alpha = 0.45) +
  ggplot2::geom_hline(yintercept = 1, linetype = "dotted") +
  ggplot2::guides(fill = "none") +
  ggplot2::labs(x = NULL, y = "Mean squared error")
```

# Cross-validation

A training-test split estimates performance using only part of the data for testing.

\pause

Cross-validation repeats this idea more systematically.

\pause

For example, in $K$-fold cross-validation:

1. Split the data into $K$ parts.
2. Fit the modelling procedure on $K-1$ parts.
3. Evaluate it on the held-out part.
4. Repeat until every part has been held out once.
5. Average the held-out errors.

\pause

The important point is that selection must be repeated inside each training fold.

# A modelling procedure, not just a model

When variable selection is involved, what we evaluate is the whole procedure:

\pause

1. Start with these candidate predictors.
2. Use this selection rule.
3. Fit the selected model.
4. Predict new observations.

\pause

Cross-validation or test-error estimation must include all four steps.

\pause

If the selected variables are chosen before cross-validation, the error estimate can still be biased downward.

# Summary

- A confidence interval for a mean response quantifies uncertainty about $E(Y_0\mid\mathbf x_0)$.

\pause

- A prediction interval for a new observation adds irreducible future-observation variation.

\pause

- Prediction intervals are wider, especially because of the extra $1$ in
$$
1+\mathbf x_0^TC\mathbf x_0.
$$

\pause

- Prediction uncertainty increases for unusual predictor vectors and extrapolation.

\pause

- Variable selection can improve simplicity and sometimes prediction, but it introduces extra uncertainty.

\pause

- Training error is optimistic, so predictive performance should be estimated using held-out data or cross-validation.

# What to take into your reports

When reporting prediction from a regression model:

- say whether the interval is for a mean response or a new observation;
- report the fitted value and interval endpoints;
- avoid extrapolating without warning;
- evaluate prediction performance on data not used to fit or select the model;
- be cautious about ordinary inference after variable selection.
