---
title: "Regression Week V: Matching the model to the data"
subtitle: "Slide deck II: Interactions"
author: "Miguel Rodo"
institute: University of Cape Town
date: today
bibliography: zotero.bib
format:
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: ../preamble.tex
    keep-tex: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

format_number <- function(x, digits = 3) {
  format(signif(x, digits), trim = TRUE, scientific = FALSE)
}

make_plot_window <- function() {
  par(
    mar = c(4.0, 4.2, 2.2, 0.8),
    mgp = c(2.3, 0.7, 0),
    tcl = -0.25,
    las = 1
  )
}

linear_combination_test <- function(fit, ell, value = 0) {
  ell <- matrix(ell, ncol = 1)
  beta_hat <- matrix(coef(fit), ncol = 1)
  estimate <- as.numeric(t(ell) %*% beta_hat)
  se <- sqrt(as.numeric(t(ell) %*% vcov(fit) %*% ell))
  t_stat <- (estimate - value) / se
  p_value <- 2 * pt(abs(t_stat), df = df.residual(fit), lower.tail = FALSE)
  data.frame(
    Estimate = estimate,
    `Std. error` = se,
    `t statistic` = t_stat,
    `p-value` = p_value,
    check.names = FALSE
  )
}

set.seed(551)
n <- 140
interaction_tbl <- data.frame(
  x = runif(n, 0, 10),
  group = factor(sample(c("A", "B"), n, replace = TRUE))
)
interaction_tbl$y <- ifelse(
  interaction_tbl$group == "A",
  3.0 + 0.45 * interaction_tbl$x,
  2.0 + 1.35 * interaction_tbl$x
) + rnorm(n, sd = 1.0)

fit_additive <- lm(y ~ x + group, data = interaction_tbl)
fit_interaction <- lm(y ~ x * group, data = interaction_tbl)
interaction_grid <- expand.grid(
  x = seq(min(interaction_tbl$x), max(interaction_tbl$x), length.out = 100),
  group = levels(interaction_tbl$group)
)
interaction_grid$additive <- predict(fit_additive, newdata = interaction_grid)
interaction_grid$interaction <- predict(fit_interaction, newdata = interaction_grid)

coef_interaction <- coef(fit_interaction)
x_bar <- mean(interaction_tbl$x)
x0 <- 6

set.seed(552)
three_tbl <- data.frame(
  x = runif(180, 0, 10),
  group = factor(sample(c("A", "B", "C"), 180, replace = TRUE))
)
three_tbl$y <- with(
  three_tbl,
  ifelse(group == "A", 4 + 0.35 * x,
    ifelse(group == "B", 3 + 1.00 * x, 5 + 1.45 * x)
  ) + rnorm(180, sd = 1.2)
)
fit_three <- lm(y ~ x * group, data = three_tbl)
```

# Why interactions matter

A regression coefficient is often described as the effect of an explanatory variable while holding the others fixed.

\pause

An interaction asks whether that effect itself depends on another explanatory variable.

\pause

For example:

- Does the association between study time and test score differ by tutorial group?
- Does the association between dose and response differ by treatment arm?
- Does the difference between two groups change as age changes?

# Lecture outline

By the end of this deck, you should be able to:

- Explain what an interaction means in a regression model
- Construct the design matrix for a continuous-by-categorical interaction
- Derive group-specific fitted lines
- Interpret main effects in the presence of an interaction
- Test whether slopes differ between groups
- Test group differences at a specified value of $X$
- Extend the ideas to several factor levels and other interaction types

# Example: different fitted slopes by group

```{r}
#| fig-width: 7.4
#| fig-height: 4.2
#| out-width: "82%"
old_par <- par(no.readonly = TRUE)
make_plot_window()
plot(
  interaction_tbl$x,
  interaction_tbl$y,
  pch = ifelse(interaction_tbl$group == "A", 19, 17),
  col = ifelse(interaction_tbl$group == "A", "#003366", "#990000"),
  cex = 0.75,
  xlab = "x",
  ylab = "y",
  main = "The slope appears to depend on group"
)
lines(
  interaction_grid$x[interaction_grid$group == "A"],
  interaction_grid$interaction[interaction_grid$group == "A"],
  lwd = 2.2,
  col = "#003366"
)
lines(
  interaction_grid$x[interaction_grid$group == "B"],
  interaction_grid$interaction[interaction_grid$group == "B"],
  lwd = 2.2,
  col = "#990000"
)
legend(
  "topleft",
  legend = c("Group A", "Group B"),
  pch = c(19, 17),
  col = c("#003366", "#990000"),
  bty = "n"
)
box()
par(old_par)
```

# Additive model versus interaction model

An additive model says the group difference is constant:

$$
E(Y\mid x,G)=\beta_0+\beta_Xx+\beta_GI(G=B).
$$

\pause

An interaction model says the slope can also differ by group:

$$
E(Y\mid x,G)=\beta_0+\beta_Xx+\beta_GI(G=B)+\beta_{X:G}xI(G=B).
$$

# Visual comparison

```{r}
#| fig-width: 9.2
#| fig-height: 4.2
#| out-width: "92%"
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
make_plot_window()
plot(
  interaction_tbl$x,
  interaction_tbl$y,
  pch = ifelse(interaction_tbl$group == "A", 19, 17),
  col = ifelse(interaction_tbl$group == "A", "#003366", "#990000"),
  cex = 0.65,
  xlab = "x",
  ylab = "y",
  main = "Additive model"
)
for (g in levels(interaction_tbl$group)) {
  idx <- interaction_grid$group == g
  lines(
    interaction_grid$x[idx],
    interaction_grid$additive[idx],
    lwd = 2.2,
    col = ifelse(g == "A", "#003366", "#990000")
  )
}
box()
make_plot_window()
plot(
  interaction_tbl$x,
  interaction_tbl$y,
  pch = ifelse(interaction_tbl$group == "A", 19, 17),
  col = ifelse(interaction_tbl$group == "A", "#003366", "#990000"),
  cex = 0.65,
  xlab = "x",
  ylab = "y",
  main = "Interaction model"
)
for (g in levels(interaction_tbl$group)) {
  idx <- interaction_grid$group == g
  lines(
    interaction_grid$x[idx],
    interaction_grid$interaction[idx],
    lwd = 2.2,
    col = ifelse(g == "A", "#003366", "#990000")
  )
}
box()
par(old_par)
```

# Model formula and R syntax

For a two-level factor with reference group A:

$$
Y_i
=
\beta_0+eta_Xx_i+eta_GI(G_i=B)+
\beta_{X:G}x_iI(G_i=B)+e_i.
$$

\pause

In R:

```{r}
#| echo: true
fit_interaction <- lm(y ~ x * group, data = interaction_tbl)
```

\pause

The formula `x * group` expands to `x + group + x:group`.

# Design-matrix rows {.smaller}

```{r}
#| results: asis
row_A <- which(interaction_tbl$group == "A")[1]
row_B <- which(interaction_tbl$group == "B")[1]
X_int <- model.matrix(fit_interaction)[c(row_A, row_B), ]
rownames(X_int) <- c("one group A row", "one group B row")
knitr::kable(X_int, format = "latex", digits = 3)
```

\pause

The interaction column is the product of $x$ and the group-B dummy variable.

# Group-specific fitted lines

For group A,

$$
E(Y\mid x,A)=\beta_0+\beta_Xx.
$$

\pause

For group B,

$$
\begin{aligned}
E(Y\mid x,B)
&=\beta_0+\beta_Xx+\beta_G+\beta_{X:G}x\\
&=(\beta_0+\beta_G)+(\beta_X+\beta_{X:G})x.
\end{aligned}
$$

# Intercepts and slopes {.smaller}

\begin{center}
\begin{tabular}{lll}
\hline
Quantity & Group A & Group B \\
\hline
Intercept & $\beta_0$ & $\beta_0+\beta_G$ \\
Slope in $x$ & $\beta_X$ & $\beta_X+\beta_{X:G}$ \\
\hline
\end{tabular}
\end{center}

\pause

So $\beta_{X:G}$ is the difference between the group-B slope and the group-A slope.

# Estimated coefficients {.smaller}

```{r}
#| results: asis
coef_tbl <- data.frame(
  Term = names(coef_interaction),
  Estimate = unname(coef_interaction),
  `Std. error` = summary(fit_interaction)$coefficients[, "Std. Error"],
  `p-value` = summary(fit_interaction)$coefficients[, "Pr(>|t|)"],
  check.names = FALSE
)
knitr::kable(coef_tbl, format = "latex", digits = 3)
```

\pause

These individual coefficient tests answer only some of the useful interaction questions.

# Interpreting the fitted slopes

For the fitted model:

$$
\widehat{\text{slope for group A}}
=
\hat\beta_X
=
`r format_number(coef_interaction["x"])`.
$$

\pause

$$
\widehat{\text{slope for group B}}
=
\hat\beta_X+\hat\beta_{X:G}
=
`r format_number(coef_interaction["x"] + coef_interaction["x:groupB"])`.
$$

\pause

The difference in slopes is

$$
\hat\beta_{X:G}
=
`r format_number(coef_interaction["x:groupB"])`.
$$

# Main effects in the presence of an interaction

In this model, $\beta_G$ is not the overall difference between groups.

\pause

It is the difference between groups when $x=0$:

$$
E(Y\mid x=0,B)-E(Y\mid x=0,A)=\beta_G.
$$

\pause

If $x=0$ is not meaningful, then the group coefficient may not answer a useful scientific question.

# Centre the continuous variable when needed

Let

$$
z_i=x_i-x_0.
$$

Then fit

$$
Y_i=\gamma_0+\gamma_Xz_i+\gamma_GI(G_i=B)+\gamma_{X:G}z_iI(G_i=B)+e_i.
$$

\pause

Now $\gamma_G$ is the group difference at $x=x_0$.

\pause

Common choices for $x_0$ include the sample mean or a scientifically meaningful value.

# Difference between groups at a specified value

At $X=x_0$, the fitted group-B minus group-A difference is

$$
\{\beta_0+\beta_G+(\beta_X+\beta_{X:G})x_0\}
-
\{\beta_0+\beta_Xx_0\}.
$$

\pause

Therefore,

$$
E(Y\mid x_0,B)-E(Y\mid x_0,A)
=
\beta_G+x_0\beta_{X:G}.
$$

# Test: do the slopes differ?

The slopes differ if

$$
\beta_{X:G}\ne0.
$$

\pause

So the test is

$$
H_0:\beta_{X:G}=0
\qquad\text{against}\qquad
H_1:\beta_{X:G}\ne0.
$$

\pause

For a two-level factor, this is the ordinary $t$ test for the interaction coefficient.

# Test: is the slope non-zero in group B?

The group-B slope is

$$
\beta_X+\beta_{X:G}.
$$

\pause

So we test

$$
H_0:\beta_X+\beta_{X:G}=0
\qquad\text{against}\qquad
H_1:\beta_X+\beta_{X:G}\ne0.
$$

\pause

This is a linear-combination test with

$$
\ell'\boldsymbol\beta
=
\begin{bmatrix}0&1&0&1\end{bmatrix}
\boldsymbol\beta.
$$

# Linear-combination test result {.smaller}

```{r}
#| results: asis
test_slope_B <- linear_combination_test(fit_interaction, c(0, 1, 0, 1))
knitr::kable(test_slope_B, format = "latex", digits = 4)
```

\pause

This uses the same linear-combination inference developed in Week III.

# Test: do groups differ at $x_0$?

At $x_0=`r format_number(x0, 2)`$, the group difference is

$$
\beta_G+x_0\beta_{X:G}.
$$

\pause

The null hypothesis is

$$
H_0:\beta_G+x_0\beta_{X:G}=0.
$$

\pause

This is tested using

$$
\ell'
=
\begin{bmatrix}0&0&1&x_0\end{bmatrix}.
$$

# Group-difference test result {.smaller}

```{r}
#| results: asis
test_diff_x0 <- linear_combination_test(fit_interaction, c(0, 0, 1, x0))
knitr::kable(test_diff_x0, format = "latex", digits = 4)
```

\pause

The test depends on the value of $x_0$ because the group difference changes with $x$.

# Several-level categorical variables {.smaller}

For an $a$-level factor with reference level 1:

$$
\begin{aligned}
E(Y\mid X,G)
&=
\beta_0+eta_XX\\
&\quad+
\sum_{j=2}^a\beta_{G_j}I(G=j)
+
\sum_{j=2}^a\beta_{X:G_j}XI(G=j).
\end{aligned}
$$

\pause

For group $j$:

$$
\text{intercept}=\beta_0+\beta_{G_j},
\qquad
\text{slope}=\beta_X+\beta_{X:G_j}.
$$

# Fitted lines for three groups

```{r}
#| fig-width: 7.4
#| fig-height: 4.2
#| out-width: "82%"
old_par <- par(no.readonly = TRUE)
make_plot_window()
cols <- c("#003366", "#990000", "#006633")
plot(
  three_tbl$x,
  three_tbl$y,
  pch = as.numeric(three_tbl$group) + 14,
  col = cols[as.numeric(three_tbl$group)],
  cex = 0.7,
  xlab = "x",
  ylab = "y",
  main = "Three group-specific fitted lines"
)
for (g in levels(three_tbl$group)) {
  x_seq <- seq(min(three_tbl$x), max(three_tbl$x), length.out = 100)
  new_dat <- data.frame(x = x_seq, group = factor(g, levels = levels(three_tbl$group)))
  lines(x_seq, predict(fit_three, newdata = new_dat), lwd = 2.2, col = cols[which(levels(three_tbl$group) == g)])
}
legend(
  "topleft",
  legend = paste("Group", levels(three_tbl$group)),
  pch = 15:17,
  col = cols,
  bty = "n"
)
box()
par(old_par)
```

# Tests with several groups {.smaller}

Common questions become linear restrictions:

\begin{center}
\begin{tabular}{ll}
\hline
Question & Null hypothesis \\
\hline
Does group $j$ have a different slope from reference? & $\beta_{X:G_j}=0$ \\
Do any slopes differ across groups? & $\beta_{X:G_2}=\cdots=\beta_{X:G_a}=0$ \\
Is the slope non-zero in group $j$? & $\beta_X+\beta_{X:G_j}=0$ \\
Does group $j$ differ from reference at $x_0$? & $\beta_{G_j}+x_0\beta_{X:G_j}=0$ \\
Does factor $G$ affect the fitted line? & all group and interaction terms are zero \\
\hline
\end{tabular}
\end{center}

# Joint tests use the $F$ test

When the null hypothesis contains several restrictions, use an $F$ test.

\pause

For example, in the three-level case, testing whether any slope differs by group uses

$$
H_0:\beta_{X:G_B}=\beta_{X:G_C}=0.
$$

\pause

Equivalently, compare the additive model to the interaction model.

# Reduced-model comparison in R {.smaller}

```{r}
#| echo: true
fit_three_additive <- lm(y ~ x + group, data = three_tbl)
fit_three_interaction <- lm(y ~ x * group, data = three_tbl)
anova(fit_three_additive, fit_three_interaction)
```

\pause

This is the same full-versus-reduced-model logic used in earlier hypothesis-testing and ANOVA slides.

# Other interaction types

The same idea extends beyond continuous-by-categorical interactions.

\pause

- Categorical by categorical:

$$
Y\sim A*B.
$$

The difference between levels of one factor may depend on the level of the other factor.

\pause

- Continuous by continuous:

$$
Y_i=\beta_0+\beta_1x_i+\beta_2z_i+\beta_3x_iz_i+e_i.
$$

The slope in $x$ is $\beta_1+\beta_3z_i$.

# Interaction workflow

When fitting an interaction model:

1. Plot the fitted relationship separately by group or value of the modifier.
2. Write the model explicitly.
3. Derive the fitted mean in each group or condition.
4. Interpret main effects only at their reference values.
5. Translate each substantive question into a linear restriction.
6. Use a $t$ test for one restriction and an $F$ test for several restrictions.

# Summary

- An interaction means that the effect of one explanatory variable depends on another.
- In a continuous-by-categorical interaction, each group can have its own fitted line.
- Main effects are conditional effects at reference values.
- Group differences depend on $x$ when slopes differ.
- Most useful interaction questions are linear-combination or joint-restriction tests.
