---
title: "Regression Week IV: Model checking"
subtitle: "Slide deck II: Model checking"
author: "Yovna Junglee, edited by Miguel Rodo"
institute: University of Cape Town
date: today
format:
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: ../preamble.tex
    keep-tex: true
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

set.seed(402)

format_number <- function(x, digits = 3) {
  format(signif(x, digits), trim = TRUE, scientific = FALSE)
}

# Polynomial example -------------------------------------------------------
n_poly <- 100
poly_tbl <- data.frame(
  x = seq(-2, 2, length.out = n_poly)
)
poly_tbl$mean <- 0.25 + 0.75 * poly_tbl$x - 0.45 * poly_tbl$x^2 +
  0.40 * poly_tbl$x^3 + 0.55 * sin(3.2 * poly_tbl$x)
poly_tbl$y <- poly_tbl$mean + rnorm(n_poly, sd = 0.45)

poly_fit <- function(degree) {
  lm(y ~ poly(x, degree, raw = TRUE), data = poly_tbl)
}
poly_degrees <- c(1, 2, 3, 10)
poly_fits <- lapply(poly_degrees, poly_fit)
poly_r2_tbl <- data.frame(
  Degree = poly_degrees,
  `R squared` = vapply(poly_fits, function(f) summary(f)$r.squared, numeric(1)),
  `Adjusted R squared` = vapply(poly_fits, function(f) summary(f)$adj.r.squared, numeric(1)),
  check.names = FALSE
)

# Diagnostic example -------------------------------------------------------
set.seed(421)
n_diag <- 48
p_diag <- 8
X_diag <- matrix(rnorm(n_diag * p_diag), nrow = n_diag, ncol = p_diag)
colnames(X_diag) <- paste0("X", seq_len(p_diag))

# Create a few high-leverage observations and outlying responses.
X_diag[1, ] <- c(4.5, 3.8, 3.2, 2.7, 2.4, 2.0, 1.7, 1.5)
X_diag[23, ] <- c(3.0, -3.1, 2.9, -2.5, 2.1, -1.8, 1.4, -1.2)
X_diag[32, ] <- c(5.5, -4.9, 4.6, -4.2, 3.7, -3.0, 2.2, -1.5)

lin_pred <- 80 + 18 * X_diag[, 1] - 13 * X_diag[, 2] + 7 * X_diag[, 3]
err_sd <- 7 + 0.015 * pmax(lin_pred - mean(lin_pred), 0)^2
Y_diag <- lin_pred + rnorm(n_diag, sd = err_sd)
Y_diag[2] <- Y_diag[2] + 42
Y_diag[10] <- Y_diag[10] - 35
Y_diag[32] <- Y_diag[32] + 70

bank_tbl <- data.frame(Y = Y_diag, X_diag)
fit_bank <- lm(Y ~ ., data = bank_tbl)
fit_bank_summary <- summary(fit_bank)

ehat <- residuals(fit_bank)
fitted_bank <- fitted(fit_bank)
hat_diag <- hatvalues(fit_bank)
stand_resid <- rstandard(fit_bank)
stud_resid <- rstudent(fit_bank)
cook_d <- cooks.distance(fit_bank)
n <- nobs(fit_bank)
k <- length(coef(fit_bank))
df_resid <- df.residual(fit_bank)
lev_cut <- 2 * k / n
cook_cut <- 4 / (n - k - 1)
mod_cook <- sqrt((n - k) / k * hat_diag / (1 - hat_diag)) * abs(stud_resid)

shapiro_res <- shapiro.test(ehat)
ks_res <- suppressWarnings(ks.test(ehat, "pnorm", mean(ehat), sd(ehat)))
normality_tbl <- data.frame(
  Test = c("Shapiro-Wilk", "Kolmogorov-Smirnov"),
  Statistic = c(unname(shapiro_res$statistic), unname(ks_res$statistic)),
  `p-value` = c(shapiro_res$p.value, ks_res$p.value),
  Comment = c(
    "Useful as a rough formal check, but sensitive to sample size.",
    "Not standard when mean and variance are estimated from the residuals."
  ),
  check.names = FALSE
)

plot_theme <- function() {
  par(
    mar = c(4.0, 4.2, 2.1, 1.0),
    mgp = c(2.4, 0.7, 0),
    tcl = -0.25,
    las = 1
  )
}
```

# So far...

1. Formulated the normal general linear model.
2. Estimated the regression parameters by least squares.
3. Built confidence intervals.
4. Tested hypotheses and compared models using ANOVA.


Next: **Model checking.**

# Lecture outline

By the end of this deck, you should be able to:

- Explain what $R^2$ and adjusted $R^2$ measure
- State the main assumptions checked after fitting a linear model
- Use residual plots to diagnose common model problems
- Distinguish outliers, high-leverage points and influential observations
- Calculate and interpret standardised residuals, Studentised residuals, leverages and Cook-type diagnostics

# Why model checking matters

- Least squares will produce estimates whenever $X^TX$ is invertible.

\pause

- That does not mean the fitted model is a good description of the data-generating process.

\pause

- Inference from the model depends on assumptions about the errors.

\pause

- Model checking asks whether those assumptions look plausible enough for the analysis we want to do.

# A motivating example: more terms can fit better

```{r}
#| fig-width: 7.6
#| fig-height: 3.8
#| out-width: "88%"
old_par <- par(no.readonly = TRUE)
plot_theme()
plot(
  poly_tbl$x, poly_tbl$y,
  pch = 16,
  col = grDevices::adjustcolor("grey35", alpha.f = 0.65),
  xlab = "x",
  ylab = "y",
  main = "One data set with a non-linear pattern"
)
lines(poly_tbl$x, poly_tbl$mean, lwd = 2.2, col = "#003366")
legend(
  "topleft",
  legend = c("Observed data", "Underlying mean pattern"),
  pch = c(16, NA),
  lty = c(NA, 1),
  lwd = c(NA, 2.2),
  col = c("grey35", "#003366"),
  bty = "n"
)
par(old_par)
```

# Polynomial fits to the same data {.smaller}

```{r}
#| fig-width: 8.5
#| fig-height: 5.0
#| out-width: "94%"
old_par <- par(no.readonly = TRUE)
par(
  mfrow = c(2, 2),
  mar = c(3.2, 3.5, 2.2, 1.0),
  mgp = c(2.0, 0.6, 0),
  tcl = -0.25,
  las = 1
)
for (i in seq_along(poly_degrees)) {
  degree <- poly_degrees[i]
  fit <- poly_fits[[i]]
  plot(
    poly_tbl$x, poly_tbl$y,
    pch = 16,
    cex = 0.55,
    col = grDevices::adjustcolor("grey35", alpha.f = 0.55),
    xlab = "x",
    ylab = "y",
    main = bquote(y == beta[0] + cdots + beta[.(degree)] * x^.(degree))
  )
  lines(poly_tbl$x, fitted(fit), lwd = 2.2, col = "#003366")
  lines(poly_tbl$x, poly_tbl$mean, lwd = 1.6, col = "#B00020", lty = 2)
}
par(old_par)
```

# Coefficient of determination

For a model with an intercept, define
$$
SST=\sum_{i=1}^n (Y_i-\bar Y)^2,
\qquad
SSE=\sum_{i=1}^n (Y_i-\hat Y_i)^2,
\qquad
SSR=\sum_{i=1}^n (\hat Y_i-\bar Y)^2.
$$

\pause

Then
$$
SST = SSR + SSE,
$$
and the coefficient of determination is
$$
R^2
=
\frac{SSR}{SST}
=
1-\frac{SSE}{SST}.
$$

# Interpreting $R^2$

- $0\le R^2\le 1$ for least-squares models with an intercept.

\pause

- $R^2$ is the proportion of the sample variation in $Y$ explained by the fitted regression on the chosen predictors.

\pause

- A larger $R^2$ means a smaller residual sum of squares relative to the total variation.

\pause

- It does not, by itself, prove that the model is appropriate.

# Why $R^2$ is not enough

Adding predictors to a least-squares model cannot increase $SSE$.

\pause

Therefore, adding predictors cannot decrease $R^2$.

\pause

This can make a larger model look better even if the additional terms are not useful for prediction or interpretation.

\pause

The issue is especially clear when we add many polynomial terms to the same data.

# $R^2$ in the polynomial example

```{r}
#| results: asis
poly_print_tbl <- poly_r2_tbl
poly_print_tbl$`R squared` <- format_number(poly_print_tbl$`R squared`, 4)
poly_print_tbl$`Adjusted R squared` <- format_number(poly_print_tbl$`Adjusted R squared`, 4)
knitr::kable(poly_print_tbl, align = "rcc")
```

\pause

The degree 10 polynomial fits the data very closely, but the curve is more complicated than the lower-degree fits.

# Adjusted $R^2$

Adjusted $R^2$ penalises the use of additional predictors:
$$
R^2_{\mathrm{adj}}
=
1-
\frac{SSE/(n-k)}{SST/(n-1)}
=
1-(1-R^2)\frac{n-1}{n-k},
$$
where $k$ is the number of regression coefficients, including the intercept.

\pause

Key points:

- $R^2_{\mathrm{adj}}\le R^2$.
- $R^2_{\mathrm{adj}}$ can decrease when a weak predictor is added.
- It is a useful summary, but not a substitute for scientific judgement or diagnostic plots.

# Example model summaries {.smaller}

For the polynomial data, compare the linear model with the degree 10 model.

```{r}
#| results: asis
summ_tbl <- data.frame(
  Model = c("Linear", "Degree 10 polynomial"),
  `Residual df` = c(df.residual(poly_fits[[1]]), df.residual(poly_fits[[4]])),
  `Residual SE` = c(summary(poly_fits[[1]])$sigma, summary(poly_fits[[4]])$sigma),
  `R squared` = c(summary(poly_fits[[1]])$r.squared, summary(poly_fits[[4]])$r.squared),
  `Adjusted R squared` = c(summary(poly_fits[[1]])$adj.r.squared, summary(poly_fits[[4]])$adj.r.squared),
  check.names = FALSE
)
summ_tbl[, 3:5] <- lapply(summ_tbl[, 3:5], format_number, digits = 4)
knitr::kable(summ_tbl)
```

\pause

A good-looking summary statistic is only the start of model checking.

# Model assumptions

In the normal general linear model, conditional on the design matrix $X$,
$$
Y=X\boldsymbol\beta+\mathbf e,
\qquad
\mathbf e\sim N_n(\mathbf 0,\sigma^2I_n).
$$

\pause

This combines four assumptions:

1. $E(\mathbf e\mid X)=\mathbf 0$.
2. $\operatorname{Var}(e_i\mid X)=\sigma^2$ for all $i$.
3. $\operatorname{Cov}(e_i,e_j\mid X)=0$ for $i\ne j$.
4. The errors are normally distributed.

# What residual plots check

We cannot observe the errors $e_i$ directly.

\pause

We therefore inspect the fitted residuals
$$
\hat e_i=Y_i-\hat Y_i.
$$

\pause

Common checks:

- residuals versus fitted values for non-linearity and unequal variance
- histogram or QQ plot for normality
- residuals versus time or order for dependence
- leverage and influence diagnostics for unusual observations

# Residuals versus fitted values {.smaller}

```{r}
#| fig-width: 5.3
#| fig-height: 4.2
#| out-width: "62%"
old_par <- par(no.readonly = TRUE)
plot_theme()
plot(
  fitted_bank, ehat,
  pch = 16,
  col = grDevices::adjustcolor("grey35", alpha.f = 0.75),
  xlab = "Fitted values",
  ylab = "Residuals",
  main = "Residuals versus fitted values"
)
abline(h = 0, lty = 3, col = "grey45")
lines(lowess(fitted_bank, ehat), lwd = 2.0, col = "#B00020")
text(fitted_bank, ehat, labels = seq_along(ehat), pos = 3, cex = 0.55)
par(old_par)
```

# Reading the residuals versus fitted plot

Look for:

- a roughly horizontal band around zero
- no clear curve in the centre of the residual cloud
- roughly constant vertical spread across fitted values
- no extreme isolated points

\pause

Warning signs:

- curvature suggests the mean structure may be wrong
- increasing or decreasing spread suggests non-constant variance
- isolated large residuals suggest possible outliers

# Histogram of residuals

```{r}
#| fig-width: 5.2
#| fig-height: 4.0
#| out-width: "60%"
old_par <- par(no.readonly = TRUE)
plot_theme()
hist(
  ehat,
  breaks = "FD",
  freq = FALSE,
  col = "grey85",
  border = "grey35",
  xlab = "Residuals",
  main = "Histogram of residuals"
)
curve(dnorm(x, mean(ehat), sd(ehat)), add = TRUE, lwd = 2.0, col = "#003366")
par(old_par)
```

\pause

A histogram is useful for seeing skewness, heavy tails and isolated residuals, but it is sensitive to bin choice.

# Normal QQ plot

```{r}
#| fig-width: 5.2
#| fig-height: 4.0
#| out-width: "60%"
old_par <- par(no.readonly = TRUE)
plot_theme()
qqnorm(ehat, pch = 16, cex = 0.75, main = "Normal QQ plot of residuals")
qqline(ehat, lwd = 2.0, col = "#003366")
text(qqnorm(ehat, plot.it = FALSE)$x, sort(ehat), labels = order(ehat), pos = 3, cex = 0.55)
par(old_par)
```

\pause

If the errors are approximately normal, the residuals should lie close to a straight line.

# Formal tests for normality {.smaller}

A common null hypothesis is
$$
H_0:\quad \text{the errors are normally distributed.}
$$

```{r}
#| results: asis
normality_print_tbl <- normality_tbl
normality_print_tbl$Statistic <- format_number(normality_print_tbl$Statistic, 4)
normality_print_tbl$`p-value` <- format_number(normality_print_tbl$`p-value`, 4)
knitr::kable(normality_print_tbl)
```

\pause

Formal tests should not replace plots. In particular, a Kolmogorov-Smirnov test against a normal distribution with parameters estimated from the same residuals does not have the usual null distribution.

# Outliers and influential observations

\begin{block}{Definitions}

- An **outlier** is an observation that is not well fitted by the model.
- A **high-leverage observation** has an unusual predictor vector.
- An **influential observation** has a relatively large effect on fitted values, coefficients or inferences.

\end{block}

\pause

These are different ideas. A point can have high leverage without being influential, and an outlier can have low leverage.

# Estimated residuals {.smaller}

Let
$$
H=X(X^TX)^{-1}X^T,
\qquad
M=I_n-H.
$$

Then
$$
\hat{\mathbf e}
=
\mathbf Y-\hat{\mathbf Y}
=
\mathbf Y-X\hat{\boldsymbol\beta}
=\{I_n-X(X^TX)^{-1}X^T\}\mathbf Y
=M\mathbf Y.
$$

Since $MX=0$ and $\mathbf Y=X\boldsymbol\beta+\mathbf e$,
$$
\hat{\mathbf e}=M\mathbf e.
$$

# Covariance matrix of the residuals {.smaller}

The matrices $H$ and $M$ are symmetric and idempotent:
$$
H^T=H,
\qquad
H^2=H,
\qquad
M^T=M,
\qquad
M^2=M.
$$

\pause

Because $\operatorname{Var}(\mathbf e\mid X)=\sigma^2I_n$,
$$
\operatorname{Var}(\hat{\mathbf e}\mid X)
=
M(\sigma^2I_n)M^T
=
\sigma^2M.
$$

\pause

Therefore residuals are generally correlated, and
$$
\operatorname{Var}(\hat e_i\mid X)=\sigma^2(1-h_i),
$$
where $h_i$ is the $i$th diagonal element of $H$.

# Internally standardised residuals

If $\sigma^2$ were known, we could use
$$
z_i=
\frac{\hat e_i}{\sigma\sqrt{1-h_i}}
$$
to scale residuals by their own standard deviations.

\pause

Since $\sigma^2$ is unknown, use
$$
r_i=
\frac{\hat e_i}{s\sqrt{1-h_i}},
\qquad
s^2=\frac{\hat{\mathbf e}^T\hat{\mathbf e}}{n-k}.
$$

\pause

The $r_i$ are internally standardised residuals. Values with $|r_i|>2$ are often treated as possible outliers needing investigation.

# Externally Studentised residuals

Let $\hat{\boldsymbol\beta}_{(i)}$ and $s^2_{(i)}$ be the least-squares estimate and residual variance estimate after deleting observation $i$.

\pause

The externally Studentised residual is
$$
t_i
=
\frac{Y_i-x_i^T\hat{\boldsymbol\beta}_{(i)}}
{s_{(i)}\sqrt{1+x_i^T(X_{(i)}^TX_{(i)})^{-1}x_i}}.
$$

\pause

Under the normal linear model,
$$
t_i\sim t_{n-k-1}.
$$

# Useful computational form

It can be shown that
$$
t_i
=
\frac{\hat e_i}{s_{(i)}\sqrt{1-h_i}}.
$$

\pause

It can also be shown that
$$
(n-k-1)s^2_{(i)}
=
(n-k)s^2-
\frac{\hat e_i^2}{1-h_i}.
$$

\pause

Combining these with
$$
r_i^2=\frac{\hat e_i^2}{s^2(1-h_i)}
$$
gives
$$
t_i
=
r_i
\sqrt{\frac{n-k-1}{n-k-r_i^2}}.
$$

# Why externally Studentised residuals are useful

- They compare observation $i$ with a model fitted without observation $i$.

\pause

- This prevents the observation from helping to hide its own unusualness.

\pause

- They have an exact $t_{n-k-1}$ distribution under the normal linear model.

\pause

- In practice, $|t_i|>2$ is a rough screening rule, while formal testing should account for checking many observations.

# Standardised and Studentised residual plots

```{r}
#| fig-width: 8.5
#| fig-height: 4.2
#| out-width: "92%"
old_par <- par(no.readonly = TRUE)
par(
  mfrow = c(1, 2),
  mar = c(4.0, 4.2, 2.4, 0.8),
  mgp = c(2.4, 0.7, 0),
  tcl = -0.25,
  las = 1
)
plot(
  seq_along(stand_resid), stand_resid,
  pch = 16,
  cex = 0.7,
  xlab = "Index",
  ylab = "Standardised residuals",
  main = "Internally standardised"
)
abline(h = c(-2, 2), lty = 2, col = "#B00020")
abline(h = 0, lty = 3, col = "grey45")
text(seq_along(stand_resid), stand_resid, labels = seq_along(stand_resid), pos = 3, cex = 0.55)
plot(
  seq_along(stud_resid), stud_resid,
  pch = 16,
  cex = 0.7,
  xlab = "Index",
  ylab = "Studentised residuals",
  main = "Externally Studentised"
)
abline(h = c(-2, 2), lty = 2, col = "#B00020")
abline(h = 0, lty = 3, col = "grey45")
text(seq_along(stud_resid), stud_resid, labels = seq_along(stud_resid), pos = 3, cex = 0.55)
par(old_par)
```

# Leverage

The fitted values can be written as
$$
\hat{\mathbf Y}=H\mathbf Y,
\qquad
H=X(X^TX)^{-1}X^T.
$$

\pause

The diagonal values
$$
h_i=x_i^T(X^TX)^{-1}x_i
$$
are called leverages or hat values.

\pause

Since $H$ is idempotent and symmetric,
$$
\operatorname{tr}(H)
=
\operatorname{tr}\{X(X^TX)^{-1}X^T\}
=
\operatorname{tr}(I_k)
=k.
$$
Thus
$$
\sum_{i=1}^n h_i=k,
\qquad
\overline h = \frac{k}{n}.
$$

# High-leverage observations

A common screening rule is
$$
h_i\ge \frac{2k}{n}.
$$

\pause

This identifies observations with unusual predictor values.

\pause

Correction to remember:

- High leverage does not automatically mean high influence.
- Influence depends on both leverage and how poorly the observation is fitted.

# Leverage plot

```{r}
#| fig-width: 5.6
#| fig-height: 4.2
#| out-width: "64%"
old_par <- par(no.readonly = TRUE)
plot_theme()
plot(
  seq_along(hat_diag), hat_diag,
  pch = 16,
  cex = 0.75,
  xlab = "Index",
  ylab = "Leverage",
  main = "Leverage values"
)
abline(h = lev_cut, lty = 2, col = "#B00020")
text(seq_along(hat_diag), hat_diag, labels = seq_along(hat_diag), pos = 3, cex = 0.55)
legend(
  "topright",
  legend = bquote(2*k/n == .(format_number(lev_cut, 3))),
  lty = 2,
  col = "#B00020",
  bty = "n"
)
par(old_par)
```

# Cook's distance

Cook's distance measures the change in the fitted model when observation $i$ is deleted.

\pause

The usual Cook's distance can be written as
$$
D_i
=
\frac{(\hat{\boldsymbol\beta}_{(i)}-\hat{\boldsymbol\beta})^T
X^TX
(\hat{\boldsymbol\beta}_{(i)}-\hat{\boldsymbol\beta})}{ks^2}
=
\left(\frac{r_i^2}{k}\right)\frac{h_i}{1-h_i}.
$$

\pause

It combines:

- residual size, through $r_i^2$
- leverage, through $h_i/(1-h_i)$

# Modified Cook statistic

The course-preferred version replaces $s^2$ by the deleted variance estimate $s^2_{(i)}$:
$$
D_i^{\ast}
=
\frac{(\hat{\boldsymbol\beta}_{(i)}-\hat{\boldsymbol\beta})^T
X^TX
(\hat{\boldsymbol\beta}_{(i)}-\hat{\boldsymbol\beta})}{ks^2_{(i)}}
=
\left(\frac{t_i^2}{k}\right)\frac{h_i}{1-h_i}.
$$

\pause

Atkinson's modified Cook statistic is
$$
A_i
=
\left\{\frac{n-k}{k}\frac{h_i}{1-h_i}\right\}^{1/2}|t_i|
=
\sqrt{(n-k)D_i^{\ast}}.
$$

Proposed cut-off values are often 2 or 3.

# Cook's distance plot

```{r}
#| fig-width: 5.6
#| fig-height: 4.2
#| out-width: "64%"
old_par <- par(no.readonly = TRUE)
plot_theme()
plot(
  seq_along(cook_d), cook_d,
  type = "h",
  lwd = 2.0,
  xlab = "Index",
  ylab = "Cook's distance",
  main = "Cook's distance"
)
points(seq_along(cook_d), cook_d, pch = 16, cex = 0.75)
abline(h = cook_cut, lty = 2, col = "#B00020")
text(seq_along(cook_d), cook_d, labels = seq_along(cook_d), pos = 3, cex = 0.55)
legend(
  "topright",
  legend = bquote(4/(n-k-1) == .(format_number(cook_cut, 3))),
  lty = 2,
  col = "#B00020",
  bty = "n"
)
par(old_par)
```

# Leverage and residual size together

```{r}
#| fig-width: 5.8
#| fig-height: 4.2
#| out-width: "66%"
old_par <- par(no.readonly = TRUE)
plot_theme()
plot(
  hat_diag, abs(stud_resid),
  pch = 16,
  cex = 0.8,
  xlab = "Leverage",
  ylab = expression(abs(t[i])),
  main = "Leverage and Studentised residuals"
)
abline(v = lev_cut, lty = 2, col = "#B00020")
abline(h = 2, lty = 2, col = "#B00020")
text(hat_diag, abs(stud_resid), labels = seq_along(hat_diag), pos = 3, cex = 0.55)
par(old_par)
```

\pause

The most concerning observations tend to be those with both high leverage and large Studentised residuals.

# Modified Cook statistic plot

```{r}
#| fig-width: 5.6
#| fig-height: 4.2
#| out-width: "64%"
old_par <- par(no.readonly = TRUE)
plot_theme()
plot(
  seq_along(mod_cook), mod_cook,
  type = "h",
  lwd = 2.0,
  xlab = "Index",
  ylab = "Modified Cook statistic",
  main = "Atkinson's modified Cook statistic"
)
points(seq_along(mod_cook), mod_cook, pch = 16, cex = 0.75)
abline(h = c(2, 3), lty = 2, col = "#B00020")
text(seq_along(mod_cook), mod_cook, labels = seq_along(mod_cook), pos = 3, cex = 0.55)
par(old_par)
```

# Summary

Model checking is not a single test.

\pause

Use several pieces of evidence together:

- $R^2$ and adjusted $R^2$ describe fit, but do not validate the model.
- Residual plots check the mean structure, variance pattern and possible outliers.
- QQ plots and normality tests check the normal-error assumption.
- Leverage identifies unusual predictor vectors.
- Studentised residuals and Cook-type statistics help identify observations worth investigating.

# Corrections relative to the original PDF

- Avoided treating larger $R^2$ as automatically better.
- Corrected the distinction between high leverage and influence.
- Used conditional model assumptions given $X$.
- Corrected the deleted-case residual variance notation.
- Corrected the Sherman-Morrison-Woodbury dimensions by not relying on the misstated version in the slides.
- Treated the Kolmogorov-Smirnov normality test with caution when normal parameters are estimated from residuals.
