---
title: "Regression Week IV: ANOVA and model checking"
subtitle: "Slide deck I: ANOVA"
author: "Yovna Junglee, edited by Miguel Rodo"
institute: University of Cape Town
date: today
format:
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: ../preamble.tex
    keep-tex: true
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

set.seed(1811)
n_group <- 10
group_levels <- c("Placebo", "Treatment A", "Treatment B")
trt_tbl <- tibble::tibble(
  group = factor(rep(group_levels, each = n_group), levels = group_levels),
  bp = c(
    rnorm(n_group, mean = 124, sd = 8),
    rnorm(n_group, mean = 108, sd = 8),
    rnorm(n_group, mean = 119, sd = 8)
  )
)

fit_oneway <- lm(bp ~ group, data = trt_tbl)
fit_reduced <- lm(bp ~ 1, data = trt_tbl)

n <- nrow(trt_tbl)
a <- nlevels(trt_tbl$group)
X_oneway <- model.matrix(fit_oneway)
k_oneway <- ncol(X_oneway)
C_oneway <- solve(crossprod(X_oneway))
s2_oneway <- sigma(fit_oneway)^2

trt_means <- aggregate(bp ~ group, data = trt_tbl, FUN = mean)
n_i <- as.numeric(table(trt_tbl$group))
y_bar <- mean(trt_tbl$bp)
y_i_bar <- trt_means$bp
SST <- sum((trt_tbl$bp - y_bar)^2)
SSA <- sum(n_i * (y_i_bar - y_bar)^2)
SSE <- sum(residuals(fit_oneway)^2)
MSA <- SSA / (a - 1)
MSE <- SSE / (n - a)
F_anova <- MSA / MSE
p_anova <- pf(F_anova, df1 = a - 1, df2 = n - a, lower.tail = FALSE)

anova_tbl <- anova(fit_oneway)
model_comparison_tbl <- anova(fit_reduced, fit_oneway)

L_group <- rbind(
  c(0, 1, 0),
  c(0, 0, 1)
)
r_group <- c(0, 0)
joint_difference <- as.vector(L_group %*% coef(fit_oneway) - r_group)
F_group <- drop(
  t(joint_difference) %*%
    solve(L_group %*% C_oneway %*% t(L_group)) %*%
    joint_difference /
    (nrow(L_group) * s2_oneway)
)
p_group <- pf(F_group, df1 = nrow(L_group), df2 = df.residual(fit_oneway), lower.tail = FALSE)

trt_aug <- trt_tbl
trt_aug$obs <- ave(trt_aug$bp, trt_aug$group, FUN = seq_along)
trt_aug$group_index <- as.numeric(trt_aug$group)
trt_aug$group_mean <- ave(trt_aug$bp, trt_aug$group, FUN = mean)
trt_aug$grand_mean <- y_bar

set.seed(201)
ad_tbl <- tibble::tibble(
  spend_radio = runif(100, 0, 3)
)
ad_tbl$spend_tv <- 1.2 * ad_tbl$spend_radio + runif(100, 0, 3)
ad_tbl$revenue <- 10 + 2 * ad_tbl$spend_tv + rnorm(100, sd = 2)

fit_tv_first <- lm(revenue ~ spend_tv + spend_radio, data = ad_tbl)
fit_radio_first <- lm(revenue ~ spend_radio + spend_tv, data = ad_tbl)
fit_ad_full <- lm(revenue ~ spend_tv + spend_radio, data = ad_tbl)
fit_no_tv <- lm(revenue ~ spend_radio, data = ad_tbl)
fit_no_radio <- lm(revenue ~ spend_tv, data = ad_tbl)

df_ad <- df.residual(fit_ad_full)
coef_ad <- summary(fit_ad_full)$coefficients

tv_partial <- anova(fit_no_tv, fit_ad_full)
radio_partial <- anova(fit_no_radio, fit_ad_full)

format_number <- function(x, digits = 3) {
  format(signif(x, digits), trim = TRUE, scientific = FALSE)
}
format_p <- function(p) {
  if (p < 0.001) "<0.001" else paste0("=", format_number(p))
}
```

# From joint tests to ANOVA

- We have already derived $t$ tests for one linear restriction.

\pause

- We have also derived $F$ tests for several simultaneous linear restrictions.

\pause

- ANOVA gives an important special case of the same idea.

\pause

- Its distinctive contribution is to express the test through a **decomposition of variation**.

# Lecture outline

By the end of this deck, you should be able to:

- State the one-way ANOVA model and equal-means hypothesis
- Decompose total variation into between-group and within-group variation
- Construct and interpret the one-way ANOVA $F$ statistic
- Explain how one-way ANOVA is a regression model with a categorical predictor
- Translate the equal-means hypothesis into a joint regression restriction
- Distinguish sequential and partial sums of squares in regression ANOVA tables

# The treatment example

Suppose individuals are given one of three blood-pressure treatments:

- placebo
- treatment A
- treatment B

\pause

The response is systolic blood pressure after treatment.

\pause

A typical question is:

> Is there evidence that mean blood pressure differs across the treatment groups?

# Blood pressure by treatment group

```{r}
#| fig-width: 7.6
#| fig-height: 4.0
#| out-width: "86%"
ggplot2::ggplot(trt_tbl, ggplot2::aes(x = group, y = bp)) +
  ggplot2::geom_boxplot(width = 0.45, outlier.shape = NA) +
  ggplot2::geom_point(
    ggplot2::aes(shape = group),
    position = ggplot2::position_jitter(width = 0.08, height = 0),
    size = 2.2,
    alpha = 0.75
  ) +
  ggplot2::stat_summary(fun = mean, geom = "point", size = 4) +
  ggplot2::labs(
    x = "Treatment group",
    y = "Blood pressure",
    caption = "Large points indicate group means."
  ) +
  ggplot2::theme_minimal(base_size = 15) +
  ggplot2::theme(legend.position = "none")
```

# The cell-means model

For group $i$ and observation $j$, write
$$
Y_{ij}=\mu_i+e_{ij},
\qquad
 i=1,\ldots,a,
 \quad
 j=1,\ldots,n_i.
$$

\pause

Here:

- $\mu_i$ is the population mean for group $i$;
- $e_{ij}$ is the random error for observation $j$ in group $i$;
- $a$ is the number of groups;
- $n=\sum_{i=1}^a n_i$ is the total sample size.

# The equal-means hypothesis

The one-way ANOVA null hypothesis is
$$
H_0:\mu_1=\mu_2=\cdots=\mu_a.
$$

\pause

The alternative is
$$
H_1:\text{not all group means are equal}.
$$

\pause

For the treatment example,
$$
H_0:
\mu_{\mathrm{Placebo}}
=
\mu_{\mathrm{A}}
=
\mu_{\mathrm{B}}.
$$

\pause

The alternative does **not** say that every pair of means differs.

# Model assumptions

For the exact finite-sample ANOVA $F$ test, assume
$$
e_{ij}\overset{\mathrm{iid}}{\sim}N(0,\sigma^2).
$$

\pause

This says that:

- errors are independent;
- errors have mean zero;
- errors have a common variance across groups;
- errors are normally distributed.

\pause

The common variance assumption is why ANOVA is an analysis of **variance** used to test hypotheses about **means**.

# Why not just do several $t$ tests?

With three groups, there are three pairwise comparisons.

\pause

With $a$ groups, there are
$$
\binom{a}{2}=\frac{a(a-1)}{2}
$$
pairwise comparisons.

\pause

Doing many separate tests changes the error-rate question.

\pause

The one-way ANOVA $F$ test first asks one omnibus question:

> Do the data provide evidence that not all population group means are equal?

# Decomposing one observation

For any observation,
$$
Y_{ij}-\bar Y_{\cdot\cdot}
=
(\bar Y_{i\cdot}-\bar Y_{\cdot\cdot})
+
(Y_{ij}-\bar Y_{i\cdot}).
$$

\pause

That is:

$$
\text{distance from grand mean}
=
\text{distance of group mean from grand mean}
+
\text{distance from group mean}.
$$

# Visualising the decomposition

```{r}
#| fig-width: 7.8
#| fig-height: 4.1
#| out-width: "88%"
ggplot2::ggplot(trt_aug, ggplot2::aes(x = group_index, y = bp)) +
  ggplot2::geom_hline(
    yintercept = y_bar,
    linetype = 2,
    linewidth = 0.8
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(xend = group_index, yend = group_mean),
    alpha = 0.35
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(
      x = group_index - 0.24,
      xend = group_index - 0.24,
      y = group_mean,
      yend = grand_mean
    ),
    linewidth = 1
  ) +
  ggplot2::geom_point(size = 2, alpha = 0.75) +
  ggplot2::geom_point(
    data = trt_means,
    ggplot2::aes(x = as.numeric(group), y = bp),
    size = 4
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq_along(group_levels),
    labels = group_levels
  ) +
  ggplot2::labs(
    x = "Treatment group",
    y = "Blood pressure",
    caption = "Dashed line: grand mean. Large points: group means."
  ) +
  ggplot2::theme_minimal(base_size = 15)
```

# Sums of squares

Squaring and summing the decomposition gives
$$
\underbrace{
\sum_{i=1}^a\sum_{j=1}^{n_i}
(Y_{ij}-\bar Y_{\cdot\cdot})^2
}_{SST}
=
\underbrace{
\sum_{i=1}^a
n_i(\bar Y_{i\cdot}-\bar Y_{\cdot\cdot})^2
}_{SS_A}
+
\underbrace{
\sum_{i=1}^a\sum_{j=1}^{n_i}
(Y_{ij}-\bar Y_{i\cdot})^2
}_{SSE}.
$$

\pause

So:
$$
SST=SS_A+SSE.
$$

# What the pieces mean

::::{.columns}

:::{.column width="50%"}

\begin{block}{Between groups}
$$
SS_A=
\sum_{i=1}^a
n_i(\bar Y_{i\cdot}-\bar Y_{\cdot\cdot})^2
$$

Large when group means are far from the grand mean.
\end{block}

:::

:::{.column width="50%"}

\begin{block}{Within groups}
$$
SSE=
\sum_{i=1}^a\sum_{j=1}^{n_i}
(Y_{ij}-\bar Y_{i\cdot})^2
$$

Large when observations vary a lot within groups.
\end{block}

:::

::::

# Degrees of freedom

The total variation has
$$
n-1
$$
degrees of freedom.

\pause

The between-group variation has
$$
a-1
$$
degrees of freedom.

\pause

The within-group variation has
$$
n-a
$$
degrees of freedom.

\pause

So the degrees of freedom decompose in the same way:
$$
n-1=(a-1)+(n-a).
$$

# Mean squares and the ANOVA $F$ statistic

Define
$$
MS_A=\frac{SS_A}{a-1},
\qquad
MSE=\frac{SSE}{n-a}.
$$

\pause

The ANOVA test statistic is
$$
\boxed{
F=
\frac{MS_A}{MSE}
=
\frac{SS_A/(a-1)}{SSE/(n-a)}
}.
$$

\pause

Under $H_0$,
$$
F\sim F_{a-1,n-a}.
$$

# Why this statistic makes sense

Under $H_0$, the group means differ only because of random error.

\pause

Then both $MS_A$ and $MSE$ estimate the same error variance $\sigma^2$.

\pause

So under $H_0$, the ratio
$$
\frac{MS_A}{MSE}
$$
should be close to one.

\pause

If group means differ more than expected from within-group variation, $MS_A$ becomes large relative to $MSE$.

# ANOVA table anatomy

\begin{center}
\begin{tabular}{lcccc}
\hline
Source & Sum of squares & df & Mean square & $F$ \\
\hline
Groups & $SS_A$ & $a-1$ & $SS_A/(a-1)$ & $MS_A/MSE$ \\
Residual & $SSE$ & $n-a$ & $SSE/(n-a)$ & \\
Total & $SST$ & $n-1$ & & \\
\hline
\end{tabular}
\end{center}

\pause

The residual line estimates unexplained variation after allowing different group means.

# Calculation for the treatment example {.smaller}

```{r}
calc_tbl <- data.frame(
  Quantity = c("SST", "SS_A", "SSE", "MS_A", "MSE", "F", "p-value"),
  Value = c(SST, SSA, SSE, MSA, MSE, F_anova, p_anova),
  check.names = FALSE
)
knitr::kable(calc_tbl, format = "latex", digits = 3)
```

\pause

For these data,
$$
F=`r format_number(F_anova)`,
\qquad
p`r format_p(p_anova)`.
$$

# Using `anova()` in `R`

```{r}
#| echo: true
fit_oneway <- lm(bp ~ group, data = trt_tbl)
anova(fit_oneway)
```

\pause

The model formula `bp ~ group` fits a different mean for each treatment group.

\pause

The ANOVA table compares that fitted model with the intercept-only model.

# Interpreting the omnibus test

For these data, the ANOVA test gives
$$
p`r format_p(p_anova)`.
$$

\pause

At the 5% level, this is evidence against
$$
H_0:
\mu_{\mathrm{Placebo}}
=
\mu_{\mathrm{A}}
=
\mu_{\mathrm{B}}.
$$

\pause

A careful conclusion is:

> The data provide evidence that mean blood pressure is not the same across all treatment groups.

# What the omnibus test does not say

A significant ANOVA test does **not** by itself tell us:

- which groups differ;
- that every pair of groups differs;
- that the differences are practically important;
- that the model assumptions are adequate.

\pause

To answer pairwise or planned comparison questions, use contrasts or post-hoc comparisons with suitable multiplicity control.

# One factor as a regression model

The same one-way ANOVA can be written as a regression model with a categorical predictor.

\pause

If placebo is the reference group, then
$$
Y_i
=
\beta_0
+
\beta_A I(\text{Treatment A})
+
\beta_B I(\text{Treatment B})
+e_i.
$$

\pause

The group means are
$$
\mu_{\mathrm{Placebo}}=\beta_0,
\qquad
\mu_A=\beta_0+\beta_A,
\qquad
\mu_B=\beta_0+\beta_B.
$$

# Why not include all group indicators?

If we include an intercept and one indicator for each group, then
$$
\mathbf 1
=
I(\text{Placebo})
+
I(\text{Treatment A})
+
I(\text{Treatment B}).
$$

\pause

So the columns of the design matrix are linearly dependent.

\pause

Then $X^TX$ is not invertible, and the usual least-squares formula
$$
\hat{\boldsymbol\beta}=(X^TX)^{-1}X^TY
$$
is not available.

\pause

Reference coding fixes this by omitting one group indicator.

# The design matrix used by `lm()`

```{r}
#| echo: true
model.matrix(fit_oneway)[c(1, 11, 21), ]
```

\pause

- The first row is a placebo observation.
- The second row is a treatment A observation.
- The third row is a treatment B observation.

\pause

The intercept is the placebo mean. The other coefficients are differences from placebo.

# Equal means as a joint restriction

In the reference-coded regression model,
$$
\mu_{\mathrm{Placebo}}=\mu_A=\mu_B
\quad\Longleftrightarrow\quad
\beta_A=\beta_B=0.
$$

\pause

This is a joint hypothesis:
$$
H_0:L\boldsymbol\beta=\mathbf 0,
\qquad
L=
\begin{pmatrix}
0&1&0\\
0&0&1
\end{pmatrix}.
$$

\pause

It is exactly the type of hypothesis tested with the Week III joint $F$ statistic.

# The joint regression $F$ test

From Week III, for $q$ restrictions,
$$
F
=
\frac{
(L\hat{\boldsymbol\beta}-\mathbf r)^T
\{L(X^TX)^{-1}L^T\}^{-1}
(L\hat{\boldsymbol\beta}-\mathbf r)
}{
q s^2
}
\sim F_{q,n-k}
$$
under $H_0$.

\pause

For one-way ANOVA with $a$ groups:
$$
q=a-1,
\qquad
k=a,
\qquad
n-k=n-a.
$$

# Computing the joint regression test {.smaller}

```{r}
#| echo: true
L <- rbind(
  c(0, 1, 0),
  c(0, 0, 1)
)
r <- c(0, 0)
X <- model.matrix(fit_oneway)
C <- solve(crossprod(X))
difference <- as.vector(L %*% coef(fit_oneway) - r)
F_statistic <- drop(
  t(difference) %*%
    solve(L %*% C %*% t(L)) %*%
    difference /
    (nrow(L) * sigma(fit_oneway)^2)
)
p_value <- pf(
  F_statistic,
  df1 = nrow(L),
  df2 = df.residual(fit_oneway),
  lower.tail = FALSE
)
c("F statistic" = F_statistic, "p-value" = p_value)
```

# Full and reduced models

The same test can be written as a comparison of nested models.

\pause

Reduced model:
$$
M_R:
\quad
Y_i=\beta_0+e_i.
$$

\pause

Full model:
$$
M_F:
\quad
Y_i
=
\beta_0
+
\beta_A I(\text{Treatment A})
+
\beta_B I(\text{Treatment B})
+e_i.
$$

# Model-comparison $F$ statistic

Let $RSS_R$ be the residual sum of squares for the reduced model and $RSS_F$ the residual sum of squares for the full model.

\pause

If the full model has $k_F$ coefficients and the reduced model has $k_R$ coefficients,
$$
\boxed{
F
=
\frac{(RSS_R-RSS_F)/(k_F-k_R)}{RSS_F/(n-k_F)}
}
\sim F_{k_F-k_R,n-k_F}
$$
under $H_0$.

# Why this is the ANOVA statistic

For the one-way treatment example,
$$
RSS_R=SST,
\qquad
RSS_F=SSE.
$$

\pause

Therefore,
$$
RSS_R-RSS_F=SST-SSE=SS_A.
$$

\pause

Substituting into the model-comparison statistic gives
$$
F
=
\frac{SS_A/(a-1)}{SSE/(n-a)}.
$$

\pause

So the ANOVA $F$ statistic is the regression model-comparison $F$ statistic.

# Comparing models in `R`

```{r}
#| echo: true
fit_reduced <- lm(bp ~ 1, data = trt_tbl)
fit_full <- lm(bp ~ group, data = trt_tbl)
anova(fit_reduced, fit_full)
```

\pause

This gives the same test as `anova(lm(bp ~ group, data = trt_tbl))` for the one-way model.

# The global regression $F$ test

The `summary()` output for a fitted linear model includes a global $F$ test.

\pause

It compares:

- the intercept-only model;
- the fitted model containing all supplied predictors.

\pause

Its null hypothesis is that all non-intercept coefficients are zero.

\pause

For `bp ~ group`, that is exactly the one-way ANOVA equal-means test.

# Sequential sums of squares

For a model with more than one explanatory variable, `anova(fit)` reports sequential sums of squares.

\pause

These compare models in the order specified in the formula.

\pause

For example,

```{r}
#| echo: true
fit_tv_first <- lm(revenue ~ spend_tv + spend_radio, data = ad_tbl)
anova(fit_tv_first)
```

# Reversing the order

The first term is tested before the second term has been added.

\pause

So reversing the order can change the ANOVA table.

```{r}
#| echo: true
fit_radio_first <- lm(revenue ~ spend_radio + spend_tv, data = ad_tbl)
anova(fit_radio_first)
```

\pause

This is called Type I, or sequential, ANOVA.

# Partial tests

A partial test compares the full model with the model obtained by removing one term while keeping the others.

\pause

For TV spend, compare:
$$
\texttt{revenue \textasciitilde{} spend\_radio}
\quad\text{against}\quad
\texttt{revenue \textasciitilde{} spend\_radio + spend\_tv}.
$$

```{r}
#| echo: true
fit_no_tv <- lm(revenue ~ spend_radio, data = ad_tbl)
fit_full <- lm(revenue ~ spend_radio + spend_tv, data = ad_tbl)
anova(fit_no_tv, fit_full)
```

# Partial tests and coefficient tests

For a one-degree-of-freedom predictor,
$$
F=t^2.
$$

\pause

Thus the partial $F$ test for adding TV spend gives the same conclusion as the two-sided $t$ test for the TV coefficient in the full model.

\pause

```{r}
#| echo: true
summary(fit_full)$coefficients
```

\pause

For a factor with several levels, the partial $F$ test evaluates all dummy coefficients for that factor jointly.

# Type II ANOVA

Type II ANOVA tables test each main effect after accounting for the other main effects in the model.

\pause

For main-effects-only models, this matches the idea of a partial test for each term.

```{r}
#| echo: true
if (requireNamespace("car", quietly = TRUE)) {
  car::Anova(fit_full, type = "II")
} else {
  "Install the car package to run car::Anova(fit_full, type = 'II')."
}
```

\pause

The key point is conceptual: always ask **which model comparison** an ANOVA table is carrying out.

# Assumptions and limitations

The ANOVA $F$ test is a model-based test.

\pause

Before relying on its conclusion, assess whether the model adequately represents:

- independence of the errors;
- constant error variance;
- normality of the errors for exact finite-sample inference;
- the scientific structure of the grouping or design.

\pause

A small $p$-value does not rescue a poorly specified model.

# Summary {.smaller}

- One-way ANOVA tests
$$
H_0:\mu_1=\cdots=\mu_a.
$$

- The variation decomposition is
$$
SST=SS_A+SSE.
$$

- The one-way ANOVA statistic is
$$
F=
\frac{SS_A/(a-1)}{SSE/(n-a)}
\sim F_{a-1,n-a}
\quad\text{under }H_0.
$$

- The same test is a regression joint $F$ test and a full-versus-reduced-model $F$ test.

- In larger models, ANOVA tables must be interpreted as specific model comparisons.

# Transition

ANOVA focuses on how much variation is explained by fitted model terms.

\pause

Next, we use the same residual quantities to ask whether the fitted model is adequate.

\pause

That leads to:

- $R^2$ and adjusted $R^2$;
- residual plots;
- outliers, leverage and influence;
- model checking.
