---
title: "Regression Week II: The general linear model"
subtitle: "Slide deck I: Matrix formulation, estimation and sampling distributions"
author: "Yovna Junglee, edited by Miguel Rodo"
institute: University of Cape Town
date: today
format:
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: ../preamble.tex
    keep-tex: true
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

load_advertising <- function() {
  if (requireNamespace("ISLR2", quietly = TRUE) && exists("Advertising", asNamespace("ISLR2"))) {
    return(get("Advertising", asNamespace("ISLR2")))
  }
  if (requireNamespace("ISLR", quietly = TRUE) && exists("Advertising", asNamespace("ISLR"))) {
    return(get("Advertising", asNamespace("ISLR")))
  }
  data.frame(
    TV = c(230.1, 44.5, 17.2, 151.5, 180.8, 8.7, 57.5, 120.2, 8.6, 199.8),
    radio = c(37.8, 39.3, 45.9, 41.3, 10.8, 48.9, 32.8, 19.6, 2.1, 2.6),
    newspaper = c(69.2, 45.1, 69.3, 58.5, 58.4, 75.0, 23.5, 11.6, 1.0, 21.2),
    sales = c(22.1, 10.4, 9.3, 18.5, 12.9, 7.2, 11.8, 13.2, 4.8, 10.6)
  )
}

advertising <- load_advertising()
```

# Why we now need the general linear model

- In Week I, we developed the normal and multivariate normal results needed for regression inference.

\pause

- We now return to regression itself.

\pause

- The main move is from a model with one predictor to a model with several predictors.

\pause

- Matrix notation lets us write, estimate and analyse all of these models at once.

# Lecture outline

By the end of this deck, you should be able to:

- Write simple and multiple linear regression models in scalar and matrix form
- Explain the roles of $Y$, $X$, $\boldsymbol\beta$ and $\mathbf e$ in the general linear model
- Derive the least-squares estimator
$$
\hat{\boldsymbol\beta}=(X^TX)^{-1}X^TY
$$
- State the sampling distribution of $\hat{\boldsymbol\beta}$ under normal errors
- Define fitted values, residuals and the residual variance estimator
- Interpret regression coefficients in a multiple regression model

# The Advertising data set

The `Advertising` data set records sales across 200 markets, together with advertising budgets for three media:

- TV
- radio
- newspaper

\pause

The response variable is `sales`, measured in units sold. The predictors are advertising budgets, measured in thousands of dollars.

\pause

Typical modelling questions include:

- How are advertising budgets related to sales?
- How much do mean sales change when one advertising budget increases?
- Can we predict sales from the advertising budgets?

# A modelling workflow

A standard statistical modelling workflow is:

1. Explore and understand the data
2. Formulate a model
3. Estimate the model parameters
4. Check residuals and assumptions
5. Interpret the model and draw inference

\pause

In this deck, we focus on steps 2 and 3, and set up the distributional results needed for step 5.

# Exploratory data analysis {.smaller}

::::{.columns}

:::{.column width=44%}

```{r}
#| fig-width: 4.2
#| fig-height: 3.6
#| out-width: "100%"
old_par <- par(
  mar = c(4.0, 4.2, 2.2, 0.6),
  mgp = c(2.3, 0.7, 0),
  tcl = -0.25
)
hist(
  advertising$sales,
  breaks = 8,
  probability = TRUE,
  col = "grey85",
  border = "grey35",
  xlab = "sales",
  main = "Sales distribution",
  las = 1
)
lines(density(advertising$sales), lwd = 2.2, col = "#003366")
box()
par(old_par)
```

:::

:::{.column width=56%}

```{r}
#| fig-width: 5.2
#| fig-height: 3.6
#| out-width: "100%"
old_par <- par(
  mar = c(3.5, 3.6, 2.2, 0.8),
  mgp = c(2.2, 0.7, 0),
  tcl = -0.25
)
plot(
  advertising$TV,
  advertising$sales,
  pch = 19,
  cex = 0.7,
  xlab = "TV budget",
  ylab = "sales",
  main = "Sales against TV budget",
  las = 1
)
abline(lm(sales ~ TV, data = advertising), lwd = 2.2, col = "#003366")
box()
par(old_par)
```

:::

::::

# Simple linear regression model

\begin{block}{Model formulation}
For one predictor, the simple linear regression model is
$$
y_i=\beta_0+\beta_1x_i+e_i,
\qquad i=1,\ldots,n.
$$
A common distributional assumption is
$$
e_i \overset{\text{iid}}{\sim} N(0,\sigma^2).
$$
\end{block}

\pause

- $y_i$ is the response for observation $i$.
- $x_i$ is the predictor value for observation $i$.
- $\beta_0$ is the intercept and $\beta_1$ is the slope.
- $e_i$ is the random error term.

# Simple linear regression in matrix form

The simple linear regression model can be written as
$$
Y=\beta_0\mathbf 1_n+\beta_1X_1+\mathbf e,
$$
where
$$
Y=\begin{pmatrix}y_1\\ \vdots\\ y_n\end{pmatrix},
\qquad
X_1=\begin{pmatrix}x_1\\ \vdots\\ x_n\end{pmatrix},
\qquad
\mathbf e=\begin{pmatrix}e_1\\ \vdots\\ e_n\end{pmatrix}.
$$

\pause

Equivalently,
$$
Y=
\begin{pmatrix}
1 & x_1\\
\vdots & \vdots\\
1 & x_n
\end{pmatrix}
\begin{pmatrix}\beta_0\\\beta_1\end{pmatrix}
+
\mathbf e.
$$

# Ordinary least squares

The least-squares idea is to choose $\beta_0$ and $\beta_1$ so that the fitted line is close to the observed data.

\pause

For simple linear regression, minimise
$$
g(\beta_0,\beta_1)
=
\sum_{i=1}^n (y_i-\beta_0-\beta_1x_i)^2.
$$

\pause

This is the sum of squared vertical distances between the observed points and the fitted line.

# OLS: sales and TV {.smaller}

```{r}
#| fig-width: 6.8
#| fig-height: 4.1
#| out-width: "72%"
fit_tv <- lm(sales ~ TV, data = advertising)
old_par <- par(
  mar = c(4.0, 4.2, 2.2, 0.6),
  mgp = c(2.3, 0.7, 0),
  tcl = -0.25
)
plot(
  advertising$TV,
  advertising$sales,
  pch = 19,
  cex = 0.65,
  xlab = "TV",
  ylab = "sales",
  main = "Observed sales, fitted line and residuals",
  las = 1
)
abline(fit_tv, lwd = 2.3, col = "#003366")
for (i in seq_len(nrow(advertising))) {
  segments(
    advertising$TV[i], fitted(fit_tv)[i],
    advertising$TV[i], advertising$sales[i],
    lty = 3,
    col = "grey55"
  )
}
legend(
  "topleft",
  legend = c("Fitted line", "Residuals"),
  lty = c(1, 3),
  lwd = c(2.3, 1.2),
  col = c("#003366", "grey55"),
  bty = "n",
  cex = 0.85
)
box()
par(old_par)
```

# From simple to multiple regression

The Advertising example has more than one possible predictor:

- TV budget
- radio budget
- newspaper budget

\pause

A multiple regression model lets us study the relationship between sales and all three advertising budgets at the same time.

\pause

This is a special case of the **general linear model**.

# The general linear model

\begin{block}{Model formulation}
With $p$ predictors, the model is
$$
Y
=
\beta_0\mathbf 1_n+
\beta_1X_1+\beta_2X_2+\cdots+\beta_pX_p+
\mathbf e.
$$
Equivalently,
$$
Y=X\boldsymbol\beta+\mathbf e.
$$
\end{block}

\pause

Here $X$ contains an intercept column and $p$ predictor columns. Therefore $X$ has $k=p+1$ columns.

# Matrix notation {.smaller}

The model is
$$
Y_{n\times 1}=X_{n\times k}\boldsymbol\beta_{k\times 1}+\mathbf e_{n\times 1},
\qquad k=p+1.
$$

\pause

where
$$
Y=
\begin{pmatrix}Y_1\\\vdots\\Y_n\end{pmatrix},
\quad
X=
\begin{pmatrix}
1 & X_{11} & X_{12} & \cdots & X_{1p}\\
\vdots & \vdots & \vdots & \ddots & \vdots\\
1 & X_{n1} & X_{n2} & \cdots & X_{np}
\end{pmatrix},
\quad
\boldsymbol\beta=
\begin{pmatrix}\beta_0\\\beta_1\\\vdots\\\beta_p\end{pmatrix}.
$$

# Error assumptions

A standard version of the general linear model assumes
$$
\mathbf e\sim N_n(\mathbf 0,\sigma^2I_n).
$$

\pause

This means:

- the errors have mean $0$
- the errors have common variance $\sigma^2$
- the errors are uncorrelated
- under the normal assumption, the errors are independent

\pause

It follows that
$$
Y\sim N_n(X\boldsymbol\beta,\sigma^2I_n).
$$

# Estimating the coefficients

The least-squares estimator chooses $\boldsymbol\beta$ to minimise the residual sum of squares:
$$
RSS(\boldsymbol\beta)
=
(Y-X\boldsymbol\beta)^T(Y-X\boldsymbol\beta).
$$

\pause

Under normal errors, this is also the maximum likelihood estimator of $\boldsymbol\beta$.

\pause

For estimation, the normality assumption is not what gives the formula. The formula comes from minimising the squared residuals.

# The normal equations {.smaller}

Expand the residual sum of squares:
$$
\begin{aligned}
RSS(\boldsymbol\beta)
&=(Y-X\boldsymbol\beta)^T(Y-X\boldsymbol\beta)\\
&=Y^TY-2\boldsymbol\beta^TX^TY+\boldsymbol\beta^TX^TX\boldsymbol\beta.
\end{aligned}
$$

\pause

Differentiating with respect to $\boldsymbol\beta$ and setting the derivative equal to zero gives
$$
-2X^TY+2X^TX\hat{\boldsymbol\beta}=0.
$$

\pause

Therefore
$$
X^TX\hat{\boldsymbol\beta}=X^TY.
$$

These are called the **normal equations**.

# The least-squares estimator

If $X$ has full column rank, then $X^TX$ is invertible and the normal equations have the solution

\begin{block}{Least-squares estimator}
$$
\boxed{
\hat{\boldsymbol\beta}
=
(X^TX)^{-1}X^TY
}.
$$
\end{block}

\pause

- Full column rank means the columns of $X$ are linearly independent.
- This rules out exact collinearity among the intercept and predictors.
- If $X^TX$ is not invertible, the fitted model is not uniquely identified by this formula.

# Fitted values and residuals

The fitted values are
$$
\hat Y=X\hat{\boldsymbol\beta}.
$$

\pause

The residuals are
$$
\hat{\mathbf e}=Y-X\hat{\boldsymbol\beta}=Y-\hat Y.
$$

\pause

So each residual is
$$
\hat e_i = y_i-\hat y_i.
$$

Residuals are observed deviations from the fitted model, while errors are unobserved random quantities in the model.

# The sampling distribution of $\hat{\boldsymbol\beta}$ {.smaller}

\begin{block}{Result}
Under the model
$$
Y=X\boldsymbol\beta+\mathbf e,
\qquad
\mathbf e\sim N_n(\mathbf 0,\sigma^2I_n),
$$
with $X$ full column rank,
$$
\boxed{
\hat{\boldsymbol\beta}
\sim
N_k\left(
\boldsymbol\beta,
\sigma^2(X^TX)^{-1}
\right)
}.
$$
\end{block}

\pause

This follows because $Y$ is multivariate normal and $\hat{\boldsymbol\beta}$ is a linear transformation of $Y$.

# Why $\hat{\boldsymbol\beta}$ is random

The formula
$$
\hat{\boldsymbol\beta}=(X^TX)^{-1}X^TY
$$
contains $Y$.

\pause

If we repeated the same study under the same conditions, the random errors would change, so the observed response vector $Y$ would change.

\pause

Therefore the estimated coefficients would also change.

\pause

The sampling distribution of $\hat{\boldsymbol\beta}$ describes this variation over repeated samples.

# Visualising sampling variation {.smaller}

```{r}
#| fig-width: 8.0
#| fig-height: 3.6
#| out-width: "88%"
set.seed(2005)
n_sim <- 2000
n <- 60
x <- seq(0, 1, length.out = n)
beta0 <- 2
beta1 <- 3
sigma <- 0.7
sim_coef <- replicate(n_sim, {
  y <- beta0 + beta1 * x + rnorm(n, sd = sigma)
  coef(lm(y ~ x))
})
old_par <- par(
  mfrow = c(1, 2),
  mar = c(3.6, 3.8, 2.2, 0.6),
  mgp = c(2.2, 0.7, 0),
  tcl = -0.25
)
hist(
  sim_coef[1, ],
  breaks = 28,
  probability = TRUE,
  col = "grey85",
  border = "grey45",
  xlab = expression(hat(beta)[0]),
  main = expression("Sampling distribution of " * hat(beta)[0]),
  las = 1
)
lines(density(sim_coef[1, ]), lwd = 2.2, col = "#003366")
hist(
  sim_coef[2, ],
  breaks = 28,
  probability = TRUE,
  col = "grey85",
  border = "grey45",
  xlab = expression(hat(beta)[1]),
  main = expression("Sampling distribution of " * hat(beta)[1]),
  las = 1
)
lines(density(sim_coef[2, ]), lwd = 2.2, col = "#003366")
par(old_par)
```

# Linear functions of coefficients {.smaller}

If
$$
\hat{\boldsymbol\beta}
\sim
N_k\left(\boldsymbol\beta,\sigma^2(X^TX)^{-1}\right),
$$
and $A$ is a fixed $q\times k$ matrix, then

\begin{block}{Linear transformation result}
$$
A\hat{\boldsymbol\beta}
\sim
N_q\left(
A\boldsymbol\beta,
\sigma^2A(X^TX)^{-1}A^T
\right).
$$
\end{block}

\pause

This lets us obtain distributions for individual coefficients, differences of coefficients and fitted means.

# Example: linear functions of coefficients {.smaller}

Suppose
$$
(X^TX)^{-1}=\begin{pmatrix}5&-1\\-1&2\end{pmatrix},
\qquad
A=
\begin{pmatrix}
1&1\\
4&-1\\
1&0\\
0&1
\end{pmatrix}.
$$

\pause

Then
$$
A\hat{\boldsymbol\beta}
=
\begin{pmatrix}
\hat\beta_0+\hat\beta_1\\
4\hat\beta_0-\hat\beta_1\\
\hat\beta_0\\
\hat\beta_1
\end{pmatrix}
$$
and
$$
A\hat{\boldsymbol\beta}
\sim
N_4\left(
A\boldsymbol\beta,
\sigma^2A(X^TX)^{-1}A^T
\right).
$$

# Residual variance: MLE and unbiased estimator

The maximum likelihood estimator of $\sigma^2$ is
$$
\hat\sigma^2
=
\frac{1}{n}(Y-X\hat{\boldsymbol\beta})^T(Y-X\hat{\boldsymbol\beta})
=
\frac{RSS}{n}.
$$

\pause

This estimator is biased because fitting $k$ coefficients uses $k$ degrees of freedom.

\pause

We therefore use
$$
\boxed{
s^2
=
\frac{1}{n-k}(Y-X\hat{\boldsymbol\beta})^T(Y-X\hat{\boldsymbol\beta})
=
\frac{RSS}{n-k}
}.
$$

# Distribution of the residual variance

\begin{block}{Result}
Under the normal general linear model,
$$
\frac{(n-k)s^2}{\sigma^2}
=
\frac{RSS}{\sigma^2}
\sim
\chi^2_{n-k}.
$$
\end{block}

\pause

The $n-k$ degrees of freedom arise because:

- there are $n$ observed responses
- $k$ regression coefficients are estimated
- the residual vector is constrained to lie in an $(n-k)$-dimensional residual space

# The residual-maker matrix {.smaller}

Define
$$
H=X(X^TX)^{-1}X^T,
\qquad
M=I_n-H.
$$

\pause

Then
$$
\hat Y=HY,
\qquad
\hat{\mathbf e}=MY,
\qquad
RSS=Y^TMY.
$$

\pause

The matrix $H$ projects $Y$ onto the column space of $X$.

\pause

The matrix $M$ removes that projection, leaving the residual component.

# Why projection matrices matter

The residual-maker matrix
$$
M=I_n-X(X^TX)^{-1}X^T
$$
has three key properties:

- $M$ is symmetric
- $M$ is idempotent, so $M^2=M$
- $MX=0$ and $X^TM=0$

\pause

These properties imply
$$
RSS=Y^TMY=\mathbf e^TM\mathbf e
$$
and, under normal errors,
$$
RSS/\sigma^2\sim\chi^2_{n-k}.
$$

# Advertising model

For the Advertising data set, the multiple regression model is
$$
\text{sales}
=
\beta_0+
\beta_1\,\text{TV}
+
\beta_2\,\text{radio}
+
\beta_3\,\text{newspaper}
+e.
$$

\pause

Here:

- $Y$ is `sales`
- $p=3$ predictors
- $k=4$ coefficients, including the intercept

# Advertising coefficients {.smaller}

```{r}
#| results: asis
fit_adv <- lm(sales ~ TV + radio + newspaper, data = advertising)
coefs <- coef(summary(fit_adv))[, c("Estimate", "Std. Error"), drop = FALSE]
knitr::kable(
  round(coefs, 4),
  col.names = c("Estimate", "Std. Error"),
  caption = "Regression coefficients for the Advertising data set"
)
```

\pause

A coefficient in a multiple regression model is interpreted **holding the other predictors constant**.

\pause

For example, the TV coefficient estimates the change in mean sales associated with a one-unit increase in TV advertising budget, holding radio and newspaper budgets fixed.

# Interpreting the Advertising model

The advertising budgets are measured in thousands of dollars.

\pause

So, if the TV coefficient is approximately $0.0458$, then increasing the TV budget by $1{,}000$ dollars is associated with an increase of about $0.0458$ units of sales, holding radio and newspaper spending fixed.

\pause

Equivalently, increasing the TV budget by $10{,}000$ dollars is associated with an increase of about $0.458$ units of sales, holding the other budgets fixed.

\pause

The phrase **holding fixed** is essential: this is not the same as changing all advertising budgets together.

# Summary

In this deck, we introduced the general linear model
$$
Y=X\boldsymbol\beta+\mathbf e,
\qquad
\mathbf e\sim N_n(\mathbf 0,\sigma^2I_n).
$$

\pause

The key estimates and distributions are
$$
\hat{\boldsymbol\beta}=(X^TX)^{-1}X^TY,
$$
$$
\hat{\boldsymbol\beta}
\sim
N_k\left(\boldsymbol\beta,\sigma^2(X^TX)^{-1}\right),
$$
$$
s^2=\frac{RSS}{n-k},
\qquad
\frac{(n-k)s^2}{\sigma^2}\sim\chi^2_{n-k}.
$$
