---
title: "Regression Week III: Inference"
subtitle: "Slide deck II: Hypothesis testing"
author: "Yovna Junglee, edited by Miguel Rodo"
institute: University of Cape Town
date: today
format:
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: ../preamble.tex
    keep-tex: true
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

set.seed(201)
ad_tbl <- tibble::tibble(
  spend_radio = runif(100, 0, 3)
)
ad_tbl$spend_tv <- 1.2 * ad_tbl$spend_radio + runif(100, 0, 3)
ad_tbl$revenue <- 10 + 2 * ad_tbl$spend_tv + rnorm(100, sd = 2)

fit <- lm(revenue ~ spend_tv + spend_radio, data = ad_tbl)
X <- model.matrix(fit)
C <- solve(crossprod(X))
df_residual <- df.residual(fit)
coef_summary <- summary(fit)$coefficients
coef_ci <- confint(fit)

format_number <- function(x, digits = 3) {
  format(signif(x, digits), trim = TRUE, scientific = FALSE)
}
format_p <- function(p) {
  if (p < 0.001) "<0.001" else paste0("=", format_number(p))
}

ell_difference <- c(0, 1, -1)
difference_estimate <- drop(ell_difference %*% coef(fit))
difference_se <- sqrt(drop(ell_difference %*% vcov(fit) %*% ell_difference))
difference_t <- difference_estimate / difference_se
difference_p <- 2 * pt(abs(difference_t), df = df_residual, lower.tail = FALSE)
difference_ci <- difference_estimate +
  c(-1, 1) * qt(0.975, df = df_residual) * difference_se

L_joint <- rbind(
  c(0, 1, 0),
  c(0, 0, 1)
)
r_joint <- c(0, 0)
joint_difference <- as.vector(L_joint %*% coef(fit) - r_joint)
joint_F <- drop(
  t(joint_difference) %*%
    solve(L_joint %*% C %*% t(L_joint)) %*%
    joint_difference /
    (nrow(L_joint) * sigma(fit)^2)
)
joint_p <- pf(
  joint_F,
  df1 = nrow(L_joint),
  df2 = df_residual,
  lower.tail = FALSE
)
```

# Confidence intervals and hypothesis tests

A confidence interval asks:

> Which parameter values are reasonably compatible with the observed data and model?

\pause

A hypothesis test asks:

> Are the observed data sufficiently incompatible with a specified null value that we should reject it?

\pause

The two procedures use the same estimate and standard error, and are closely connected.

# Lecture outline

By the end of this deck, you should be able to:

- Formulate null and alternative hypotheses about regression coefficients
- Derive and use a $t$ test for one linear combination of coefficients
- Explain the connection between a two-sided test and a confidence interval
- Derive and use an $F$ test for several linear restrictions
- Distinguish a joint test of zero coefficients from a test that two coefficients are equal
- Interpret and report hypothesis-test results correctly

# Set-up and notation {.smaller}

We use the same normal general linear model and notation as in the Week II proofs:
$$
Y=X\boldsymbol\beta+\mathbf e,
\qquad
\mathbf e\sim N_n(\mathbf 0,\sigma^2I_n),
$$
where $X$ is an $n\times k$ design matrix with full column rank.

\pause

Let
$$
C=(X^TX)^{-1},
\qquad
s^2=\frac{RSS}{n-k}.
$$
The Week II proofs establish
$$
\hat{\boldsymbol\beta}
\sim
N_k\left(\boldsymbol\beta,\sigma^2C\right),
\qquad
\frac{(n-k)s^2}{\sigma^2}
=
\frac{RSS}{\sigma^2}
\sim\chi^2_{n-k},
$$
and $\hat{\boldsymbol\beta}$ is independent of $s^2$.

\pause

These are the results used below: distribution of $\hat{\boldsymbol\beta}$, distribution of the residual variance estimator, and independence of $\hat{\boldsymbol\beta}$ and $s^2$ in `NotesRegW2Proofs_referenced.qmd`.

# Null and alternative hypotheses

A hypothesis is a statement about an unknown population parameter.

\pause

For a two-sided test of one regression coefficient,
$$
H_0:\beta_j=0
\qquad\text{against}\qquad
H_1:\beta_j\neq 0.
$$

\pause

More generally, for a linear combination,
$$
H_0:\ell^T\boldsymbol\beta=r
\qquad\text{against}\qquad
H_1:\ell^T\boldsymbol\beta\neq r,
$$
where $r$ is the value specified by the null hypothesis.

# What a hypothesis test does

The test begins by assuming that $H_0$ is true.

\pause

It then measures how far the observed estimate lies from the null value, relative to its standard error.

\pause

A large absolute test statistic indicates that the data would be unusual under $H_0$.

\pause

The $p$-value is the probability, **under $H_0$**, of observing a test statistic at least as extreme as the one obtained.

# Test statistic for one linear combination {.smaller}

Under
$$
H_0:\ell^T\boldsymbol\beta=r,
$$
we have
$$
\ell^T\hat{\boldsymbol\beta}
\sim
N\left(
r,
\sigma^2\ell^TC\ell
\right).
$$

Because $\hat{\boldsymbol\beta}$ and $s^2$ are independent,

\begin{block}{$t$ statistic}
$$
\boxed{
T
=
\frac{
\ell^T\hat{\boldsymbol\beta}-r
}{
s\sqrt{\ell^TC\ell}
}
\sim t_{n-k}
\quad\text{under }H_0
}
$$
\end{block}

# Two-sided $p$-value and decision

For an observed statistic $t_{\mathrm{obs}}$, the two-sided $p$-value is
$$
p
=
P\left(
|T_{n-k}|\geq |t_{\mathrm{obs}}|
\right)
=
2P\left(
T_{n-k}\geq |t_{\mathrm{obs}}|
\right).
$$

\pause

At significance level $\alpha$:

- reject $H_0$ if $p<\alpha$
- equivalently, reject if
$$
|t_{\mathrm{obs}}|>t_{1-\alpha/2,n-k}.
$$

\pause

Failing to reject $H_0$ is not the same as proving that $H_0$ is true.

# Connection to a confidence interval

The $100(1-\alpha)\%$ confidence interval for $\ell^T\boldsymbol\beta$ is
$$
\ell^T\hat{\boldsymbol\beta}
\pm
t_{1-\alpha/2,n-k}
s\sqrt{\ell^TC\ell}.
$$

\pause

For the two-sided test
$$
H_0:\ell^T\boldsymbol\beta=r,
$$
we reject at level $\alpha$ exactly when $r$ lies outside this interval.

\pause

Thus a two-sided 5% test and a 95% confidence interval give the same reject or fail-to-reject conclusion.

# The simulated advertising example

We fit
$$
E(Y_i\mid X)
=
\beta_0
+
\beta_{\mathrm{TV}}x_{i,\mathrm{TV}}
+
\beta_{\mathrm{Radio}}x_{i,\mathrm{Radio}}.
$$

The simulated data were generated using
$$
\beta_0=10,
\qquad
\beta_{\mathrm{TV}}=2,
\qquad
\beta_{\mathrm{Radio}}=0.
$$

\pause

We first test each slope coefficient individually.

# Individual coefficient tests from `summary` {.smaller}

```{r}
#| results: asis
individual_table <- data.frame(
  Parameter = c("Intercept", "TV spend", "Radio spend"),
  Estimate = coef_summary[, "Estimate"],
  `Standard error` = coef_summary[, "Std. Error"],
  `$t$ statistic` = coef_summary[, "t value"],
  `$p$-value` = vapply(
    coef_summary[, "Pr(>|t|)"],
    function(p) if (p < 0.001) "$<0.001$" else format_number(p),
    character(1)
  ),
  check.names = FALSE
)
knitr::kable(
  individual_table,
  format = "latex",
  digits = 3,
  escape = FALSE
)
```

Each row tests the null hypothesis that the corresponding coefficient is zero, holding the other predictors in the model fixed.

# Manual test of the TV coefficient {.smaller}

To test
$$
H_0:\beta_{\mathrm{TV}}=0,
$$
choose
$$
\ell^T=\begin{pmatrix}0&1&0\end{pmatrix}.
$$

Then
$$
t
=
\frac{\hat\beta_{\mathrm{TV}}}{s\sqrt{C_{22}}}.
$$

```{r}
#| echo: true
ell_tv <- c(0, 1, 0)
t_tv <- drop(ell_tv %*% coef(fit)) /
  sqrt(drop(ell_tv %*% vcov(fit) %*% ell_tv))
p_tv <- 2 * pt(abs(t_tv), df = df.residual(fit), lower.tail = FALSE)
c("t statistic" = t_tv, "p-value" = p_tv)
```

This reproduces the TV row in `summary(fit)`.

# Reporting the TV result {.smaller}

A suitable report is:

> Holding radio spend constant, a one-unit increase in TV spend was associated with an estimated `r format_number(coef(fit)["spend_tv"])`-unit increase in mean revenue (95% CI: `r format_number(coef_ci["spend_tv", 1])` to `r format_number(coef_ci["spend_tv", 2])`; $p`r format_p(coef_summary["spend_tv", "Pr(>|t|)"])`$).

\pause

- The estimate describes the magnitude and direction of the association.
- The confidence interval describes its precision.
- The small $p$-value gives strong evidence against $H_0:\beta_{\mathrm{TV}}=0$.

# Reporting the radio result {.smaller}

A suitable report is:

> Holding TV spend constant, the estimated association between radio spend and mean revenue was `r format_number(coef(fit)["spend_radio"])` units (95% CI: `r format_number(coef_ci["spend_radio", 1])` to `r format_number(coef_ci["spend_radio", 2])`; $p=`r format_number(coef_summary["spend_radio", "Pr(>|t|)"])`$).

\pause

- The confidence interval includes zero.
- We therefore fail to reject $H_0:\beta_{\mathrm{Radio}}=0$ at the 5% level.
- This does not establish that the true coefficient is exactly zero.

# Testing whether two effects are equal

To ask whether TV and radio have the same coefficient, test
$$
H_0:
\beta_{\mathrm{TV}}-\beta_{\mathrm{Radio}}=0
$$
against
$$
H_1:
\beta_{\mathrm{TV}}-\beta_{\mathrm{Radio}}\neq 0.
$$

\pause

Choose
$$
\ell^T
=
\begin{pmatrix}0&1&-1\end{pmatrix}.
$$

Then
$$
T
=
\frac{
\hat\beta_{\mathrm{TV}}-\hat\beta_{\mathrm{Radio}}
}{
s\sqrt{C_{22}+C_{33}-2C_{23}}
}
\sim t_{n-k}
$$
under $H_0$.

# Implementing the coefficient-difference test {.smaller}

```{r}
#| echo: true
ell <- c(0, 1, -1)
estimate <- drop(ell %*% coef(fit))
standard_error <- sqrt(drop(ell %*% vcov(fit) %*% ell))
t_statistic <- estimate / standard_error
p_value <- 2 * pt(
  abs(t_statistic),
  df = df.residual(fit),
  lower.tail = FALSE
)
c(
  "estimated difference" = estimate,
  "standard error" = standard_error,
  "t statistic" = t_statistic,
  "p-value" = p_value
)
```

# Interpreting the difference test

The estimated difference is
$$
\hat\beta_{\mathrm{TV}}-\hat\beta_{\mathrm{Radio}}
=
\text{`r format_number(difference_estimate)`}.
$$

Its 95% confidence interval is
$$
\left(
\text{`r format_number(difference_ci[1])`},
\text{`r format_number(difference_ci[2])`}
\right),
$$
with $p`r format_p(difference_p)`$.

\pause

The data provide evidence that the two regression coefficients differ.

\pause

This is a different question from asking whether the two coefficients are jointly zero.

# Several restrictions at once

Suppose we wish to test $q$ linear restrictions simultaneously:
$$
H_0:L\boldsymbol\beta=\mathbf r,
$$
where

- $L$ is a fixed $q\times k$ matrix,
- the rows of $L$ are linearly independent, and
- $\mathbf r$ is a fixed $q\times 1$ vector.

\pause

A single $t$ statistic is not sufficient when $q>1$.

# Distribution under the joint null {.smaller}

Because $L\hat{\boldsymbol\beta}$ is a linear transformation,
$$
L\hat{\boldsymbol\beta}
\sim
N_q\left(
L\boldsymbol\beta,
\sigma^2LCL^T
\right).
$$

Under $H_0:L\boldsymbol\beta=\mathbf r$,
$$
L\hat{\boldsymbol\beta}-\mathbf r
\sim
N_q\left(
\mathbf 0,
\sigma^2LCL^T
\right).
$$

\pause

Therefore,
$$
Q
=
\frac{
(L\hat{\boldsymbol\beta}-\mathbf r)^T
(LCL^T)^{-1}
(L\hat{\boldsymbol\beta}-\mathbf r)
}{
\sigma^2
}
\sim\chi^2_q.
$$

# Constructing the $F$ statistic {.smaller}

We also have
$$
\frac{(n-k)s^2}{\sigma^2}\sim\chi^2_{n-k}.
$$

The statistic $Q$ is a function of $\hat{\boldsymbol\beta}$, and the Week II proofs show that $\hat{\boldsymbol\beta}$ is independent of $s^2$. Hence $Q$ is independent of $(n-k)s^2/\sigma^2$, so
$$
\frac{Q/q}{(n-k)s^2/\{\sigma^2(n-k)\}}
\sim F_{q,n-k}.
$$

After simplifying,

\begin{block}{$F$ statistic for linear restrictions}
$$
\boxed{
F
=
\frac{
(L\hat{\boldsymbol\beta}-\mathbf r)^T
(LCL^T)^{-1}
(L\hat{\boldsymbol\beta}-\mathbf r)
}{
q s^2
}
\sim F_{q,n-k}
}
$$
\end{block}

# Joint test of both advertising slopes

To test
$$
H_0:
\beta_{\mathrm{TV}}=0
\quad\text{and}\quad
\beta_{\mathrm{Radio}}=0,
$$
write
$$
L
=
\begin{pmatrix}
0&1&0\\
0&0&1
\end{pmatrix},
\qquad
\mathbf r
=
\begin{pmatrix}0\\0\end{pmatrix}.
$$

\pause

Here $q=2$. The null hypothesis states that **both** slope coefficients are zero.

# Implementing the joint $F$ test {.smaller}

```{r}
#| echo: true
L <- rbind(
  c(0, 1, 0),
  c(0, 0, 1)
)
r <- c(0, 0)
difference <- as.vector(L %*% coef(fit) - r)
C <- solve(crossprod(model.matrix(fit)))
F_statistic <- drop(
  t(difference) %*%
    solve(L %*% C %*% t(L)) %*%
    difference /
    (nrow(L) * sigma(fit)^2)
)
p_value <- pf(
  F_statistic,
  df1 = nrow(L),
  df2 = df.residual(fit),
  lower.tail = FALSE
)
c("F statistic" = F_statistic, "p-value" = p_value)
```

# Interpreting the joint $F$ test

For these data,
$$
F
=
\text{`r format_number(joint_F)`},
\qquad
p`r format_p(joint_p)`.
$$

\pause

We reject
$$
H_0:\beta_{\mathrm{TV}}=\beta_{\mathrm{Radio}}=0.
$$

\pause

The correct conclusion is:

> At least one of TV spend and radio spend is associated with mean revenue, after accounting for the other predictor.

\pause

The joint test does **not** test whether the two coefficients differ from one another.

# Two different questions {.smaller}

::::{.columns}

:::{.column width="50%"}

\begin{block}{Are the coefficients equal?}
$$
H_0:
\beta_{\mathrm{TV}}-\beta_{\mathrm{Radio}}=0
$$

- One linear restriction
- Use a $t$ test
- Equivalent to an $F$ test with one numerator degree of freedom
\end{block}

:::

:::{.column width="50%"}

\begin{block}{Are both coefficients zero?}
$$
H_0:
\beta_{\mathrm{TV}}=0,
\quad
\beta_{\mathrm{Radio}}=0
$$

- Two linear restrictions
- Use an $F$ test with two numerator degrees of freedom
- Rejection means at least one coefficient is non-zero
\end{block}

:::

::::

# Relationship between $t$ and $F$

When there is only one restriction, $q=1$.

\pause

The $F$ statistic then equals the square of the corresponding $t$ statistic:
$$
F_{1,n-k}=T_{n-k}^2.
$$

\pause

Thus, for a two-sided test of one linear combination, the $t$ and $F$ tests give the same $p$-value and the same decision.

# Interpreting $p$-values carefully

- A $p$-value is calculated assuming the null hypothesis is true.
- It is not the probability that the null hypothesis is true.
- A small $p$-value is evidence against the null hypothesis.
- A large $p$-value may reflect:
  - a small or absent effect
  - an imprecise estimate
  - insufficient data to detect the effect
- Statistical significance does not imply practical importance.

\pause

Always interpret the estimate and confidence interval alongside the $p$-value.

# Reporting hypothesis tests {.smaller}

A useful report usually includes:

- the estimated effect or contrast
- a confidence interval
- the test statistic when relevant
- the degrees of freedom when relevant
- the $p$-value
- a conclusion stated in the language of the scientific question

\pause

Avoid:

- writing that the null hypothesis has been proved
- reporting only “significant” or “not significant”
- interpreting a joint test as answering a different contrast question
- using excessive decimal places

# Assumptions

The exact $t$ and $F$ reference distributions rely on the normal general linear model.

\pause

Before interpreting the results, assess whether the model adequately represents:

- linearity of the conditional mean
- independence of the errors
- constant error variance
- normality of the errors for exact small-sample inference

\pause

Hypothesis tests do not correct a poorly specified model.

# Summary {.smaller}

For one linear restriction,
$$
H_0:\ell^T\boldsymbol\beta=r,
$$
use
$$
\boxed{
T
=
\frac{
\ell^T\hat{\boldsymbol\beta}-r
}{
s\sqrt{\ell^T(X^TX)^{-1}\ell}
}
\sim t_{n-k}
}.
$$

For $q$ simultaneous restrictions,
$$
H_0:L\boldsymbol\beta=\mathbf r,
$$
use
$$
\boxed{
F
=
\frac{
(L\hat{\boldsymbol\beta}-\mathbf r)^T
\{L(X^TX)^{-1}L^T\}^{-1}
(L\hat{\boldsymbol\beta}-\mathbf r)
}{
q s^2
}
\sim F_{q,n-k}
}.
$$
