---
title: "Regression Week III: Inference"
subtitle: "Slide deck I: Confidence intervals"
author: "Yovna Junglee, edited by Miguel Rodo"
institute: University of Cape Town
date: today
format:
  beamer:
    embed-resources: true
    aspectratio: 169
    urlcolor: cyan
    linkcolor: blue
    filecolor: magenta
    include-in-header:
      file: ../preamble.tex
    keep-tex: true
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

set.seed(201)
ad_tbl <- tibble::tibble(
  spend_radio = runif(100, 0, 3)
)
ad_tbl$spend_tv <- 1.2 * ad_tbl$spend_radio + runif(100, 0, 3)
ad_tbl$revenue <- 10 + 2 * ad_tbl$spend_tv + rnorm(100, sd = 2)

fit <- lm(revenue ~ spend_tv + spend_radio, data = ad_tbl)
X <- model.matrix(fit)
C <- solve(crossprod(X))
df_residual <- df.residual(fit)
alpha <- 0.05
critical_t <- qt(1 - alpha / 2, df = df_residual)

coef_ci <- confint(fit, level = 1 - alpha)
coef_plot_tbl <- data.frame(
  term = c("TV spend", "Radio spend"),
  estimate = coef(fit)[c("spend_tv", "spend_radio")],
  lower = coef_ci[c("spend_tv", "spend_radio"), 1],
  upper = coef_ci[c("spend_tv", "spend_radio"), 2]
)

format_number <- function(x, digits = 3) {
  format(signif(x, digits), trim = TRUE, scientific = FALSE)
}
```

# Why point estimates are not enough

- The least-squares estimate gives one value for each unknown regression coefficient.

\pause

- A different random sample would generally give different estimates.

\pause

- We therefore need to quantify the sampling uncertainty around an estimate.

\pause

- A confidence interval gives a range of parameter values that are reasonably compatible with the data and model.

# Lecture outline

By the end of this deck, you should be able to:

- Construct a confidence interval for a linear combination of regression coefficients
- Explain why replacing $\sigma$ by $s$ leads to a $t$ distribution
- Construct confidence intervals for individual coefficients and mean responses
- Distinguish a confidence interval for a mean response from a prediction interval for a new response
- Interpret and report a confidence interval correctly

# Set-up and notation {.smaller}

We use the same normal general linear model and notation as in the Week II proofs:
$$
Y=X\boldsymbol\beta+\mathbf e,
\qquad
\mathbf e\sim N_n(\mathbf 0,\sigma^2I_n),
$$
where $X$ is an $n\times k$ design matrix with full column rank.

\pause

Let
$$
C=(X^TX)^{-1},
\qquad
s^2=\frac{RSS}{n-k}.
$$
The Week II proofs establish
$$
\hat{\boldsymbol\beta}
\sim
N_k\left(\boldsymbol\beta,\sigma^2C\right),
\qquad
\frac{(n-k)s^2}{\sigma^2}
=
\frac{RSS}{\sigma^2}
\sim\chi^2_{n-k},
$$
and $\hat{\boldsymbol\beta}$ is independent of $s^2$.

\pause

These are the results used below: distribution of $\hat{\boldsymbol\beta}$, distribution of the residual variance estimator, and independence of $\hat{\boldsymbol\beta}$ and $s^2$ in `NotesRegW2Proofs_referenced.qmd`.

# The simulated advertising example

Suppose mean revenue is modelled as
$$
E(Y_i\mid X)
=
\beta_0
+
\beta_{\mathrm{TV}}x_{i,\mathrm{TV}}
+
\beta_{\mathrm{Radio}}x_{i,\mathrm{Radio}}.
$$

The simulated data were generated using
$$
\beta_0=10,
\qquad
\beta_{\mathrm{TV}}=2,
\qquad
\beta_{\mathrm{Radio}}=0,
\qquad
\sigma=2.
$$

\pause

The fitted model is
```{r}
#| results: asis
cat(
  "$$\n",
  "\\widehat{E(Y_i\\mid X)}",
  "=", format_number(coef(fit)[1]),
  "+", format_number(coef(fit)[2]), "x_{i,\\mathrm{TV}}",
  if (coef(fit)[3] >= 0) "+" else "-",
  format_number(abs(coef(fit)[3])), "x_{i,\\mathrm{Radio}}.",
  "\n$$"
)
```

# Confidence intervals from `R` {.smaller}

```{r}
#| results: asis
coef_table <- data.frame(
  Parameter = c("Intercept", "TV spend", "Radio spend"),
  Estimate = unname(coef(fit)),
  `Lower limit` = coef_ci[, 1],
  `Upper limit` = coef_ci[, 2],
  check.names = FALSE
)
knitr::kable(coef_table, format = "latex", digits = 3)
```

\pause

- The TV-spend interval is centred on its estimated coefficient.
- The radio-spend interval includes zero.
- We now derive where these intervals come from.

# Visualising coefficient intervals {.smaller}

```{r}
#| fig-width: 7.2
#| fig-height: 4.0
#| out-width: "76%"
old_par <- par(
  mar = c(3.7, 7.0, 1.0, 0.6),
  mgp = c(2.3, 0.7, 0),
  tcl = -0.25
)
y_pos <- c(2, 1)
plot(
  coef_plot_tbl$estimate,
  y_pos,
  xlim = range(c(coef_plot_tbl$lower, coef_plot_tbl$upper, 0)),
  ylim = c(0.5, 2.5),
  pch = 19,
  cex = 1.1,
  yaxt = "n",
  ylab = "",
  xlab = "Estimated coefficient",
  las = 1
)
segments(
  coef_plot_tbl$lower, y_pos,
  coef_plot_tbl$upper, y_pos,
  lwd = 2.2,
  col = "#003366"
)
segments(
  coef_plot_tbl$lower, y_pos - 0.08,
  coef_plot_tbl$lower, y_pos + 0.08,
  lwd = 1.6,
  col = "#003366"
)
segments(
  coef_plot_tbl$upper, y_pos - 0.08,
  coef_plot_tbl$upper, y_pos + 0.08,
  lwd = 1.6,
  col = "#003366"
)
abline(v = 0, lty = 3, col = "grey45")
axis(2, at = y_pos, labels = coef_plot_tbl$term, las = 1)
box()
par(old_par)
```

# Linear combinations of coefficients {.smaller}

Many quantities of interest can be written as
$$
\ell^T\boldsymbol\beta,
$$
where $\ell$ is a fixed $k\times 1$ vector.

\pause

Examples include:

- An individual coefficient:
$$
\beta_j=\ell^T\boldsymbol\beta
$$
for a vector $\ell$ that selects coefficient $j$.

- A difference between two coefficients:
$$
\beta_1-\beta_2
=
\begin{pmatrix}0&1&-1\end{pmatrix}\boldsymbol\beta.
$$

- The mean response at predictor vector $\mathbf x_0$:
$$
E(Y_0\mid \mathbf x_0)=\mathbf x_0^T\boldsymbol\beta.
$$

# Distribution of a linear combination

Because $\ell^T\hat{\boldsymbol\beta}$ is a linear transformation of $\hat{\boldsymbol\beta}$,
$$
\ell^T\hat{\boldsymbol\beta}
\sim
N\left(
\ell^T\boldsymbol\beta,
\sigma^2\ell^TC\ell
\right).
$$

\pause

If $\sigma$ were known, then
$$
Z
=
\frac{
\ell^T\hat{\boldsymbol\beta}-\ell^T\boldsymbol\beta
}{
\sigma\sqrt{\ell^TC\ell}
}
\sim N(0,1).
$$

# If $\sigma$ were known

From
$$
P\left(
-z_{1-\alpha/2}
\leq Z\leq
z_{1-\alpha/2}
\right)
=1-\alpha,
$$
we would obtain
$$
\ell^T\boldsymbol\beta
\in
\ell^T\hat{\boldsymbol\beta}
\pm
z_{1-\alpha/2}
\sigma\sqrt{\ell^TC\ell}.
$$

\pause

However, $\sigma$ is unknown in an ordinary regression analysis.

# Why simply plugging in $s$ is not enough

It is tempting to replace $\sigma$ by $s$ and continue using the normal critical value:
$$
\ell^T\hat{\boldsymbol\beta}
\pm
z_{1-\alpha/2}
s\sqrt{\ell^TC\ell}.
$$

\pause

This ignores the extra uncertainty introduced by estimating $\sigma$.

\pause

The resulting interval is too narrow in finite samples because
$$
t_{1-\alpha/2,n-k}>z_{1-\alpha/2}.
$$

# Constructing the $t$ pivot {.smaller}

Recall that
$$
\frac{
\ell^T\hat{\boldsymbol\beta}-\ell^T\boldsymbol\beta
}{
\sigma\sqrt{\ell^TC\ell}
}
\sim N(0,1)
$$
and
$$
\frac{(n-k)s^2}{\sigma^2}\sim\chi^2_{n-k}.
$$

\pause

Under the normal general linear model, $\hat{\boldsymbol\beta}$ and $s^2$ are independent. Therefore,
$$
T
=
\frac{
\ell^T\hat{\boldsymbol\beta}-\ell^T\boldsymbol\beta
}{
s\sqrt{\ell^TC\ell}
}
\sim t_{n-k}.
$$

# Why independence matters

The form alone does not guarantee a $t$ distribution.

\pause

For example, let $Z\sim N(0,1)$ and use $Z^2\sim\chi^2_1$ in the denominator:
$$
\frac{Z}{\sqrt{Z^2}}
=
\frac{Z}{|Z|}
=
\operatorname{sign}(Z).
$$

\pause

This random variable takes only the values $-1$ and $1$, so it is not $t_1$-distributed.

\pause

This is why the independence theorem in `NotesRegW2Proofs_referenced.qmd` matters: it is what turns the normal-over-square-root-chi-square ratio into a $t$ random variable.

# General confidence interval

The $100(1-\alpha)\%$ confidence interval for $\ell^T\boldsymbol\beta$ is

\begin{block}{Confidence interval for a linear combination}
$$
\boxed{
\ell^T\hat{\boldsymbol\beta}
\pm
t_{1-\alpha/2,n-k}
\,s\sqrt{\ell^TC\ell}
}
$$
\end{block}

\pause

The interval has the familiar form
$$
\text{estimate}
\pm
\text{critical value}\times\text{standard error}.
$$

# Special case: an individual coefficient

For coefficient $\beta_j$, choose $\ell$ to select the $j$th coefficient. Then
$$
\operatorname{SE}(\hat\beta_j)
=
s\sqrt{C_{jj}},
$$
and
$$
\boxed{
\beta_j
\in
\hat\beta_j
\pm
t_{1-\alpha/2,n-k}
s\sqrt{C_{jj}}
}.
$$

\pause

When the intercept is indexed as $\beta_0$, take care when translating mathematical indices into `R` matrix indices.

# Special case: a mean response

At a fixed predictor vector $\mathbf x_0$,
$$
\mu_0
=
E(Y_0\mid\mathbf x_0)
=
\mathbf x_0^T\boldsymbol\beta.
$$

Choose $\ell=\mathbf x_0$. Then
$$
\hat\mu_0=\mathbf x_0^T\hat{\boldsymbol\beta}
$$
and
$$
\boxed{
\mu_0
\in
\hat\mu_0
\pm
t_{1-\alpha/2,n-k}
s\sqrt{\mathbf x_0^TC\mathbf x_0}
}.
$$

\pause

This interval describes uncertainty about the **mean response**, not about one new observation.

# Prediction interval for a new response

A new response at $\mathbf x_0$ is
$$
Y_0=\mathbf x_0^T\boldsymbol\beta+e_0.
$$

It varies because:

- the mean response is estimated, and
- the new observation has its own random error $e_0$.

\pause

Therefore,
$$
\boxed{
Y_0
\in
\hat\mu_0
\pm
t_{1-\alpha/2,n-k}
s\sqrt{1+\mathbf x_0^TC\mathbf x_0}
}.
$$

The extra $1$ makes a prediction interval wider than a confidence interval for the mean response.

# Mean-response and prediction intervals {.smaller}

```{r}
#| fig-width: 7.8
#| fig-height: 4.3
#| out-width: "82%"
tv_grid <- seq(min(ad_tbl$spend_tv), max(ad_tbl$spend_tv), length.out = 250)
new_tbl <- data.frame(
  spend_tv = tv_grid,
  spend_radio = mean(ad_tbl$spend_radio)
)
mean_interval <- predict(fit, newdata = new_tbl, interval = "confidence")
pred_interval <- predict(fit, newdata = new_tbl, interval = "prediction")

old_par <- par(
  mar = c(4.0, 4.2, 1.2, 0.8),
  mgp = c(2.4, 0.7, 0),
  tcl = -0.25
)
plot(
  tv_grid,
  mean_interval[, "fit"],
  type = "n",
  ylim = range(pred_interval[, c("lwr", "upr")]),
  xlab = "TV spend",
  ylab = "Revenue",
  las = 1
)
polygon(
  c(tv_grid, rev(tv_grid)),
  c(pred_interval[, "lwr"], rev(pred_interval[, "upr"])),
  col = grDevices::adjustcolor("#647C9C", alpha.f = 0.20),
  border = NA
)
polygon(
  c(tv_grid, rev(tv_grid)),
  c(mean_interval[, "lwr"], rev(mean_interval[, "upr"])),
  col = grDevices::adjustcolor("#F4A261", alpha.f = 0.40),
  border = NA
)
lines(tv_grid, mean_interval[, "fit"], lwd = 2.2, col = "#003366")
points(ad_tbl$spend_tv, ad_tbl$revenue, pch = 19, cex = 0.45, col = "grey45")
legend(
  "topleft",
  legend = c("Fitted mean", "95% mean-response CI", "95% prediction interval"),
  lty = c(1, NA, NA),
  lwd = c(2.2, NA, NA),
  pch = c(NA, 15, 15),
  pt.cex = c(NA, 1.5, 1.5),
  col = c("#003366", "#F4A261", "#647C9C"),
  bty = "n",
  cex = 0.82
)
box()
par(old_par)
```

Radio spend is held at its sample mean along the displayed fitted line.

# Why the $t$ critical value matters {.smaller}

Use only ten observations, so that estimating $\sigma$ has a visible effect.

```{r}
#| include: false
small_tbl <- ad_tbl[1:10, c("spend_tv", "spend_radio")]
X_small <- model.matrix(~ spend_tv + spend_radio, data = small_tbl)
C_small <- solve(crossprod(X_small))
se_factor_tv <- sqrt(C_small["spend_tv", "spend_tv"])

set.seed(3)
coverage <- replicate(5000, {
  sim_tbl <- small_tbl
  sim_tbl$revenue <- 10 + 2 * sim_tbl$spend_tv + rnorm(10, sd = 2)
  sim_fit <- lm(revenue ~ spend_tv + spend_radio, data = sim_tbl)
  estimate <- coef(sim_fit)["spend_tv"]
  standard_error <- sigma(sim_fit) * se_factor_tv
  z_interval <- estimate + c(-1, 1) * qnorm(0.975) * standard_error
  t_interval <- estimate + c(-1, 1) * qt(0.975, df = 7) * standard_error
  c(
    z = z_interval[1] <= 2 && 2 <= z_interval[2],
    t = t_interval[1] <= 2 && 2 <= t_interval[2]
  )
})
coverage_pct <- 100 * rowMeans(coverage)
```

```{r}
#| results: asis
coverage_table <- data.frame(
  Interval = c(
    "Normal critical value with estimated standard error",
    "$t_7$ critical value"
  ),
  `Estimated coverage` = paste0(round(coverage_pct, 1), "\\%"),
  check.names = FALSE
)
knitr::kable(coverage_table, format = "latex", escape = FALSE)
```

\pause

The $t$ interval restores the intended 95% repeated-sampling coverage.

# Interpreting a confidence interval

A 95% confidence interval does **not** mean that there is a 95% probability that the fixed parameter lies in this realised interval.

\pause

The repeated-sampling interpretation is:

> If we repeatedly sampled data under the same conditions and constructed an interval in the same way, 95% of those intervals would contain the true parameter.

\pause

For the observed data, the interval either contains the parameter or it does not.

# Reporting confidence intervals {.smaller}

For TV spend, a suitable report is:

> Holding radio spend constant, a one-unit increase in TV spend was associated with an estimated `r format_number(coef(fit)["spend_tv"])`-unit increase in mean revenue (95% CI: `r format_number(coef_ci["spend_tv", 1])` to `r format_number(coef_ci["spend_tv", 2])`).

\pause

For radio spend:

> Holding TV spend constant, the estimated association between radio spend and mean revenue was `r format_number(coef(fit)["spend_radio"])` units (95% CI: `r format_number(coef_ci["spend_radio", 1])` to `r format_number(coef_ci["spend_radio", 2])`).

\pause

- Report enough digits to communicate the result, but not unnecessary precision.
- Interpret the interval in the units and context of the fitted model.

# What the interval does not rescue

Confidence intervals rely on the fitted model and its assumptions.

\pause

In particular, check whether the model adequately represents:

- linearity of the conditional mean
- independence of the errors
- constant error variance
- normality of the errors for exact small-sample inference

\pause

A precisely calculated interval from a badly specified model can still be misleading.

# Summary

Under the normal general linear model,
$$
\frac{
\ell^T\hat{\boldsymbol\beta}-\ell^T\boldsymbol\beta
}{
s\sqrt{\ell^T(X^TX)^{-1}\ell}
}
\sim t_{n-k}.
$$

Therefore,
$$
\boxed{
\ell^T\boldsymbol\beta
\in
\ell^T\hat{\boldsymbol\beta}
\pm
t_{1-\alpha/2,n-k}
s\sqrt{\ell^T(X^TX)^{-1}\ell}
}.
$$

\pause

The same result gives intervals for individual coefficients and mean responses. A prediction interval for a new response contains an additional variance term.
